---
title: "AdvR 04 -- Control Flow"
---

```{r setup, include=FALSE}
source("_common.R")
library(tidyverse)
```

# Choices

## `if` / `else` Basics

```{r}
#| eval: false
if (condition) true_action
if (condition) true_action else false_action
```

Typically with compound statements:

```{r}
#| output.lines: 3
grade <- function(x) {
  if (x > 90) "A"
  else if (x > 80) "B"
  else if (x > 50) "C"
  else "F"
}
grade(85)
```

`if` **returns a value** --- you can assign the result:

```{r}
#| output.lines: 3
x1 <- if (TRUE) 1 else 2
x2 <- if (FALSE) 1 else 2
c(x1, x2)
```

## `if` Returns `NULL` Without `else`

Without `else`, `if` invisibly returns `NULL` when `FALSE`. Since \comm{c()} and \comm{paste0()} drop `NULL` inputs, this enables a compact idiom:

```{r}
#| output.lines: 3
greet <- function(name, birthday = FALSE) {
  paste0("Hi ", name,
    if (birthday) " and HAPPY BIRTHDAY")
}
greet("Jaime", TRUE)
```

```{r}
#| output.lines: 3
greet("Maria", FALSE)
```

## Invalid Inputs

The condition must be a **single** `TRUE` or `FALSE`:

```{r}
#| error: true
#| output.lines: 3
if ("x") 1
```

```{r}
#| error: true
#| output.lines: 3
if (logical()) 1
```

```{r}
#| error: true
#| output.lines: 3
if (NA) 1
```

A vector of length > 1 also errors (as of R 4.2):

```{r}
#| error: true
#| output.lines: 3
if (c(TRUE, FALSE)) 1
```

## Vectorised `if`: `ifelse()` and `case_when()`

`if` works with scalars. For **vectors**, use \comm{ifelse()}:

```{r}
#| output.lines: 3
x <- 1:10
ifelse(x %% 2 == 0, "even", "odd")
```

Note: `NA` in test propagates to output. Best when `yes`/`no` are the same type.

\medskip

For **multiple conditions**, use \comm{dplyr::case\_when()}:

```{r}
#| output.lines: 3
dplyr::case_when(
  x %% 35 == 0 ~ "fizz buzz",
  x %% 5 == 0  ~ "fizz",
  x %% 7 == 0  ~ "buzz",
  .default = as.character(x)
)
```

## `switch()` Statement

Compact alternative to chained `if`/`else if`:

```{r}
#| output.lines: 3
x_option <- function(x) {
  switch(x,
    a = "option 1",
    b = "option 2",
    c = "option 3",
    stop("Invalid `x` value")
  )
}
x_option("b")
```

**Always include a default** (last unnamed argument) --- otherwise unmatched inputs silently return `NULL`.

## `switch()` Fall-Through

Empty RHS passes to the next match (like C's `switch`):

```{r}
#| output.lines: 3
legs <- function(x) {
  switch(x,
    cow = , horse = , dog = 4,
    human = , chicken = 2,
    plant = 0,
    stop("Unknown input"))
}
legs("cow")
```

`cow`, `horse`, and `dog` all fall through to `4`.

\medskip

Use `switch()` only with **character** inputs (numeric `x` is error-prone).

## Exercises (Choices)

1. What type of vector does each call to \comm{ifelse()} return?

```{r}
#| eval: false
ifelse(TRUE, 1, "no")
ifelse(FALSE, 1, "no")
ifelse(NA, 1, "no")
```

Read the documentation and write down the rules in your own words.

2. Why does the following code work?

```{r}
#| eval: false
x <- 1:10
if (length(x)) "not empty" else "empty"

x <- numeric()
if (length(x)) "not empty" else "empty"
```

## Solutions (Choices)

1. \comm{ifelse()} returns the type of the selected value:

```{r}
#| output.lines: 5
ifelse(TRUE, 1, "no")    # double (from yes)
ifelse(FALSE, 1, "no")   # character (from no)
ifelse(NA, 1, "no")      # logical NA
```

Rule: returns entry from `yes` when `TRUE`, from `no` when `FALSE`, `NA` when `NA`. Output type comes from whichever branch is selected.

2. `if()` accepts numeric input: `0` is treated as `FALSE`, all other numbers as `TRUE`. `length(1:10)` is `10` (truthy); `length(numeric())` is `0` (falsy).

# Loops

## `for` Loop Basics

```{r}
#| eval: false
for (item in vector) perform_action
```

```{r}
#| output.lines: 5
for (i in 1:3) {
  print(i)
}
```

`for` assigns `item` to the **current environment**, overwriting existing variables:

```{r}
#| output.lines: 3
i <- 100
for (i in 1:3) {}
i   # 3, not 100
```

## `next` and `break`

Two ways to exit early:

- \comm{next} --- skip current iteration
- \comm{break} --- exit entire loop

```{r}
#| output.lines: 5
for (i in 1:10) {
  if (i < 3) next
  print(i)
  if (i >= 5) break
}
```

## Common Pitfalls

**1. Preallocate output** --- otherwise O(n^2):

```{r}
#| output.lines: 3
means <- c(1, 50, 20)
out <- vector("list", length(means))
for (i in seq_along(means)) {
  out[[i]] <- rnorm(10, means[[i]])
}
```

**2. Never use `1:length(x)`** --- fails when `x` has length 0:

```{r}
#| output.lines: 3
1:length(numeric())   # c(1, 0) --- wrong!
seq_along(numeric())  # integer(0) --- correct
```

## Common Pitfalls: S3 Vectors

**3. Loops strip S3 attributes** from dates and factors:

```{r}
#| output.lines: 3
xs <- as.Date(c("2020-01-01", "2010-01-01"))
for (x in xs) print(x)   # prints raw doubles!
```

**Fix:** index with `[[` instead:

```{r}
#| output.lines: 3
for (i in seq_along(xs)) print(xs[[i]])
```

## `while` and `repeat`

Use when the number of iterations is **unknown**:

- \comm{while(condition) action} --- runs while `condition` is `TRUE`
- \comm{repeat action} --- runs forever until \comm{break}

```{r}
#| eval: false
# Flip coins until heads
flips <- 0
while (sample(c("H", "T"), 1) != "H") {
  flips <- flips + 1
}
```

\medskip

**Flexibility hierarchy:** `for` $\subset$ `while` $\subset$ `repeat`.

Use the **least flexible** tool that works. In practice, prefer `for` (or `map()`/`apply()`) for data analysis.

## Exercises (Loops)

1. Why does this code succeed without errors or warnings?

```{r}
#| eval: false
x <- numeric()
out <- vector("list", length(x))
for (i in 1:length(x)) {
  out[i] <- x[i] ^ 2
}
out
```

2. When this code is evaluated, what can you say about the vector being iterated?

```{r}
#| eval: false
xs <- c(1, 2, 3)
for (x in xs) {
  xs <- c(xs, x * 2)
}
xs
```

3. What does this code tell you about when the index is updated?

```{r}
#| eval: false
for (i in 1:3) {
  i <- i * 2
  print(i)
}
```

## Solutions (Loops) --- 1

1. `1:length(x)` creates `c(1, 0)` when `x` is length 0. Iteration 1: `x[1]` returns `NA` (out-of-bounds), assigned to `out[1]`. Iteration 2: `x[0]` returns `numeric(0)` (zero-index), assigned to `out[0]` (no-op). No error because each step is valid R --- just not what you intended. Use \comm{seq\_along()} instead.

## Solutions (Loops) --- 2--3

2. `for` evaluates the vector **once** at the start. Modifying `xs` inside the loop doesn't change the iteration. The loop runs 3 times (original values 1, 2, 3):

```{r}
#| output.lines: 3
xs <- c(1, 2, 3)
for (x in xs) xs <- c(xs, x * 2)
xs
```

3. The index is **reset at the start** of each iteration. Reassigning `i` inside the loop doesn't affect the next iteration:

```{r}
#| output.lines: 5
for (i in 1:3) {
  i <- i * 2
  print(i)
}
```

Prints 2, 4, 6 --- not 2, 8, 32.

<!--
# Summary

## Key Takeaways

- **`if`/`else`:** works with single `TRUE`/`FALSE` only
- **`if` returns a value:** can assign `x <- if (cond) a else b`
- **Without `else`:** `if` returns `NULL` when `FALSE`
- **Invalid inputs:** character, zero-length, `NA`, length > 1 all error
- **\comm{ifelse()}:** vectorised `if` for element-wise branching
- **\comm{case\_when()}:** multiple conditions, cleaner than nested `ifelse()`
- **\comm{switch()}:** compact dispatch on character values; always include default
- **Fall-through:** empty RHS in `switch()` passes to next match
- **`for` loops:** preallocate output, use \comm{seq\_along()} not `1:length(x)`
- **S3 pitfall:** `for` strips attributes; index with `[[` for dates/factors
- **`while`/`repeat`:** use when iteration count is unknown
- **Prefer `for`** over `while` over `repeat` (least flexible = best)
- **Prefer functionals** (`map()`, `apply()`) over explicit loops
-->
