---
title: "AdvR 14 -- Evaluation"
---

```{r setup, include=FALSE}
source("_common.R")
library(rlang)
library(purrr)
```

# Evaluation Basics

## Evaluation Overview

**Quotation** lets users selectively evaluate (unquote).
**Evaluation** lets developers run quoted code in custom environments.

Together with quosures and data masks, they form **tidy evaluation**:

| Pillar | Purpose |
|--------|---------|
| Quasiquotation | Build code from templates (`!!`, `!!!`) |
| Quosures | Bundle expression + environment |
| Data masks | Evaluate with data frame columns as variables |

## `eval()` Basics

\comm{eval()} takes an expression and an environment:

```{r}
#| output.lines: 3
x <- 10; y <- 2
eval(expr(x + y))
```

Override the environment to change what symbols mean:

```{r}
#| output.lines: 3
eval(expr(x + y), env(x = 1000))
```

**Caution:** `eval()` doesn't quote its first argument:

```{r}
#| output.lines: 3
eval(print(x + 1), env(x = 1000))   # x+1 evaluated FIRST
```

```{r}
#| output.lines: 3
eval(expr(print(x + 1)), env(x = 1000))  # correct
```

## Application: `local()`

\comm{local()} evaluates code in a temporary environment:

```{r}
#| error: true
#| output.lines: 5
foo <- local({ x <- 10; y <- 200; x + y })
foo
x   # not found -- x was local
```

Reimplemented in a few lines:

```{r}
#| output.lines: 3
local2 <- function(expr) {
  env <- env(caller_env())
  eval(enexpr(expr), env)
}
local2({ a <- 10; b <- 200; a + b })
```

## Application: `source()`

Combine \comm{eval()} with \comm{parse\_exprs()} to execute a file:

```{r}
#| eval: false
source2 <- function(path, env = caller_env()) {
  file <- paste(readLines(path, warn = FALSE),
    collapse = "\n")
  exprs <- parse_exprs(file)
  res <- NULL
  for (i in seq_along(exprs))
    res <- eval(exprs[[i]], env)
  invisible(res)
}
```

Pattern: read $\to$ parse $\to$ evaluate each expression in turn.

## Exercises (Eval Basics)

1. Predict the results:

```{r}
#| eval: false
eval(expr(eval(expr(eval(expr(2 + 2))))))
eval(eval(expr(eval(expr(eval(expr(2 + 2)))))))
expr(eval(expr(eval(expr(eval(expr(2 + 2)))))))
```

2. Re-implement `get()` using `sym()` and `eval()`:

```{r}
#| eval: false
get2 <- function(name, env) {}
```

## Solutions (Eval Basics)

1. First: `4` (eval peels each layer). Second: `4` (same). Third: returns the **expression** `eval(expr(eval(expr(eval(expr(2 + 2))))))` because the outer `expr()` quotes everything.

2.

```{r}
#| output.lines: 3
get2 <- function(name, env = caller_env()) {
  eval(sym(name), env)
}
get2("mean")
```

# Quosures

## What Are Quosures?

A **quosure** = expression + environment. Bundles captured code with **where** it was written:

```{r}
#| output.lines: 3
foo <- function(x) enquo(x)
foo(a + b)
```

Three ways to create: \comm{enquo()} (most common), \comm{quo()}, \comm{new\_quosure()}:

```{r}
#| output.lines: 3
new_quosure(expr(x + y), env(x = 1, y = 10))
```

## Evaluating Quosures

\comm{eval\_tidy()} evaluates a quosure in its bundled environment:

```{r}
#| output.lines: 3
q1 <- new_quosure(expr(x + y), env(x = 1, y = 10))
eval_tidy(q1)
```

Essential for `...` where each argument may have a **different** environment:

```{r}
#| output.lines: 3
f <- function(...) { x <- 1; g(..., f = x) }
g <- function(...) enquos(...)
x <- 0
qs <- f(global = x)
map_dbl(qs, eval_tidy)
```

## Quosures Under the Hood

Quosures are a subclass of **formulas** (`~`):

```{r}
#| output.lines: 5
q4 <- new_quosure(expr(x + y + z))
class(q4)
is_call(q4)
```

Extract parts with \comm{get\_expr()} and \comm{get\_env()}:

```{r}
#| output.lines: 3
get_expr(q4)
get_env(q4)
```

## Exercises (Quosures)

1. Predict the result of evaluating each quosure:

```{r}
#| eval: false
q1 <- new_quosure(expr(x), env(x = 1))
q2 <- new_quosure(expr(x + !!q1), env(x = 10))
q3 <- new_quosure(expr(x + !!q2), env(x = 100))
eval_tidy(q3)
```

## Solutions (Quosures)

1. `q1` evaluates to 1. `q2` = `x + q1` in env where `x = 10`, so 10 + 1 = 11. `q3` = `x + q2` in env where `x = 100`, so 100 + 11 = **111**.

```{r}
#| output.lines: 3
q1 <- new_quosure(expr(x), env(x = 1))
q2 <- new_quosure(expr(x + !!q1), env(x = 10))
q3 <- new_quosure(expr(x + !!q2), env(x = 100))
eval_tidy(q3)
```

# Data Masks

## Data Mask Basics

A **data mask** = data frame where `eval_tidy()` looks for variables first:

```{r}
#| output.lines: 3
q1 <- new_quosure(expr(x * y), env(x = 100))
df <- data.frame(y = 1:5)
eval_tidy(q1, df)
```

Wrap it into a `with()`-like function:

```{r}
#| output.lines: 3
with2 <- function(data, expr) {
  expr <- enquo(expr)
  eval_tidy(expr, data)
}
x <- 100
with2(df, x * y)
```

## Pronouns: `.data` and `.env`

Data masks create **ambiguity** --- is `x` from the data or the environment?

Resolve with **pronouns**:

- `.data$x` --- always from the data mask
- `.env$x` --- always from the environment

```{r}
#| output.lines: 3
x <- 1
df <- data.frame(x = 2)
c(with2(df, .data$x), with2(df, .env$x))
```

```{r}
#| error: true
#| output.lines: 3
with2(df, .data$y)   # error: y not in data
```

## Application: `subset2()`

Tidy eval version of \comm{base::subset()}:

```{r}
#| output.lines: 5
subset2 <- function(data, rows) {
  rows <- enquo(rows)
  rows_val <- eval_tidy(rows, data)
  stopifnot(is.logical(rows_val))
  data[rows_val, , drop = FALSE]
}
sample_df <- data.frame(a = 1:5, b = 5:1,
  c = c(5, 3, 1, 4, 1))
subset2(sample_df, b == c)
```

## Application: `transform2()`

Tidy eval version of \comm{base::transform()}:

```{r}
#| output.lines: 5
transform2 <- function(.data, ...) {
  dots <- enquos(...)
  for (i in seq_along(dots)) {
    name <- names(dots)[[i]]
    .data[[name]] <- eval_tidy(dots[[i]], .data)
  }
  .data
}
df <- data.frame(x = c(2, 3, 1), y = runif(3))
transform2(df, x2 = x * 2, y = -y)
```

Uses a for loop (not `map()`) so each column sees previous columns.

# Tidy Evaluation in Practice

## Quote and Unquote Pattern

When wrapping a quoting function, **quote** the user's input, then **unquote** when passing it on:

```{r}
#| output.lines: 3
subsample <- function(df, cond, n = nrow(df)) {
  cond <- enquo(cond)
  df <- subset2(df, !!cond)
  df[sample(nrow(df), n, replace = TRUE), ,
    drop = FALSE]
}
df <- data.frame(x = c(1,1,1,2,2), y = 1:5)
subsample(df, x == 1)
```

**Pattern:** `enquo()` $\to$ `!!` when calling downstream.

## Handling Ambiguity with Pronouns

Without pronouns, bugs arise silently:

```{r}
#| output.lines: 3
threshold_x <- function(df, val) {
  subset2(df, x >= val)
}
has_val <- data.frame(x = 1:3, val = 9:11)
threshold_x(has_val, 2)  # uses df$val, not val=2!
```

Fix with `.data` and `.env`:

```{r}
#| output.lines: 3
threshold_x <- function(df, val) {
  subset2(df, .data$x >= .env$val)
}
threshold_x(has_val, 2)  # correct
```

**Rule:** `.data$` for data columns; `.env$` (or `!!`) for function arguments.

## Exercises (Tidy Eval)

1. Why use a for loop in `transform2()` instead of `map()`? Consider `transform2(df, x = x * 2, x = x * 2)`.

2. What does this alternative `threshold_var()` do differently?

```{r}
#| eval: false
threshold_var <- function(df, var, val) {
  var <- as_string(ensym(var))
  subset2(df, .data[[var]] >= !!val)
}
```

## Solutions (Tidy Eval)

1. A for loop lets each column see **previous modifications**. `transform2(df, x = x * 2, x = x * 2)` quadruples `x` with a for loop (first pass doubles, second doubles again). With `map()`, both would see the original `x`, so `x` would only double.

2. `ensym()` captures `var` as a symbol, `as_string()` converts to a character. `.data[[var]]` uses string indexing --- more robust than `.data$var` which would literally look for a column named `var`. `!!val` unquotes `val` early, avoiding ambiguity if `val` is a column name.

<!--
# Summary

## Key Takeaways

- **`eval(expr, env)`** evaluates an expression in an environment
- **`local()`** evaluates in a temporary environment; easy to reimplement
- **`source()`** = parse file → evaluate each expression
- **Quosures** = expression + environment; created with `enquo()` / `enquos()`
- **`eval_tidy()`** evaluates quosures; supports data masks and pronouns
- **Data masks** let column names act as variables
- **Pronouns** `.data$x` and `.env$x` resolve ambiguity
- **`subset2()`**: `enquo()` → `eval_tidy(rows, data)` → logical subsetting
- **`transform2()`**: `enquos(...)` → for-loop evaluating each column
- **Quote-unquote pattern**: `enquo(cond)` then `subset2(df, !!cond)`
- **Ambiguity rule**: `.data$` for data columns, `.env$` or `!!` for arguments
- Tidy eval = quasiquotation + quosures + data masks
-->
