---
title: "AdvR 09 -- Function Factories & Operators"
---

```{r setup, include=FALSE}
source("_common.R")
library(rlang)
library(scales)
```

# Function Factories

## What Is a Function Factory?

A **function factory** makes functions. The manufactured function's enclosing environment is an execution environment of the factory:

```{r}
#| output.lines: 5
power1 <- function(exp) {
  function(x) x ^ exp
}
square <- power1(2)
cube   <- power1(3)
c(square(3), cube(3))
```

Building blocks: first-class functions + function environments + execution environments.

## How It Works

Each call to the factory creates a **separate** execution environment:

```{r}
#| output.lines: 5
fn_env(square)$exp
fn_env(cube)$exp
```

`square` and `cube` have identical bodies but different enclosing environments --- that's what makes them behave differently.

## Forcing Evaluation

Lazy evaluation causes a subtle bug:

```{r}
#| output.lines: 3
x <- 2
square <- power1(x)
x <- 3
square(2)   # returns 8, not 4!
```

Fix: **force** evaluation in the factory:

```{r}
#| output.lines: 3
power2 <- function(exp) {
  force(exp)
  function(x) x ^ exp
}
x <- 2; square <- power2(x); x <- 3
square(2)   # correct: 4
```

**Rule:** always \comm{force()} arguments used only by the manufactured function.

## Stateful Functions

Factories + `<<-` = functions that **remember** across calls:

```{r}
#| output.lines: 5
new_counter <- function() {
  i <- 0
  function() { i <<- i + 1; i }
}
counter_one <- new_counter()
counter_two <- new_counter()
c(counter_one(), counter_one(), counter_two())
```

Each manufactured function has its own enclosing environment, so counters are **independent**.

\medskip

**Tip:** when state grows complex, switch to R6.

## Garbage Collection in Factories

Manufactured functions retain their enclosing environment. Large temp objects persist:

```{r}
#| output.lines: 5
f1 <- function(n) {
  x <- runif(n); m <- mean(x)
  function() m
}
f2 <- function(n) {
  x <- runif(n); m <- mean(x); rm(x)
  function() m
}
c(lobstr::obj_size(f1(1e6)),
  lobstr::obj_size(f2(1e6)))
```

**Rule:** \comm{rm()} large temp objects before returning the closure.

## Exercises (Factory Fundamentals)

1. Why is `force(x)` better than just writing `x` in a factory?

2. Create `pick(i)` that returns a function subsetting `x` with `i`:

```{r}
#| eval: false
pick(1)(x)       # equivalent to x[[1]]
lapply(mtcars, pick(5))
```

3. What happens without a closure? Predict, then verify:

```{r}
#| eval: false
i <- 0
new_counter2 <- function() { i <<- i + 1; i }
```

4. What if you use `<-` instead of `<<-`?

```{r}
#| eval: false
new_counter3 <- function() {
  i <- 0
  function() { i <- i + 1; i }
}
```

## Solutions (Factory Fundamentals) --- 1--2

1. `force(x)` clearly signals intent --- you're forcing evaluation, not accidentally typing a bare `x`. Both do the same thing, but `force()` is self-documenting.

2.

```{r}
#| output.lines: 3
pick <- function(i) {
  force(i)
  function(x) x[[i]]
}
identical(mtcars[[5]][[1]], pick(5)(mtcars[[1]]))
```

## Solutions (Factory Fundamentals) --- 3--4

3. Without a closure, `i` lives in the **global environment** --- it can be overwritten or interfere with other code. No encapsulation:

```{r}
#| output.lines: 3
i <- 0
new_counter2 <- function() { i <<- i + 1; i }
c(new_counter2(), new_counter2())
i   # global i is modified
```

4. With `<-` instead of `<<-`, the counter **always returns 1** --- each call creates a fresh local `i` and never modifies the enclosing `i`:

```{r}
#| output.lines: 3
new_counter3 <- function() {
  i <- 0
  function() { i <- i + 1; i }
}
c3 <- new_counter3()
c(c3(), c3(), c3())
```

# Factory Applications

## ggplot2 Labelling Factories

The \comm{scales} package uses factories for axis label formatters:

```{r}
#| output.lines: 3
y <- c(12345, 123456, 1234567)
comma_format()(y)
number_format(scale = 1e-3, suffix = " K")(y)
```

Works with ggplot2's `labels` argument:

```{r}
#| eval: false
ggplot(df, aes(x, y)) +
  scale_y_continuous(labels = comma_format())
```

**Pattern:** factory configures formatting rules $\to$ manufactured function applies them.

## Box-Cox Transformation

Re-formulating as a factory makes it easy to explore:

```{r}
#| output.lines: 3
boxcox2 <- function(lambda) {
  if (lambda == 0) function(x) log(x)
  else function(x) (x ^ lambda - 1) / lambda
}
c(boxcox2(1)(4), boxcox2(0)(4))
```

The factory fixes `lambda`; the manufactured function takes `x`. Clean separation of roles.

## Maximum Likelihood with Factories

Factory precomputes constants, speeding up optimisation:

```{r}
#| output.lines: 5
ll_poisson <- function(x) {
  n <- length(x); sum_x <- sum(x)
  c <- sum(lfactorial(x))
  function(lambda) {
    log(lambda) * sum_x - n * lambda - c
  }
}
x1 <- c(41, 30, 31, 38, 29, 24, 30, 29, 31, 38)
optimise(ll_poisson(x1), c(0, 100), maximum = TRUE)
```

**Pattern:** data $\to$ factory $\to$ likelihood function $\to$ \comm{optimise()}.

## Bootstrap Generators

Factory fits model once; manufactured function resamples:

```{r}
#| output.lines: 3
boot_model <- function(df, formula) {
  mod <- lm(formula, data = df)
  fitted <- unname(fitted(mod))
  resid <- unname(resid(mod))
  rm(mod)
  function() fitted + sample(resid)
}
boot_mtcars <- boot_model(mtcars, mpg ~ wt)
head(boot_mtcars())
```

No need to \comm{force()} --- `df` and `formula` are used by `lm()` before the closure is returned.

## Exercises (Factory Applications)

1. Why might you formulate Box-Cox with `x` fixed and `lambda` varying?

```{r}
#| eval: false
boxcox3 <- function(x) {
  function(lambda) {
    if (lambda == 0) log(x)
    else (x ^ lambda - 1) / lambda
  }
}
```

2. Why doesn't `boot_permute()` store a copy of the data?

```{r}
#| eval: false
boot_permute <- function(df, var) {
  n <- nrow(df); force(var)
  function() {
    col <- df[[var]]
    col[sample(n, replace = TRUE)]
  }
}
```

## Solutions (Factory Applications)

1. Fixing `x` lets you try many `lambda` values on the **same data** --- useful for finding the best transformation. You can give it a descriptive name like `boxcox_airpassengers <- boxcox3(AirPassengers)`.

2. R's **copy-on-modify** semantics mean `df` inside the closure is just a name pointing to the same object in memory --- no copy is made unless `df` is modified (which it isn't):

```{r}
#| output.lines: 3
boot_permute <- function(df, var) {
  n <- nrow(df); force(var)
  function() df[[var]][sample(n, replace = TRUE)]
}
bp <- boot_permute(mtcars, "mpg")
lobstr::obj_sizes(mtcars, bp)
```

# Factories + Functionals

## Generating Function Families

Combine \comm{map()} with a factory to create many functions at once:

```{r}
#| output.lines: 5
names <- list(
  square = 2, cube = 3,
  root = 1/2, reciprocal = -1
)
funs <- purrr::map(names, power2)
funs$root(64)
funs$reciprocal(4)
```

Access without `funs$` prefix:

```{r}
#| output.lines: 3
with(funs, root(100))
```

## Exercises (Factories + Functionals)

1. Which is equivalent to `with(x, f(z))`?

   (a) `x$f(x$z)`
   (b) `f(x$z)`
   (c) `x$f(z)`
   (d) `f(z)`
   (e) It depends

## Solutions (Factories + Functionals)

1. **(e) It depends.** With a data frame, `with()` evaluates in a temporary environment where columns are visible, so `f(z)` looks up `f` and `z` in that environment, then the calling environment. The answer depends on what `x` contains:

```{r}
#| output.lines: 5
f <- mean; z <- 1
x <- list(f = mean, z = 1)
identical(with(x, f(z)), x$f(x$z))   # TRUE
identical(with(x, f(z)), f(x$z))     # TRUE
identical(with(x, f(z)), f(z))       # TRUE
```

All four are equivalent here because `f` and `z` in `x` match the global ones.

# Function Operators

## What Is a Function Operator?

A **function operator** takes a function as input and returns a modified function:

```{r}
#| output.lines: 5
chatty <- function(f) {
  force(f)
  function(x, ...) {
    message("Processing ", x)
    f(x, ...)
  }
}
s <- c(3, 2, 1)
purrr::map_dbl(s, chatty(\(x) x^2))
```

Function operators are function factories that take a **function** as input. Python calls them **decorators**.

## `safely()` --- Capturing Errors

\comm{purrr::safely()} wraps a function to return `result` + `error`:

```{r}
#| output.lines: 3
x <- list(1, 10, "oops")
out <- purrr::map(x, purrr::safely(log))
str(out[[3]])
```

Use \comm{list\_transpose()} to restructure:

```{r}
#| output.lines: 3
out <- purrr::list_transpose(purrr::map(x, purrr::safely(log)))
ok <- purrr::map_lgl(out$error, is.null)
out$result[ok]
```

Related: \comm{possibly()} (default on error), \comm{quietly()} (captures messages/warnings).

## `memoise()` --- Caching Results

\comm{memoise::memoise()} remembers previous inputs $\to$ instant on repeat calls:

```{r}
#| eval: false
slow_fn <- function(x) { Sys.sleep(1); x * 10 }
fast_fn <- memoise::memoise(slow_fn)
system.time(fast_fn(1))  # ~1 second
system.time(fast_fn(1))  # instant (cached)
```

**Classic example:** memoised Fibonacci is O(n) instead of exponential:

```{r}
#| eval: false
fib <- memoise::memoise(function(n) {
  if (n < 2) return(n)
  fib(n - 2) + fib(n - 1)
})
```

**Warning:** only memoise **pure** functions (output depends solely on input).

## Building Your Own: `delay_by()` and `dot_every()`

```{r}
#| output.lines: 3
delay_by <- function(f, amount) {
  force(f); force(amount)
  function(...) { Sys.sleep(amount); f(...) }
}
```

```{r}
#| output.lines: 3
dot_every <- function(f, n) {
  force(f); force(n)
  i <- 0
  function(...) {
    i <<- i + 1
    if (i %% n == 0) cat(".")
    f(...)
  }
}
```

**Compose** them with the pipe:

```{r}
#| eval: false
walk2(urls, paths,
  download.file %>% dot_every(10) %>% delay_by(0.1))
```

## Exercises (Function Operators)

1. What does `base::Vectorize()` do? When might you use it?

2. Weigh the pros and cons of:

```{r}
#| eval: false
download.file %>% dot_every(10) %>% delay_by(0.1)
# versus
download.file %>% delay_by(0.1) %>% dot_every(10)
```

3. Should you memoise `download.file()`? Why or why not?

## Solutions (Function Operators)

1. `Vectorize()` wraps a non-vectorised function so it iterates over vector arguments (via `mapply()`). It does **not** speed up the function --- it just changes the input interface. Use when you need to pass vector arguments to a scalar function.

2. Both take the same total time. **First:** dot is printed, then delay occurs before download starts. **Second:** delay occurs, then download starts, then dot is printed after 10th download completes. The difference is timing of the dot relative to the delay.

3. Generally **no**. Files at a URL may change, so memoising would serve stale content. Also, downloaded files can be large, consuming lots of memory. Only memoise if files are small, immutable, and downloaded repeatedly.

<!--
# Summary

## Key Takeaways

- **Function factory:** a function that returns a function; enclosing env = execution env of factory
- **`force()`:** always force arguments used only by the manufactured function
- **Stateful functions:** `<<-` modifies the enclosing environment; each manufactured function gets its own
- **Garbage collection:** `rm()` large temp objects before returning closures
- **Applications:** ggplot2 labellers, Box-Cox, MLE, bootstrap generators
- **Factories + functionals:** `map(params, factory)` generates function families
- **Function operator:** takes function in, returns modified function out (= decorator)
- **`safely()`:** wraps function to return `list(result, error)` instead of throwing
- **`memoise()`:** caches results; only for pure functions
- **`delay_by()` / `dot_every()`:** custom FOs compose with pipe for readable code
- **`Vectorize()`:** base R FO; wraps scalar functions for vector input
-->
