---
title: "AdvR 07 -- Conditions"
---

```{r setup, include=FALSE}
source("_common.R")
library(rlang)
```

# Signalling Conditions

## Three Types of Conditions

R has three conditions you can signal in code:

| Condition | Function | Severity | Prefix |
|-----------|----------|----------|--------|
| Error | `stop()` / `abort()` | Fatal | `Error` |
| Warning | `warning()` / `warn()` | Recoverable | `Warning` |
| Message | `message()` / `inform()` | Informational | (none) |

```{r}
#| error: true
#| output.lines: 5
stop("This is an error")
```

```{r}
#| output.lines: 3
warning("This is a warning")
```

```{r}
#| output.lines: 3
message("This is a message")
```

## Errors

Use `call. = FALSE` to suppress the call in errors (best practice):

```{r}
#| error: true
#| output.lines: 3
h <- function() stop("Bad!", call. = FALSE)
h()
```

\comm{rlang::abort()} suppresses the call by default and stores a backtrace:

```{r}
#| error: true
#| output.lines: 3
h <- function() abort("Bad!")
h()
```

## Warnings

Warnings are weaker than errors --- execution continues:

```{r}
#| eval: false
fw <- function() {
  cat("1\n"); warning("W1")
  cat("2\n"); warning("W2")
}
fw()
#> 1
#> 2
#> Warning messages:
#> 1: In fw() : W1
#> 2: In fw() : W2
```

Control with `options(warn = ...)`:

- `warn = 0` (default): cached, displayed at end
- `warn = 1`: displayed immediately
- `warn = 2`: **promoted to errors** (useful for debugging)

## Messages

Messages are for informational side-effects:

```{r}
#| output.lines: 5
fm <- function() {
  cat("1\n"); message("M1")
  cat("2\n"); message("M2")
}
fm()
```

**`cat()` vs. `message()`:** Use `cat()` when printing is the *purpose*. Use `message()` when it's a *side-effect* (user can suppress with `suppressMessages()`).

## Exercises (Signalling)

1. What does `appendLF` in \comm{message()} do? How does it relate to \comm{cat()}?

2. Write a wrapper around \comm{file.remove()} that throws an error if the file doesn't exist.

## Solutions (Signalling)

1. `appendLF = TRUE` (default) appends a newline `\n` to the message. `cat()` never appends a newline --- you must add `"\n"` explicitly.

2.

```{r}
#| error: true
#| output.lines: 3
file_remove_strict <- function(path) {
  if (!file.exists(path)) {
    stop("Can't delete \"", path,
      "\": doesn't exist.", call. = FALSE)
  }
  file.remove(path)
}
file_remove_strict("nonexistent.csv")
```

# Ignoring Conditions

## Suppressing Conditions

Three heavy-handed tools:

```{r}
#| output.lines: 3
# Ignore errors --- execution continues
try(log("x"))
```

```{r}
#| output.lines: 3
# Ignore warnings
suppressWarnings(as.numeric(c("1", "a")))
```

```{r}
#| output.lines: 3
# Ignore messages
suppressMessages(message("Hello"))
```

**Assignment inside `try()`** --- set a default on failure:

```{r}
#| eval: false
default <- NULL
try(default <- read.csv("maybe-bad.csv"), silent = TRUE)
```

# Handling Conditions

## Condition Objects

Conditions are S3 objects with `message` and `call` components:

```{r}
#| output.lines: 5
cnd <- catch_cnd(stop("An error"))
str(cnd)
```

\comm{rlang::catch\_cnd()} captures a condition object for inspection. \comm{conditionMessage(cnd)} extracts the message text.

## `tryCatch()` --- Exiting Handlers

\comm{tryCatch()} registers **exiting** handlers --- control leaves the protected code:

```{r}
#| output.lines: 3
f3 <- function(x) {
  tryCatch(
    error = function(cnd) NA,
    log(x)
  )
}
f3("x")
```

The handler **exits** the block --- remaining code is skipped:

```{r}
#| output.lines: 3
tryCatch(
  message = function(cnd) "Caught!",
  {
    message("Here")
    stop("Never reached!")
  }
)
```

## `tryCatch()` Details

Extract the error message inside the handler:

```{r}
#| output.lines: 3
tryCatch(
  error = function(cnd) {
    paste0("--", conditionMessage(cnd), "--")
  },
  stop("This is an error")
)
```

`finally` runs cleanup code regardless of success/failure:

```{r}
#| eval: false
path <- tempfile()
tryCatch(
  { writeLines("Hi!", path) },
  finally = { unlink(path) }
)
```

## `withCallingHandlers()` --- Calling Handlers

**Calling** handlers: code execution **continues** after the handler:

```{r}
#| output.lines: 5
withCallingHandlers(
  message = function(cnd) cat("Caught!\n"),
  {
    message("Someone there?")
    message("Why, yes!")
  }
)
```

Compare: `tryCatch()` exits on first catch. `withCallingHandlers()` catches but **keeps going**.

## Muffling Conditions

Use \comm{cnd\_muffle()} to prevent a condition from propagating:

```{r}
#| output.lines: 3
withCallingHandlers(
  message = function(cnd) {
    cat("Caught:", conditionMessage(cnd))
    cnd_muffle(cnd)
  },
  message("Hello")
)
```

Without muffling, the condition bubbles up to the default handler (which prints it).

## Exercises (Handling)

1. Predict the results:

```{r}
#| eval: false
show_condition <- function(code) {
  tryCatch(
    error = function(cnd) "error",
    warning = function(cnd) "warning",
    message = function(cnd) "message",
    { code; NULL }
  )
}
show_condition(stop("!"))
show_condition(10)
show_condition(warning("?!"))
show_condition({10; message("?"); warning("?!")})
```

2. Explain the output:

```{r}
#| eval: false
withCallingHandlers(
  message = function(cnd) message("b"),
  withCallingHandlers(
    message = function(cnd) message("a"),
    message("c")
  )
)
```

## Solutions (Handling) --- 1

1. Results:

```{r}
#| include: false
show_condition <- function(code) {
  tryCatch(
    error = function(cnd) "error",
    warning = function(cnd) "warning",
    message = function(cnd) "message",
    { code; NULL }
  )
}
```

```{r}
#| output.lines: 5
show_condition(stop("!"))          # "error"
show_condition(10)                 # NULL (no condition)
show_condition(warning("?!"))      # "warning"
show_condition({10; message("?"); warning("?!")})  # "message"
```

The last returns `"message"` because `tryCatch()` **exits** on the first caught condition. The `warning("?!")` is never reached.

## Solutions (Handling) --- 2

2. `message("c")` triggers inner handler $\to$ calls `message("a")` $\to$ outer handler catches it $\to$ calls `message("b")` $\to$ not caught, prints **b**. Then `"a"` propagates to default $\to$ prints **a**. Original `"c"` bubbles up to outer handler $\to$ calls `message("b")` again $\to$ prints **b**. Then `"c"` reaches default $\to$ prints **c**.

Output: `b`, `a`, `b`, `c`.

# Custom Conditions

## Why Custom Conditions?

Built-in conditions only have a text `message`. Matching errors by string is fragile:

```{r}
#| error: true
#| output.lines: 3
log(letters)
```

Better: attach **metadata** so handlers can dispatch on structure, not text.

## Creating Custom Conditions with `abort()`

\comm{rlang::abort()} accepts a subclass and extra fields:

```{r}
#| error: true
#| output.lines: 3
abort_bad_argument <- function(arg, must, not = NULL) {
  msg <- glue::glue("`{arg}` must {must}")
  if (!is.null(not)) {
    msg <- glue::glue("{msg}; not {typeof(not)}.")
  }
  abort("error_bad_argument",
    message = msg, arg = arg, must = must, not = not
  )
}
abort_bad_argument("x", "be numeric", not = "a")
```

## Handling Custom Conditions

Use the custom class in \comm{tryCatch()} for targeted handling:

```{r}
#| output.lines: 5
my_log <- function(x, base = exp(1)) {
  if (!is.numeric(x))
    abort_bad_argument("x", "be numeric", not = x)
  base::log(x, base = base)
}

tryCatch(
  error_bad_argument = function(cnd) "bad_argument",
  error = function(cnd) "other error",
  my_log("a")
)
```

**Order matters:** put the most specific handler first. `error` would catch everything before `error_bad_argument`.

# Applications

## Useful `tryCatch()` Patterns

**Default on failure:**

```{r}
#| output.lines: 3
fail_with <- function(expr, value = NULL) {
  tryCatch(error = function(cnd) value, expr)
}
fail_with(log(10), NA_real_)
fail_with(log("x"), NA_real_)
```

## More `tryCatch()` Patterns

**Does it error?**

```{r}
#| output.lines: 3
does_error <- function(expr) {
  tryCatch(error = function(cnd) TRUE,
    { expr; FALSE })
}
does_error(1 + 1)
does_error(stop("!"))
```

## Safe Execution

**Safe result + error** (related to \comm{purrr::safely()}):

```{r}
#| output.lines: 5
safety <- function(expr) {
  tryCatch(
    error = function(cnd) list(result = NULL, error = cnd),
    list(result = expr, error = NULL)
  )
}
str(safety(1 + 10))
```

**Promote warnings to errors** (useful for debugging):

```{r}
#| eval: false
warning2error <- function(expr) {
  withCallingHandlers(
    warning = function(cnd) abort(conditionMessage(cnd)),
    expr
  )
}
```

## Exercises (Applications)

1. How could you rewrite `show_condition()` to use a **single** handler?

## Solutions (Applications)

1. Use the `condition` class (parent of all conditions):

```{r}
#| output.lines: 5
show_condition2 <- function(code) {
  tryCatch(
    condition = function(cnd) {
      if (inherits(cnd, "error"))   return("error")
      if (inherits(cnd, "warning")) return("warning")
      if (inherits(cnd, "message")) return("message")
    },
    { code; NULL }
  )
}
show_condition2(stop("!"))
show_condition2(warning("?"))
```

<!--
# Summary

## Key Takeaways

- **Three conditions:** errors (fatal), warnings (recoverable), messages (informational)
- **`stop()`/`abort()`:** signal errors; use `call. = FALSE` or `abort()` for clean messages
- **`warning()`/`warn()`:** signal warnings; `options(warn = 2)` promotes to errors
- **`message()`/`inform()`:** informational; use for side-effects, not primary output
- **`try()`:** ignore errors; `suppressWarnings()`/`suppressMessages()` for others
- **Condition objects:** lists with `message`, `call`, and `class`
- **`tryCatch()`:** exiting handlers --- control leaves on catch; best for errors
- **`withCallingHandlers()`:** calling handlers --- execution continues; best for warnings/messages
- **`cnd_muffle()`:** prevent condition from propagating to default handler
- **Custom conditions:** `abort("subclass", ...)` attaches metadata for structured handling
- **Handler order:** most specific class first in `tryCatch()`
- **`finally`:** cleanup code that always runs (like `on.exit()`)
- **Useful patterns:** `fail_with()`, `does_error()`, `safety()`, warning-to-error
-->
