---
title: "AdvR 10 -- OOP Overview & S3"
---

```{r setup, include=FALSE}
source("_common.R")
library(sloop)
```

# OOP in R

## Why OOP?

**Polymorphism:** same function, different behaviour for different types:

```{r}
#| output.lines: 3
summary(c(1, 2, 3, 4, 5))
```

```{r}
#| output.lines: 3
summary(factor(c("a", "b", "a")))
```

Without OOP, `summary()` would need endless `if-else` chains --- and only the original author could add new types.

## Two Paradigms

| | Encapsulated | Functional |
|------|-------------|-----------|
| Call style | `object.method(arg)` | `generic(object, arg)` |
| Methods belong to | Objects/classes | Generic functions |
| R systems | R6, RC | **S3**, S4 |
| Other languages | Python, Java, C++ | Julia |

R is unusual: **functional OOP** is the default.

## R's OOP Systems

| System | Style | Source | Use when |
|--------|-------|--------|----------|
| **S3** | Functional | Base R | Simple, flexible dispatch |
| **S4** | Functional | Base R | Formal, rigorous (Bioconductor) |
| **RC** | Encapsulated | Base R | Mutable objects |
| **R6** | Encapsulated | CRAN | Modern RC replacement |

\medskip

**Priority:** S3 > R6 > S4 for most tasks. S3 is used throughout base R and the tidyverse.

## The sloop Package

\comm{sloop} ("sail the seas of OOP") helps inspect OOP objects:

```{r}
#| output.lines: 5
otype(1:10)         # base object
otype(mtcars)       # S3 object
```

```{r}
#| output.lines: 3
ftype(print)        # S3 generic
ftype(str)          # S3 generic
ftype(t.test)       # S3 generic (not a method!)
```

# Base Types

## Base vs. OO Objects

OO objects have a **class attribute**; base objects don't:

```{r}
#| output.lines: 3
is.object(1:10)
attr(1:10, "class")
```

```{r}
#| output.lines: 3
is.object(mtcars)
attr(mtcars, "class")
```

Every object has a **base type** (via \comm{typeof()}), but only OO objects have a class.

## Common Base Types

```{r}
#| output.lines: 5
typeof(1L)          # integer
typeof(1.5)         # double
typeof(TRUE)        # logical
typeof("hello")     # character
typeof(mean)        # closure (function)
```

25 base types total. Key groups: vectors, functions, environments, language components.

\medskip

**Caution:** "numeric" means three different things in R --- alias for `double`, shorthand for integer-or-double, or "behaves like a number" (\comm{is.numeric()}).

# S3 Basics

## What Is an S3 Object?

A base type + a `class` attribute:

```{r}
#| output.lines: 3
f <- factor(c("a", "b", "c"))
typeof(f)
attributes(f)
```

Strip the class with \comm{unclass()}:

```{r}
#| output.lines: 2
unclass(f)
```

## Generics and Method Dispatch

A **generic** dispatches to a **method** based on the object's class:

```{r}
#| output.lines: 3
ftype(print)   # "generic" = dispatches
```

```{r}
#| output.lines: 3
s3_dispatch(print(f))
```

- `=>` = method called; `*` = method exists but not called
- Methods are named `generic.class()` (e.g., `print.factor()`)

\medskip

**Never** call `print.factor()` directly --- always use the generic.

## The `generic.class()` Trap

The `.` in function names is ambiguous:

```{r}
#| output.lines: 3
ftype(t.test)        # generic, not a method!
ftype(t.data.frame)  # method for t() on data frames
```

\comm{t.test()} is a **generic** function for t-tests. \comm{t.data.frame()} is a **method** for the `t()` generic.

\medskip

When unsure, check with \comm{ftype()}.

## Exercises (Basics)

1. Describe the difference in behaviour:

```{r}
#| eval: false
set.seed(1014)
some_days <- as.Date("2017-01-31") + sample(10, 5)
mean(some_days)
mean(unclass(some_days))
```

2. What class, base type, and attributes does this return?

```{r}
#| eval: false
x <- table(rpois(100, 5))
```

## Solutions (Basics)

1. `mean(some_days)` dispatches to `mean.Date()`, returning a Date. `mean(unclass(some_days))` strips the class, so `mean.default()` returns a raw numeric (days since 1970-01-01).

```{r}
#| output.lines: 3
set.seed(1014)
some_days <- as.Date("2017-01-31") + sample(10, 5)
c(mean(some_days), mean(unclass(some_days)))
```

2. Class `table`, base type `integer`, with `dimnames` attribute:

```{r}
#| output.lines: 5
x <- table(rpois(100, 5))
c(class(x), typeof(x))
names(attributes(x))
```

# Creating S3 Classes

## Three Functions for Every Class

1. **Constructor** `new_myclass()` --- creates objects with correct structure
2. **Validator** `validate_myclass()` --- checks values are valid
3. **Helper** `myclass()` --- user-friendly creation

## Constructor Example: Date

A `Date` is a double with class `"Date"`:

```{r}
#| output.lines: 3
new_Date <- function(x = double()) {
  stopifnot(is.double(x))
  structure(x, class = "Date")
}
new_Date(c(-1, 0, 1))
```

A `difftime` also has a `units` attribute:

```{r}
#| output.lines: 3
new_difftime <- function(x = double(), units = "secs") {
  stopifnot(is.double(x))
  units <- match.arg(units,
    c("secs", "mins", "hours", "days", "weeks"))
  structure(x, class = "difftime", units = units)
}
new_difftime(52, "weeks")
```

## Validators

Constructors check **types**; validators check **values**:

```{r}
#| error: true
#| output.lines: 5
new_factor <- function(x = integer(), levels = character()) {
  stopifnot(is.integer(x)); stopifnot(is.character(levels))
  structure(x, levels = levels, class = "factor")
}
validate_factor <- function(x) {
  values <- unclass(x); levels <- attr(x, "levels")
  if (!all(!is.na(values) & values > 0))
    stop("All values must be > 0", call. = FALSE)
  if (length(levels) < max(values))
    stop("Not enough levels", call. = FALSE)
  x
}
validate_factor(new_factor(1:5, "a"))
```

## Helpers

Helpers provide a **user-friendly** interface:

```{r}
#| output.lines: 3
factor2 <- function(x = character(), levels = unique(x)) {
  ind <- match(x, levels)
  validate_factor(new_factor(ind, levels))
}
factor2(c("a", "a", "b"))
```

**Pattern:** helper $\to$ coerce inputs $\to$ call constructor $\to$ call validator.

# Generics and Methods

## Creating a Generic

A generic calls \comm{UseMethod()} to dispatch:

```{r}
#| output.lines: 3
my_generic <- function(x, ...) {
  UseMethod("my_generic")
}
my_generic.numeric <- function(x, ...) {
  paste("Numeric:", sum(x))
}
my_generic.character <- function(x, ...) {
  paste("Character:", paste(x, collapse = ", "))
}
my_generic(1:3)
my_generic(c("a", "b"))
```

## Method Dispatch Details

\comm{UseMethod()} builds `paste0("generic", ".", class(x))` and searches:

```{r}
#| output.lines: 3
s3_dispatch(print(Sys.Date()))
```

The **default** pseudo-class provides a fallback:

```{r}
#| output.lines: 3
s3_dispatch(mean(1:5))
```

## Finding Methods

```{r}
#| output.lines: 5
s3_methods_generic("mean")
```

```{r}
#| output.lines: 5
s3_methods_class("ordered")
```

## Exercises (Generics)

1. Read the source of `t()` and `t.test()`. Confirm that `t.test()` is a generic, not a method for `t()`.

2. What generics does the `table` class have methods for?

## Solutions (Generics)

1. `t()` is a generic that transposes. `t.test()` is a **separate generic** for t-tests (not the `test` method of `t()`):

```{r}
#| output.lines: 3
ftype(t)        # generic
ftype(t.test)   # generic (not t.method!)
```

2.

```{r}
#| output.lines: 5
s3_methods_class("table")
```

# Object Styles and Inheritance

## Four Object Styles

| Style | Length means | Example |
|-------|------------|---------|
| Vector | # observations | `Date`, `factor` |
| Record | # components | `POSIXlt` (list of 11) |
| Data frame | # columns | `data.frame`, `tibble` |
| Scalar | 1 thing | `lm` (list of 12) |

```{r}
#| output.lines: 3
mod <- lm(mpg ~ wt, data = mtcars)
length(mod)   # 12 components, but 1 model
```

## Inheritance via Class Vectors

A class can be a **character vector** --- R searches methods left to right:

```{r}
#| output.lines: 5
class(ordered("x"))
s3_dispatch(print(ordered("x")))
```

- `ordered` is a **subclass** of `factor`
- `factor` is a **superclass** of `ordered`

## `NextMethod()`

\comm{NextMethod()} delegates to the next method in the class hierarchy:

```{r}
#| output.lines: 5
new_secret <- function(x = double()) {
  stopifnot(is.double(x))
  structure(x, class = "secret")
}
print.secret <- function(x, ...) {
  print(strrep("x", nchar(x)))
  invisible(x)
}
`[.secret` <- function(x, i) {
  new_secret(NextMethod())
}
x <- new_secret(c(15, 1, 456))
x[1]
```

```{r}
#| output.lines: 3
s3_dispatch(x[1])
```

## Group Generics

Four **group generics** override behaviour for entire families:

| Group | Functions |
|-------|-----------|
| Math | `abs()`, `sqrt()`, `log()`, `cos()`, ... |
| Ops | `+`, `-`, `*`, `/`, `==`, `<`, `&`, `\|`, ... |
| Summary | `sum()`, `min()`, `max()`, `range()`, ... |
| Complex | `Re()`, `Im()`, `Mod()`, ... |

```{r}
#| output.lines: 3
s3_dispatch(abs(as.difftime(10, units = "mins")))
```

Ops group uses **double dispatch** (dispatches on both arguments).

## Exercises (Inheritance)

1. What does this code return? Why?

```{r}
#| eval: false
generic2 <- function(x) UseMethod("generic2")
generic2.a1 <- function(x) "a1"
generic2.a2 <- function(x) "a2"
generic2.b <- function(x) {
  class(x) <- "a1"
  NextMethod()
}
generic2(structure(list(), class = c("b", "a2")))
```

## Solutions (Inheritance)

1. Returns `"a2"`. Even though `generic2.b` changes the class to `"a1"`, \comm{NextMethod()} ignores the modified class. It uses the **original** class vector `c("b", "a2")` and dispatches to the next in line: `generic2.a2`.

```{r}
#| output.lines: 3
generic2 <- function(x) UseMethod("generic2")
generic2.a1 <- function(x) "a1"
generic2.a2 <- function(x) "a2"
generic2.b <- function(x) {
  class(x) <- "a1"
  NextMethod()
}
generic2(structure(list(), class = c("b", "a2")))
```

<!--
# Summary

## Key Takeaways

- **Two OOP paradigms:** encapsulated (R6/RC) vs functional (S3/S4)
- **S3** is R's simplest, most-used OOP system --- functional style
- **Base types:** `typeof()` returns the C-level type; 25 total
- **S3 object:** base type + `class` attribute
- **Generics** dispatch via `UseMethod()`: `generic.class()` naming
- **`s3_dispatch()`:** shows which method is called
- **Three class functions:** constructor (`new_`), validator (`validate_`), helper
- **Constructor:** checks types, uses `structure()`
- **Four object styles:** vector, record, data frame, scalar
- **Inheritance:** class vectors; `NextMethod()` delegates to parent
- **Group generics:** Math, Ops, Summary, Complex --- override families
- **Double dispatch:** Ops group dispatches on both arguments
- **`sloop`:** `otype()`, `ftype()`, `s3_dispatch()`, `s3_methods_generic()`
-->
