---
title: "AdvR 08 -- Functionals"
---

```{r setup, include=FALSE}
source("_common.R")
library(purrr)
```

# `map()`

## What Is a Functional?

A **functional** takes a function as input and returns a vector:

```{r}
#| output.lines: 3
randomise <- function(f) f(runif(1e3))
randomise(mean)
randomise(sum)
```

Functionals replace `for` loops with **intention-revealing** code. Each functional is tailored for a specific task.

\medskip

**Rule:** if a functional fits, use it. If not, keep the `for` loop.

## `map()` Basics

\comm{purrr::map()} calls a function on each element, returns a **list**:

```{r}
#| output.lines: 5
triple <- function(x) x * 3
map(1:3, triple)
```

**Type-stable variants** return atomic vectors:

```{r}
#| output.lines: 3
map_dbl(1:3, triple)
map_chr(1:3, ~ paste0("Number: ", .x))
map_lgl(1:3, ~ .x > 1)
```

## Anonymous Functions and Shortcuts

**Anonymous function** (formula interface):

```{r}
#| output.lines: 3
map_dbl(1:3, ~ .x ^ 2)
```

**Extraction shortcuts** --- select elements by name, position, or both:

```{r}
#| output.lines: 3
x <- list(list(a = 1, b = 2), list(a = 3, b = 4))
map_dbl(x, "a")        # by name
map_dbl(x, 2)           # by position
```

Behind the scenes, shortcuts use \comm{as\_mapper()} + \comm{pluck()}.

## Passing Arguments

Pass extra arguments via `...`:

```{r}
#| output.lines: 3
x <- list(c(1, NA, 3), c(4, 5, 6))
map_dbl(x, mean, na.rm = TRUE)
```

Or via formula (safer --- avoids partial matching):

```{r}
#| output.lines: 3
map_dbl(x, ~ mean(.x, na.rm = TRUE))
```

**Key difference:** arguments in `...` are evaluated **once** and passed to every call. Formulas are re-evaluated each iteration.

## Exercises (`map()`)

1. What is the difference?

```{r}
#| eval: false
map(1:3, ~ runif(2))   # (a)
map(1:3, runif(2))      # (b)
```

2. Use `map()` and the relevant string/numeric/logical variant to:
   a. Compute the mean of every column in `mtcars`
   b. Determine the type of each column in `nycflights13::flights`
   c. Compute the number of unique values in each column of `iris`

## Solutions (`map()`) --- 1

1. **(a)** \comm{\~{} runif(2)} is a formula $\to$ function: each call generates a **new** pair of random numbers. Returns 3 different length-2 vectors.

**(b)** `runif(2)` is evaluated **once** (producing a numeric vector), then used as an extractor via \comm{pluck()}. Returns `NULL`s.

## Solutions (`map()`) --- 2

2.

```{r}
#| output.lines: 3
map_dbl(mtcars, mean)
```

```{r}
#| output.lines: 3
map_chr(mtcars, typeof)
```

```{r}
#| output.lines: 3
map_int(iris, ~ length(unique(.x)))
```

# Purrr Style

## Chaining Functionals

`map()` returns a list $\to$ chain with `|>` or `%>%`:

```{r}
#| output.lines: 5
mtcars |>
  split(mtcars$cyl) |>
  map(\(df) lm(mpg ~ wt, data = df)) |>
  map(summary) |>
  map_dbl("r.squared")
```

**Pipeline pattern:** data $\to$ split $\to$ fit models $\to$ extract summaries $\to$ pull metric.

# Map Variants

## Overview of Variants

Two dimensions: **(1) output type** and **(2) number of inputs**:

| | List | Atomic | Same type | Side effect |
|------|------|--------|-----------|-------------|
| 1 arg | `map()` | `map_*()` | `modify()` | `walk()` |
| 2 args | `map2()` | `map2_*()` | `modify2()` | `walk2()` |
| n args | `pmap()` | `pmap_*()` | --- | `pwalk()` |

Plus: \comm{imap()} (iterates with index/name), \comm{modify\_if()} (conditional).

## `modify()` --- Same Type Out

\comm{modify()} returns the **same type** as its input:

```{r}
#| output.lines: 5
df <- data.frame(x = 1:3, y = 6:8)
modify(df, ~ .x * 2)
```

Compare: `map(df, ~ .x * 2)` returns a **list**, not a data frame.

## `map2()` and `pmap()`

\comm{map2()} iterates over two inputs in parallel:

```{r}
#| output.lines: 3
map2_dbl(1:3, 4:6, ~ .x + .y)
```

\comm{pmap()} takes a **list** of inputs (any number):

```{r}
#| output.lines: 3
pmap_dbl(list(1:3, 4:6, 7:9), sum)
```

Useful with data frames (each row = one call):

```{r}
#| output.lines: 3
params <- tibble::tibble(n = 1:3, min = c(0, 10, 100))
pmap(params, runif)
```

## `imap()` --- Iterate with Index

\comm{imap(x, f)} is shorthand for `map2(x, names(x), f)` (or indices if unnamed):

```{r}
#| output.lines: 3
imap_chr(mtcars[1:3], ~ paste(.y, "mean:", round(mean(.x), 1)))
```

## `walk()` --- Side Effects Only

\comm{walk()} calls a function for its **side effects**, returns input invisibly:

```{r}
#| output.lines: 3
walk(1:3, ~ cat("Step", .x, "\n"))
```

Use `walk2()` for two inputs (e.g., writing multiple files):

```{r}
#| eval: false
walk2(data_frames, paths, write.csv)
```

## Exercises (Variants)

1. Explain the results:

```{r}
#| eval: false
modify(mtcars, 1)
```

2. Rewrite this code to use `iwalk()`:

```{r}
#| eval: false
cyls <- split(mtcars, mtcars$cyl)
paths <- paste0("cyl-", names(cyls), ".csv")
walk2(cyls, paths, write.csv)
```

## Solutions (Variants)

1. \comm{modify(mtcars, 1)} extracts the **first element** of each column, returning a 1-row data frame (same type as input).

2.

```{r}
#| eval: false
mtcars |>
  split(mtcars$cyl) |>
  iwalk(~ write.csv(.x, paste0("cyl-", .y, ".csv")))
```

# Reduce

## `reduce()` and `accumulate()`

\comm{reduce()} collapses a vector by successively applying a binary function:

```{r}
#| output.lines: 3
reduce(1:4, `+`)   # ((1+2)+3)+4 = 10
```

Always supply `.init` for robustness (handles empty input):

```{r}
#| output.lines: 3
reduce(1:4, `+`, .init = 0)
```

\comm{accumulate()} keeps all intermediate results:

```{r}
#| output.lines: 3
accumulate(1:4, `+`)
```

## `reduce()` Use Case: Merging

```{r}
#| output.lines: 5
dfs <- list(
  data.frame(id = 1:2, a = c("x", "y")),
  data.frame(id = 1:2, b = c(10, 20)),
  data.frame(id = 1:2, c = c(TRUE, FALSE))
)
reduce(dfs, merge, by = "id")
```

## Exercises (Reduce)

1. Explain how the result changes when you switch from `reduce()` to `accumulate()`.

2. Implement a `span()` function that, given a list `x` and a predicate `p`, returns the longest sequential run of elements where `p` is `TRUE`.

## Solutions (Reduce)

1. `reduce()` returns only the final result; `accumulate()` returns a vector of **all intermediate results** (same length as input).

```{r}
#| output.lines: 3
reduce(1:4, `*`)       # 24
accumulate(1:4, `*`)   # 1, 2, 6, 24
```

2. Use \comm{rle()} on the logical vector to find the longest `TRUE` run:

```{r}
#| output.lines: 3
span <- function(x, p) {
  idx <- unlist(map(x, p))
  rle_val <- rle(idx)
  good <- rle_val$values & rle_val$lengths == max(rle_val$lengths[rle_val$values])
  x[rep(good, rle_val$lengths)]
}
```

# Predicate Functionals

## Predicates: Logical Tests

A **predicate** returns a single `TRUE` or `FALSE`.

**Three families:**

| Task | Function | Result |
|------|----------|--------|
| Any match? | `some(x, p)` | Single logical |
| All match? | `every(x, p)` | Single logical |
| First match | `detect(x, p)` | Value |
| First index | `detect_index(x, p)` | Integer |
| Keep matches | `keep(x, p)` | Subset |
| Drop matches | `discard(x, p)` | Subset |

## Predicate Functionals in Action

```{r}
#| output.lines: 3
df <- data.frame(a = 1:3, b = c("x", "y", "z"), c = 4:6)
keep(df, is.numeric)
```

```{r}
#| output.lines: 3
some(df, is.character)
every(df, is.numeric)
```

```{r}
#| output.lines: 3
detect(list(1, "a", 3), is.character)
detect_index(list(1, "a", 3), is.character)
```

## `map_if()` and `modify_if()`

Apply a function **only where a predicate is true**:

```{r}
#| output.lines: 5
str(map_if(iris[1:3, ], is.numeric, mean))
```

\comm{modify\_if()} preserves the input type:

```{r}
#| output.lines: 5
modify_if(iris[1:3, ], is.numeric, round, digits = 0)
```

# Base R Functionals

## `apply()` Family

\comm{apply(X, MARGIN, FUN)} for matrices/arrays:

```{r}
#| output.lines: 3
m <- matrix(1:12, nrow = 3)
apply(m, 1, sum)   # row sums
apply(m, 2, sum)   # col sums
```

**Caution:** never use `apply()` on data frames (coerces to matrix first).

\medskip

**Mathematical functionals** (no purrr equivalent):

```{r}
#| output.lines: 3
integrate(sin, 0, pi)$value
```

```{r}
#| output.lines: 3
uniroot(\(x) x^2 - 4, c(0, 3))$root
```

<!--
# Summary

## Key Takeaways

- **Functional:** takes a function as input; prefer over `for` loops
- **`map()`:** applies function to each element, returns list
- **`map_dbl/chr/lgl/int()`:** type-stable variants returning atomic vectors
- **Shortcuts:** `~ .x + 1` (formula), `"name"` (extractor), `1` (position)
- **`modify()`:** returns same type as input (data frame in $\to$ data frame out)
- **`map2()`/`pmap()`:** iterate over 2 or n inputs in parallel
- **`imap()`:** iterate with names/indices
- **`walk()`:** for side effects (printing, writing files)
- **`reduce()`:** collapse vector with binary function; use `.init`
- **`accumulate()`:** like `reduce()` but keeps intermediates
- **Predicate functionals:** `keep()`, `discard()`, `detect()`, `every()`, `some()`
- **`map_if()`/`modify_if()`:** conditional application
- **`apply()`:** base R for matrices; never on data frames
- **Math functionals:** `integrate()`, `uniroot()`, `optimise()`
-->
