---
title: "AdvR 15 -- Translating R Code"
---

```{r setup, include=FALSE}
source("_common.R")
library(rlang)
library(purrr)

# --- HTML DSL helpers (not shown on slides) ---
html <- function(x) structure(x, class = "advr_html")
print.advr_html <- function(x, ...) {
  out <- paste0("<HTML> ", x)
  cat(paste(strwrap(out), collapse = "\n"), "\n", sep = "")
}
escape <- function(x) UseMethod("escape")
escape.character <- function(x) {
  x <- gsub("&", "&amp;", x)
  x <- gsub("<", "&lt;", x)
  x <- gsub(">", "&gt;", x)
  html(x)
}
escape.advr_html <- function(x) x

dots_partition <- function(...) {
  dots <- list2(...)
  if (is.null(names(dots))) {
    is_named <- rep(FALSE, length(dots))
  } else {
    is_named <- names(dots) != ""
  }
  list(named = dots[is_named], unnamed = dots[!is_named])
}

html_attributes <- function(list) {
  if (length(list) == 0) return("")
  attr <- map2_chr(names(list), list, html_attribute)
  paste0(" ", unlist(attr), collapse = "")
}
html_attribute <- function(name, value = NULL) {
  if (length(value) == 0) return(name)
  if (length(value) != 1) stop("`value` must be NULL or length 1")
  if (is.logical(value)) value <- tolower(value)
  else value <- escape_attr(value)
  paste0(name, "='", value, "'")
}
escape_attr <- function(x) {
  x <- escape.character(x)
  x <- gsub("\'", '&#39;', x)
  x <- gsub("\"", '&quot;', x)
  x <- gsub("\r", '&#13;', x)
  x <- gsub("\n", '&#10;', x)
  x
}

tag <- function(tag) {
  new_function(exprs(... = ), expr({
    dots <- dots_partition(...)
    attribs <- html_attributes(dots$named)
    children <- map_chr(dots$unnamed, escape)
    html(paste0(!!paste0("<", tag), attribs, ">",
      paste(children, collapse = ""),
      !!paste0("</", tag, ">")))
  }), caller_env())
}
void_tag <- function(tag) {
  new_function(exprs(... = ), expr({
    dots <- dots_partition(...)
    if (length(dots$unnamed) > 0)
      abort(!!paste0("<", tag, "> must not have unnamed arguments"))
    attribs <- html_attributes(dots$named)
    html(paste0(!!paste0("<", tag), attribs, " />"))
  }), caller_env())
}

tags <- c("a","abbr","address","article","aside","audio","b","bdi",
  "bdo","blockquote","body","button","canvas","caption","cite",
  "code","colgroup","data","datalist","dd","del","details","dfn",
  "div","dl","dt","em","eventsource","fieldset","figcaption",
  "figure","footer","form","h1","h2","h3","h4","h5","h6","head",
  "header","hgroup","html","i","iframe","ins","kbd","label",
  "legend","li","mark","map","menu","meter","nav","noscript",
  "object","ol","optgroup","option","output","p","pre","progress",
  "q","ruby","rp","rt","s","samp","script","section","select",
  "small","span","strong","style","sub","summary","sup","table",
  "tbody","td","textarea","tfoot","th","thead","time","title","tr",
  "u","ul","var","video")
void_tags <- c("area","base","br","col","command","embed","hr","img",
  "input","keygen","link","meta","param","source","track","wbr")
html_tags <- c(
  tags %>% set_names() %>% map(tag),
  void_tags %>% set_names() %>% map(void_tag))
with_html <- function(code) {
  code <- enquo(code)
  eval_tidy(code, html_tags)
}

# --- LaTeX DSL helpers ---
expr_type <- function(x) {
  if (rlang::is_syntactic_literal(x)) "constant"
  else if (is.symbol(x)) "symbol"
  else if (is.call(x)) "call"
  else if (is.pairlist(x)) "pairlist"
  else typeof(x)
}
switch_expr <- function(x, ...) {
  switch(expr_type(x), ...,
    stop("Unknown type: ", typeof(x), call. = FALSE))
}
flat_map_chr <- function(.x, .f, ...) {
  unlist(purrr::map(.x, .f, ...), use.names = FALSE)
}

latex <- function(x) structure(x, class = "advr_latex")
print.advr_latex <- function(x) cat("<LATEX> ", x, "\n", sep = "")

all_names_rec <- function(x) {
  switch_expr(x, constant = character(),
    symbol = as.character(x),
    call = flat_map_chr(as.list(x[-1]), all_names))
}
all_names <- function(x) unique(all_names_rec(x))

all_calls_rec <- function(x) {
  switch_expr(x, constant = , symbol = character(), call = {
    fname <- as.character(x[[1]])
    children <- flat_map_chr(as.list(x[-1]), all_calls)
    c(fname, children)
  })
}
all_calls <- function(x) unique(all_calls_rec(x))

greek <- c("alpha","theta","tau","beta","vartheta","pi","upsilon",
  "gamma","varpi","phi","delta","kappa","rho","varphi","epsilon",
  "lambda","varrho","chi","varepsilon","mu","sigma","psi","zeta",
  "nu","varsigma","omega","eta","xi","Gamma","Lambda","Sigma",
  "Psi","Delta","Xi","Upsilon","Omega","Theta","Pi","Phi")
greek_list <- set_names(paste0("\\", greek), greek)
greek_env <- as_environment(greek_list)

unary_op <- function(left, right) {
  new_function(exprs(e1 = ),
    expr(paste0(!!left, e1, !!right)), caller_env())
}
binary_op <- function(sep) {
  new_function(exprs(e1 = , e2 = ),
    expr(paste0(e1, !!sep, e2)), caller_env())
}

f_env <- child_env(.parent = empty_env(),
  `+` = binary_op(" + "), `-` = binary_op(" - "),
  `*` = binary_op(" * "), `/` = binary_op(" / "),
  `^` = binary_op("^"), `[` = binary_op("_"),
  `{` = unary_op("\\left{ ", " \\right}"),
  `(` = unary_op("\\left( ", " \\right)"),
  paste = paste,
  sqrt = unary_op("\\sqrt{", "}"),
  sin = unary_op("\\sin(", ")"),
  log = unary_op("\\log(", ")"),
  abs = unary_op("\\left| ", "\\right| "),
  frac = function(a, b) paste0("\\frac{", a, "}{", b, "}"),
  hat = unary_op("\\hat{", "}"),
  tilde = unary_op("\\tilde{", "}"))

unknown_op <- function(op) {
  new_function(exprs(... = ), expr({
    contents <- paste(..., collapse = ", ")
    paste0(!!paste0("\\mathrm{", op, "}("), contents, ")")
  }))
}

to_math <- function(x) {
  expr <- enexpr(x)
  out <- eval_bare(expr, latex_env(expr))
  latex(out)
}

latex_env <- function(expr) {
  calls <- all_calls(expr)
  call_list <- map(set_names(calls), unknown_op)
  call_env <- as_environment(call_list)
  f_env2 <- env_clone(f_env, call_env)
  names <- all_names(expr)
  symbol_env <- as_environment(set_names(names), parent = f_env2)
  env_clone(greek_env, parent = symbol_env)
}
```

# Motivation

## Why Translate R Code?

Metaprogramming + lexical scoping + first-class environments $\to$ translate R code into **other languages**.

**dbplyr** powers this for SQL:

```{r}
#| eval: false
library(dbplyr)
con <- simulate_postgres()
translate_sql(x ^ 2, con = con)
#> <SQL> POWER(`x`, 2.0)
translate_sql(x < 5 & !is.na(x), con = con)
#> <SQL> `x` < 5.0 AND NOT((`x` IS NULL))
```

We'll build two simpler DSLs:

1. **HTML** --- quasiquotation + data masks
2. **LaTeX** --- tidy eval + AST walking

# HTML DSL

## Goal: R Code $\to$ HTML

Generate this HTML from R code that **mirrors its structure**:

```{r}
#| eval: false
with_html(body(
  h1("A heading", id = "first"),
  p("Some text &", b("some bold text.")),
  img(src = "myimg.png", width = 100, height = 100)
))
```

\medskip

Three DSL properties:

- Nesting of function calls $=$ nesting of tags
- Unnamed args $\to$ content; named args $\to$ attributes
- Special characters (`&`, `<`, `>`) auto-escaped

## Escaping with S3

An S3 class separates **raw text** (needs escaping) from **safe HTML**:

```{r}
#| output.lines: 3
html <- function(x) structure(x, class = "advr_html")
escape <- function(x) UseMethod("escape")
escape.character <- function(x) {
  x <- gsub("&", "&amp;", x)
  x <- gsub("<", "&lt;", x)
  x <- gsub(">", "&gt;", x)
  html(x)
}
escape.advr_html <- function(x) x  # already safe
```

```{r}
#| output.lines: 3
c(escape("x > 1 & y < 2"),
  escape(html("<hr />")))   # not double-escaped
```

## Separating Arguments: \comm{dots\_partition()}

Split `...` into named (attributes) and unnamed (children):

```{r}
#| output.lines: 5
dots_partition <- function(...) {
  dots <- list2(...)
  is_named <- if (is.null(names(dots)))
    rep(FALSE, length(dots))
  else names(dots) != ""
  list(named = dots[is_named],
    unnamed = dots[!is_named])
}
str(dots_partition(a = 1, 2, b = 3, 4))
```

## Building a Tag Function: \comm{p()}

Unnamed args become content, named become HTML attributes:

```{r}
#| output.lines: 3
p <- function(...) {
  dots <- dots_partition(...)
  attribs <- html_attributes(dots$named)
  children <- map_chr(dots$unnamed, escape)
  html(paste0("<p", attribs, ">",
    paste(children, collapse = ""), "</p>"))
}
p("Some text", id = "myid")
```

\comm{html\_attributes()} converts a named list to HTML attribute strings (e.g., ` id='myid'`). Implementation omitted --- doesn't introduce new ideas.

## Tag Factory: \comm{tag()}

Generalise `p()` to **any** tag with \comm{rlang::new\_function()} and \comm{!!}:

```{r}
#| output.lines: 3
tag <- function(tag) {
  new_function(
    exprs(... = ),
    expr({
      dots <- dots_partition(...)
      attribs <- html_attributes(dots$named)
      children <- map_chr(dots$unnamed, escape)
      html(paste0(
        !!paste0("<", tag), attribs, ">",
        paste(children, collapse = ""),
        !!paste0("</", tag, ">")))
    }), caller_env())
}
p <- tag("p"); b <- tag("b"); i <- tag("i")
p("Text. ", b(i("bold italic")), class = "x")
```

## Void Tags and \comm{with\_html()}

**Void tags** (`<img>`, `<br>`) have no children --- \comm{void\_tag()} factory enforces attributes only.

Generate functions for all 100+ tags, store in a list, use as **data mask**:

```{r}
#| output.lines: 5
html_tags <- c(
  tags %>% set_names() %>% map(tag),
  void_tags %>% set_names() %>% map(void_tag))
with_html <- function(code) {
  code <- enquo(code)
  eval_tidy(code, html_tags)  # list as data mask!
}
with_html(body(
  h1("A heading", id = "first"),
  p("Some text &", b("some bold text.")),
  img(src = "myimg.png", width = 100,
    height = 100)))
```

## Exercises (HTML)

1. Predict the output. Will each call work or fail? Why?

```{r}
#| eval: false
greeting <- "Hello!"
with_html(p(greeting))

p <- function() "p"
address <- "123 anywhere street"
with_html(p(address))
```

2. How would you modify `escape()` and `tag()` so that `<script>` tags escape `</script>` instead of `&`, `<`, `>`?

## Solutions (HTML)

1. First call **works**: `p` comes from `html_tags` (data mask has precedence over env), `greeting` is found in the caller env. Second call **errors**: `address` is **also** an HTML tag in `html_tags`. Data mask has precedence, so `address` resolves to the tag function, not the string `"123 anywhere street"`.

```{r}
#| error: true
#| output.lines: 5
greeting <- "Hello!"
with_html(p(greeting))
p <- function() "p"
address <- "123 anywhere street"
with_html(p(address))
```

---

2. Add a `script` argument to `escape()`: when `TRUE`, use `gsub("</script>", "<\\/script>", x)` instead of escaping `&`, `<`, `>`. Pass `!!script` into `tag()` so the generated function calls `escape(..., script = TRUE)`.

# LaTeX DSL

## LaTeX Math Basics and Goal

LaTeX math notation: `x * y`, `x_1` (subscript), `\pi` ($\pi$), `\frac{a}{b}` ($\frac{a}{b}$), `\sqrt{x}` ($\sqrt{x}$).

\medskip

**Goal** --- four translation rules:

| Input | Output | Rule |
|-------|--------|------|
| `pi` | `\pi` | Known symbol |
| `x` | `x` | Unknown symbol (leave as-is) |
| `sqrt(frac(a, b))` | `\sqrt{\frac{a}{b}}` | Known function |
| `f(a)` | `\textrm{f}(a)` | Unknown function |

Unlike the HTML DSL, we need to **walk the AST** to handle unknown symbols and functions.

## \comm{to\_math()} Wrapper

Capture the expression, evaluate in a **custom environment**:

```{r}
#| output.lines: 3
latex <- function(x) structure(x, class = "advr_latex")
print.advr_latex <- function(x)
  cat("<LATEX> ", x, "\n", sep = "")

to_math <- function(x) {
  expr <- enexpr(x)
  out <- eval_bare(expr, latex_env(expr))
  latex(out)
}
```

Key difference from HTML DSL: the evaluation environment **varies** with the input expression (must handle unknown symbols/functions dynamically).

## Known Symbols: Greek Letters

Bind each Greek name to its LaTeX string in an environment:

```{r}
#| output.lines: 3
greek <- c("alpha", "theta", "tau", "beta",
  "pi", "gamma", "phi", "delta", "sigma",
  "omega", "Gamma", "Lambda", "Sigma", "Psi")
  # ... 40 total
greek_list <- set_names(paste0("\\", greek), greek)
greek_env <- as_environment(greek_list)
```

When `eval_bare()` looks up `pi`, it finds `"\\pi"` in `greek_env`.

```{r}
#| output.lines: 3
to_math(pi)
to_math(beta)
```

## Unknown Symbols: Walking the AST

Walk the AST to find all symbol names, create an environment mapping each to its string representation:

```{r}
#| output.lines: 3
all_names_rec <- function(x) {
  switch_expr(x,
    constant = character(),
    symbol = as.character(x),
    call = flat_map_chr(as.list(x[-1]),
      all_names))
}
all_names <- function(x) unique(all_names_rec(x))
all_names(expr(x + y + f(a, b, 10)))
```

Chain: `greek_env` (parent: `symbol_env`) $\to$ Greek has priority over defaults. So `to_math(pi)` gives `\pi`, not `"pi"`.

## Function Factories: \comm{unary\_op} and \comm{binary\_op}

Helpers that manufacture LaTeX operator functions:

```{r}
#| output.lines: 3
unary_op <- function(left, right) {
  new_function(exprs(e1 = ),
    expr(paste0(!!left, e1, !!right)),
    caller_env())
}
binary_op <- function(sep) {
  new_function(exprs(e1 = , e2 = ),
    expr(paste0(e1, !!sep, e2)),
    caller_env())
}
unary_op("\\sqrt{", "}")
```

These are **function factories**: each call returns a new function with the operator baked in via `!!`.

## Known Functions: \comm{f\_env}

Map R operators and functions to LaTeX equivalents:

```{r}
#| eval: false
f_env <- child_env(.parent = empty_env(),
  `+` = binary_op(" + "),
  `-` = binary_op(" - "),
  `*` = binary_op(" * "),
  `/` = binary_op(" / "),
  `^` = binary_op("^"),
  `[` = binary_op("_"),          # subscript
  `(` = unary_op("\\left( ", " \\right)"),
  sqrt = unary_op("\\sqrt{", "}"),
  sin  = unary_op("\\sin(", ")"),
  frac = function(a, b)
    paste0("\\frac{", a, "}{", b, "}"),
  hat = unary_op("\\hat{", "}"))
```

Lexical scoping lets us redefine `+`, `*`, `(`, etc. within this environment.

## Unknown Functions and the Complete DSL

Walk the AST for function names; wrap unknowns in `\textrm{}`:

```{r}
#| output.lines: 3
unknown_op <- function(op) {
  new_function(exprs(... = ), expr({
    contents <- paste(..., collapse = ", ")
    paste0(!!paste0("\\mathrm{", op, "}("),
      contents, ")")
  }))
}
```

The **environment chain** (child $\to$ parent):

`greek_env` $\to$ `symbol_env` $\to$ `f_env` $\to$ `call_env`

```{r}
#| output.lines: 3
to_math(sin(x + pi))
to_math(sin(sin))
to_math(hat(beta))
```

## Exercises (LaTeX)

1. What does `to_math(sin(sin))` return? Trace the environment lookup to explain why.

2. The special characters `\`, `$`, `%` need escaping in LaTeX. Sketch how you'd add escaping using the same S3 pattern as the HTML DSL.

## Solutions (LaTeX)

1. Returns `\sin(sin)`. The function position `sin` matches `sin` in `f_env` $\to$ `unary_op("\\sin(", ")")`. The **argument** `sin` matches the string `"sin"` in `symbol_env` (unknown symbols map to their own name). Functions take precedence because `f_env` is an ancestor of `symbol_env` --- the argument is looked up in `symbol_env` first (closer to `greek_env`), where it finds the string.

2. Create an `advr_latex` S3 class with `escape_latex()` generic: `escape_latex.character()` prepends `\` to special chars, `escape_latex.advr_latex()` returns input unchanged (prevents double-escaping). Apply `escape_latex()` in `latex_env()` when creating the symbol environment and inside `unknown_op()`.

<!--
# Summary

## Key Takeaways

- **DSLs** translate R code to other languages using metaprogramming
- **dbplyr** is the prime example: R → SQL via `translate_sql()`
- **HTML DSL:** S3 escaping class + tag factory via `new_function()` + data mask via `eval_tidy()`
- **Escaping pattern:** S3 class distinguishes raw text from safe output
- **`tag()` factory:** `new_function()` + `!!` generates a function for each HTML tag
- **`with_html()`:** list of tag functions as data mask — `eval_tidy()` trick
- **Data mask precedence:** tag names like `address` shadow env variables
- **LaTeX DSL:** custom evaluation environment chain + AST walking
- **`to_math()`:** `enexpr()` → `eval_bare()` in dynamic environment
- **Environment chain:** `greek_env` → `symbol_env` → `f_env` → `call_env`
- **AST walking** discovers unknown symbols (`all_names`) and functions (`all_calls`)
- **Function factories** `unary_op()` / `binary_op()` manufacture LaTeX operator functions
- Both DSLs combine environments, lexical scoping, quasiquotation, and evaluation
-->
