---
title: "AdvR 01 -- Names and Values"
---

```{r setup, include=FALSE}
source("_common.R")
library(lobstr)
```

# Binding Basics

## Names Have Values, Not the Other Way Around

A common simplification: "the object `x` stores the value `1:3`."

More accurate: **the name `x` is bound to the object `1:3`**.

\medskip

```{r}
#| output.lines: 3
x <- c(1, 2, 3)
```

Think of `<-` as an **arrow** pointing from name to value.

\medskip

**Multiple names** can point to the same object:

```{r}
#| output.lines: 5
y <- x
obj_addr(x)
obj_addr(y)   # same address --- no copy!
```

## Non-Syntactic Names

**Syntactic names** may contain letters, digits, `.` and `_`. Must start with a letter (or `.` not followed by a digit). Cannot be reserved words.

\medskip

**Non-syntactic names** need backticks:

```{r}
#| output.lines: 5
`_abc` <- 1
`_abc`

`if` <- 10
`if`
```

\comm{make.names()} converts to syntactic names:

```{r}
#| output.lines: 3
make.names(c("_abc", "if", "1st"))
```

## Exercises (Binding)

1. Explain the relationship between `a`, `b`, `c`, and `d`:

```{r}
#| eval: false
a <- 1:10
b <- a
c <- b
d <- 1:10
```

2. Do these all point to the same function object? Verify with \comm{obj\_addr()}.

```{r}
#| eval: false
mean
base::mean
get("mean")
evalq(mean)
match.fun("mean")
```

3. Why does \comm{read.csv()} auto-converting non-syntactic names cause problems? What option suppresses it?

4. Why is `.123e1` not a syntactic name?

## Solutions (Binding)

1. `a`, `b`, `c` all point to the **same object**. `d` points to a **different object** with the same value.

```{r}
#| output.lines: 5
a <- 1:10; b <- a; c <- b; d <- 1:10
obj_addrs(list(a, b, c, d))
```

2. Yes, all five expressions return the same object:

```{r}
#| output.lines: 3
obj_addrs(list(mean, base::mean, get("mean"),
  evalq(mean), match.fun("mean"))) |> unique()
```

3. Column names are data; \comm{make.names()} is non-invertible (information loss). Use `check.names = FALSE`.

4. `.123e1` parses as the double `1.23` (dot followed by digit).

# Copy-on-Modify

## The Core Rule

When you modify an object with **multiple bindings**, R creates a copy:

```{r}
#| output.lines: 5
x <- c(1, 2, 3)
y <- x
obj_addr(x)  # both point to same object

y[[3]] <- 4  # modifying y triggers a copy
obj_addr(y)  # y now points to new object
```

`x` is **unchanged** --- the copy was modified, not the original.

## Detecting Copies with `tracemem()`

\comm{tracemem()} prints a message whenever an object is copied:

```{r}
#| eval: false
x <- c(1, 2, 3)
cat(tracemem(x), "\n")
#> <0x7f80c0e0ffc8>

y <- x
y[[3]] <- 4  # triggers copy
#> tracemem[0x7f80c0e0ffc8 -> 0x7f80c4427f40]

untracemem(x)
```

Turn off with \comm{untracemem()}. Useful for finding unexpected copies in performance-sensitive code.

## Copy-on-Modify in Function Calls

Function arguments create **temporary bindings** --- no copy if the object is only read:

```{r}
#| output.lines: 5
f <- function(a) a   # just returns a

x <- c(1, 2, 3)
z <- f(x)
obj_addr(x)
obj_addr(z)   # same address --- no copy
```

Inside `f()`, the parameter `a` binds to the same object as `x`. Since `f()` doesn't modify anything, no copy occurs.

\medskip

**Rule of thumb:** Reading is free. Copying only happens on **modification**.

## Shallow Copies: Lists

Lists use **shallow copies** --- the list structure is copied, but elements still reference the same objects:

```{r}
#| output.lines: 6
l1 <- list(1, 2, 3)
l2 <- l1
l2[[3]] <- 4
```

After modification: `l2` has a new list structure, but `l2[[1]]` and `l2[[2]]` still point to the same values as `l1[[1]]` and `l1[[2]]`.

\medskip

Verify with \comm{ref()}:

```{r}
#| output.lines: 8
ref(l1, l2)
```

## Shallow Copies: Data Frames

Data frames are lists of columns. **Column modification** is efficient:

```{r}
#| eval: false
d1 <- data.frame(x = c(1, 5, 6), y = c(2, 4, 3))
d2 <- d1
d2[, 2] <- d2[, 2] * 2
# Only column 2 is copied; column 1 still shared
```

**Row modification** copies every column:

```{r}
#| eval: false
d3 <- d1
d3[1, ] <- d3[1, ] * 3
# Every column must be copied (all contain the modified row)
```

\medskip

**Implication:** Prefer column-wise operations over row-wise for performance.

## Character Vectors and the String Pool

R uses a **global string pool** --- each unique string is stored once:

```{r}
#| output.lines: 8
x <- c("a", "a", "abc", "d")
ref(x, character = TRUE)
```

Repeated strings point to the same location in the pool. This saves memory for character-heavy data.

## Exercises (Copy-on-Modify)

1. Why is `tracemem(1:10)` not useful?

2. Why does \comm{tracemem()} show two copies here?

```{r}
#| eval: false
x <- c(1L, 2L, 3L)
tracemem(x)
x[[3]] <- 4
```

3. Sketch the relationships:

```{r}
#| eval: false
a <- 1:10
b <- list(a, a)
c <- list(b, a, 1:10)
```

4. What happens here? Why isn't the result circular?

```{r}
#| eval: false
x <- list(1:10)
x[[2]] <- x
```

## Solutions (Copy-on-Modify) --- 1--2

1. `1:10` creates an object not bound to any name. It can't be modified, so tracking copies is pointless.

2. **Two copies** because `4` is a double but `x` is integer. R must:
   - Copy `x` (copy-on-modify, since \comm{tracemem} holds a reference)
   - Coerce integer to double

Avoid the extra copy by assigning `4L` (integer):

```{r}
#| eval: false
x <- c(1L, 2L, 3L)
tracemem(x)
x[[3]] <- 4L   # no type coercion needed
```

## Solutions (Copy-on-Modify) --- 3--4

3. `a`, `b[[1]]`, `b[[2]]`, and `c[[2]]` all reference the **same** object. `c[[3]]` is a **different** object with the same value. `c[[1]]` references `b`.

```{r}
#| output.lines: 8
a <- 1:10; b <- list(a, a); c <- list(b, a, 1:10)
ref(c)
```

4. Not circular because **copy-on-modify** fires. `x[[2]] <- x` copies the old list into a new one. The old list becomes `x[[2]]`; `x` itself is the new copy.

```{r}
#| output.lines: 5
x <- list(1:10)
x[[2]] <- x
ref(x)
```

# Object Size

## Measuring with `obj_size()`

Use \comm{lobstr::obj\_size()}, not \comm{utils::object.size()} (which ignores shared references):

```{r}
#| output.lines: 5
x <- runif(1e6)
obj_size(x)

y <- list(x, x, x)
obj_size(y)          # only ~80 bytes more!
```

All three list elements reference the same vector. The list overhead is just 80 bytes (empty 3-element list).

\medskip

**Combined size** accounts for sharing:

```{r}
#| output.lines: 3
obj_size(x, y)       # same as obj_size(y)
```

## ALTREP and String Pool

**ALTREP** (alternative representation) stores sequences compactly:

```{r}
#| output.lines: 5
obj_size(1:3)
obj_size(1:1e3)
obj_size(1:1e6)
obj_size(1:1e9)      # all the same!
```

Only the start and end are stored --- no matter how long the sequence.

\medskip

**String pool** --- each unique string stored once; repeating is cheap:

```{r}
#| output.lines: 2
obj_size(rep("bananas bananas bananas", 100))
```

## Exercises (Object Size)

1. Why are \comm{object.size(y)} and \comm{obj\_size(y)} so different?

```{r}
#| eval: false
y <- rep(list(runif(1e4)), 100)
object.size(y)   # ~8 MB
obj_size(y)      # ~81 KB
```

2. Why is this list's size misleading?

```{r}
#| eval: false
funs <- list(mean, sd, var)
obj_size(funs)
```

3. Predict the output of each line (see next slide for answer).

## Solutions (Object Size) --- 1--2

1. \comm{object.size()} doesn't account for shared references. All 100 elements point to the same vector, so true size is ~80 KB, not ~8 MB.

2. These functions exist in base R regardless --- they're always loaded. Measuring their "size" doesn't reflect freeable memory.

## Solutions (Object Size) --- 3a

3. Key predictions --- initial state:

```{r}
#| output.lines: 5
a <- runif(1e6)
cat("a:", obj_size(a), "\n")
b <- list(a, a)
cat("b:", obj_size(b), "\n")     # a + 64B list
cat("a,b:", obj_size(a, b), "\n") # same as b (shared)
```

Both list elements share `a` $\to$ `b` is only 64 bytes larger than `a`.

## Solutions (Object Size) --- 3b

After modifications --- sharing breaks:

```{r}
#| output.lines: 5
b[[1]][[1]] <- 10  # copy-on-modify: b[[1]] diverges
cat("b:", obj_size(b), "\n")       # ~16 MB (two copies)
cat("a,b:", obj_size(a, b), "\n")  # b[[2]] still shares a
```

```{r}
#| output.lines: 4
b[[2]][[1]] <- 10  # b[[2]] also diverges from a
cat("b:", obj_size(b), "\n")       # still ~16 MB
cat("a,b:", obj_size(a, b), "\n")  # ~24 MB (no sharing)
```

# Modify-in-Place

## Single Binding Optimization

If an object has only **one binding**, R modifies it in place (no copy):

```{r}
#| eval: false
v <- c(1, 2, 3)
# v is the only name bound to this object

v[[3]] <- 4
# Modified in place --- no copy
```

\medskip

**Complications:**

- R tracks references as: 0, 1, or "many"
- Once "many", it **never** goes back to 1 (even after unbinding)
- Most functions create temporary references, preventing in-place modification
- Exception: **primitive** C functions (like `[[<-`) don't increment

\medskip

**Practical implication:** In-place modification is hard to predict. Focus on algorithmic efficiency instead.

## Environments: Always Modified in Place

Environments use **reference semantics** --- never copied:

```{r}
#| output.lines: 5
e1 <- rlang::env(a = 1, b = 2, c = 3)
e2 <- e1

e1$c <- 4
e2$c           # also 4! same environment
```

Environments can even contain **themselves**:

```{r}
#| output.lines: 5
e <- rlang::env()
e$self <- e
ref(e)
```

This is impossible with lists (copy-on-modify prevents circularity).

## Case Study: Loop Performance

**Problem:** Modifying a data frame in a loop copies it every iteration:

```{r}
#| eval: false
x <- data.frame(matrix(runif(5 * 1e4), ncol = 5))
medians <- vapply(x, median, numeric(1))

for (i in seq_along(medians)) {
  x[[i]] <- x[[i]] - medians[[i]]
}
# Each iteration copies the ENTIRE data frame!
```

**Fix:** Convert to list first (uses internal C code, no extra references):

```{r}
#| eval: false
y <- as.list(x)
for (i in seq_along(medians)) {
  y[[i]] <- y[[i]] - medians[[i]]
}
# Much fewer copies --- list modification is efficient
x <- list2DF(y)
```

Data frame approach is **O(n^2)**; list approach is **O(n)**.

## Exercises (Modify-in-Place)

1. Why doesn't this create a circular list?

```{r}
#| eval: false
x <- list()
x[[1]] <- x
```

2. Wrap the data frame and list median-subtraction methods into functions. Use \comm{bench::mark()} to compare. How does performance change with number of columns?

3. What happens if you use \comm{tracemem()} on an environment?

## Solutions (Modify-in-Place)

1. Copy-on-modify prevents it. `x[[1]] <- x` copies the old (empty) list. `x` now points to a new list containing the old one --- not itself.

```{r}
#| output.lines: 5
x <- list()
x[[1]] <- x
ref(x)   # x and x[[1]] have different addresses
```

2. The data frame approach is O(n^2) in columns (each iteration copies all columns). The list approach is O(n). The list version wins decisively above ~300 columns.

3. Error: environments are not duplicated, so tracing is not useful.

```{r}
#| error: true
#| output.lines: 3
tracemem(new.env())
```

# Unbinding and the Garbage Collector

## How R Frees Memory

When no name points to an object, it becomes **unreachable**:

```{r}
#| eval: false
x <- 1:3
x <- 2:4   # old object (1:3) is now unbound
rm(x)      # new object (2:4) is now unbound
```

The **garbage collector (GC)** periodically frees unreachable objects.

\medskip

**Key facts:**

- GC runs **automatically** when R needs memory
- Traces from global environment + call stack to find reachable objects
- Everything else is deleted
- You rarely need to call \comm{gc()} manually
- \comm{gc()} is useful mainly for its memory usage report

```{r}
#| output.lines: 5
gc()
```

<!--
# Summary

## Key Takeaways

- **Names have values:** `x <- 1:3` binds the name `x` to an object
- **Multiple bindings:** `y <- x` creates a second reference, not a copy
- **Copy-on-modify:** copies happen only when a shared object is modified
- **\comm{tracemem()}:** tracks when objects are copied
- **Shallow copies:** lists/data frames copy structure but share elements
- **\comm{obj\_size()}:** use instead of \comm{object.size()} (accounts for sharing)
- **ALTREP:** sequences like `1:1e9` stored compactly (start + end only)
- **String pool:** repeated strings share memory
- **Single binding:** objects with one name are modified in place (when possible)
- **Environments:** always modified in place (reference semantics)
- **Loop optimization:** convert data frames to lists before looping
- **Garbage collector:** automatically frees unreachable objects
-->
