---
title: "AdvR 05 -- Functions"
---

```{r setup, include=FALSE}
source("_common.R")
```

# Function Fundamentals

## Three Components

Every R function has three parts:

- \comm{formals()}: list of arguments
- \comm{body()}: code inside the function
- \comm{environment()}: where to look up names

```{r}
#| output.lines: 5
f <- function(x, y) { x + y }
formals(f)
body(f)
environment(f)
```

## Primitive Functions

Exception: **primitive functions** call C directly --- all three components are `NULL`:

```{r}
#| output.lines: 5
sum
typeof(sum)
```

```{r}
#| output.lines: 3
formals(sum)
body(sum)
```

Two types: `builtin` (e.g., `sum`) and `special` (e.g., `[`). Only in the base package.

## First-Class Functions and `do.call()`

Functions are **objects** --- no special syntax needed:

```{r}
#| eval: false
f <- function(x) sin(1 / x ^ 2)  # bind to name
lapply(mtcars, function(x) length(unique(x)))  # anonymous
```

R 4.1+ shorthand: `\(x) x + 1` is equivalent to `function(x) x + 1`.

\medskip

**Lists of functions:**

```{r}
#| output.lines: 3
funs <- list(half = \(x) x / 2, double = \(x) x * 2)
funs$double(10)
```

\comm{do.call()} calls a function with arguments from a list:

```{r}
#| output.lines: 3
do.call(mean, list(1:10, na.rm = TRUE))
```

## Exercises (Fundamentals)

1. Given a name like `"mean"`, \comm{match.fun()} finds a function. Given a function, can you find its name? Why doesn't that make sense in R?

2. Which of these is correct? Why?

```{r}
#| eval: false
function(x) 3()
(function(x) 3)()
```

3. What function tells you if an object is a function? What about a primitive function?

4. What are the three important components of a function?

## Solutions (Fundamentals)

1. No one-to-one mapping: a function can have **zero, one, or many** names. Multiple names can bind to the same function; some functions are anonymous.

2. The second: `(function(x) 3)()`. Parentheses separate the anonymous function from the call `()`. The first creates a function with invalid body `3()`.

3. \comm{is.function()} and \comm{is.primitive()}.

4. \comm{formals()} (arguments), \comm{body()} (code), \comm{environment()} (lookup context). Primitives have all three as `NULL`.

# Lexical Scoping

## Four Scoping Rules

R uses **lexical scoping**: looks up values based on where a function is **defined**, not where it is **called**.

\medskip

**1. Name masking** --- inner definitions shadow outer:

```{r}
#| output.lines: 3
x <- 10
g <- function() { x <- 20; x }
g()
```

**2. Functions vs. variables** --- R ignores non-functions in function-call position:

```{r}
#| output.lines: 3
g09 <- function(x) x + 100
g10 <- function() { g09 <- 10; g09(g09) }
g10()
```

## Four Scoping Rules (continued)

**3. Fresh start** --- each call creates a new environment:

```{r}
#| output.lines: 3
g11 <- function() {
  if (!exists("a")) a <- 1
  else a <- a + 1
  a
}
c(g11(), g11())   # both 1
```

**4. Dynamic lookup** --- values looked up at **run time**, not definition time:

```{r}
#| output.lines: 3
g12 <- function() x + 1
x <- 15; g12()
```

Use \comm{codetools::findGlobals()} to detect external dependencies.

## Exercises (Scoping)

1. What does this return? How is each `c` interpreted?

```{r}
#| eval: false
c <- 10
c(c = c)
```

2. What are the four principles governing how R looks for values?

3. Predict the result:

```{r}
#| eval: false
f <- function(x) {
  f <- function(x) {
    f <- function() x ^ 2
    f() + 1
  }
  f(x) * 2
}
f(10)
```

## Solutions (Scoping)

1. Returns `c(c = 10)`. First `c` = the \comm{c()} function; second `c` = a quoted name; third `c` = the value `10`.

2. Name masking, functions vs. variables, fresh start, dynamic lookup.

3. `((10^2) + 1) * 2` = **202**. Innermost `f()` returns `100`, middle adds `1` $\to$ `101`, outer multiplies by `2`.

# Lazy Evaluation

## Arguments Are Evaluated Only When Accessed

```{r}
#| output.lines: 3
h01 <- function(x) 10
h01(stop("This is an error!"))  # no error!
```

Powered by **promises**: expression + environment + cached value. Evaluated at most **once**.

\medskip

**Default arguments** can reference other arguments or variables defined later:

```{r}
#| output.lines: 3
h04 <- function(x = 1, y = x * 2) c(x, y)
h04()
```

Default arguments are evaluated **inside** the function; user-supplied arguments in the **calling** environment.

## Missing Arguments

\comm{missing()} detects whether argument was user-supplied:

```{r}
#| output.lines: 5
h06 <- function(x = 10) {
  list(missing(x), x)
}
str(h06())
str(h06(10))
```

## The `%||%` Operator

The `%||%` operator provides a cleaner pattern for defaults:

```{r}
#| output.lines: 3
`%||%` <- function(lhs, rhs) {
  if (!is.null(lhs)) lhs else rhs
}
NULL %||% 42
```

## Exercises (Lazy Evaluation)

1. What important property of `&&` makes this work?

```{r}
#| eval: false
x_ok <- function(x) {
  !is.null(x) && length(x) == 1 && x > 0
}
```

What happens if you replace `&&` with `&`?

2. What does this return? Which principle?

```{r}
#| eval: false
f2 <- function(x = z) { z <- 100; x }
f2()
```

## Solutions (Lazy Evaluation)

1. `&&` **short-circuits**: if the LHS is `FALSE`, the RHS is never evaluated. This prevents `length(NULL) == 1` from running when `x` is `NULL`. With `&`, all conditions are evaluated (vectorized), causing errors or unexpected results for `NULL` or length > 1 inputs.

2. Returns **100**. Lazy evaluation: `x = z` is not evaluated until `x` is accessed, by which time `z <- 100` has already run inside the function.

# `...` (Dot-Dot-Dot)

## Passing Extra Arguments

`...` lets a function accept and pass on **any number** of extra arguments:

```{r}
#| output.lines: 5
i02 <- function(x, ...) {
  list(x = x, extras = list(...))
}
str(i02(1, a = 2, b = 3))
```

**Two main uses:**

1. Pass arguments to inner functions (e.g., `na.rm` through `lapply`)
2. Allow S3 methods to take different arguments

\medskip

**Downside:** misspelled arguments silently absorbed:

```{r}
#| output.lines: 3
sum(1, 2, NA, na_rm = TRUE)  # na_rm, not na.rm!
```

## Exercises (`...`)

1. Explain these results:

```{r}
#| output.lines: 5
sum(1, 2, 3)
mean(1, 2, 3)
sum(1, 2, 3, na.omit = TRUE)
mean(1, 2, 3, na.omit = TRUE)
```

## Solutions (`...`)

1. \comm{sum()} takes `...` --- all arguments are summed. `na.omit = TRUE` is coerced to `1`, giving `7` instead of `6`.

\medskip

\comm{mean()} has `x`, `trim`, `na.rm`, `...`. Arguments matched by position: `x = 1`, `trim = 2`, `na.rm = 3`. `na.omit` goes to `...` (ignored). Only `x = 1` matters for the mean, so both calls return `1`.

# Exiting a Function

## Return Values and Invisibility

**Implicit return:** last evaluated expression:

```{r}
#| output.lines: 3
j01 <- function(x) if (x < 10) 0 else 10
j01(5)
```

**Explicit return:** via \comm{return()}. Use for early exits only.

\medskip

**Invisible return** --- suppresses auto-printing:

```{r}
#| output.lines: 3
j04 <- function() invisible(1)
j04()           # nothing printed
print(j04())    # force printing
```

`<-` returns invisibly, enabling chained assignment: `a <- b <- c <- 1`.

## Exit Handlers with `on.exit()`

\comm{on.exit()} runs cleanup code **regardless** of how the function exits:

```{r}
#| error: true
#| output.lines: 5
j06 <- function(x) {
  cat("Hello\n")
  on.exit(cat("Goodbye!\n"), add = TRUE)
  if (x) return(10) else stop("Error")
}
j06(TRUE)
```

```{r}
#| error: true
#| output.lines: 3
j06(FALSE)
```

## `on.exit()` Pattern

**Restore global state** (working directory, options, connections):

```{r}
#| eval: false
with_dir <- function(dir, code) {
  old <- setwd(dir)
  on.exit(setwd(old), add = TRUE)
  force(code)
}
```

Always use `add = TRUE` to avoid overwriting previous handlers.

## Exercises (Exiting)

1. What does \comm{load()} return? Why don't you normally see it?

2. Write a function that opens a graphics device, runs code, and closes the device (always, even on error).

## Solutions (Exiting)

1. \comm{load()} invisibly returns a character vector of names of loaded objects. You don't see it because the return is invisible.

2.

```{r}
#| eval: false
plot_pdf <- function(code) {
  pdf("test.pdf")
  on.exit(dev.off(), add = TRUE)
  force(code)
}
```

# Function Forms

## Four Calling Styles

> Everything that exists is an object. Everything that happens is a function call. --- John Chambers

| Form | Example | Prefix equivalent |
|------|---------|-------------------|
| Prefix | `mean(x)` | `mean(x)` |
| Infix | `x + y` | `` `+`(x, y) `` |
| Replacement | `names(x) <- y` | `` `names<-`(x, y) `` |
| Special | `for (i in x) ...` | `` `for`(i, x, ...) `` |

**Everything** can be rewritten in prefix form:

```{r}
#| output.lines: 3
`+`(1, `*`(2, 3))   # 1 + (2 * 3)
```

## Argument Matching and Infix Functions

Arguments are matched by: (1) **exact name**, (2) **partial match**, (3) **position**.

```{r}
#| eval: false
# Confusing but valid:
x <- sample(replace = TRUE, 20, x = c(1:10, NA))
# Clear version:
x <- sample(c(1:10, NA), size = 20, replace = TRUE)
```

\medskip

**User-defined infix functions** use `%name%`:

```{r}
#| output.lines: 3
`%xor%` <- function(a, b) xor(a, b)
TRUE %xor% TRUE
FALSE %xor% TRUE
```

**Replacement functions** must return the modified object:

```{r}
#| eval: false
`random<-` <- function(x, value) {
  x[sample(length(x), 1)] <- value
  x
}
```

## Exercises (Function Forms)

1. Rewrite in prefix form: `1 + 2 + 3`

2. Clarify these scrambled function calls:

```{r}
#| eval: false
x <- sample(replace = TRUE, 20, x = c(1:10, NA))
y <- runif(min = 0, max = 1, 20)
cor(m = "k", y = y, u = "p", x = x)
```

## Solutions (Function Forms)

1. `` `+`(`+`(1, 2), 3) ``

2. Rewritten with explicit matching:

```{r}
#| eval: false
x <- sample(c(1:10, NA), size = 20, replace = TRUE)
y <- runif(20, min = 0, max = 1)
cor(x, y, use = "pairwise.complete.obs",
  method = "kendall")
```

Arguments matched: exact names first, then partial matching (`m` $\to$ `method`, `u` $\to$ `use`), then by position.

<!--
# Summary

## Key Takeaways

- **Three components:** \comm{formals()}, \comm{body()}, \comm{environment()} (all `NULL` for primitives)
- **First-class:** functions are objects; can be anonymous, in lists, passed as arguments
- **\comm{do.call()}:** call a function with args from a list
- **Lexical scoping:** name masking, functions vs. variables, fresh start, dynamic lookup
- **Lazy evaluation:** arguments evaluated only when accessed (promises)
- **Default args:** evaluated inside the function environment
- **`%||%`:** `NULL`-coalescing operator (left if non-`NULL`, else right)
- **`...`:** variadic arguments; enables pass-through but hides typos
- **Invisible return:** \comm{invisible()} suppresses auto-printing
- **\comm{on.exit()}:** cleanup handler; always use `add = TRUE`
- **Four forms:** prefix, infix (`%name%`), replacement (`name<-`), special
- **Everything** can be rewritten in prefix form
-->
