---
title: "AdvR 11 -- R6, S4 & OO Trade-offs"
---

```{r setup, include=FALSE}
source("_common.R")
library(R6)
library(sloop)
```

# R6

## What Is R6?

R6 is **encapsulated OOP** with **reference semantics**:

- Methods belong to **objects**, called with `object$method()`
- Objects are **mutable** --- modified in place, not copied

```{r}
#| output.lines: 3
Accumulator <- R6Class("Accumulator", list(
  sum = 0,
  add = function(x = 1) {
    self$sum <- self$sum + x
    invisible(self)
  }
))
x <- Accumulator$new()
x$add(4)
x$sum
```

One function: \comm{R6::R6Class()}. That's all you need.

## `$initialize()` and `$print()`

\comm{\$initialize()} validates inputs on construction. \comm{\$print()} customises display:

```{r}
#| error: true
#| output.lines: 5
Person <- R6Class("Person", list(
  name = NULL, age = NA,
  initialize = function(name, age = NA) {
    stopifnot(is.character(name), length(name) == 1)
    stopifnot(is.numeric(age), length(age) == 1)
    self$name <- name; self$age <- age
  },
  print = function(...) {
    cat("Person:", self$name,
      "(age", self$age, ")\n")
    invisible(self)
  }
))
hadley <- Person$new("Hadley", age = 38)
hadley
```

## Method Chaining

Side-effect methods return `invisible(self)` $\to$ enables chaining:

```{r}
#| output.lines: 3
x <- Accumulator$new()
x$add(10)$add(20)$add(30)$sum
```

Equivalent to piping in functional style. Readable with line breaks:

```{r}
#| eval: false
x$
  add(10)$
  add(20)$
  sum
```

## Inheritance

Use `inherit` to derive from an existing class. Override methods, access parent with `super$`:

```{r}
#| output.lines: 3
AccumulatorChatty <- R6Class("AccumulatorChatty",
  inherit = Accumulator,
  public = list(
    add = function(x = 1) {
      cat("Adding", x, "\n")
      super$add(x = x)
    }
  )
)
x2 <- AccumulatorChatty$new()
x2$add(10)$add(1)$sum
```

`super$` is analogous to `NextMethod()` in S3.

## Private Fields and Methods

`private` hides internals --- accessible only via `private$`:

```{r}
#| output.lines: 5
Person <- R6Class("Person",
  public = list(
    initialize = function(name, age = NA) {
      private$name <- name; private$age <- age
    },
    print = function(...) {
      cat("Person:", private$name,
        "(age", private$age, ")\n")
      invisible(self)
    }
  ),
  private = list(name = NULL, age = NA)
)
p <- Person$new("Hadley", 38)
p
```

```{r}
#| output.lines: 3
p$name   # NULL -- can't access private field
```

## Active Fields

**Active fields** look like fields but are computed via functions:

```{r}
#| output.lines: 3
Rando <- R6Class("Rando", active = list(
  random = function(value) {
    if (missing(value)) runif(1)
    else stop("Can't set `$random`", call. = FALSE)
  }
))
x <- Rando$new()
c(x$random, x$random, x$random)
```

Useful for **read-only** fields or fields with validation on write.

## Reference Semantics

R6 objects are **not** copied --- all references point to the same object:

```{r}
#| output.lines: 3
x <- Accumulator$new()
y <- x       # y and x are the SAME object
y$add(10)
x$sum         # x is also modified!
```

Use `$clone()` for an independent copy:

```{r}
#| output.lines: 3
x <- Accumulator$new()
y <- x$clone()
y$add(10)
x$sum         # x is unaffected
```

For objects containing other R6 objects, use `$clone(deep = TRUE)`.

## Exercises (R6)

1. Create a bank account R6 class with `$deposit()` and `$withdraw()`. Create a subclass that throws an error on overdraft.

2. Why can't you model a bank account with an S3 class?

3. What base type are R6 objects built on? What attributes do they have?

## Solutions (R6)

1.

```{r}
#| error: true
#| output.lines: 5
BankAccount <- R6Class("BankAccount", list(
  balance = 0,
  deposit  = function(x) {
    self$balance <- self$balance + x; invisible(self)
  },
  withdraw = function(x) {
    self$balance <- self$balance - x; invisible(self)
  }
))
SafeAccount <- R6Class("SafeAccount",
  inherit = BankAccount,
  public = list(
    withdraw = function(x) {
      if (self$balance - x < 0) stop("Overdraft!")
      super$withdraw(x)
    }
  )
)
sa <- SafeAccount$new(); sa$deposit(100)$withdraw(150)
```

## Solutions (R6) --- continued

2. S3 uses copy-on-modify: `$withdraw()` must return a new object and the caller must reassign (`s <- withdraw(s, 50)`). This is awkward ("threading state"). R6 modifies in place, so `s$withdraw(50)` just works.

3. R6 objects are **environments** (base type `environment`) with a `class` attribute containing the R6 class hierarchy:

```{r}
#| output.lines: 3
x <- Person$new("Test", 30)
typeof(x)
class(x)
```

# S4

## S4 Overview

S4 is **formal functional OOP** --- like S3 but stricter:

```{r}
#| output.lines: 5
methods::setClass("S4Person",
  slots = c(name = "character", age = "numeric")
)
john <- methods::new("S4Person",
  name = "John", age = 30)
john@name   # @ accesses slots (like $ for S3)
```

Three key functions: \comm{setClass()}, \comm{setGeneric()}, \comm{setMethod()}.

## S4 Generics and Methods

```{r}
#| include: false
setGeneric <- function(...) invisible(methods::setGeneric(...))
setMethod <- function(...) invisible(methods::setMethod(...))
```

```{r}
#| output.lines: 5
setGeneric("greet", function(x) {
  standardGeneric("greet")
})
setMethod("greet", "S4Person", function(x) {
  paste0("Hello, ", x@name, "!")
})
greet(john)
```

S4 supports **multiple inheritance** and **multiple dispatch** (dispatch on multiple arguments).

## S4 Key Differences from S3

| Feature | S3 | S4 |
|---------|----|----|
| Define class | `structure()` + attribute | `setClass()` with slots |
| Access fields | `$` | `@` |
| Create generic | `UseMethod()` | `setGeneric()` + `standardGeneric()` |
| Define method | `generic.class()` | `setMethod()` |
| Inheritance | Class vector | `contains` argument |
| Multiple dispatch | No | Yes |
| Validation | Manual | `setValidity()` |

\medskip

S4 is used extensively by **Bioconductor**.

# OO Trade-offs

## S4 vs. S3

**S3:** simple, flexible, good for individuals and small teams.

**S4:** formal, strict, good for large teams and complex systems.

\medskip

| | S3 | S4 |
|---|---|---|
| Learning curve | Low | High |
| Upfront design | Minimal | Required |
| Guarantees | Conventions | Enforced |
| Best for | Most R code | Bioconductor, Matrix |

\medskip

**Default to S3** unless you have a compelling reason for S4.

## R6 vs. S3: Namespacing

**S3 generics** live in the **global** namespace $\to$ naming conflicts:

```{r}
#| eval: false
plot(data)       # plot some data
plot(bank_heist) # plot a crime?
```

**R6 methods** live in the **object** $\to$ no conflict:

```{r}
#| eval: false
data$plot()
bank_heist$plot()
```

R6 methods can also have more specific argument names (no need for `x`, `...`).

## R6 vs. S3: Threading State

**S3 problem:** returning a value AND modifying an object:

```{r}
#| eval: false
# S3 stack: awkward
out <- pop(s)    # returns list(item, stack)
s <- out$x       # must reassign!
```

**R6 solution:** modify in place, return only the value:

```{r}
#| eval: false
# R6 stack: natural
value <- s$pop() # modifies s AND returns value
```

Reference semantics solve "threading state" --- but make code harder to reason about.

## When to Use Which System

| Scenario | Recommended |
|----------|-------------|
| Extending base R / tidyverse | **S3** |
| Most R packages | **S3** |
| Mutable state (connections, caches) | **R6** |
| Web API wrappers, databases | **R6** |
| Bioconductor, complex hierarchies | **S4** |

\medskip

**General rule:** start with S3. Use R6 when you need mutability. Use S4 when formality pays off.

## Exercises (Trade-offs)

1. How does method chaining in R6 relate to the pipe (`|>`) in S3?

2. Name two advantages of S3's global namespace for generics.

## Solutions (Trade-offs)

1. Both allow sequential operations. **Method chaining** (`x$add(1)$add(2)$sum`) reads left-to-right like a pipe. **Pipe** (`x |> add(1) |> add(2)`) does the same for functional code. The key difference: chaining requires `invisible(self)` returns; piping works with any function that returns a new object.

2. (a) **Uniform API:** `summary()`, `print()`, `predict()` work on any object --- users learn one verb for many types. (b) **Discoverability:** tab-completion on a generic shows all available methods; with R6, you must know the object's class first.

<!--
# Summary

## Key Takeaways

- **R6** = encapsulated OOP + reference semantics (modified in place)
- **R6Class()** is the only function you need: `public`, `private`, `active`
- **`$initialize()`** validates on creation; **`$print()`** customises display
- **Method chaining:** return `invisible(self)` from side-effect methods
- **Inheritance:** `inherit` argument + `super$` for parent delegation
- **Private fields:** hidden from users; accessed via `private$`
- **Active fields:** computed properties (read-only, validation)
- **Reference semantics:** `y <- x` shares the object; use `$clone()` for copies
- **S4** = formal functional OOP: `setClass()`, `setGeneric()`, `setMethod()`
- **S4** supports slots (`@`), multiple inheritance, multiple dispatch, `setValidity()`
- **S4 vs S3:** more guarantees but steeper learning curve
- **R6 vs S3:** local namespacing, solves threading state, but reference semantics add complexity
- **Default to S3;** R6 for mutability; S4 for formality
-->
