---
title: "AdvR 17 -- Rcpp"
---

```{r setup, include=FALSE}
source("_common.R")
```

# Getting Started

## Why Rcpp?

R is slow at three types of tasks:

1. **Recursive problems** --- each step depends on the previous (MCMC, dynamic programming)
2. **Tight loops** --- millions of iterations that can't be vectorised
3. **Advanced data structures** --- balanced trees, hash maps beyond R's built-in types

\medskip

**Rcpp** lets you write C++ that plugs into R seamlessly:

- Typical speedups: **10--1000x** for loop-heavy code
- Handles R types (vectors, lists, data frames) natively
- Auto-generates R wrapper functions

\medskip

**Rule of thumb:** optimise R first (vectorise, avoid copies). Use C++ only for remaining bottlenecks.

## R vs C++ Syntax

| Feature | R | C++ |
|---------|---|-----|
| Assignment | `x <- 1` | `double x = 1;` |
| Indexing | `x[1]` (1-based) | `x[0]` (0-based) |
| Scalar types | just `numeric` | `int`, `double`, `bool`, `String` |
| Vectors | `c(1, 2, 3)` | `NumericVector x = {1, 2, 3};` |
| If statement | `if (x > 0) {}` | `if (x > 0) {}` (same) |
| For loop | `for (i in 1:n) {}` | `for (int i = 0; i < n; i++) {}` |
| Function | `f <- function(x) {}` | `double f(double x) {}` |
| Return | implicit last value | explicit `return` |
| Line ending | newline | **semicolon** `;` |
| Comments | `# comment` | `// comment` |

Every C++ statement **must** end with `;`. Forgetting it is the #1 beginner error.

## \comm{cppFunction()}: Inline C++

Write C++ directly in R with \comm{Rcpp::cppFunction()}:

```{r}
#| eval: false
library(Rcpp)
cppFunction('
  int one() {
    return 1;
  }
')
one()
#> [1] 1
```

\medskip

Key differences from R equivalent (`one <- function() 1L`):

- Must **declare** return type (`int`)
- Must use explicit `return`
- Every statement ends with `;`
- Scalars are typed: `int`, `double`, `bool`, `String`

## Example: \comm{signR()} --- Scalar I/O

**Scalar input/output** --- sign function:

```{r}
#| eval: false
cppFunction('
  int signR(int x) {
    if (x > 0) return 1;
    else if (x == 0) return 0;
    else return -1;
  }
')
signR(-5)
#> [1] -1
```

\medskip

Note how similar this is to the R version --- main differences are type declarations (`int`), semicolons, and explicit `return`.

## Example: \comm{sumC()} --- Vector Input

**Vector input, scalar output** --- sum:

```{r}
#| eval: false
cppFunction('
  double sumC(NumericVector x) {
    int n = x.size();
    double total = 0;
    for (int i = 0; i < n; ++i)
      total += x[i];
    return total;
  }
')
sumC(1:10)
#> [1] 55
```

**Pattern:** `NumericVector` accepts R vectors. Use `.size()` for length, `x[i]` for 0-based access.

## Example: \comm{pdistC()} --- Vector Output

Euclidean distance from each element to a point:

```{r}
#| eval: false
cppFunction('
  NumericVector pdistC(double x,
                       NumericVector ys) {
    int n = ys.size();
    NumericVector out(n);  // pre-allocate
    for (int i = 0; i < n; ++i)
      out[i] = sqrt(pow(ys[i] - x, 2.0));
    return out;
  }
')
pdistC(5, 1:10)
#> [1] 4 3 2 1 0 1 2 3 4 5
```

**Pattern:** create output vector with `NumericVector out(n)`, fill in loop, return.

\medskip

C++ equivalent of R's `abs(ys - x)` --- same result, but the C++ loop runs **much** faster for large vectors.

## \comm{sourceCpp()}: Standalone Files

For real projects, put C++ in a `.cpp` file:

```cpp
// mean_c.cpp
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
double meanC(NumericVector x) {
  int n = x.size();
  double total = 0;
  for (int i = 0; i < n; ++i)
    total += x[i];
  return total / n;
}
```

```{r}
#| eval: false
sourceCpp("mean_c.cpp")
meanC(1:10)
#> [1] 5.5
```

**Key ingredients:** `#include <Rcpp.h>`, `using namespace Rcpp;`, and `// [[Rcpp::export]]` before each function to expose to R.

## Exercises (Getting Started)

1. With no arguments, `mean.default(x)` vs `meanC(x)` --- which is faster for `x <- runif(1e5)`? Why?

2. Convert this R function to C++:

```{r}
#| eval: false
f4 <- function(a, b) {
  if (a > b) a else b   # max of two values
}
```

## Solutions (Getting Started)

1. `meanC()` is faster: skips S3 dispatch and `na.rm` checking. `mean.default()` makes two passes (for numerical accuracy); `meanC()` makes one.

2.

```{r}
#| eval: false
cppFunction('
  double f4C(double a, double b) {
    if (a > b) return a;
    else return b;
  }
')
f4C(3.5, 2.1)
#> [1] 3.5
```

# Rcpp in Practice

## Rcpp Types

| R type | Rcpp scalar | Rcpp vector | Rcpp matrix |
|--------|-------------|-------------|-------------|
| `logical` | `bool` | `LogicalVector` | `LogicalMatrix` |
| `integer` | `int` | `IntegerVector` | `IntegerMatrix` |
| `double` | `double` | `NumericVector` | `NumericMatrix` |
| `character` | `String` | `CharacterVector` | `CharacterMatrix` |
| `list` | --- | `List` | --- |
| `data.frame` | --- | `DataFrame` | --- |

\medskip

Vectors have useful methods: `.size()`, `.push_back()`, `.begin()`, `.end()`.

Use `as<type>()` to convert R objects: `as<double>(x)`.

## Missing Values

**Scalars:** R's `NA` maps to type-specific sentinels in C++:

```{r}
#| eval: false
# Use these constants in C++:
#   NA_REAL      (double)
#   NA_INTEGER   (int)
#   NA_LOGICAL   (int)
#   NA_STRING    (String)
```

\medskip

**Vectors:** use `NumericVector::is_na()`:

```{r}
#| eval: false
cppFunction('
  LogicalVector is_naC(NumericVector x) {
    int n = x.size();
    LogicalVector out(n);
    for (int i = 0; i < n; ++i)
      out[i] = NumericVector::is_na(x[i]);
    return out;
  }
')
is_naC(c(1, NA, 3))
#> [1] FALSE  TRUE FALSE
```

## Missing Values: Gotchas

**`int` has no real NA** --- `NA_INTEGER` is just a special int value:

```{r}
#| eval: false
cppFunction('
  int missing_sampler() { return NA_INTEGER; }
')
missing_sampler()
#> [1] NA
```

\medskip

**Danger with `double`:** `NA_REAL` is `NaN`, so C++ math propagates correctly.
But `int` NA doesn't propagate --- silent wrong answers:

```{r}
#| eval: false
# In C++: NA_INTEGER + 1 = -2147483647 (not NA!)
```

**Best practice:** use `NumericVector` and `double` when NAs are possible. Check with `is_na()` before arithmetic.

## Rcpp Sugar

**Rcpp sugar** provides vectorised C++ equivalents of common R functions:

- **Arithmetic:** `+`, `-`, `*`, `/`, `pow`, `sqrt`, `abs`
- **Logical:** `<`, `>`, `==`, `!=`, `!`, `&`, `|`
- **Summary:** `sum()`, `mean()`, `min()`, `max()`, `range()`
- **Finding values:** `match()`, `duplicated()`, `which_min()`, `which_max()`
- **d/p/q/r** for distributions: `dnorm()`, `pnorm()`, `rnorm()`, etc.

\medskip

Sugar lets you write **vectorised** C++ that looks like R:

```{r}
#| eval: false
cppFunction('
  NumericVector pdistSugar(double x,
                           NumericVector ys) {
    return abs(ys - x);  // Rcpp sugar!
  }
')
```

Same result as the manual `pdistC()` loop, but more readable.

## Exercises (Practice)

1. Rewrite `cumsum()` in C++ as `cumsumC()`. What happens when given `NA` input?

2. Write an `any_naC()` function that returns `TRUE` if a `NumericVector` contains any `NA`.

## Solutions (Practice) --- 1

1.

```{r}
#| eval: false
cppFunction('
  NumericVector cumsumC(NumericVector x) {
    int n = x.size();
    NumericVector out(n);
    out[0] = x[0];
    for (int i = 1; i < n; ++i)
      out[i] = out[i - 1] + x[i];
    return out;
  }
')
cumsumC(1:5)
#> [1]  1  3  6 10 15
```

With `NA` input, the `NA` (as `NaN`) propagates through all subsequent sums --- same behavior as R's `cumsum()`.

## Solutions (Practice) --- 2

2.

```{r}
#| eval: false
cppFunction('
  bool any_naC(NumericVector x) {
    int n = x.size();
    for (int i = 0; i < n; ++i)
      if (NumericVector::is_na(x[i])) return true;
    return false;
  }
')
any_naC(c(1, 2, NA))
#> [1] TRUE
```

# STL

## The Standard Template Library

The **STL** provides battle-tested C++ data structures and algorithms:

- **Iterators** --- generic pointers for traversing containers
- **Algorithms** --- `std::sort`, `std::accumulate`, `std::unique`, ...
- **Data structures** --- `vector`, `set`, `map`, `unordered_map`, ...

\medskip

**Why use STL with Rcpp?**

- R lacks efficient ordered sets, hash maps, priority queues
- STL algorithms often faster than hand-written loops
- Rcpp converts STL containers $\leftrightarrow$ R types automatically

## Iterators and Algorithms

**Iterators** point to container elements: `.begin()` $\to$ first, `.end()` $\to$ past-the-last.

```{r}
#| eval: false
cppFunction('
  double sum_stl(NumericVector x) {
    return std::accumulate(x.begin(),
                           x.end(), 0.0);
  }
')
sum_stl(1:10)
#> [1] 55
```

\medskip

Useful algorithms:

| Algorithm | Does |
|-----------|------|
| `std::accumulate(b, e, init)` | Sum/fold |
| `std::sort(b, e)` | In-place sort |
| `std::upper_bound(b, e, val)` | Binary search |
| `std::unique(b, e)` | Remove adjacent duplicates |
| `std::distance(b, it)` | Position of iterator |

## STL Data Structures

| Structure | R equivalent | Use case |
|-----------|-------------|----------|
| `vector<T>` | `c()` | Growable array |
| `set<T>` | `unique(sort(x))` | Unique sorted elements |
| `unordered_set<T>` | (no direct) | Fast membership testing |
| `map<K,V>` | `names(x)` | Key-value (sorted) |
| `unordered_map<K,V>` | environment | Key-value (hashed, fast) |

\medskip

**Example:** `set` for unique sorted values:

```{r}
#| eval: false
cppFunction('
  NumericVector uniqueC(NumericVector x) {
    std::set<double> seen;
    for (int i = 0; i < x.size(); ++i)
      seen.insert(x[i]);
    return Rcpp::wrap(
      std::vector<double>(seen.begin(),
                          seen.end()));
  }
')
```

## Case Study: \comm{tableC()} with \comm{unordered\_map}

Reimplement `table()` using a hash map:

```{r}
#| eval: false
cppFunction('
  std::map<int, int> tableC(IntegerVector x) {
    std::map<int, int> counts;
    for (int i = 0; i < x.size(); ++i)
      counts[x[i]]++;
    return counts;
  }
')
tableC(c(1L, 2L, 1L, 3L, 2L, 1L))
#> 1 2 3
#> 3 2 1
```

\medskip

**Pattern:** `counts[key]++` auto-initialises to 0 on first access.

For large inputs, `unordered_map` is $O(1)$ lookup vs `map`'s $O(\log n)$.

## Case Study: \comm{rleC()}

Run-length encoding --- classic loop that R can't vectorise:

```{r}
#| eval: false
cppFunction('
  List rleC(NumericVector x) {
    int n = x.size();
    std::vector<int> lengths;
    std::vector<double> values;
    int i = 0;
    while (i < n) {
      int j = i + 1;
      while (j < n && x[j] == x[i]) j++;
      lengths.push_back(j - i);
      values.push_back(x[i]);
      i = j;
    }
    return List::create(
      _["lengths"] = lengths,
      _["values"]  = values);
  }
')
rleC(c(1, 1, 2, 2, 2, 3))
```

## Exercises (STL)

1. `median()` is slow for large vectors. Implement a `medianC()` using `std::nth_element()` (partial sort, $O(n)$ instead of full $O(n \log n)$ sort).

2. Implement `duplicatedC()` that returns a `LogicalVector` indicating which elements have been seen before. Use `unordered_set`.

## Solutions (STL) --- 1

1.

```{r}
#| eval: false
cppFunction('
  double medianC(NumericVector x) {
    int n = x.size();
    NumericVector y = clone(x);  // dont modify input
    std::nth_element(y.begin(),
      y.begin() + n/2, y.end());
    if (n % 2 == 1) return y[n/2];
    std::nth_element(y.begin(),
      y.begin() + n/2 - 1, y.end());
    return (y[n/2 - 1] + y[n/2]) / 2.0;
  }
')
```

## Solutions (STL) --- 2

2.

```{r}
#| eval: false
cppFunction('
  LogicalVector duplicatedC(IntegerVector x) {
    int n = x.size();
    LogicalVector out(n);
    std::unordered_set<int> seen;
    for (int i = 0; i < n; ++i) {
      if (seen.count(x[i])) out[i] = true;
      else { out[i] = false; seen.insert(x[i]); }
    }
    return out;
  }
')
duplicatedC(c(1L, 2L, 1L, 3L, 2L))
#> [1] FALSE FALSE  TRUE FALSE  TRUE
```

<!--
# Summary

## Key Takeaways

- **Rcpp** bridges R and C++ for loop-heavy, recursive, or data-structure bottlenecks
- **`cppFunction()`** for inline C++; **`sourceCpp()`** for `.cpp` files
- C++ requires **type declarations**, **semicolons**, **0-based indexing**, explicit `return`
- Rcpp types mirror R: `NumericVector`, `IntegerVector`, `List`, `DataFrame`
- **Missing values:** use `NA_REAL`/`NA_INTEGER`; beware `int` NA doesn't propagate
- **Rcpp sugar** provides vectorised C++ equivalents of R functions
- **STL iterators** + algorithms: `accumulate`, `sort`, `nth_element`, `upper_bound`
- **STL data structures:** `vector` (growable), `set` (unique sorted), `map`/`unordered_map` (key-value)
- **Pattern:** `counts[key]++` auto-initialises in maps
- **Rule of thumb:** optimise R first, then C++ for remaining bottlenecks
-->
