---
title: "AdvR 16 -- Debugging & Performance"
---

```{r setup, include=FALSE}
source("_common.R")
library(bench)
```

# Debugging

## Debugging Strategy

> Finding your bug is a process of confirming the many things that you believe are true --- until you find one which is not true.
> --- Norm Matloff

\medskip

Four-step approach:

1. **Google!** --- search the error message (remove variable names/values)
2. **Make it repeatable** --- minimal reproducible example, simplify data
3. **Figure out where** --- use `traceback()`, `browser()`, systematic experiments
4. **Fix it and test it** --- write a test case to prevent regression

## \comm{traceback()} and Call Stacks

\comm{traceback()} shows the sequence of calls leading to the error:

```{r}
#| eval: false
f <- function(a) g(a)
g <- function(b) h(b)
h <- function(c) i(c)
i <- function(d) {
  if (!is.numeric(d)) stop("`d` must be numeric")
  d + 10
}
f("a")
#> Error: `d` must be numeric
traceback()
#> 4: i(c) at #1
#> 3: h(b) at #1
#> 2: g(a) at #1
#> 1: f("a")
```

Read **bottom $\to$ top**: `f()` $\to$ `g()` $\to$ `h()` $\to$ `i()` $\to$ error. Use \comm{rlang::last\_trace()} for tree-structured output.

## Interactive Debugger: \comm{browser()}

Insert \comm{browser()} to **pause execution** and explore interactively:

```{r}
#| eval: false
g <- function(b) {
  browser()    # execution pauses here
  h(b)
}
```

**Commands** inside the debugger:

| Command | Action |
|---------|--------|
| `n` (Next) | Execute next step |
| `s` (Step into) | Step into the next function call |
| `f` (Finish) | Finish current loop or function |
| `c` (Continue) | Resume normal execution |
| `Q` (Quit) | Stop debugging, return to global |

Conditional: `if (b < 0) browser()`.

## Alternatives and Non-interactive Debugging

**Alternatives to `browser()`:**

- **Breakpoints** --- click left of line number in RStudio
- `options(error = recover)` --- choose any frame to inspect
- `debug(f)` / `debugonce(f)` --- auto-pause on function entry

\medskip

**Non-interactive debugging** (batch jobs, pipelines):

- `dump.frames(to.file = TRUE)` $\to$ load + `debugger()` later
- **Print debugging** --- insert `cat()` / `print()` statements
- Unexpected warnings $\to$ `options(warn = 2)` (converts to errors)
- Unexpected messages $\to$ `rlang::with_abort(f(), "message")`
- R crashes $\to$ bug in compiled (C/C++) code

# Measuring Performance

## Profiling with \comm{profvis}

**Profile** code to find bottlenecks --- don't guess:

```{r}
#| eval: false
library(profvis)
profvis({
  f <- function() { pause(0.1); g(); h() }
  g <- function() { pause(0.1); h() }
  h <- function() pause(0.1)
  f()
})
```

profvis output has two panes:

- **Top**: source code + time/memory bars per line
- **Bottom**: **flame graph** --- full call stack over time

`<GC>` in the flame graph = garbage collector running $\to$ sign of many short-lived objects (e.g., growing a vector in a loop).

## Microbenchmarking: \comm{bench::mark()}

Compare small code snippets precisely:

```{r}
#| output.lines: 5
x <- runif(100)
bench::mark(
  sqrt(x),
  x ^ 0.5
)[c("expression", "min", "median", "itr/sec")]
```

Auto-checks all expressions return the **same value** (set `check = FALSE` to override).

\medskip

| Time | Calls per second |
|------|-----------------|
| 1 ms | 1,000 |
| 1 $\mu$s | 1,000,000 |
| 1 ns | 1,000,000,000 |

## Exercises (Measuring)

1. Which is fastest? Predict, then benchmark:

```{r}
#| eval: false
x <- runif(100)
sqrt(x);  x ^ 0.5;  x ^ (1/2);  exp(log(x) / 2)
```

2. How do `system.time()` estimates compare to `bench::mark()`? Why might they differ?

## Solutions (Measuring)

1.

```{r}
#| output.lines: 5
x <- runif(100)
bench::mark(
  sqrt(x), x ^ 0.5,
  x ^ (1/2), exp(log(x) / 2)
)[c("expression", "min", "median", "itr/sec")]
```

`sqrt()` is fastest (dedicated C function). `exp(log(x)/2)` is slowest (two function calls).

2. `system.time()` has ~ms precision; `bench::mark()` uses a high-precision timer (~ns). `system.time()` also includes process overhead and can't isolate individual expressions.

# Improving Performance

## Strategy: Code Organisation

1. Wrap each approach in a **function**
2. Generate a **representative test case** (big enough to matter, small enough to iterate)
3. Benchmark with \comm{bench::mark()} --- auto-checks correctness

```{r}
#| output.lines: 3
x <- runif(1e5)
mean1 <- function(x) mean(x)
mean2 <- function(x) sum(x) / length(x)
bench::mark(mean1(x), mean2(x)
  )[c("expression", "min", "median")]
```

`sum(x)/length(x)` is faster --- `mean()` makes two passes for numerical accuracy.

## Do Less Work

Use **specialised** functions tailored to your input:

- `rowSums()` / `colMeans()` instead of `apply()`
- `vapply()` instead of `sapply()` (pre-specifies output)
- `any(x == 10)` instead of `10 %in% x`
- `readr::read_csv()` instead of `read.csv()`

\medskip

Skip **method dispatch** in tight loops:

```{r}
#| output.lines: 3
x <- runif(1e2)
bench::mark(
  mean(x), mean.default(x)
)[c("expression", "min", "median", "itr/sec")]
```

Trade-off: `mean.default()` is faster but fails if `x` isn't numeric.

## Vectorise

**Vectorisation** = whole-object approach; loops run in C, not R.

```{r}
#| output.lines: 3
x <- matrix(runif(1e5), nrow = 1000)
bench::mark(
  apply(x, 1, sum), rowSums(x)
)[c("expression", "min", "median", "itr/sec")]
```

Key vectorised functions: `rowSums()`, `colMeans()`, `cumsum()`, `diff()`, `pmin()`, `pmax()`, `crossprod()`.

\medskip

Vectorised subsetting is also fast: `x[is.na(x)] <- 0` replaces all NAs in one step.

## Avoid Copies

**Growing objects in loops** is $O(n^2)$ --- each iteration copies everything:

```{r}
#| output.lines: 3
collapse <- function(xs) {
  out <- ""
  for (x in xs) out <- paste0(out, x)
  out
}
s <- replicate(100, paste(sample(letters, 50,
  replace = TRUE), collapse = ""))
bench::mark(
  loop = collapse(s),
  vectorised = paste(s, collapse = ""),
  check = FALSE
)[c("expression", "min", "median")]
```

**Rule:** pre-allocate output or use a vectorised function.

## Case Study: Faster t-test

1000 experiments $\times$ 50 individuals. Step 1 --- **do less work**:

```{r}
#| output.lines: 3
m <- 1000; n <- 50
X <- matrix(rnorm(m * n, 10, 3), nrow = m)
grp <- rep(1:2, each = n / 2)

my_t <- function(x, grp) {
  t_stat <- function(x) {
    m <- mean(x); n <- length(x)
    list(m = m, n = n,
      var = sum((x - m)^2) / (n - 1))
  }
  g1 <- t_stat(x[grp == 1])
  g2 <- t_stat(x[grp == 2])
  (g1$m - g2$m) /
    sqrt(g1$var/g1$n + g2$var/g2$n)
}
```

Strips formatting, p-values, etc. from `t.test()` --- ~6x faster.

## Case Study: Vectorised t-test

Step 2 --- **vectorise**: `mean()` $\to$ `rowMeans()`, `sum()` $\to$ `rowSums()`:

```{r}
#| output.lines: 3
rowtstat <- function(X, grp) {
  t_stat <- function(X) {
    m <- rowMeans(X); n <- ncol(X)
    var <- rowSums((X - m)^2) / (n - 1)
    list(m = m, n = n, var = var)
  }
  g1 <- t_stat(X[, grp == 1])
  g2 <- t_stat(X[, grp == 2])
  se <- sqrt(g1$var/g1$n + g2$var/g2$n)
  (g1$m - g2$m) / se
}
bench::mark(
  loop = purrr::map_dbl(1:m, ~my_t(X[.,], grp)),
  vectorised = rowtstat(X, grp)
)[c("expression", "min", "median")]
```

Vectorised version: ~100x faster than loop, ~1000x faster than `t.test()`.

## Exercises (Improving)

1. Compare `apply(x, 1, sum)` vs `rowSums(x)` for a 10000 $\times$ 100 matrix. How large is the speedup?

2. What's the difference between `rowSums()` and `.rowSums()`?

## Solutions (Improving)

1.

```{r}
#| output.lines: 3
x <- matrix(runif(1e6), nrow = 10000)
bench::mark(
  apply(x, 1, sum), rowSums(x)
)[c("expression", "min", "median", "itr/sec")]
```

`rowSums()` is ~10--20x faster: implemented in C, avoids R-level loop.

2. `.rowSums(x, m, n)` is the **internal** function --- skips dimension checks and dispatching. Slightly faster but requires you to pass `nrow` and `ncol` explicitly. Use `rowSums()` unless profiling shows it's a bottleneck.

<!--
# Summary

## Key Takeaways

- **Debugging strategy:** Google → reproduce → locate → fix & test
- **`traceback()`** shows the call stack leading to an error (read bottom → top)
- **`browser()`** pauses execution for interactive exploration (n/s/f/c/Q)
- **Alternatives:** breakpoints (RStudio), `recover()`, `debug()`
- **Non-interactive:** `dump.frames()`, print debugging, convert warnings to errors
- **Profiling** with `profvis`: flame graphs show call stacks over time
- **`<GC>`** in profiles = garbage collector = too many short-lived objects
- **`bench::mark()`** for precise microbenchmarking (auto-checks equality)
- **Strategy:** wrap in functions, representative test case, benchmark
- **Do less:** specialised functions, skip method dispatch
- **Vectorise:** whole-object approach, loops in C (rowSums, colMeans, etc.)
- **Avoid copies:** pre-allocate or use vectorised functions; growing objects is O(n²)
- **Case study:** t-test ~1000x faster by doing less work + vectorising
-->
