---
title: "AdvR 12 -- Metaprogramming & Expressions"
---

```{r setup, include=FALSE}
source("_common.R")
library(rlang)
library(lobstr)
```

# Big Picture

## Why Metaprogramming?

**Metaprogramming** = code that inspects, modifies, or generates **other code**.

R exposes this more than most languages:

- `dplyr::filter(df, x > 5)` --- `x > 5` is **captured**, not evaluated
- `ggplot2::aes(x = mpg, y = hp)` --- column names are **quoted**, not looked up
- `~` formulas --- store expressions + environments

\medskip

Three pillars: **expressions** (code as data), **quasiquotation** (code templates), **evaluation** (running captured code).

## Code Is Data

\comm{rlang::expr()} captures code without evaluating it:

```{r}
#| output.lines: 3
expr(mean(x, na.rm = TRUE))
expr(10 + 100 + 1000)
```

Inside a function, use \comm{enexpr()} to capture the **caller's** code:

```{r}
#| output.lines: 3
capture_it <- function(x) enexpr(x)
capture_it(a + b + c)
```

Captured code is called an **expression** (call, symbol, constant, or pairlist).

## Modifying Expressions

Expressions behave like **lists** --- use `[[` and `$`:

```{r}
#| output.lines: 5
f <- expr(f(x = 1, y = 2))
f$z <- 3
f
f[[2]] <- NULL
f
```

The first element is the function; arguments start at position 2.

## Code Is a Tree (AST)

Code has a **tree** structure --- the **abstract syntax tree**:

```{r}
#| output.lines: 3
lobstr::ast(f(a, "b"))
```

Nested calls create deeper trees. Operators are also calls:

```{r}
#| output.lines: 5
lobstr::ast(1 + 2 * 3)
```

## Code Can Generate Code

\comm{rlang::call2()} builds calls from components:

```{r}
#| output.lines: 3
call2("+", 1, call2("*", 2, 3))
```

\comm{!!} (bang-bang) **unquotes** --- inserts a stored expression into a template:

```{r}
#| output.lines: 3
xx <- expr(x + x)
yy <- expr(y + y)
expr(!!xx / !!yy)
```

Wrap in a function for reusable code generators:

```{r}
#| output.lines: 3
cv <- function(var) {
  var <- enexpr(var)
  expr(sd(!!var) / mean(!!var))
}
cv(x + y)
```

## Evaluation Runs Code

\comm{base::eval()} executes an expression in an environment:

```{r}
#| output.lines: 3
eval(expr(x + y), env(x = 1, y = 10))
eval(expr(x + y), env(x = 2, y = 100))
```

Two powerful uses:

1. **Override functions** (DSLs)
2. **Data masks** (refer to columns as variables)

## Custom Evaluation: Functions

Bind **functions** in the evaluation environment to change behaviour:

```{r}
#| output.lines: 3
string_math <- function(x) {
  e <- env(caller_env(),
    `+` = function(x, y) paste0(x, y),
    `*` = function(x, y) strrep(x, y))
  eval(enexpr(x), e)
}
name <- "Hadley"
string_math("Hello " + name)
string_math(("x" * 2 + "-y") * 3)
```

This is how `dbplyr` translates R to SQL.

## Custom Evaluation: Data Masks

\comm{rlang::eval\_tidy()} adds a **data mask** --- columns act like variables:

```{r}
#| output.lines: 3
df <- data.frame(x = 1:5, y = sample(5))
eval_tidy(expr(x + y), df)
```

Wrap it into a `with()`-like function:

```{r}
#| output.lines: 3
with2 <- function(df, expr) {
  eval_tidy(enexpr(expr), df)
}
with2(df, x + y)
```

**Problem:** if `with2()` defines an internal variable `a`, it can shadow the caller's `a`.

## Quosures

A **quosure** bundles an expression + its environment:

```{r}
#| output.lines: 3
with2 <- function(df, expr) {
  a <- 1000
  eval_tidy(enquo(expr), df)
}
df <- data.frame(x = 1:3)
a <- 10
with2(df, x + a)
```

`enquo()` captures the expression **and** where it was written. Result: `a` is 10 (caller's value), not 1000.

\medskip

**Rule:** with data masks, always use \comm{enquo()} instead of `enexpr()`.

## Exercises (Big Picture)

1. What does `expr()` return when called inside a function? What about `enexpr()`?

2. Use `call2()` to construct the call `mean(1:10, na.rm = TRUE)`.

## Solutions (Big Picture)

1. `expr()` returns the **literal code** you typed (e.g., `expr(x)` returns the symbol `x` regardless of what `x` is). `enexpr()` captures the **caller's** expression --- it "enriches" a promise into an expression.

2.

```{r}
#| output.lines: 3
call2("mean", 1:10, na.rm = TRUE)
```

# Expressions

## Abstract Syntax Trees

**Expressions** = ASTs. Conventions for `lobstr::ast()`:

- **Leaves** = symbols (e.g., `f`, `x`) or constants (`1`, `"y"`)
- **Branches** = call objects (function calls)

```{r}
#| output.lines: 5
lobstr::ast(f(x, "y", 1))
```

ASTs are **abstract** --- they drop whitespace and comments:

```{r}
#| output.lines: 3
ast(f(x,  y)  # this comment is dropped
)
```

## Infix Calls in the AST

Every infix operator is a **function call** in the AST:

```{r}
#| output.lines: 5
lobstr::ast(y <- x * 10)
```

These are equivalent:

```{r}
#| eval: false
y <- x * 10
`<-`(y, `*`(x, 10))
```

R prints expressions in infix form even if constructed with prefix:

```{r}
#| output.lines: 3
expr(`<-`(y, `*`(x, 10)))
```

## Expression Types

Four types form expressions:

| Type | Example | `str()` | `typeof()` |
|------|---------|---------|------------|
| Constant | `TRUE`, `1`, `"x"` | `logi`/`int`/... | `logical`/... |
| Symbol | `x`, `mean` | `symbol` | `symbol` |
| Call | `f(x)` | `language` | `language` |
| Pairlist | (function args) | Dotted pair list | `pairlist` |

\medskip

Constants are **self-quoting**: `expr(TRUE)` is identical to `TRUE`.

## Symbols

A **symbol** represents a name (like `x` or `mtcars`):

```{r}
#| output.lines: 5
expr(x)
sym("x")          # string -> symbol
as_string(expr(x)) # symbol -> string
is.symbol(expr(x))
```

Symbols are always length 1. For multiple symbols, use \comm{rlang::syms()}.

## Call Objects

A **call object** = captured function call. Behaves like a list:

```{r}
#| output.lines: 5
x <- expr(read.table("important.csv", row.names = FALSE))
x[[1]]           # function (symbol)
x[[2]]           # first argument
x$row.names      # named argument
length(x) - 1    # number of arguments
```

## Call Objects: Standardise and Modify

\comm{rlang::call\_standardise()} expands partial argument names:

```{r}
#| output.lines: 3
x <- expr(read.table("important.csv", row.names = FALSE))
rlang::call_standardise(x)
```

Calls can be modified like lists:

```{r}
#| output.lines: 3
x$header <- TRUE
x
```

## Constructing Calls

\comm{rlang::call2()} builds calls from parts:

```{r}
#| output.lines: 5
call2("mean", x = expr(x), na.rm = TRUE)
call2(expr(base::mean), x = expr(x), na.rm = TRUE)
call2("<-", expr(x), 10)
```

## Parsing and Deparsing

**Parsing:** string $\to$ expression. **Deparsing:** expression $\to$ string.

```{r}
#| output.lines: 5
x1 <- "y <- x + 10"
x2 <- rlang::parse_expr(x1)
is.call(x2)
rlang::expr_text(x2)
```

For multiple expressions, use \comm{parse\_exprs()}:

```{r}
#| output.lines: 3
rlang::parse_exprs("a <- 1; a + 1")
```

Parsing is lossy: comments and whitespace are dropped.

## Operator Precedence

Infix operators have **precedence rules** --- `ast()` reveals them:

```{r}
#| output.lines: 3
lobstr::ast(1 + 2 * 3)
```

`*` binds tighter than `+`. Surprising: `!` has **low** precedence:

```{r}
#| output.lines: 3
lobstr::ast(!x %in% y)
```

Use parentheses when in doubt. R has 30+ infix operators in 18 precedence groups.

## Associativity

Most operators are **left-associative**: `1 + 2 + 3` = `(1 + 2) + 3`:

```{r}
#| output.lines: 3
lobstr::ast(1 + 2 + 3)
```

Two exceptions --- **right-associative**: exponentiation and assignment:

```{r}
#| output.lines: 3
lobstr::ast(2^2^3)
```

```{r}
#| output.lines: 3
lobstr::ast(x <- y <- z)
```

## Walking ASTs: Recursive Functions

ASTs are trees $\to$ natural fit for **recursion**:

```{r}
#| output.lines: 5
expr_type <- function(x) {
  if (rlang::is_syntactic_literal(x)) "constant"
  else if (is.symbol(x)) "symbol"
  else if (is.call(x)) "call"
  else if (is.pairlist(x)) "pairlist"
  else typeof(x)
}
c(expr_type(expr("a")),
  expr_type(expr(x)),
  expr_type(expr(f(1, 2))))
```

**Base cases:** constants, symbols. **Recursive cases:** calls, pairlists.

## Example: Finding `T` and `F`

Detect logical abbreviations `T`/`F` in code:

```{r}
#| output.lines: 5
switch_expr <- function(x, ...) {
  switch(expr_type(x), ...,
    stop("Unknown type: ", typeof(x)))
}
logical_abbr_rec <- function(x) {
  switch_expr(x,
    constant = FALSE,
    symbol = as_string(x) %in% c("F", "T"),
    call = , pairlist = purrr::some(x, logical_abbr_rec))
}
logical_abbr <- function(x) logical_abbr_rec(enexpr(x))
c(logical_abbr(TRUE), logical_abbr(mean(x, na.rm = T)))
```

## Specialised Data Structures

**Pairlists:** store function formals. Treat like lists:

```{r}
#| output.lines: 3
f <- expr(function(x, y = 10) x + y)
typeof(f[[2]])
```

**Missing arguments:** empty symbol for missing defaults:

```{r}
#| output.lines: 3
is_missing(missing_arg())
args <- f[[2]]
is_missing(args[[1]])   # x has no default
```

**Expression vectors:** returned by `parse()` and `expression()`. Use lists of expressions instead.

## Exercises (Expressions)

1. Which two atomic vector types can't appear in an expression? Why?

2. What happens when you subset a call to remove the first element?

```{r}
#| eval: false
expr(read.csv("foo.csv", header = TRUE))[-1]
```

3. Does `-2^2` yield 4 or -4? Why?

## Solutions (Expressions)

1. **Raw** and **complex** vectors can't appear directly --- you need a function call to create them (e.g., `as.raw()`, `complex()`), which makes them calls, not constants. Similarly, vectors of length > 1 require `c()`.

2. Removing the first element shifts the first argument into the function position:

```{r}
#| output.lines: 3
expr(read.csv("foo.csv", header = TRUE))[-1]
```

So `"foo.csv"` becomes the "function" --- produces `"foo.csv"(header = TRUE)`.

3. Returns **-4**. Exponentiation `^` has higher precedence than unary `-`:

```{r}
#| output.lines: 3
lobstr::ast(-2^2)
```

The AST shows `-(2^2)`, not `(-2)^2`.

<!--
# Summary

## Key Takeaways

- **Metaprogramming** = code that inspects/modifies/generates code
- **`expr()`** captures code you type; **`enexpr()`** captures caller's code
- Expressions behave like **lists** --- subsettable and modifiable
- Code is a **tree** (AST): branches = calls, leaves = symbols/constants
- **`!!` (bang-bang)** unquotes --- inserts expressions into templates
- **`eval()`** runs expressions; supply an environment to customize
- **Data masks** let columns act as variables (`eval_tidy()`)
- **Quosures** = expression + environment; use `enquo()` with data masks
- Four expression types: **constants**, **symbols**, **calls**, **pairlists**
- Call objects: first element = function, rest = arguments
- **Parsing** = string → AST; **deparsing** = AST → string
- Operator precedence: `*` before `+`; `^` and `<-` are right-associative
- Walk ASTs with **recursive functions**: base cases (constants/symbols) + recursive cases (calls/pairlists)
- Pairlists store function formals; missing args use empty symbols
-->
