---
title: "AdvR 02 -- Vectors"
---

```{r setup, include=FALSE}
source("_common.R")
library(lobstr)
library(tibble)
```

# Atomic Vectors

## Vector Types

Two flavors of vectors in R:

1. **Atomic vectors** --- all elements same type
2. **Lists** --- elements can be any type

\medskip

**Four common atomic types:**

| Type | Example | `typeof()` |
|------|---------|-----------|
| Logical | `TRUE`, `FALSE` | `"logical"` |
| Integer | `1L`, `6L` | `"integer"` |
| Double | `1.5`, `1.23e4` | `"double"` |
| Character | `"hello"` | `"character"` |

Two rare types: `complex` and `raw`.

\medskip

"Numeric" = integer + double collectively.

## Creating Vectors with `c()`

```{r}
#| output.lines: 5
lgl_var <- c(TRUE, FALSE)
int_var <- c(1L, 6L, 10L)
dbl_var <- c(1, 2.5, 4.5)
chr_var <- c("these are", "some strings")

typeof(int_var)
length(dbl_var)
```

\comm{c()} **flattens** nested calls:

```{r}
#| output.lines: 3
c(c(1, 2), c(3, 4))
```

Special doubles: `Inf`, `-Inf`, `NaN`.

## Missing Values

`NA` is infectious --- most operations with `NA` return `NA`:

```{r}
#| output.lines: 3
NA > 5
10 * NA
```

Exceptions (identity holds for all inputs): `NA ^ 0` $\to$ `1`, `NA | TRUE` $\to$ `TRUE`, `NA & FALSE` $\to$ `FALSE`.

\medskip

**Test** with \comm{is.na()}, never `==`. Logical $\to$ numeric is useful:

```{r}
#| output.lines: 3
x <- c(FALSE, FALSE, TRUE, TRUE, FALSE)
sum(x)      # count TRUEs
mean(x)     # proportion of TRUEs
```

Four typed NAs: `NA` (logical), `NA_integer_`, `NA_real_`, `NA_character_`.

## Coercion

When types are mixed, R coerces to the most flexible type:

$$\text{logical} \to \text{integer} \to \text{double} \to \text{character}$$

```{r}
#| output.lines: 5
c(1, FALSE)        # double: c(1, 0)
c("a", 1)          # character: c("a", "1")
c(TRUE, 1L)        # integer: c(1L, 1L)
```

Manual coercion with \comm{as.*()}:

```{r}
#| output.lines: 3
as.integer("42")
as.double(TRUE)
```

Test with \comm{is.na()}, not `== NA` (which returns `NA`).

## Exercises (Atomic Vectors)

1. How do you create raw and complex scalars? (Hint: `?raw` and `?complex`.)

2. Predict the output:

```{r}
#| eval: false
c(1, FALSE)
c("a", 1)
c(TRUE, 1L)
```

3. Why is `1 == "1"` true? Why is `-1 < FALSE` true? Why is `"one" < 2` false?

4. Why is the default missing value `NA` a logical vector? (Hint: think about \comm{c(FALSE, NA\_character\_)}.)

## Solutions (Atomic Vectors)

1. `as.raw(42)` or `charToRaw("A")` for raw. `complex(real = 1, imaginary = 1)` or `1 + 1i` for complex.

2.

```{r}
#| output.lines: 5
str(c(1, FALSE))     # double: c(1, 0)
str(c("a", 1))       # character: c("a", "1")
str(c(TRUE, 1L))     # integer: c(1L, 1L)
```

3. Comparison operators coerce to common type. `1` $\to$ `"1"`, `FALSE` $\to$ `0`, `2` $\to$ `"2"` (and `"one" > "2"` lexicographically).

4. Logical is the least flexible type. `NA` coerces **up** (to `NA_integer_`, etc.) without affecting the result type. If `NA` were character, everything would become character.

# Attributes

## Getting and Setting Attributes

Attributes are **name-value pairs** attached to any object:

```{r}
#| output.lines: 5
a <- 1:3
attr(a, "x") <- "abcdef"
attr(a, "y") <- 4:6
str(attributes(a))
```

Or use \comm{structure()}:

```{r}
#| output.lines: 5
a <- structure(1:3, x = "abcdef", y = 4:6)
str(attributes(a))
```

**Most attributes are lost** by operations:

```{r}
#| output.lines: 3
attributes(a[1])   # NULL
```

Two exceptions routinely preserved: **names** and **dim**.

## Names and Dimensions

**Names** --- three ways to create:

```{r}
#| eval: false
x <- c(a = 1, b = 2, c = 3)         # at creation
names(x) <- c("a", "b", "c")         # assignment
x <- setNames(1:3, c("a", "b", "c")) # setNames()
```

Remove with `names(x) <- NULL` or \comm{unname(x)}.

\medskip

**Dimensions** turn vectors into matrices/arrays:

```{r}
#| output.lines: 5
x <- matrix(1:6, nrow = 2, ncol = 3)
x
```

| Vector | Matrix | Array |
|--------|--------|-------|
| `names()` | `rownames()`, `colnames()` | `dimnames()` |
| `length()` | `nrow()`, `ncol()` | `dim()` |
| `c()` | `rbind()`, `cbind()` | `abind::abind()` |

## Exercises (Attributes)

1. How is \comm{setNames()} implemented? How is \comm{unname()} implemented?

2. What does \comm{dim()} return for a 1-d vector? When might you use \comm{NROW()} or \comm{NCOL()}?

3. Describe these objects. What makes them different from `1:5`?

```{r}
#| eval: false
x1 <- array(1:5, c(1, 1, 5))
x2 <- array(1:5, c(1, 5, 1))
x3 <- array(1:5, c(5, 1, 1))
```

## Solutions (Attributes)

1. \comm{setNames()} assigns `names(object) <- nm` and returns the object. \comm{unname()} sets `names(obj) <- NULL` (and optionally `dimnames`).

2. `dim()` returns `NULL` for a plain vector. Use \comm{NROW()}/\comm{NCOL()} to treat vectors uniformly with matrices:

```{r}
#| output.lines: 5
x <- 1:10
c(nrow(x), ncol(x))     # NULL, NULL
c(NROW(x), NCOL(x))     # 10, 1
```

3. All three are 3-d arrays with a `dim` attribute (1-by-1-by-5, 1-by-5-by-1, 5-by-1-by-1). Unlike `1:5`, they have dimensionality.

# S3 Atomic Vectors

## Factors

**Factors** store categorical data. Built on integer vectors with `levels`:

```{r}
#| output.lines: 5
x <- factor(c("a", "b", "b", "a"))
typeof(x)
levels(x)
```

Factors record **all possible** values, even unobserved:

```{r}
#| output.lines: 5
sex_char <- c("m", "m", "m")
sex_factor <- factor(sex_char, levels = c("m", "f"))
table(sex_char)
table(sex_factor)
```

**Ordered factors** have meaningful level order:

```{r}
#| eval: false
grade <- ordered(c("b", "b", "a", "c"),
  levels = c("c", "b", "a"))
```

## Dates and Date-Times

**Dates** --- doubles representing days since 1970-01-01:

```{r}
#| output.lines: 5
today <- Sys.Date()
typeof(today)
unclass(as.Date("1970-02-01"))  # 31 days
```

**Date-times (POSIXct)** --- doubles representing seconds since epoch:

```{r}
#| output.lines: 3
now_ct <- as.POSIXct("2018-08-01 22:00", tz = "UTC")
typeof(now_ct)
```

**Durations (difftime)** --- doubles with a `units` attribute:

```{r}
#| output.lines: 2
as.difftime(1, units = "weeks")
```

## Exercises (S3 Vectors)

1. What sort of object does \comm{table()} return? What is its type? What attributes does it have?

2. What happens to a factor when you modify its levels?

```{r}
#| eval: false
f1 <- factor(letters)
levels(f1) <- rev(levels(f1))
```

3. How do `f2` and `f3` differ from `f1`?

```{r}
#| eval: false
f2 <- rev(factor(letters))
f3 <- factor(letters, levels = rev(letters))
```

## Solutions (S3 Vectors)

1. \comm{table()} returns an integer array with class `"table"`, a `dim` attribute, and `dimnames`. Dimensions correspond to unique values in each input.

2. The **underlying integers stay the same** but the level labels change. This makes it *look* like the data changed, but only the mapping changed:

```{r}
#| output.lines: 5
f1 <- factor(letters[1:5])
as.integer(f1)
levels(f1) <- rev(levels(f1))
as.integer(f1)   # still 1 2 3 4 5
f1               # but displays as e d c b a
```

3. `f2 <- rev(factor(letters))` reverses **element order** (integers change, levels stay). `f3 <- factor(letters, levels = rev(letters))` reverses **level order** (integers change, levels reversed).

# Lists

## Lists: Heterogeneous Vectors

Each element can be **any type**, including other lists:

```{r}
#| output.lines: 6
l1 <- list(1:3, "a", c(TRUE, FALSE), c(2.3, 5.9))
str(l1)
```

Lists store **references** (not copies) and are **recursive**:

```{r}
#| output.lines: 3
lobstr::obj_size(mtcars)
lobstr::obj_size(list(mtcars, mtcars, mtcars))
```

Three copies of `mtcars` in a list costs only ~80 extra bytes. Lists of lists: `list(list(list(1)))`.

## Combining and Coercing Lists

\comm{c()} with lists combines them --- \comm{list()} nests, \comm{c()} flattens:

```{r}
#| output.lines: 5
str(list(list(1, 2), c(3, 4)))   # nested
```

```{r}
#| output.lines: 5
str(c(list(1, 2), c(3, 4)))     # flat
```

\comm{as.list()} converts atomic to list. \comm{unlist()} does the reverse.

## Exercises (Lists)

1. List all ways a list differs from an atomic vector.

2. Why do you need \comm{unlist()} to convert a list to an atomic vector? Why doesn't \comm{as.vector()} work?

3. Compare \comm{c()} and \comm{unlist()} when combining a date and a date-time.

## Solutions (Lists)

1. Key differences:
   - **Heterogeneous** vs. homogeneous types
   - Elements are **references** to separate objects
   - Can be **recursive** (lists of lists)
   - Out-of-bounds subsetting returns `NULL`, not `NA`

2. A list **is already a vector** (just not atomic). \comm{as.vector()} won't change that. \comm{unlist()} recursively extracts and coerces elements into an atomic vector.

3. \comm{c()} converts to a common S3 type (POSIXct):

```{r}
#| output.lines: 3
c(as.Date("1970-01-02"),
  as.POSIXct("1970-01-01 01:00", tz = "UTC"))
```

\comm{unlist()} strips **all attributes** --- returns raw doubles (losing date meaning).

# Data Frames and Tibbles

## Data Frame Basics

Data frames are **named lists of equal-length vectors**:

```{r}
#| output.lines: 5
df <- data.frame(x = 1:3, y = c("a", "b", "c"))
typeof(df)
attributes(df) |> str()
```

Key properties:

```{r}
#| output.lines: 3
names(df)
nrow(df)
ncol(df)
```

## Tibbles vs. Data Frames

Tibbles are a modern reimagining with stricter behavior:

| Feature | `data.frame` | `tibble` |
|---------|-------------|----------|
| String coercion | Converts to factor (pre-4.0) | Never |
| Non-syntactic names | Transforms | Preserves |
| Recycling | Integer multiples | Length-1 only |
| `$` partial matching | Yes (silent) | No (warns) |
| `[, "col"]` returns | Vector | Tibble |

```{r}
#| output.lines: 3
names(data.frame(`1` = 1))  # "X1" (transformed)
names(tibble(`1` = 1))      # "1" (preserved)
```

Tibbles allow **column references** during creation: `tibble(x = 1:3, y = x * 2)`.

## Row Names and List Columns

**Row names** are discouraged --- use a column instead:

```{r}
#| eval: false
df3 <- data.frame(age = c(35, 27),
  row.names = c("Bob", "Sue"))
as_tibble(df3, rownames = "name")
# Converts row names to a regular column
```

**List columns** store complex data per row:

```{r}
#| output.lines: 5
tibble(x = 1:3, y = list(1:2, 1:3, 1:4))
```

In base R, list columns need \comm{I()}: `data.frame(x = 1:3, y = I(list(...)))`.

## Exercises (Data Frames)

1. Can you have a data frame with zero rows? Zero columns?

2. What happens if you set rownames that are not unique?

3. If `df` is a data frame, what can you say about `t(df)` and `t(t(df))`?

4. What does \comm{as.matrix()} do with mixed column types? How does \comm{data.matrix()} differ?

## Solutions (Data Frames) --- 1--2

1. Yes to both:

```{r}
#| output.lines: 5
data.frame(a = integer(), b = logical()) # zero rows
data.frame()                             # zero columns
mtcars[0, 0]                             # both zero
```

2. Data frames **error** on duplicate row names. But subsetting with `[` can create duplicates (auto-deduplicated with suffixes).

```{r}
#| error: true
#| output.lines: 3
df <- data.frame(x = 1:3)
row.names(df) <- c("x", "y", "y")
```

## Solutions (Data Frames) --- 3--4

3. `t(df)` coerces to matrix first (\comm{as.matrix()}), then transposes. With mixed types, everything becomes the most flexible type (usually character). `t(t(df))` transposes back but remains a matrix.

```{r}
#| output.lines: 5
df <- data.frame(x = 1:3, y = letters[1:3])
t(df)  # character matrix
```

4. \comm{as.matrix()} coerces to the most flexible type (character if any character columns). \comm{data.matrix()} always returns a **numeric** matrix --- factors become integer codes, characters become `NA`.

# NULL

## NULL: The Empty/Absent Vector

`NULL` is a special type of length zero with no attributes:

```{r}
#| output.lines: 3
typeof(NULL)
length(NULL)
```

Two uses: (1) **empty vector** (`c()` returns `NULL`) and (2) **absent vector** (default argument when no value makes sense).

\medskip

Contrast with `NA`:

- `NULL` = the **vector** is absent (disappears in `c()`)
- `NA` = an **element** is absent (stays in `c()`)

```{r}
#| output.lines: 3
c(NULL, 1, 2)      # NULL disappears
c(NA, 1, 2)        # NA stays
```

<!--
# Summary

## Key Takeaways

- **Two vector types:** atomic (homogeneous) and list (heterogeneous)
- **Four common types:** logical, integer, double, character
- **Coercion:** logical $\to$ integer $\to$ double $\to$ character
- **NA is infectious:** use \comm{is.na()}, not `== NA`
- **Attributes:** name-value metadata; most lost on operations
- **Names and dim:** the two preserved attributes
- **Factors:** integer + levels; record all possible values
- **Dates:** doubles (days since epoch); POSIXct (seconds)
- **Lists:** heterogeneous, recursive, store references
- **Data frames:** named lists of equal-length vectors
- **Tibbles:** stricter subsetting, no coercion, better printing
- **List columns:** store complex data; easy with tibbles
- **NULL:** absent vector vs. NA (absent element)
-->
