---
title: "AdvR 13 -- Quasiquotation"
---

```{r setup, include=FALSE}
source("_common.R")
library(rlang)
library(lobstr)
```

# Motivation

## Why Quasiquotation?

**Quotation** = capturing code without evaluating it.
**Unquotation** = selectively evaluating parts of quoted code.
Together: **quasiquotation**.

```{r}
#| output.lines: 3
cement <- function(...) {
  args <- ensyms(...)
  paste(purrr::map(args, as_string), collapse = " ")
}
cement(Good, morning, Hadley)
```

Problem: how to use a **variable** inside a quoting function?

```{r}
#| output.lines: 3
cement(Good, time, name)  # quotes literally
```

## `!!` to the Rescue

\comm{!!} (bang-bang) **unquotes** --- tells the quoting function to evaluate:

```{r}
#| output.lines: 3
name <- "Hadley"; time <- "morning"
cement(Good, !!time, !!name)
```

Compare:

```{r}
#| eval: false
paste("Good", time, name)   # paste evaluates -> quote where needed
cement(Good, !!time, !!name) # cement quotes -> unquote where needed
```

## Quoted vs. Evaluated Arguments

- **Evaluated** argument: obeys normal R evaluation rules
- **Quoted** argument: captured by the function, processed in a custom way

Test: run the argument **outside** the function. If it fails, it's quoted:

```{r}
#| error: true
#| output.lines: 3
library(MASS)   # works: MASS is quoted
MASS             # fails: not a regular object
```

## Exercises (Motivation)

1. For each function below, which arguments are quoted and which evaluated?

```{r}
#| eval: false
library(MASS)
mtcars2 <- subset(mtcars, cyl == 4)
with(mtcars2, sum(vs))
sum(mtcars2$am)
rm(mtcars2)
```

## Solutions (Motivation)

1.

| Call | Quoted | Evaluated |
|------|--------|-----------|
| `library(MASS)` | `MASS` | --- |
| `subset(mtcars, cyl == 4)` | `cyl == 4` | `mtcars` |
| `with(mtcars2, sum(vs))` | `sum(vs)` | `mtcars2` |
| `sum(mtcars2$am)` | `am` (by `$`) | `mtcars2$am` |
| `rm(mtcars2)` | `mtcars2` | --- |

# Quoting

## Capturing Expressions

Four rlang quoting functions:

```{r}
#| output.lines: 3
expr(x + y)           # capture YOUR code
f2 <- function(x) enexpr(x)
f2(a + b + c)          # capture CALLER's code
```

```{r}
#| output.lines: 3
f <- function(...) enexprs(...)
f(x = 1, y = 10 * z)  # capture all ... args
```

Summary:

|      | Developer | User |
|------|-----------|------|
| One  | `expr()`  | `enexpr()` |
| Many | `exprs()` | `enexprs()` |

## Capturing Symbols

\comm{ensym()} / \comm{ensyms()} --- like `enexpr()` but **only** accept symbols or strings:

```{r}
#| output.lines: 3
f <- function(...) ensyms(...)
f(x)
f("x")
```

Useful when you want to restrict input to variable names.

## Base R Equivalents

| rlang | Base R |
|-------|--------|
| `expr()` | `quote()` |
| `enexpr()` | `substitute()` |
| `exprs()` | `alist()` |
| `enexprs()` | `as.list(substitute(...()))` |

\medskip

Key difference: base R functions **don't support unquoting** (`!!`).

## `substitute()` Does Substitution

\comm{substitute()} both **quotes** and **substitutes** values:

```{r}
#| output.lines: 3
f4 <- function(x) substitute(x * 2)
f4(a + b + c)
```

Use the second argument to make substitution explicit:

```{r}
#| output.lines: 3
substitute(x * y * z, list(x = 10, y = quote(a + b)))
```

## Exercises (Quoting)

1. Compare `f1` and `f2`. What does each return when called with `f(a + b, c + d)`?

```{r}
#| eval: false
f1 <- function(x, y) exprs(x = x, y = y)
f2 <- function(x, y) enexprs(x = x, y = y)
```

## Solutions (Quoting)

1. `f1` uses `exprs()` (developer quoting) --- returns the literal symbols `x` and `y`. `f2` uses `enexprs()` (user quoting) --- captures the caller's expressions:

```{r}
#| output.lines: 5
f1 <- function(x, y) exprs(x = x, y = y)
f2 <- function(x, y) enexprs(x = x, y = y)
str(f1(a + b, c + d))
str(f2(a + b, c + d))
```

# Unquoting

## `!!` --- Unquoting One Argument

\comm{!!} evaluates a single expression and inlines the result into the AST:

```{r}
#| output.lines: 3
x <- expr(-1)
expr(f(!!x, y))
```

Works with symbols, constants, and calls:

```{r}
#| output.lines: 3
a <- sym("y"); b <- 1
expr(f(!!a, !!b))
```

## `!!` Preserves Precedence

Because `!!` works with ASTs, not strings, operator precedence is correct:

```{r}
#| output.lines: 3
x1 <- expr(x + 1)
x2 <- expr(x + 2)
expr(!!x1 / !!x2)
```

Result: `(x + 1) / (x + 2)`. If we pasted strings, we'd get `x + 1 / x + 2` --- wrong precedence.

## Unquoting a Function

To unquote in the **function position**, use extra parentheses:

```{r}
#| output.lines: 3
f <- expr(pkg::foo)
expr((!!f)(x, y))
```

Or use \comm{call2()} for clarity:

```{r}
#| output.lines: 3
call2(f, expr(x), expr(y))
```

## `!!!` --- Unquoting Many Arguments

\comm{!!!} (bang-bang-bang) **splices** a list into a call:

```{r}
#| output.lines: 5
xs <- exprs(1, a, -b)
expr(f(!!!xs, y))
ys <- set_names(xs, c("a", "b", "c"))
expr(f(!!!ys, d = 4))
```

Works in any rlang function that takes `...`, including \comm{call2()}:

```{r}
#| output.lines: 3
call2("f", !!!xs, expr(y))
```

## The Polite Fiction of `!!`

From R's perspective, `!!` is just double negation:

```{r}
#| output.lines: 3
!!TRUE
!!!TRUE
```

`!!` and `!!!` behave as real operators **only inside rlang quoting functions**. Outside, you can get silent errors:

```{r}
#| output.lines: 3
df <- data.frame(x = 1:5)
y <- 100
with(df, x + !!y)   # !!100 = 1, not 100!
```

## Exercises (Unquoting)

1. Given these components, use `!!` and `!!!` to construct the calls below:

```{r}
#| eval: false
xy <- expr(x + y); xz <- expr(x + z)
yz <- expr(y + z); abc <- exprs(a, b, c)

(x + y) / (y + z)             # use !! twice
sum(a, b, c)                   # use !!!
mean(c(a, b, c), na.rm = TRUE) # use !!!
```

## Solutions (Unquoting)

1.

```{r}
#| output.lines: 5
xy <- expr(x + y); xz <- expr(x + z)
yz <- expr(y + z); abc <- exprs(a, b, c)
expr(!!xy / !!yz)
expr(sum(!!!abc))
expr(mean(c(!!!abc), na.rm = TRUE))
```

# Non-quoting and Tidy Dots

## Base R Non-quoting Techniques

Base R disables quoting rather than using `!!`:

| Technique | Example |
|-----------|---------|
| Quoting/non-quoting pair | `$` (quoted) vs `[[` (evaluated) |
| Character argument | `rm(list = "x")` instead of `rm(x)` |
| Switch argument | `library(pkg, character.only = TRUE)` |
| Eval-on-fail | `help(var)` tries eval, then quotes |

\medskip

```{r}
#| output.lines: 3
x <- list(var = 1, y = 2); var <- "y"
c(x$var, x[[var]])  # $ quotes; [[ evaluates
```

## `!!!` with `...` (Tidy Dots)

**Problem 1:** elements already in a list:

```{r}
#| output.lines: 3
dfs <- list(
  a = data.frame(x = 1, y = 2),
  b = data.frame(x = 3, y = 4))
dplyr::bind_rows(!!!dfs)
```

**Problem 2:** dynamic argument names --- use \comm{:=}:

```{r}
#| output.lines: 3
var <- "x"; val <- c(4, 3, 9)
tibble::tibble(!!var := val)
```

`:=` is needed because R's grammar doesn't allow `!!var = val`.

## `exec()` and `do.call()`

\comm{rlang::exec()} calls a function with `!!!` and `:=` support:

```{r}
#| output.lines: 3
params <- list(na.rm = TRUE, trim = 0.1)
exec("mean", x = 1:10, !!!params)
```

Base R equivalent: \comm{do.call()}:

```{r}
#| output.lines: 3
do.call("mean", c(list(x = 1:10), params))
```

`exec()` also supports dynamic argument names:

```{r}
#| output.lines: 3
arg_name <- "na.rm"
exec("mean", 1:10, !!arg_name := TRUE)
```

## Exercises (Tidy Dots)

1. One way to implement `exec()`:

```{r}
#| eval: false
exec <- function(f, ..., .env = caller_env()) {
  args <- list2(...)
  do.call(f, args, envir = .env)
}
```

Describe how it works. What are the key ideas?

## Solutions (Tidy Dots)

1. `list2(...)` collects arguments with tidy dots support (`!!!` for splicing, `:=` for dynamic names). Then `do.call()` calls `f` with those arguments in the specified environment. Key ideas: (a) `list2()` gives any function tidy dot semantics, (b) `do.call()` provides the actual function call, (c) `.env` ensures correct scoping.

<!--
# Summary

## Key Takeaways

- **Quotation** captures code; **unquotation** (`!!`) selectively evaluates parts
- **`expr()`** / **`exprs()`** = developer quoting; **`enexpr()`** / **`enexprs()`** = user quoting
- **`ensym()`** restricts to symbols/strings only
- Base equivalents: `quote()`, `substitute()`, `alist()` --- no unquoting support
- **`!!` (bang-bang)** unquotes one expression; preserves operator precedence
- **`!!!` (bang-bang-bang)** splices a list of expressions into a call
- **`:=` (colon-equals)** allows dynamic argument names on the LHS
- `!!` is a fiction --- it's double negation outside rlang quoting functions
- Base R uses non-quoting techniques: `[[` vs `$`, `character.only`, `do.call()`
- **Tidy dots** (`list2()`) gives `!!!` and `:=` to any function taking `...`
- **`exec()`** = `do.call()` with tidy dots; **`call2()`** builds calls without evaluating
-->
