---
title: "AdvR 06 -- Environments"
---

```{r setup, include=FALSE}
source("_common.R")
library(rlang)
library(lobstr)
```

# Environment Basics

## Four Key Differences from Lists

An environment is like a named list, with four exceptions:

1. Every name must be **unique**
2. Names are **not ordered**
3. Every environment has a **parent**
4. Environments have **reference semantics** (modify in place, not copy-on-modify)

\medskip

Environments can even **contain themselves**:

```{r}
#| output.lines: 3
e1 <- env(a = 1, b = 2)
e1$self <- e1
env_names(e1)
```

## Creating and Inspecting Environments

Create with \comm{rlang::env()} (like \comm{list()}):

```{r}
#| output.lines: 5
e1 <- env(a = FALSE, b = "a", c = 2.3, d = 1:3)
e1   # just prints memory address
```

Use \comm{env\_print()} for useful info:

```{r}
#| output.lines: 5
env_print(e1)
```

```{r}
#| output.lines: 3
env_names(e1)
```

## Important Environments

- \comm{current\_env()}: where code is currently executing
- \comm{global\_env()}: the interactive workspace

\medskip

Compare with \comm{identical()}, **not** `==`:

```{r}
#| error: true
#| output.lines: 3
identical(global_env(), current_env())
```

```{r}
#| error: true
#| output.lines: 3
global_env() == current_env()
```

Base equivalents: `globalenv()`, `environment()`.

## Parents

Every environment has a **parent** (used for lexical scoping):

```{r}
#| output.lines: 3
e2a <- env(d = 4, e = 5)
e2b <- env(e2a, a = 1, b = 2, c = 3)
env_parent(e2b)
```

The only environment without a parent is the **empty environment**:

```{r}
#| output.lines: 3
env_parents(e2b)
```

```{r}
#| output.lines: 3
env_parents(e2b, last = empty_env())
```

Base equivalent: `parent.env()`.

## Super Assignment `<<-`

`<-` creates a variable in the **current** environment. `<<-` modifies an existing variable in a **parent** environment:

```{r}
#| output.lines: 3
x <- 0
f <- function() { x <<- 1 }
f()
x
```

If `<<-` doesn't find the variable, it creates one in the **global** environment --- usually undesirable. Best used with function factories (closures).

## Getting and Setting

Use `$` and `[[` like lists, but **no** numeric indexing or `[`:

```{r}
#| output.lines: 3
e3 <- env(x = 1, y = 2)
e3$x
e3[["y"]]
```

```{r}
#| error: true
#| output.lines: 3
e3[[1]]
```

\comm{env\_get()} errors on missing names (safer than `$`):

```{r}
#| error: true
#| output.lines: 3
env_get(e3, "xyz")
```

## Binding and Inspecting

\comm{env\_bind()} sets multiple bindings at once:

```{r}
#| output.lines: 3
env_bind(e3, a = 10, b = 20)
env_names(e3)
```

## Removing Bindings

Setting to `NULL` does **not** remove --- use \comm{env\_unbind()}:

```{r}
#| output.lines: 3
e3$a <- NULL
env_has(e3, "a")   # still there!
```

```{r}
#| output.lines: 3
env_unbind(e3, "a")
env_has(e3, "a")   # now gone
```

## Advanced Bindings

**Delayed bindings** --- evaluated only on first access (like promises):

```{r}
#| eval: false
env_bind_lazy(current_env(), b = {Sys.sleep(1); 1})
system.time(print(b))  # slow (evaluates)
system.time(print(b))  # fast (cached)
```

Used by \comm{autoload()} to defer loading package datasets.

\medskip

**Active bindings** --- re-computed on every access:

```{r}
#| output.lines: 3
env_bind_active(current_env(), z1 = function(val) runif(1))
c(z1, z1, z1)   # different each time
```

Used by R6 active fields.

## Exercises (Basics)

1. List three ways an environment differs from a list.

2. Explain why `e[[1]]` and `e[c("a", "b")]` don't make sense for environments.

3. What does this function do? How does it differ from `<<-`?

```{r}
#| eval: false
rebind <- function(name, value, env = caller_env()) {
  if (identical(env, empty_env())) {
    stop("Can't find `", name, "`", call. = FALSE)
  } else if (env_has(env, name)) {
    env_poke(env, name, value)
  } else {
    rebind(name, value, env_parent(env))
  }
}
```

## Solutions (Basics) --- 1--2

1. Key differences:
   - **Reference semantics** (no copy-on-modify)
   - Every environment has a **parent**
   - All names must be **unique**
   - Names are **not ordered**
   - Can contain themselves

2. `e[[1]]` --- elements are not ordered, so numeric indexing is meaningless. `e[c("a", "b")]` --- `[` is not defined for environments; subsetting would need to return something, but environments are not subsettable structures.

## Solutions (Basics) --- 3

3. `rebind()` walks up parent environments looking for an existing binding to reassign. Unlike `<<-`, it **never creates** a new variable --- it throws an error if the name doesn't exist anywhere. This is safer because `<<-` silently creates global variables when no binding is found.

```{r}
#| include: false
rebind <- function(name, value, env = caller_env()) {
  if (identical(env, empty_env())) {
    stop("Can't find `", name, "`", call. = FALSE)
  } else if (env_has(env, name)) {
    env_poke(env, name, value)
  } else {
    rebind(name, value, env_parent(env))
  }
}
```

```{r}
#| error: true
#| output.lines: 5
a <- 5
rebind("a", 10)
a
rebind("zzz", 10)   # error: doesn't exist
```

# Recursing over Environments

## The `where()` Function

A template for **recursive environment traversal**:

```{r}
#| output.lines: 3
where <- function(name, env = caller_env()) {
  if (identical(env, empty_env())) {
    stop("Can't find ", name, call. = FALSE)
  } else if (env_has(env, name)) {
    env
  } else {
    where(name, env_parent(env))
  }
}
x <- 5
where("x")
```

```{r}
#| output.lines: 3
where("mean")
```

## Recursive Template

The general pattern for computing over environments:

```{r}
#| eval: false
f <- function(..., env = caller_env()) {
  if (identical(env, empty_env())) {
    # base case
  } else if (success) {
    # success case
  } else {
    # recursive case
    f(..., env = env_parent(env))
  }
}
```

Can also use iteration:

```{r}
#| eval: false
f <- function(..., env = caller_env()) {
  while (!identical(env, empty_env())) {
    if (success) return()
    env <- env_parent(env)
  }
  # base case
}
```

## Exercises (Recursion)

1. Modify `where()` to return **all** environments that contain a binding for `name`.

2. Write `fget()` that finds only **function** objects, skipping non-function bindings with the same name.

## Solutions (Recursion) --- 1

1. Recurse to the empty environment regardless of matches; collect all hits:

```{r}
#| output.lines: 5
where2 <- function(name, env = caller_env(), results = list()) {
  if (identical(env, empty_env())) {
    results
  } else {
    if (env_has(env, name)) results <- c(results, env)
    where2(name, env_parent(env), results)
  }
}
```

## Solutions (Recursion) --- 2

2. Check if the found object is a function; if not, recurse:

```{r}
#| output.lines: 3
fget <- function(name, env = caller_env()) {
  if (env_has(env, name)) {
    obj <- env_get(env, name)
    if (is.function(obj)) return(obj)
  }
  if (identical(env, empty_env()))
    stop("No function called \"", name, "\"", call. = FALSE)
  fget(name, env_parent(env))
}
mean <- 10
fget("mean")   # finds base::mean, not the value 10
```

# Special Environments

## Package Environments and Search Path

Each `library()` call adds a package to the **search path** --- a chain of parent environments above global:

```{r}
#| output.lines: 5
search()
```

- **Last attached** = immediate parent of global
- Bottom of chain: `Autoloads` $\to$ `package:base`

\medskip

The search path is why you can call `mean()` without `base::mean()`.

## Function Environment

A function captures the environment where it was **defined** --- its **function environment** (= **enclosing** environment):

```{r}
#| output.lines: 3
y <- 1
f <- function(x) x + y
fn_env(f)
```

This is why R uses **lexical scoping**: the function looks up `y` where it was defined, not where it was called.

\medskip

Functions that capture their environment are called **closures**.

\medskip

Base equivalent: `environment(f)`.

## Namespaces

**Problem:** If the search path determines lookups, wouldn't different `library()` orders break packages?

\medskip

**Solution: namespaces.** Every package has two environments:

- **Package environment** --- external interface (what users see via `::`)
- **Namespace environment** --- internal interface (how functions find variables)

\medskip

The namespace chain: imports $\to$ base namespace $\to$ global env.

\medskip

This ensures `sd()` always finds `var()` from `stats`, regardless of what's on the search path.

## Execution Environments

Each function call creates a **fresh execution environment**:

```{r}
#| output.lines: 3
g <- function(x) {
  if (!env_has(current_env(), "a")) {
    message("Defining a"); a <- 1
  } else a <- a + 1
  a
}
c(g(10), g(10))   # both 1 (fresh start)
```

- Parent of execution env = **function environment**
- Usually **ephemeral** (garbage collected after return)
- Persists if you return it or return a closure that captures it

## Execution Environments and Closures

Returning a function captures the execution environment:

```{r}
#| output.lines: 3
plus <- function(x) {
  function(y) x + y
}
plus_one <- plus(1)
plus_one(2)
```

The enclosing environment of `plus_one` is the (now-persistent) **execution environment** of `plus(1)`, which binds `x = 1`.

## Exercises (Special Environments)

1. How is \comm{search\_envs()} different from \comm{env\_parents(global\_env())}?

2. What is the enclosing environment of each nested function?

```{r}
#| eval: false
f1 <- function(x1) {
  f2 <- function(x2) {
    f3 <- function(x3) { x1 + x2 + x3 }
    f3(3)
  }
  f2(2)
}
f1(1)
```

## Solutions (Special Environments)

1. \comm{search\_envs()} includes the global env and all attached packages (ends at base). \comm{env\_parents(global\_env())} excludes the global env itself and includes the empty env at the end.

2. `f1` is defined in the global env $\to$ encloses global. `f2` is defined inside `f1`'s execution $\to$ encloses that execution env. `f3` is defined inside `f2`'s execution $\to$ encloses that execution env. Each nested function captures variables from all enclosing scopes (`x1`, `x2`, `x3`).

# Call Stacks

## Call Stacks and Frames

The **call stack** tracks the sequence of function calls:

```{r}
#| eval: false
f <- function(x) g(x = 2)
g <- function(x) h(x = 3)
h <- function(x) lobstr::cst()

f(x = 1)
#> x
#> +-f(x = 1)
#> | \-g(x = 2)
#> |   \-h(x = 3)
#> |     \-lobstr::cst()
```

Each element is a **frame**: an expression, an environment, and a parent frame.

\medskip

The **caller environment** (\comm{caller\_env()} or base \comm{parent.frame()}) is the environment from which the function was **called**.

## Dynamic Scope

**Lexical scoping:** look up values where the function was **defined**.

**Dynamic scoping:** look up values where the function was **called**.

\medskip

R uses lexical scoping by default. Dynamic scoping is used for:

- Interactive data analysis tools (tidy evaluation)
- `parent.frame()` / `caller_env()` patterns

\medskip

Few languages use dynamic scoping (Emacs Lisp is a notable exception).

## Exercises (Call Stacks)

1. Write a function that lists all variables defined in the environment in which it was **called** (same result as `ls()`).

## Solutions (Call Stacks)

1. Use \comm{caller\_env()} to access the calling environment:

```{r}
#| output.lines: 5
ls2 <- function(env = caller_env()) {
  sort(env_names(env))
}
e1 <- env(a = 1, b = 2)
ls2(e1)
```

# Environments as Data Structures

## Three Use Cases

Environments have **reference semantics** --- useful beyond scoping:

\medskip

**1. Avoiding copies of large data** --- no accidental copy-on-modify. Better yet, use R6 objects (built on environments).

\medskip

**2. Managing state within a package:**

```{r}
#| eval: false
my_env <- new.env(parent = emptyenv())
my_env$a <- 1
get_a <- function() my_env$a
set_a <- function(value) {
  old <- my_env$a; my_env$a <- value
  invisible(old)
}
```

**3. As a hashmap** --- O(1) lookup by name (constant time regardless of size).

<!--
# Summary

## Key Takeaways

- **Environments vs. lists:** unique names, unordered, have parents, reference semantics
- **Creating:** \comm{env()}, inspecting with \comm{env\_print()}, \comm{env\_names()}
- **Parents:** every env has one; empty env is the root
- **`<<-`:** modifies variable in parent env; creates global if not found
- **Getting/setting:** `$`, `[[`, \comm{env\_get()}, \comm{env\_bind()}; no `[` or numeric index
- **Removing:** \comm{env\_unbind()}, not `NULL` assignment
- **Recursive traversal:** base case (empty env) $\to$ success $\to$ recurse on parent
- **Search path:** chain of package environments above global
- **Function environment:** where defined, powers lexical scoping (closures)
- **Namespace:** insulates package internals from search path changes
- **Execution environment:** fresh per call, parent = function env, usually ephemeral
- **Caller environment:** \comm{caller\_env()} / \comm{parent.frame()} for dynamic scoping
- **Data structure uses:** avoiding copies, managing state, hashmaps
-->
