---
title: "AdvR 03 -- Subsetting"
---

```{r setup, include=FALSE}
source("_common.R")
library(tidyverse)
```

# Selecting Multiple Elements

## Six Ways to Subset Atomic Vectors (1--3)

```{r}
#| output.lines: 3
x <- c(2.1, 4.2, 3.3, 5.4)
x[c(3, 1)]         # 1. Positive integers: select
x[c(1, 1)]         # Duplicates allowed
```

```{r}
#| output.lines: 3
x[-c(3, 1)]        # 2. Negative integers: exclude
```

```{r}
#| output.lines: 3
x[c(TRUE, TRUE, FALSE, FALSE)]   # 3. Logical vector
x[x > 3]           # Most common use
```

## Six Ways to Subset Atomic Vectors (4--6)

```{r}
#| output.lines: 3
x[]                 # 4. Nothing: returns all
x[0]                # 5. Zero: returns empty
```

```{r}
#| output.lines: 3
y <- setNames(x, letters[1:4])
y[c("d", "c", "a")]   # 6. Character: by name
```

## Matrices and Data Frames

**Matrices** --- subset with `[rows, cols]`:

```{r}
#| output.lines: 5
a <- matrix(1:9, nrow = 3)
colnames(a) <- c("A", "B", "C")
a[1:2, ]
```

```{r}
#| output.lines: 3
a[c(TRUE, FALSE, TRUE), c("B", "A")]
```

**Data frames** --- single index = columns (list behavior); two indices = rows, columns (matrix behavior):

```{r}
#| output.lines: 5
df <- data.frame(x = 1:3, y = 3:1, z = letters[1:3])
df[c("x", "z")]     # list-style: data frame
```

## Preserving Dimensionality

By default, `[` simplifies output (drops dimensions):

```{r}
#| output.lines: 3
a <- matrix(1:4, nrow = 2)
str(a[1, ])               # vector (simplified)
str(a[1, , drop = FALSE]) # matrix (preserved)
```

Same for data frames:

```{r}
#| output.lines: 3
df <- data.frame(a = 1:2, b = 1:2)
str(df[, "a"])             # vector
str(df[, "a", drop = FALSE]) # data frame
```

**Use `drop = FALSE` in functions** to avoid surprising simplification.

\medskip

Tibbles default to `drop = FALSE` --- they always return tibbles.

## Exercises (Selecting Multiple)

1. Fix each error:

```{r}
#| eval: false
mtcars[mtcars$cyl = 4, ]          # a
mtcars[-1:4, ]                     # b
mtcars[mtcars$cyl <= 5]            # c
mtcars[mtcars$cyl == 4 | 6, ]     # d
```

2. Why does `x[NA]` (where `x <- 1:5`) return five `NA`s?

3. What does \comm{upper.tri()} return? How does subsetting a matrix with it work?

4. Why does `mtcars[1:20]` error but `mtcars[1:20, ]` doesn't?

5. Implement your own function to extract diagonal entries from a matrix.

## Solutions (Selecting Multiple) --- 1

```{r}
#| eval: false
# a. Use == not =
mtcars[mtcars$cyl == 4, ]

# b. Use -(1:4) not -1:4
mtcars[-(1:4), ]

# c. Missing comma (selects columns, not rows)
mtcars[mtcars$cyl <= 5, ]

# d. Need full comparison on both sides
mtcars[mtcars$cyl == 4 | mtcars$cyl == 6, ]
# Or: mtcars[mtcars$cyl %in% c(4, 6), ]
```

## Solutions (Selecting Multiple) --- 2--5

2. `NA` is logical type $\to$ recycled to `c(NA, NA, NA, NA, NA)`. Each `NA` index produces `NA` output.

3. \comm{upper.tri()} returns a logical matrix (`TRUE` above diagonal). Subsetting with a logical matrix selects all `TRUE` positions as a vector.

4. Single index on data frame = column selection. `mtcars[1:20]` tries to select 20 columns, but only 11 exist. Two-index `mtcars[1:20, ]` selects rows.

5.

```{r}
#| output.lines: 3
diag2 <- function(x) {
  n <- min(nrow(x), ncol(x))
  x[cbind(seq_len(n), seq_len(n))]
}
diag2(matrix(1:9, 3))
```

# Selecting a Single Element

## `[[` --- Extract One Element

`[` returns a **sub-structure** (list $\to$ list). `[[` extracts the **element itself**:

\medskip

**Train car metaphor:**

- `x[4:6]` = a train of cars 4--6
- `x[[5]]` = the **contents** of car 5

\medskip

```{r}
#| output.lines: 5
x <- list(1:3, "a", 4:6)
str(x[1])      # list containing the vector
str(x[[1]])    # the vector itself
```

`[[` can only take a **single** positive integer or string.

\medskip

**Recursive indexing:** `x[[c(1, 2)]]` = `x[[1]][[2]]` (dig into nested structures).

## `$` --- Shorthand for Named Elements

`x$y` is roughly `x[["y"]]`, but with **partial matching**:

```{r}
#| output.lines: 5
x <- list(abc = 1, def = 2)
x$a          # matches "abc" --- dangerous!
x[["a"]]     # NULL (exact match only)
```

**Common mistake** with data frames:

```{r}
#| output.lines: 3
var <- "cyl"
# mtcars$var   # wrong! looks for column named "var"
mtcars[[var]]   # correct! uses the value of var
```

Disable partial matching: `options(warnPartialMatchDollar = TRUE)`.

## Missing and Out-of-Bounds with `[[`

| `row[[col]]` | Zero-length | OOB (int) | OOB (chr) | Missing |
|--------------|-------------|-----------|-----------|---------|
| Atomic | Error | Error | Error | Error |
| List | Error | Error | `NULL` | `NULL` |
| `NULL` | `NULL` | `NULL` | `NULL` | `NULL` |

\medskip

Use \comm{purrr::pluck()} for safe extraction with defaults:

```{r}
#| output.lines: 5
x <- list(a = list(1, 2, 3), b = list(3, 4, 5))
purrr::pluck(x, "a", 1)    # 1
purrr::pluck(x, "c", 1)    # NULL (no error)
purrr::pluck(x, "c", .default = NA)  # NA
```

## Exercises (Single Element)

1. Brainstorm as many ways as possible to extract the third value from `cyl` in `mtcars`.

2. Given `mod <- lm(mpg ~ wt, data = mtcars)`, extract the residual degrees of freedom. Then extract R-squared from \comm{summary(mod)}.

## Solutions (Single Element) --- 1

1. Many approaches:

```{r}
#| output.lines: 5
# Column first, then element
mtcars$cyl[[3]]
mtcars[["cyl"]][[3]]

# Row first, then column
mtcars[3, "cyl"]
mtcars[3, ]$cyl
```

Also: `mtcars[[c(2, 3)]]`, `with(mtcars, cyl[[3]])`.

## Solutions (Single Element) --- 2

2. Models are lists:

```{r}
#| output.lines: 3
mod <- lm(mpg ~ wt, data = mtcars)
mod$df.residual
summary(mod)$r.squared
```

# Subsetting and Assignment

## Subassignment

Combine subsetting operators with assignment:

```{r}
#| output.lines: 3
x <- 1:5
x[c(1, 2)] <- c(101, 102)
x
```

**With lists** --- `NULL` removes, `list(NULL)` adds literal NULL:

```{r}
#| output.lines: 5
x <- list(a = 1, b = 2)
x[["b"]] <- NULL     # removes b
str(x)
```

```{r}
#| output.lines: 5
y <- list(a = 1, b = 2)
y["b"] <- list(NULL)  # b exists but is NULL
str(y)
```

## Subsetting with Nothing Preserves Structure

`x[]` keeps the original structure:

```{r}
#| output.lines: 5
mtcars[] <- lapply(mtcars, as.integer)
is.data.frame(mtcars)  # TRUE --- still a data frame
```

Without `[]`:

```{r}
#| output.lines: 3
data(mtcars)  # reset
mtcars2 <- lapply(mtcars, as.integer)
is.data.frame(mtcars2)  # FALSE --- now a list!
```

**Replace NAs:** `df[is.na(df)] <- 0` replaces all `NA`s with `0`.

# Applications

## Lookup Tables

A **named vector** acts as a simple lookup table:

```{r}
#| output.lines: 3
x <- c("m", "f", "u", "f", "f", "m", "m")
lookup <- c(m = "Male", f = "Female", u = NA)
unname(lookup[x])
```

For complex lookups, use \comm{match()}:

```{r}
#| output.lines: 4
grades <- c(1, 2, 2, 3, 1)
info <- data.frame(grade = 3:1,
  desc = c("Excellent", "Good", "Poor"),
  fail = c(FALSE, FALSE, TRUE))
info[match(grades, info$grade), ]
```

For multiple columns, use \comm{merge()} or \comm{dplyr::left\_join()}.

## Random Samples and Ordering

**Random sampling** with \comm{sample()}:

```{r}
#| output.lines: 5
df <- data.frame(x = c(1, 2, 3, 1, 2), y = 5:1)
df[sample(nrow(df), 3), ]   # 3 random rows
```

**Ordering** with \comm{order()}:

```{r}
#| output.lines: 3
x <- c("b", "c", "a")
x[order(x)]
```

```{r}
#| eval: false
# Sort data frame by column
df[order(df$x), ]

# Sort columns alphabetically
df[order(names(df))]
```

## Expanding and Deduplicating

**Expanding aggregated counts** with \comm{rep()}:

```{r}
#| output.lines: 5
df <- data.frame(x = c(2, 4, 1), n = c(3, 5, 1))
df[rep(1:nrow(df), df$n), ]
```

**Deduplicating** with \comm{duplicated()} or `!duplicated()`:

```{r}
#| eval: false
df[!duplicated(df$x), ]
```

## Removing Columns and Filtering Rows

**Remove columns** --- three ways:

```{r}
#| eval: false
df$z <- NULL                      # set to NULL
df[c("x", "y")]                   # keep wanted
df[setdiff(names(df), "z")]       # drop unwanted
```

**Filter rows** with logical subsetting:

```{r}
#| eval: false
mtcars[mtcars$gear == 5, ]
mtcars[mtcars$gear == 5 & mtcars$cyl == 4, ]
```

Use `&` and `|` (not `&&` and `||`).

\medskip

**De Morgan's laws:** `!(X & Y)` = `!X | !Y` and `!(X | Y)` = `!X & !Y`.

## Boolean Algebra vs. Sets

| Boolean (logical) | Set (integer via \comm{which()}) |
|-------------------|-------------------------------|
| `x & y` | `intersect(x, y)` |
| `x \| y` | `union(x, y)` |
| `x & !y` | `setdiff(x, y)` |

\comm{which()} converts logical to integer indices:

```{r}
#| output.lines: 3
x <- c(FALSE, TRUE, FALSE, TRUE, TRUE)
which(x)
```

**Key difference:** logical subsetting preserves `NA`s; \comm{which()} drops them.

Prefer integer subsetting when you need the first/last `TRUE` or have very few `TRUE`s in a long vector.

## Exercises (Applications)

1. How would you randomly permute the columns of a data frame? Can you permute rows and columns simultaneously?

2. How would you select a random sample of `m` rows? What about a contiguous sample?

3. How could you put columns in alphabetical order?

## Solutions (Applications)

1.

```{r}
#| eval: false
mtcars[sample(ncol(mtcars))]         # permute cols
mtcars[sample(nrow(mtcars)),
  sample(ncol(mtcars))]              # both
```

2. Random sample: `mtcars[sample(nrow(mtcars), m), ]`. Contiguous:

```{r}
#| eval: false
start <- sample(nrow(mtcars) - m + 1, 1)
mtcars[start:(start + m - 1), ]
```

3. `mtcars[order(names(mtcars))]` or `mtcars[sort(names(mtcars))]`.

<!--
# Summary

## Key Takeaways

- **Six subsetting modes:** positive, negative, logical, nothing, zero, character
- **`[`** selects sub-structures (list $\to$ list, df $\to$ df)
- **`[[`** extracts a single element (digs into the structure)
- **`$`** is shorthand for `[[` with partial matching (beware!)
- **`drop = FALSE`:** preserves dimensionality; use in functions
- **Tibbles** always return tibbles (no silent simplification)
- **Subassignment:** `x[i] <- value` modifies in place
- **`x[]` vs `x`:** subsetting with nothing preserves structure
- **Lookup tables:** named vectors + character subsetting
- **Sampling:** `sample()` + integer subsetting
- **Ordering:** `order()` returns indices, not sorted values
- **Filtering:** logical subsetting with `&`/`|` (not `&&`/`||`)
- **\comm{which()}:** converts logical to integer; drops `NA`s
-->
