---
title: "R4DS 13 -- Functions"
---

```{r setup, include=FALSE}
source("_common.R")
library(tidyverse)
library(nycflights13)
```

# Vector Functions

## Why Write Functions?

Spot the copy-paste bug:

```{r}
#| eval: false
df |> mutate(
  a = (a - min(a, na.rm = TRUE)) /
    (max(a, na.rm = TRUE) - min(a, na.rm = TRUE)),
  b = (b - min(a, na.rm = TRUE)) /   # bug: a!
    (max(b, na.rm = TRUE) - min(b, na.rm = TRUE)),
  c = (c - min(c, na.rm = TRUE)) /
    (max(c, na.rm = TRUE) - min(c, na.rm = TRUE)))
```

**Rule of thumb:** write a function when you've copied code more than twice.

Three useful types: **vector** (for `mutate`/`filter`), **data frame**, and **plot** functions.

## Writing a Function

**Template:** `name <- function(arguments) { body }`

\medskip

**Example** --- rescale to $[0, 1]$:

```{r}
#| output.lines: 4
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
rescale01(c(-10, 0, 10))
rescale01(c(1, 2, 3, NA, 5))
```

Now the repetitive code becomes clean:

```{r}
#| eval: false
df |> mutate(a = rescale01(a), b = rescale01(b),
             c = rescale01(c), d = rescale01(d))
```

## Improving: Handle Edge Cases

Infinite values cause problems:

```{r}
#| output.lines: 3
rescale01(c(1:10, Inf))
```

Fix: add \comm{finite = TRUE} to \comm{range()}:

```{r}
#| output.lines: 4
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
rescale01(c(1:10, Inf))
```

**Strategy:** get the basics right, then handle edge cases iteratively.

## Mutate Functions

Return same-length output --- use in \comm{mutate()} and \comm{filter()}:

```{r}
#| output.lines: 3
clamp <- function(x, min, max) {
  case_when(x < min ~ min, x > max ~ max,
            .default = x)
}
clamp(1:10, min = 3, max = 7)
```

```{r}
#| output.lines: 3
first_upper <- function(x) {
  str_sub(x, 1, 1) <-
    str_to_upper(str_sub(x, 1, 1))
  x
}
first_upper("hello")
```

Other useful patterns: \comm{z\_score()}, \comm{clean\_number()}, \comm{fix\_na()}.

## Summary Functions

Return a single value --- use in \comm{summarize()}:

```{r}
#| output.lines: 3
commas <- function(x) {
  str_flatten(x, collapse = ", ",
              last = " and ")
}
commas(c("cat", "dog", "pigeon"))
```

Pass \comm{na.rm} through to the underlying function:

```{r}
#| eval: false
cv <- function(x, na.rm = FALSE) {
  sd(x, na.rm = na.rm) / mean(x, na.rm = na.rm)
}

n_missing <- function(x) sum(is.na(x))

mape <- function(actual, predicted) {
  sum(abs((actual - predicted) / actual)) /
    length(actual)
}
```

## Exercises (Vector Functions)

1. Turn these into functions. What would you name each?

```{r}
#| eval: false
mean(is.na(x))           # snippet 1
x / sum(x, na.rm = TRUE) # snippet 2
round(x / sum(x, na.rm = TRUE) * 100, 1)
```

2. Rewrite \comm{rescale01()} so \comm{-Inf} $\to$ 0 and \comm{Inf} $\to$ 1.
3. Write a function to compute age in years from a vector of birthdates.
4. Write functions for variance and skewness of a numeric vector.
5. Write \comm{both\_na()}: takes two vectors, returns count of positions with \comm{NA} in both.
6. What do these do? Why are they useful despite being short?

```{r}
#| eval: false
is_directory <- function(x) file.info(x)$isdir
is_readable <- function(x) file.access(x, 4) == 0
```

## Solutions (Vector Functions) --- 1--3

1.

```{r}
#| eval: false
prop_na <- function(x) mean(is.na(x))
prop_of_total <- function(x)
  x / sum(x, na.rm = TRUE)
pct_of_total <- function(x)
  round(x / sum(x, na.rm = TRUE) * 100, 1)
```

2.

```{r}
#| output.lines: 3
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  y <- (x - rng[1]) / (rng[2] - rng[1])
  y[x == -Inf] <- 0; y[x == Inf] <- 1; y
}
rescale01(c(-Inf, 1:5, Inf))
```

3.

```{r}
#| eval: false
age_years <- function(birthdate) {
  as.numeric(difftime(Sys.Date(), birthdate,
    units = "days")) / 365.25
}
```

## Solutions (Vector Functions) --- 4--6

4.

```{r}
#| eval: false
variance <- function(x, na.rm = FALSE) {
  if (na.rm) x <- x[!is.na(x)]
  n <- length(x)
  sum((x - mean(x))^2) / (n - 1)
}
skewness <- function(x, na.rm = FALSE) {
  if (na.rm) x <- x[!is.na(x)]
  n <- length(x); m <- mean(x); s <- sd(x)
  (n / ((n-1) * (n-2))) * sum(((x-m) / s)^3)
}
```

5.

```{r}
#| output.lines: 3
both_na <- function(x, y) {
  sum(is.na(x) & is.na(y))
}
both_na(c(1, NA, 3, NA), c(NA, NA, 3, NA))
```

6. They wrap low-level system calls into readable names. Short functions are valuable when they **improve readability**.

# Data Frame Functions

## The Indirection Problem

This fails --- dplyr looks for a column literally named \comm{group\_var}:

```{r}
#| eval: false
grouped_mean <- function(df, group_var, mean_var) {
  df |>
    group_by(group_var) |>
    summarize(mean(mean_var))
}
diamonds |> grouped_mean(cut, carat)
#> Error: column `group_var` not found
```

The issue: dplyr uses **tidy evaluation** --- it looks for column names in the data, not in R's environment.

## Embracing with `{{ }}`

**Solution:** wrap arguments in \comm{\{\{ \}\}} ("embrace") to tell dplyr to look inside:

```{r}
#| output.lines: 7
#| message: false
grouped_mean <- function(df, group_var, mean_var) {
  df |>
    group_by({{ group_var }}) |>
    summarize(mean({{ mean_var }}))
}
diamonds |> grouped_mean(cut, carat)
```

**Rule:** if an argument is passed to a dplyr verb, embrace it with \comm{\{\{ \}\}}.

## Common Use Cases

Flexible summary function:

```{r}
#| output.lines: 5
summary6 <- function(data, var) {
  data |> summarize(
    min = min({{ var }}, na.rm = TRUE),
    mean = mean({{ var }}, na.rm = TRUE),
    median = median({{ var }}, na.rm = TRUE),
    max = max({{ var }}, na.rm = TRUE),
    n = n(), n_miss = sum(is.na({{ var }})),
    .groups = "drop")
}
diamonds |> summary6(carat)
```

Works on grouped data and with expressions:

```{r}
#| eval: false
diamonds |> group_by(cut) |> summary6(carat)
diamonds |> group_by(cut) |> summary6(log10(carat))
```

## More Use Cases

Count with proportions:

```{r}
#| output.lines: 5
count_prop <- function(df, var, sort = FALSE) {
  df |>
    count({{ var }}, sort = sort) |>
    mutate(prop = n / sum(n))
}
diamonds |> count_prop(clarity)
```

Unique values under a condition:

```{r}
#| output.lines: 4
unique_where <- function(df, condition, var) {
  df |> filter({{ condition }}) |>
    distinct({{ var }}) |> arrange({{ var }})
}
flights |> unique_where(month == 12, dest)
```

## Data-Masking vs. Tidy-Selection

Two types of tidy evaluation:

- **Data-masking** (\comm{filter()}, \comm{mutate()}, \comm{summarize()}): works with expressions like \comm{x + 1}
- **Tidy-selection** (\comm{select()}, \comm{rename()}, \comm{across()}): works with selections like \comm{starts\_with("x")}

\medskip

Problem: \comm{group\_by()} is data-masking, but you want to pass multiple variables. Use \comm{pick()}:

```{r}
#| output.lines: 5
count_missing <- function(df, group_vars, x_var) {
  df |>
    group_by(pick({{ group_vars }})) |>
    summarize(n_miss = sum(is.na({{ x_var }})),
              .groups = "drop")
}
flights |>
  count_missing(c(year, month, day), dep_time)
```

## Exercises (Data Frame Functions)

1. Using \comm{nycflights13}, write functions that:
   a. Find cancelled or >1 hour delayed flights
   b. Count cancelled and >1 hour delayed by group
   c. Like (a) but with user-supplied hours
   d. Summarize weather: min, mean, max of any variable
   e. Convert clock time to decimal hours

2. Which arguments use tidy evaluation in: \comm{distinct()}, \comm{count()}, \comm{group\_by()}, \comm{rename\_with()}, \comm{slice\_min()}, \comm{slice\_sample()}?

3. Generalize \comm{count\_prop()} to accept any number of variables.

## Solutions (Data Frame Functions) --- 1a--c

```{r}
#| eval: false
filter_severe <- function(df, hours = 1) {
  df |> filter(is.na(arr_time) |
    dep_delay > hours * 60)
}

summarize_severe <- function(df) {
  df |> summarize(
    n_cancelled = sum(is.na(arr_time)),
    n_delayed = sum(dep_delay > 60, na.rm = TRUE))
}
```

\comm{hours} parameter in \comm{filter\_severe()} handles both 1a and 1c.

## Solutions (Data Frame Functions) --- 1d--e

```{r}
#| eval: false
summarize_weather <- function(df, var) {
  df |> summarize(
    min = min({{ var }}, na.rm = TRUE),
    mean = mean({{ var }}, na.rm = TRUE),
    max = max({{ var }}, na.rm = TRUE))
}

standardize_time <- function(df, var) {
  df |> mutate({{ var }} :=
    ({{ var }} %/% 100) + ({{ var }} %% 100) / 60)
}
```

Note: \comm{standardize\_time()} uses the walrus operator \comm{:=} to modify the column in place.

## Solutions (Data Frame Functions) --- 2--3

2. **Data-masking:** \comm{distinct(...)}, \comm{count(...)}, \comm{group\_by(...)}, \comm{slice\_min(order\_by)}. **Tidy-selection:** \comm{rename\_with(.cols)}. \comm{slice\_sample()} uses neither for column selection.

3. Use \comm{...} instead of a single \comm{var}:

```{r}
#| output.lines: 6
count_prop <- function(df, ..., sort = FALSE) {
  df |>
    count(..., sort = sort) |>
    mutate(prop = n / sum(n))
}
diamonds |> count_prop(clarity, cut)
```

# Plot Functions

## Building Plot Functions

Same \comm{\{\{ \}\}} pattern --- embracing works inside \comm{aes()}:

```{r}
#| eval: false
histogram <- function(df, var, binwidth = NULL) {
  df |>
    ggplot(aes(x = {{ var }})) +
    geom_histogram(binwidth = binwidth)
}

diamonds |> histogram(carat, 0.1)

# Can still add layers:
diamonds |> histogram(carat, 0.1) +
  labs(x = "Size (in carats)")
```

## More Variables and Labels

```{r}
#| eval: false
linearity_check <- function(df, x, y) {
  df |> ggplot(aes(x = {{ x }}, y = {{ y }})) +
    geom_point() +
    geom_smooth(method = "loess", formula = y ~ x,
      color = "red", se = FALSE) +
    geom_smooth(method = "lm", formula = y ~ x,
      color = "blue", se = FALSE)
}
starwars |> filter(mass < 1000) |>
  linearity_check(mass, height)
```

Dynamic labels with \comm{rlang::englue()} --- \comm{\{\{var\}\}} inserts the variable name:

```{r}
#| eval: false
histogram <- function(df, var, binwidth) {
  label <- rlang::englue(
    "A histogram of {{var}} with binwidth {binwidth}")
  df |> ggplot(aes(x = {{ var }})) +
    geom_histogram(binwidth = binwidth) +
    labs(title = label)
}
```

## Combining with Tidyverse

Use the **walrus operator** \comm{:=} to create variable names with \comm{\{\{ \}\}} on the left side:

```{r}
#| eval: false
sorted_bars <- function(df, var) {
  df |>
    mutate({{ var }} :=
      fct_rev(fct_infreq({{ var }}))) |>
    ggplot(aes(y = {{ var }})) +
    geom_bar()
}
diamonds |> sorted_bars(clarity)
```

\comm{:=} works like \comm{=} but allows embracing on the **left** side of assignment.

## Exercises (Plot Functions)

Build up a rich plotting function incrementally:

1. Draw a scatterplot given a dataset and \comm{x} and \comm{y} variables.
2. Add a line of best fit (linear model, no standard errors).
3. Add a title that includes the variable names.

## Solutions (Plot Functions)

Build incrementally --- final version includes all three:

```{r}
#| eval: false
# 1. Start: scatterplot
scatter <- function(df, x, y) {
  df |> ggplot(aes(x = {{ x }}, y = {{ y }})) +
    geom_point()
}

# 2+3. Add line of best fit + dynamic title
scatter <- function(df, x, y) {
  df |> ggplot(aes(x = {{ x }}, y = {{ y }})) +
    geom_point() +
    geom_smooth(method = "lm", formula = y ~ x,
      se = FALSE) +
    labs(title = rlang::englue("{{x}} vs {{y}}"))
}
```

# Style

## Function Naming and Formatting

**Names:** use verbs, be descriptive:

```{r}
#| eval: false
# Too vague           # Better
f()                    impute_missing()
my_function()          collapse_years()
```

\medskip

**Formatting:** extra indent inside function body:

```{r}
#| eval: false
# Good --- body indented extra 2 spaces
density <- function(color, facets,
                    binwidth = 0.1) {
  diamonds |>
    ggplot(aes(x = carat,
      y = after_stat(density),
      color = {{ color }})) +
    geom_freqpoly(binwidth = binwidth) +
    facet_wrap(vars({{ facets }}))
}
```

## Exercises (Style)

1. What do these do? Suggest better names:

```{r}
#| eval: false
f1 <- function(string, prefix) {
  str_sub(string, 1, str_length(prefix)) == prefix
}
f3 <- function(x, y) {
  rep(y, length.out = length(x))
}
```

2. Make a case for \comm{norm\_r()}, \comm{norm\_d()} over \comm{rnorm()}, \comm{dnorm()}. Then make the opposite case.

## Solutions (Style)

1. \comm{f1} checks if a string starts with a given prefix $\to$ \comm{has\_prefix()} or \comm{str\_has\_prefix()}. \comm{f3} repeats \comm{y} to match the length of \comm{x} $\to$ \comm{match\_length()} or \comm{recycle()}.

2. **For \comm{norm\_r()}:** common prefix groups all normal functions together --- autocomplete shows \comm{norm\_...}. Aids discovery. **Against:** R's convention (\comm{rnorm}, \comm{dnorm}, \comm{pnorm}) groups by **operation** across distributions (\comm{r...} = random, \comm{d...} = density). Muscle memory and decades of convention matter.

<!--
# Summary

## Key Takeaways

- **Rule of thumb:** write a function after copying code more than twice
- **Template:** `name <- function(args) { body }`
- **Vector functions:** same-length output for \comm{mutate()}/\comm{filter()}; single value for \comm{summarize()}
- **Iterative improvement:** get basics right, then handle edge cases (\comm{NA}, \comm{Inf})
- **Data frame functions:** use \comm{\{\{ \}\}} (embracing) for arguments passed to dplyr verbs
- **Tidy evaluation:** data-masking (expressions) vs. tidy-selection (columns)
- **\comm{pick()}:** bridges tidy-selection inside data-masking functions
- **Plot functions:** \comm{\{\{ \}\}} works inside \comm{aes()}; return ggplot objects for \comm{+} layering
- **Labeling:** \comm{rlang::englue()} for dynamic titles with variable names
- **Walrus operator:** \comm{:=} allows \comm{\{\{ \}\}} on the left side
- **Style:** verb names, descriptive arguments, extra indent in function body
-->
