---
title: "R4DS 05 -- Layers"
---

```{r setup, include=FALSE}
source("_common.R")
library(tidyverse)
```

# Aesthetic Mappings

## Mapping vs. Setting Aesthetics

**Map** a variable to an aesthetic inside \comm{aes()}:

```{r}
#| fig.height: 2.5
#| warning: false
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

## Shape Aesthetic

```{r}
#| fig.height: 2.5
#| warning: false
ggplot(mpg, aes(x = displ, y = hwy, shape = class)) +
  geom_point()
```

Only 6 shapes available by default --- SUVs (7th group) go unplotted.

## Size and Alpha

```{r}
#| eval: false
# Size: implies ranking --- not ideal for unordered categories
ggplot(mpg, aes(x = displ, y = hwy, size = class)) +
  geom_point()

# Alpha: same issue
ggplot(mpg, aes(x = displ, y = hwy, alpha = class)) +
  geom_point()
```

Mapping unordered categorical variables to ordered aesthetics (\comm{size}, \comm{alpha}) implies a ranking that doesn't exist.

## Setting Aesthetics (Outside `aes()`)

```{r}
#| fig.height: 2.5
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "blue")
```

Fixed values go **outside** \comm{aes()}: \comm{color = "blue"}, \comm{size = 3}, \comm{shape = 17}.

## Exercises (Aesthetics)

1. Scatterplot of \comm{hwy} vs. \comm{displ} with pink filled-in triangles.
2. Why doesn't this produce blue points?

```{r}
#| eval: false
ggplot(mpg) +
  geom_point(aes(x = displ, y = hwy, color = "blue"))
```

3. What does the \comm{stroke} aesthetic do? Which shapes does it work with?
4. What happens with \comm{aes(color = displ < 5)}?

## Solutions (Aesthetics)

1.
```{r}
#| fig.height: 2
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "pink", shape = 24, fill = "pink")
```

## Solutions (Aesthetics) --- cont.

2. \comm{color = "blue"} is **inside** \comm{aes()}, so ggplot maps the string \comm{"blue"} as a categorical variable (red by default). Move it outside: \comm{geom\_point(color = "blue")}.

3. \comm{stroke} controls border thickness for shapes 21--25 (filled shapes with separate border and fill).

4. Creates a logical variable (\comm{TRUE}/\comm{FALSE}), coloring points by whether \comm{displ < 5}.

```{r}
#| fig.height: 1.8
ggplot(mpg, aes(x = displ, y = hwy, color = displ < 5)) +
  geom_point()
```

# Geometric Objects

## Different Geoms, Same Data

```{r}
#| eval: false
# Scatterplot
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()

# Smooth curve
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth()
```

Same data + different geom = different plot. Every geom has its own set of aesthetics.

## Linetype Aesthetic

```{r}
#| fig.height: 2.5
#| message: false
ggplot(mpg, aes(x = displ, y = hwy, linetype = drv)) +
  geom_smooth()
```

\comm{geom\_smooth()} uses \comm{linetype} (not \comm{shape}) to distinguish groups.

## Grouping

```{r}
#| fig.height: 2.2
#| message: false
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(aes(group = drv))
```

\comm{group} splits data without adding a legend. Mapping to \comm{color} or \comm{linetype} adds a legend automatically.

## Local vs. Global Aesthetics

```{r}
#| fig.height: 2.2
#| message: false
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth()
```

\comm{color = class} in \comm{geom\_point()} only $\to$ one overall smooth line.

## Different Data per Layer

```{r}
#| fig.height: 2.2
#| message: false
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_point(
    data = mpg |> filter(class == "2seater"),
    color = "red", size = 3
  )
```

Override \comm{data} in a layer to highlight a subset.

## Geom Gallery

| Geom | Plot Type | Key Aesthetic |
|------|-----------|---------------|
| `geom_point()` | Scatterplot | `shape` |
| `geom_smooth()` | Smooth curve | `linetype` |
| `geom_bar()` | Bar chart | `fill` |
| `geom_histogram()` | Histogram | `binwidth` |
| `geom_density()` | Density | `fill`, `alpha` |
| `geom_boxplot()` | Boxplot | `fill` |
| `geom_line()` | Line chart | `linetype` |
| `geom_jitter()` | Jittered scatter | `width`, `height` |

40+ geoms available; extension packages add more.

## Exercises (Geoms)

1. What geom for: line chart? boxplot? histogram? area chart?
2. What does \comm{show.legend = FALSE} do? What if you remove it?
3. What does the \comm{se} argument to \comm{geom\_smooth()} do?
4. Recreate these six plots (categorical variable is \comm{drv}):

```{r}
#| eval: false
# (a) points + smooth
# (b) grouped smooth + points
# (c) color = drv on both geoms
# (d) color = drv on points only, one smooth
# (e) color = drv on points, linetype = drv on smooth
# (f) white border points (layer trick)
```

## Solutions (Geoms) --- 1--3

1. \comm{geom\_line()}, \comm{geom\_boxplot()}, \comm{geom\_histogram()}, \comm{geom\_area()}.

2. \comm{show.legend = FALSE} hides the legend for that layer. Removing it shows the legend. Used to keep plots clean when legend is redundant.

3. \comm{se = FALSE} removes the confidence interval ribbon around the smooth curve.

## Solutions (Geoms) --- 4

```{r}
#| eval: false
# (a)
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() + geom_smooth(se = FALSE)
# (b)
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(aes(group = drv), se = FALSE) + geom_point()
# (c)
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point() + geom_smooth(se = FALSE)
# (d)
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) + geom_smooth(se = FALSE)
# (e)
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) +
  geom_smooth(aes(linetype = drv), se = FALSE)
# (f)
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(size = 4, color = "white") +
  geom_point(aes(color = drv))
```

# Facets

## `facet_wrap()` --- One Variable

```{r}
#| fig.height: 2.5
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~cyl)
```

Splits plot into subplots by one categorical variable.

## `facet_grid()` --- Two Variables

```{r}
#| fig.height: 2.5
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ cyl)
```

\comm{rows \~{} cols}: drive train in rows, cylinders in columns.

## Free Scales

```{r}
#| fig.height: 2.5
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ cyl, scales = "free")
```

\comm{scales = "free"} lets each panel have its own axis range. Also: \comm{"free\_x"}, \comm{"free\_y"}.

## Exercises (Facets)

1. What happens if you facet on a continuous variable?
2. What do empty cells in \comm{facet\_grid(drv \~{} cyl)} mean?
3. What does \comm{.} do in \comm{facet\_grid(drv \~{} .)} and \comm{facet\_grid(. \~{} cyl)}?
4. Faceting vs. color aesthetic: advantages/disadvantages? Effect of dataset size?
5. Read \comm{?facet\_wrap}. What do \comm{nrow}/\comm{ncol} do? Why doesn't \comm{facet\_grid()} have them?
6. When placing a faceting variable across rows vs. columns, which makes comparison easier?
7. Recreate \comm{facet\_grid(drv \~{} .)} using \comm{facet\_wrap()}. How do labels change?

## Solutions (Facets) --- 1--4

1. One facet per unique value --- many panels, often not useful.

2. Empty cells = no observations for that combination (e.g., no 5-cyl front-wheel-drive cars).

3. \comm{.} means "don't facet" on that dimension. \comm{drv \~{} .} = facet rows only; \comm{. \~{} cyl} = facet columns only.

4. Faceting avoids overplotting and is easier to read for many groups. Color is better for direct comparison (overlaid). With large data, faceting handles many groups better; color becomes cluttered.

## Solutions (Facets) --- 5--7

5. \comm{nrow}/\comm{ncol} control the layout grid for \comm{facet\_wrap()}. \comm{facet\_grid()} doesn't need them --- rows and columns are determined by the variables.

6. Facet rows share the $x$-axis $\to$ easier to compare horizontal distributions. Facet columns share the $y$-axis $\to$ easier to compare vertical distributions.

7.

```{r}
#| eval: false
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~drv, ncol = 1)
```

Labels move from the right side (strip labels in \comm{facet\_grid}) to the top.

# Statistical Transformations

## Stats Behind Geoms

```{r}
#| fig.height: 2.5
ggplot(diamonds, aes(x = cut)) +
  geom_bar()
```

The $y$-axis shows \comm{count} --- but \comm{count} isn't in the data! \comm{geom\_bar()} uses \comm{stat\_count()} internally.

## `geom_bar()` vs. `geom_col()`

```{r}
#| eval: false
# geom_bar(): stat = "count" (counts rows)
ggplot(diamonds, aes(x = cut)) +
  geom_bar()

# geom_col(): stat = "identity" (uses y values directly)
diamonds |>
  count(cut) |>
  ggplot(aes(x = cut, y = n)) +
  geom_col()
```

Use \comm{geom\_bar()} for counting; \comm{geom\_col()} when you already have summary values.

## Proportion Bar Chart

```{r}
#| fig.height: 2.5
ggplot(diamonds, aes(x = cut, y = after_stat(prop), group = 1)) +
  geom_bar()
```

\comm{after\_stat(prop)} accesses computed proportions. \comm{group = 1} forces proportions across the whole dataset.

## `stat_summary()`

```{r}
#| fig.height: 2.5
ggplot(diamonds) +
  stat_summary(
    aes(x = cut, y = depth),
    fun.min = min, fun.max = max, fun = median
  )
```

Use stats directly to emphasize the transformation. Default geom: \comm{pointrange}.

## Exercises (Stats)

1. Default geom for \comm{stat\_summary()}? Rewrite the previous plot using that geom.
2. What does \comm{geom\_col()} do? How is it different from \comm{geom\_bar()}?
3. List geom-stat pairs. What do they have in common?
4. What variables does \comm{stat\_smooth()} compute? What controls its behavior?
5. Why do we need \comm{group = 1} in the proportion bar chart?

## Solutions (Stats)

1. Default geom is \comm{geom\_pointrange()}. Rewrite:

```{r}
#| eval: false
ggplot(diamonds, aes(x = cut, y = depth)) +
  geom_pointrange(stat = "summary",
    fun.min = min, fun.max = max, fun = median)
```

2. \comm{geom\_col()} uses \comm{stat = "identity"} (bars from data values); \comm{geom\_bar()} uses \comm{stat = "count"} (counts rows).

3. Common pairs: \comm{geom\_bar}/\comm{stat\_count}, \comm{geom\_histogram}/\comm{stat\_bin}, \comm{geom\_smooth}/\comm{stat\_smooth}, \comm{geom\_boxplot}/\comm{stat\_boxplot}. Each pair shares a default.

## Solutions (Stats) --- cont.

4. \comm{stat\_smooth()} computes \comm{y} (predicted), \comm{ymin}, \comm{ymax}, \comm{se}. Controlled by \comm{method}, \comm{formula}, \comm{se}, \comm{n}, \comm{span}.

5. Without \comm{group = 1}, proportions are computed **within each group** (each bar = 100\%). With \comm{group = 1}, proportions are relative to the total.

# Position Adjustments

## Bar Chart Fill

```{r}
#| fig.height: 2.5
ggplot(mpg, aes(x = drv, fill = class)) +
  geom_bar()
```

Mapping \comm{fill} to a variable $\to$ stacked bars automatically.

## Position Options for Bars

```{r}
#| eval: false
# Stacked (default)
ggplot(mpg, aes(x = drv, fill = class)) + geom_bar()

# Proportional
ggplot(mpg, aes(x = drv, fill = class)) +
  geom_bar(position = "fill")

# Side-by-side
ggplot(mpg, aes(x = drv, fill = class)) +
  geom_bar(position = "dodge")
```

- \comm{"fill"}: scales bars to 100\% (compare proportions)
- \comm{"dodge"}: places bars side by side (compare counts)

## `position = "fill"` Example

```{r}
#| fig.height: 2.5
ggplot(mpg, aes(x = drv, fill = class)) +
  geom_bar(position = "fill") +
  labs(y = "proportion")
```

## `position = "dodge"` Example

```{r}
#| fig.height: 2.5
ggplot(mpg, aes(x = drv, fill = class)) +
  geom_bar(position = "dodge")
```

## Overplotting and Jitter

```{r}
#| fig.height: 2
#| layout-ncol: 2
#| fig-width: 4
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_jitter()
```

\comm{geom\_jitter()} adds random noise to reveal overlapping points. Shorthand for \comm{geom\_point(position = "jitter")}.

## Exercises (Position)

1. What's the problem with this plot? How to improve it?

```{r}
#| eval: false
ggplot(mpg, aes(x = cty, y = hwy)) + geom_point()
```

2. Difference between \comm{geom\_point()} and \comm{geom\_point(position = "identity")}?
3. What parameters to \comm{geom\_jitter()} control the amount of jittering?
4. Compare \comm{geom\_jitter()} and \comm{geom\_count()}.
5. Default position for \comm{geom\_boxplot()}?

## Solutions (Position)

1. Overplotting --- many points overlap. Fix: use \comm{geom\_jitter()} or set \comm{alpha}.

2. No difference --- \comm{"identity"} is the default position for \comm{geom\_point()}.

3. \comm{width} and \comm{height} control horizontal and vertical jitter amount.

4. \comm{geom\_jitter()} adds random noise; \comm{geom\_count()} maps point size to the number of overlapping observations. \comm{geom\_count()} preserves exact positions.

5. \comm{position = "dodge2"}. Example:

```{r}
#| fig.height: 2
ggplot(mpg, aes(x = drv, y = hwy, fill = factor(cyl))) +
  geom_boxplot()
```

# Coordinate Systems

## `coord_flip()` and `coord_fixed()`

```{r}
#| eval: false
# Flip axes (useful for long category names)
ggplot(mpg, aes(x = class, y = hwy)) +
  geom_boxplot() + coord_flip()

# Fixed aspect ratio (1:1 scale)
ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point() + geom_abline() + coord_fixed()
```

## `coord_polar()` --- Polar Coordinates

```{r}
#| fig.height: 1.8
#| layout-ncol: 2
#| fig-width: 4
bar <- ggplot(diamonds) +
  geom_bar(aes(x = clarity, fill = clarity),
           show.legend = FALSE, width = 1)

bar + coord_flip()
bar + coord_polar()
```

Polar coordinates turn a bar chart into a Coxcomb chart.

## Exercises (Coordinates)

1. Turn a stacked bar chart into a pie chart using \comm{coord\_polar()}.
2. Difference between \comm{coord\_quickmap()} and \comm{coord\_map()}?
3. What does this plot tell you? Why is \comm{coord\_fixed()} important? What does \comm{geom\_abline()} do?

```{r}
#| eval: false
ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point() + geom_abline() + coord_fixed()
```

## Solutions (Coordinates)

1.

```{r}
#| fig.height: 1.5
ggplot(diamonds, aes(x = "", fill = cut)) +
  geom_bar(width = 1) + coord_polar(theta = "y") +
  theme(legend.key.size = unit(0.3, "cm"))
```

## Solutions (Coordinates) --- cont.

2. \comm{coord\_map()} uses Mercator (or other) projection (more accurate, slower). \comm{coord\_quickmap()} uses a fast approximation that works well for small areas.

3. \comm{hwy} is always higher than \comm{cty}. \comm{coord\_fixed()} ensures a 1:1 aspect ratio so the 45-degree reference line (\comm{geom\_abline()}) is visually meaningful.

# The Layered Grammar

## The Full ggplot2 Template

```{r}
#| eval: false
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(
    mapping = aes(<MAPPINGS>),
    stat = <STAT>,
    position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```

Seven parameters: data, mappings, geom, stat, position, coordinate system, facets.

In practice, ggplot2 provides sensible defaults for everything except data, mappings, and geom.

<!--
# Summary

## Key Takeaways

- **Aesthetics:** \comm{aes()} maps variables; fixed values set outside \comm{aes()}
- **Geoms:** each geom has its own aesthetics; layer multiple geoms with \comm{+}
- **Local vs. global:** \comm{ggplot(aes(...))} inherited; \comm{geom\_*(aes(...))} local only
- **Facets:** \comm{facet\_wrap(\~var)} for one variable; \comm{facet\_grid(rows \~{} cols)} for two
- **Stats:** geoms compute behind the scenes (\comm{geom\_bar} $\to$ \comm{stat\_count})
- **\comm{geom\_bar()} vs. \comm{geom\_col()}:** counting vs. pre-computed values
- **Position:** \comm{"stack"}, \comm{"fill"}, \comm{"dodge"}, \comm{"jitter"}
- **Coordinates:** \comm{coord\_polar()}, \comm{coord\_fixed()}, \comm{coord\_flip()}
-->
