---
title: "R4DS 08 -- Logicals & Numbers"
---

```{r setup, include=FALSE}
source("_common.R")
library(tidyverse)
library(nycflights13)
```

# Comparisons

## Numeric Comparisons

Create logical vectors with `<`, `<=`, `>`, `>=`, `!=`, `==`:

```{r}
#| output.lines: 5
flights |>
  mutate(
    daytime = dep_time > 600 & dep_time < 2000,
    approx_ontime = abs(arr_delay) < 20,
    .keep = "used"
  )
```

Naming intermediate logical steps makes complex filters easier to read and debug.

## Floating Point Comparison

Beware of `==` with computed numbers:

```{r}
x <- c(1 / 49 * 49, sqrt(2) ^ 2)
x == c(1, 2)
```

```{r}
print(x, digits = 16)
```

Use \comm{dplyr::near()} to handle floating point issues:

```{r}
near(x, c(1, 2))
```

## Missing Values in Comparisons

\comm{NA} is contagious --- any operation involving \comm{NA} returns \comm{NA}:

```{r}
NA > 5
10 == NA
NA == NA
```

\medskip

Use \comm{is.na()} to test for missing values:

```{r}
is.na(c(TRUE, NA, FALSE))
is.na(c(1, NA, 3))
```

## Using `is.na()` with Data

```{r}
#| output.lines: 5
# Find cancelled flights
flights |> filter(is.na(dep_time))
```

\medskip

Sort NAs first with \comm{arrange()}:

```{r}
#| eval: false
flights |>
  filter(month == 1, day == 1) |>
  arrange(desc(is.na(dep_time)), dep_time)
```

## Exercises (Comparisons)

1. How does \comm{dplyr::near()} work? Type \comm{near} to see the source. Is \comm{sqrt(2)\^{}2} near 2?
2. Use \comm{mutate()}, \comm{is.na()}, and \comm{count()} to describe how \comm{dep\_time}, \comm{sched\_dep\_time}, and \comm{dep\_delay} missing values are connected.

## Solutions (Comparisons)

1. \comm{near()} checks if \comm{abs(x - y) < tol}, where \comm{tol} defaults to \comm{.Machine\$double.eps\^{}0.5}. Yes, \comm{sqrt(2)\^{}2} is near 2.

2.

```{r}
#| output.lines: 5
flights |>
  mutate(
    dep_na = is.na(dep_time),
    sched_na = is.na(sched_dep_time),
    delay_na = is.na(dep_delay)
  ) |>
  count(dep_na, sched_na, delay_na)
```

When \comm{dep\_time} is NA, \comm{dep\_delay} is also NA. \comm{sched\_dep\_time} is never missing.

# Boolean Algebra

## Boolean Operators

```{r}
#| eval: false
# & = AND, | = OR, ! = NOT, xor() = exclusive or
df |> filter(!is.na(x))         # x is not missing
df |> filter(x < -10 | x > 0)  # x outside [-10, 0]
```

\medskip

NAs in Boolean algebra follow logic rules:

```{r}
#| output.lines: 6
df <- tibble(x = c(TRUE, FALSE, NA))
df |> mutate(and = x & NA, or = x | NA)
```

\comm{TRUE | NA} is \comm{TRUE} (at least one is TRUE). \comm{FALSE \& NA} is \comm{FALSE} (at least one is FALSE).

## Order of Operations and `%in%`

Common mistake --- this does NOT filter Nov/Dec:

```{r}
#| eval: false
flights |> filter(month == 11 | 12)  # WRONG!
# R evaluates: (month == 11) | 12 → always TRUE
```

Correct approaches:

```{r}
#| eval: false
flights |> filter(month == 11 | month == 12)
flights |> filter(month %in% c(11, 12))  # preferred
```

\medskip

\comm{\%in\%} handles NAs differently from \comm{==}:

```{r}
c(1, 2, NA) == NA
c(1, 2, NA) %in% NA
```

## Exercises (Boolean Algebra)

1. Find all flights where \comm{arr\_delay} is missing but \comm{dep\_delay} is not. Find all flights where neither \comm{arr\_time} nor \comm{sched\_arr\_time} are missing, but \comm{arr\_delay} is.
2. How many flights have a missing \comm{dep\_time}? What other variables are missing? What might these rows represent?
3. Assuming missing \comm{dep\_time} = cancelled, look at cancelled flights per day. Is there a connection between proportion cancelled and average delay?

## Solutions (Boolean Algebra)

1.

```{r}
#| output.lines: 4
flights |>
  filter(is.na(arr_delay) & !is.na(dep_delay)) |>
  nrow()
```

```{r}
#| output.lines: 4
flights |>
  filter(!is.na(arr_time) & !is.na(sched_arr_time) &
         is.na(arr_delay)) |>
  nrow()
```

2. 8,255 flights. \comm{dep\_delay}, \comm{arr\_time}, \comm{arr\_delay}, \comm{air\_time} also missing --- these are cancelled flights.

3. Higher average delay days tend to have higher cancellation proportions.

# Logical Summaries

## `any()` and `all()`

```{r}
#| output.lines: 5
flights |>
  group_by(year, month, day) |>
  summarize(
    all_delayed = all(dep_delay <= 60, na.rm = TRUE),
    any_long_delay = any(arr_delay >= 300, na.rm = TRUE),
    .groups = "drop"
  )
```

\comm{any()} = equivalent of \comm{|}. \comm{all()} = equivalent of \comm{\&}.

## Numeric Summaries of Logicals

\comm{TRUE} becomes 1, \comm{FALSE} becomes 0 in numeric context. So \comm{sum()} counts TRUEs, \comm{mean()} gives proportion:

```{r}
#| output.lines: 5
flights |>
  group_by(year, month, day) |>
  summarize(
    prop_delayed = mean(dep_delay <= 60, na.rm = TRUE),
    count_long = sum(arr_delay >= 300, na.rm = TRUE),
    .groups = "drop"
  )
```

## Logical Subsetting with `[`

Use \comm{[} to filter within \comm{summarize()} --- compute different summaries for different subsets:

```{r}
#| output.lines: 5
flights |>
  group_by(year, month, day) |>
  summarize(
    behind = mean(arr_delay[arr_delay > 0], na.rm = TRUE),
    ahead = mean(arr_delay[arr_delay < 0], na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )
```

## Exercises (Logical Summaries)

1. What will \comm{sum(is.na(x))} tell you? How about \comm{mean(is.na(x))}?
2. What does \comm{prod()} return on a logical vector? What logical summary is it equivalent to? What about \comm{min()}?

## Solutions (Logical Summaries)

1. \comm{sum(is.na(x))} = count of NAs. \comm{mean(is.na(x))} = proportion of NAs.

2. \comm{prod()} on logicals: returns 1 only if all are TRUE (equivalent to \comm{all()}). Returns 0 if any is FALSE. \comm{min()} returns 0 if any is FALSE, 1 if all TRUE --- also equivalent to \comm{all()}.

```{r}
x <- c(TRUE, TRUE, FALSE)
prod(x)
min(x)
```

# Conditional Transformations

## `if_else()`

Three required arguments: condition, value if TRUE, value if FALSE:

```{r}
x <- c(-3:3, NA)
if_else(x > 0, "+ve", "-ve")
```

Optional fourth argument for NAs:

```{r}
if_else(x > 0, "+ve", "-ve", "???")
```

Vectorized --- can use variables on both sides:

```{r}
if_else(x < 0, -x, x)  # simple abs()
```

## `case_when()`

For multiple conditions, use \comm{case\_when()} with \comm{condition \textasciitilde{} output} pairs:

```{r}
case_when(
  x == 0   ~ "0",
  x < 0    ~ "-ve",
  x > 0    ~ "+ve",
  is.na(x) ~ "???"
)
```

Use \comm{.default} for catch-all:

```{r}
#| eval: false
case_when(
  x < 0 ~ "-ve",
  x > 0 ~ "+ve",
  .default = "???"
)
```

## `case_when()` with Data

```{r}
#| output.lines: 6
flights |>
  mutate(
    status = case_when(
      is.na(arr_delay)      ~ "cancelled",
      arr_delay < -30       ~ "very early",
      arr_delay < -15       ~ "early",
      abs(arr_delay) <= 15  ~ "on time",
      arr_delay < 60        ~ "late",
      arr_delay < Inf       ~ "very late",
    ),
    .keep = "used"
  )
```

First matching condition wins. Unmatched rows get \comm{NA}.

## Exercises (Conditional)

1. Use \comm{if\_else()} to determine whether each number 0--20 is even or odd. (Hint: \comm{x \%\% 2 == 0}.)
2. Given days like \comm{c("Monday", "Saturday", "Wednesday")}, label as weekends or weekdays.
3. Use \comm{if\_else()} to compute absolute value of a numeric vector.
4. Write a \comm{case\_when()} using \comm{month} and \comm{day} from \comm{flights} to label US holidays.

## Solutions (Conditional) --- 1--3

1.

```{r}
x <- 0:20
if_else(x %% 2 == 0, "even", "odd")
```

2.

```{r}
x <- c("Monday", "Saturday", "Wednesday")
if_else(x %in% c("Saturday", "Sunday"),
        "weekend", "weekday")
```

3. `if_else(x < 0, -x, x)`

## Solutions (Conditional) --- 4

```{r}
#| output.lines: 6
flights |>
  mutate(
    holiday = case_when(
      month == 1  & day == 1  ~ "New Year's Day",
      month == 7  & day == 4  ~ "Independence Day",
      month == 12 & day == 25 ~ "Christmas",
      .default = NA
    ),
    .keep = "used"
  ) |>
  filter(!is.na(holiday))
```

# Making Numbers & Counts

## Parsing Strings to Numbers

```{r}
x <- c("1.2", "5.6", "1e3")
parse_double(x)
```

\comm{parse\_number()} ignores non-numeric text:

```{r}
x <- c("$1,234", "USD 3,513", "59%")
parse_number(x)
```

## `count()` and Variants

```{r}
#| output.lines: 5
flights |> count(dest, sort = TRUE)
```

Equivalent with \comm{group\_by()} + \comm{summarize()}:

```{r}
#| output.lines: 5
flights |>
  group_by(dest) |>
  summarize(n = n(),
            delay = mean(arr_delay, na.rm = TRUE))
```

## Count Variants

```{r}
#| output.lines: 5
# n_distinct: unique values per group
flights |>
  group_by(dest) |>
  summarize(carriers = n_distinct(carrier)) |>
  arrange(desc(carriers))
```

```{r}
#| output.lines: 4
# Weighted count
flights |> count(tailnum, wt = distance, sort = TRUE)
```

## Exercises (Counts)

1. How can you use \comm{count()} to count rows with a missing value for a given variable?
2. Expand to \comm{group\_by()} + \comm{summarize()} + \comm{arrange()}:
   - \comm{flights |> count(dest, sort = TRUE)}
   - \comm{flights |> count(tailnum, wt = distance)}

## Solutions (Counts)

1. `flights |> filter(is.na(dep_time)) |> count()` or `flights |> count(is.na(dep_time))`.

2.

```{r}
#| eval: false
# (a)
flights |> group_by(dest) |>
  summarize(n = n()) |> arrange(desc(n))
# (b)
flights |> group_by(tailnum) |>
  summarize(n = sum(distance))
```

# Numeric Transformations

## Recycling Rules

R recycles shorter vectors to match longer ones:

```{r}
x <- c(1, 2, 10, 20)
x / 5          # recycles 5 to c(5, 5, 5, 5)
x * c(1, 2)    # recycles to c(1, 2, 1, 2)
```

**Danger:** silent wrong results if lengths don't match as expected:

```{r}
#| eval: false
# WRONG: recycles c(1, 2) row by row
flights |> filter(month == c(1, 2))
# RIGHT: use %in%
flights |> filter(month %in% c(1, 2))
```

## `pmin()` and `pmax()`

Parallel (element-wise) min/max --- different from \comm{min()}/\comm{max()} which return a single value:

```{r}
#| output.lines: 6
df <- tribble(
  ~x, ~y,
  1,  3,
  5,  2,
  7, NA,
)
df |> mutate(min = pmin(x, y, na.rm = TRUE),
             max = pmax(x, y, na.rm = TRUE))
```

## Modular Arithmetic

\comm{\%/\%} = integer division, \comm{\%\%} = remainder:

```{r}
1:10 %/% 3
1:10 %% 3
```

Useful for unpacking time variables:

```{r}
#| output.lines: 5
flights |>
  mutate(hour = sched_dep_time %/% 100,
         minute = sched_dep_time %% 100,
         .keep = "used")
```

## Rounding

```{r}
round(123.456, 2)   # nearest 0.01
round(123.456, 1)   # nearest 0.1
round(123.456, -1)  # nearest 10
```

Banker's rounding (round half to even):

```{r}
round(c(1.5, 2.5))
```

\comm{floor()} always rounds down, \comm{ceiling()} always rounds up.

```{r}
floor(123.456)
ceiling(123.456)
```

## Logarithms and `cut()`

Three log functions: \comm{log()} (base $e$), \comm{log2()} (base 2), \comm{log10()} (base 10).

Prefer \comm{log2()} (doubling = +1) or \comm{log10()} (easy back-transform).

\medskip

\comm{cut()} bins a numeric vector:

```{r}
x <- c(1, 2, 5, 10, 15, 20)
cut(x, breaks = c(0, 5, 10, 15, 20),
    labels = c("sm", "md", "lg", "xl"))
```

Values outside breaks become \comm{NA}.

## Cumulative Aggregates

```{r}
x <- 1:10
cumsum(x)
```

Also: \comm{cumprod()}, \comm{cummin()}, \comm{cummax()}, \comm{cummean()} (dplyr).

\medskip

Useful for creating session IDs from gaps:

```{r}
#| eval: false
events |>
  mutate(diff = time - lag(time, default = first(time)),
         has_gap = diff >= 5,
         group = cumsum(has_gap))
```

## Exercises (Numeric Transformations)

1. Explain what each line does in computing proportion of cancelled flights by hour.
2. Convert \comm{dep\_time} and \comm{sched\_dep\_time} to minutes since midnight.
3. Round \comm{dep\_time} and \comm{arr\_time} to the nearest five minutes.

## Solutions (Numeric Transformations)

1. Group by hour (extracted via \comm{\%/\% 100}), compute \comm{mean(is.na(dep\_time))} for proportion cancelled, count \comm{n()}, filter small groups, plot.

2.

```{r}
#| output.lines: 5
flights |>
  mutate(
    dep_minutes = (dep_time %/% 100) * 60 +
                   dep_time %% 100,
    .keep = "used"
  )
```

3. `round(dep_time / 5) * 5`

# General Transformations

## Ranks

```{r}
x <- c(1, 5, 5, 17, 22, NA)
min_rank(x)
min_rank(desc(x))
```

Variants: \comm{row\_number()} (no ties), \comm{dense\_rank()} (no gaps), \comm{percent\_rank()}, \comm{cume\_dist()}.

```{r}
#| output.lines: 8
tibble(x) |>
  mutate(row_number = row_number(x),
         dense_rank = dense_rank(x),
         percent_rank = percent_rank(x))
```

## Offsets: `lead()` and `lag()`

```{r}
x <- c(2, 5, 11, 11, 19, 35)
lag(x)
lead(x)
```

Compute differences and detect changes:

```{r}
x - lag(x)
x == lag(x)
```

## Exercises (General Transformations)

1. Find the 10 most delayed flights using a ranking function. How do you handle ties?
2. Which plane (\comm{tailnum}) has the worst on-time record?
3. What time of day should you fly to avoid delays?
4. For each destination, compute total delay minutes. For each flight, compute its proportion of destination's total delay.

## Solutions (General) --- 1--2

1.

```{r}
#| output.lines: 5
flights |>
  mutate(rank = min_rank(desc(arr_delay))) |>
  filter(rank <= 10) |>
  select(arr_delay, rank, carrier, flight) |>
  arrange(rank)
```

2.

```{r}
#| output.lines: 5
flights |>
  group_by(tailnum) |>
  summarize(prop_late = mean(arr_delay > 0, na.rm = TRUE),
            n = n()) |>
  filter(n >= 20) |>
  arrange(desc(prop_late))
```

## Solutions (General) --- 3--4

3.

```{r}
#| fig.height: 2
#| warning: false
flights |>
  group_by(hour = dep_time %/% 100) |>
  summarize(avg_delay = mean(dep_delay, na.rm = TRUE)) |>
  filter(!is.na(hour)) |>
  ggplot(aes(x = hour, y = avg_delay)) +
  geom_line() + geom_point()
```

Fly early morning (5--7 AM) for lowest delays.

## Solutions (General) --- 4

```{r}
#| output.lines: 5
flights |>
  filter(arr_delay > 0) |>
  group_by(dest) |>
  mutate(
    total_delay = sum(arr_delay),
    prop_delay = arr_delay / total_delay
  ) |>
  select(dest, flight, arr_delay, prop_delay) |>
  ungroup()
```

# Numeric Summaries

## Center: `mean()` vs. `median()`

```{r}
#| fig.height: 2.2
flights |>
  group_by(year, month, day) |>
  summarize(mean = mean(dep_delay, na.rm = TRUE),
            median = median(dep_delay, na.rm = TRUE),
            .groups = "drop") |>
  ggplot(aes(x = mean, y = median)) +
  geom_abline(slope = 1, intercept = 0,
              color = "white", linewidth = 2) +
  geom_point()
```

Median always $\leq$ mean (right-skewed data).

## Quantiles, Spread, and Positions

**Quantiles:** generalization of median.

```{r}
#| output.lines: 5
flights |>
  group_by(year, month, day) |>
  summarize(
    max = max(dep_delay, na.rm = TRUE),
    q95 = quantile(dep_delay, 0.95, na.rm = TRUE),
    .groups = "drop"
  )
```

**Spread:** \comm{sd()} and \comm{IQR()} (middle 50\%).

**Positions:** \comm{first()}, \comm{last()}, \comm{nth(x, n)}.

## Summaries with `mutate()`

Summary functions + \comm{mutate()} = group standardization:

```{r}
#| eval: false
# Proportion of total
x / sum(x)

# Z-score
(x - mean(x)) / sd(x)

# Normalize to [0, 1]
(x - min(x)) / (max(x) - min(x))

# Index relative to first observation
x / first(x)
```

## Exercises (Numeric Summaries)

1. Brainstorm 5+ ways to assess typical delay characteristics. When is \comm{mean()} useful? \comm{median()}?
2. Which destinations show the greatest variation in air speed?

## Solutions (Numeric Summaries)

1. Mean (symmetric data), median (skewed), proportion delayed, number > 30 min late, IQR of delays, 90th percentile. Use arrival delay for passenger impact, departure delay for operations. Join \comm{planes} for aircraft age/type effects.

2.

```{r}
#| output.lines: 5
flights |>
  mutate(speed = distance / (air_time / 60)) |>
  group_by(dest) |>
  summarize(speed_sd = sd(speed, na.rm = TRUE),
            n = n()) |>
  filter(n >= 20) |>
  arrange(desc(speed_sd))
```

<!--
# Summary

## Key Takeaways

- **Comparisons:** `<`, `>`, `==`, `!=`; use \comm{near()} for floats; \comm{is.na()} for NAs
- **Boolean:** `&`, `|`, `!`; use \comm{\%in\%} instead of `== | ==`
- **Logical summaries:** \comm{any()}/\comm{all()}, \comm{sum()}/\comm{mean()} on logicals, `[` for subsetting
- **Conditionals:** \comm{if\_else()} for two cases, \comm{case\_when()} for many
- **Parsing:** \comm{parse\_double()}, \comm{parse\_number()} for strings $\to$ numbers
- **Counts:** \comm{count()}, \comm{n()}, \comm{n\_distinct()}
- **Transforms:** \comm{\%/\%}/\comm{\%\%} (modular), \comm{round()}/\comm{floor()}/\comm{ceiling()}, \comm{log2()}/\comm{log10()}, \comm{cut()}
- **Ranks/offsets:** \comm{min\_rank()}, \comm{lag()}/\comm{lead()}, \comm{consecutive\_id()}
- **Summaries:** \comm{mean()}/\comm{median()}, \comm{quantile()}, \comm{sd()}/\comm{IQR()}, \comm{first()}/\comm{last()}
-->
