---
title: "R4DS 02 -- Data Visualization"
---

```{r setup, include=FALSE}
source("_common.R")
```

# Introduction

## The Grammar of Graphics

> "The simple graph has brought more information to the data analyst's mind than any other device." --- John Tukey

- **ggplot2** implements a *grammar of graphics*
- Learn one system, apply it everywhere
- Core idea: **map** variables in data to **aesthetic** properties (position, color, size, shape)
- Build plots **layer by layer**

## Prerequisites

```{r}
#| output.lines: 3
library(tidyverse)
```

```{r}
library(palmerpenguins)
library(ggthemes)
```

- \comm{palmerpenguins}: the \comm{penguins} dataset
- \comm{ggthemes}: colorblind-safe palettes

# First Steps

## The `penguins` Data

```{r}
#| output.lines: 8
penguins
```

- `r nrow(penguins)` observations, `r ncol(penguins)` variables
- Species: Adelie, Chinstrap, Gentoo
- Key variables: \comm{flipper\_length\_mm}, \comm{body\_mass\_g}, \comm{bill\_length\_mm}, \comm{bill\_depth\_mm}

## Key Definitions

- **Variable**: a quantity or property you can measure
- **Value**: the state of a variable when you measure it
- **Observation** (data point): a set of measurements on the same object
- **Tabular data** is *tidy* if each value has its own cell, each variable its own column, each observation its own row

## Building a ggplot --- Step 1: Data

```{r}
#| fig.height: 2
ggplot(data = penguins)
```

\comm{ggplot(data = ...)} creates an empty canvas primed with a dataset.

## Building a ggplot --- Step 2: Aesthetics

```{r}
#| fig.height: 2
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
)
```

\comm{mapping = aes(x = ..., y = ...)} maps variables to axes.

## Building a ggplot --- Step 3: Geom

```{r}
#| fig.height: 2.5
#| warning: false
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point()
```

\comm{geom\_point()} adds a **layer** of points $\to$ scatterplot.

## Adding Color by Species

```{r}
#| fig.height: 2.5
#| warning: false
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g,
                color = species)
) +
  geom_point()
```

Map a categorical variable to \comm{color} $\to$ automatic legend.

## Adding a Smooth Line

```{r}
#| fig.height: 2.5
#| warning: false
#| message: false
ggplot(penguins,
       aes(x = flipper_length_mm, y = body_mass_g,
           color = species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

\comm{color = species} in \comm{ggplot()} $\to$ **both** geoms inherit it $\to$ 3 separate lines.

## Global vs. Local Aesthetics

```{r}
#| fig.height: 2.5
#| warning: false
#| message: false
ggplot(penguins,
       aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = species)) +
  geom_smooth(method = "lm")
```

Move \comm{color} and \comm{shape} into \comm{geom\_point()} $\to$ **one** overall line.

## The Polished Plot

```{r}
#| fig.height: 2
#| warning: false
#| message: false
ggplot(penguins,
       aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = species)) +
  geom_smooth(method = "lm") +
  labs(title = "Body mass and flipper length",
       subtitle = "Adelie, Chinstrap, and Gentoo",
       x = "Flipper length (mm)", y = "Body mass (g)",
       color = "Species", shape = "Species") +
  scale_color_colorblind()
```

## Exercises 2.2.1 (1--5)

1. How many rows are in \comm{penguins}? How many columns?
2. What does \comm{bill\_depth\_mm} describe? (Hint: \comm{?penguins})
3. Scatterplot of \comm{bill\_depth\_mm} vs. \comm{bill\_length\_mm}. Describe the relationship.
4. Scatterplot of \comm{species} vs. \comm{bill\_depth\_mm}. What might be a better geom?
5. Why does this give an error?

```{r}
#| eval: false
ggplot(data = penguins) +
  geom_point()
```

## Exercises 2.2.1 (6--10)

6. What does \comm{na.rm} do in \comm{geom\_point()}? Default value? Use it set to \comm{TRUE}.
7. Add caption: "Data come from the palmerpenguins package." (Hint: \comm{labs()})
8. Recreate: scatterplot with \comm{bill\_depth\_mm} mapped to color, smooth curve overlaid. Global or local?
9. Predict the output, then run:

```{r}
#| eval: false
ggplot(penguins, aes(x = flipper_length_mm,
    y = body_mass_g, color = island)) +
  geom_point() + geom_smooth(se = FALSE)
```

10. Will these two plots look different? Why?

## Solutions 2.2.1 (1--5)

1. `r nrow(penguins)` rows, `r ncol(penguins)` columns.
2. Bill depth in mm --- vertical distance from top to bottom of bill.
3.
```{r}
#| fig.height: 1.8
#| warning: false
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point()
```

Negative overall; Simpson's paradox --- positive within each species.

## Solutions 2.2.1 (4--5)

4.
```{r}
#| fig.height: 1.8
#| warning: false
ggplot(penguins, aes(x = species, y = bill_depth_mm)) +
  geom_boxplot()
```

Categorical $\times$ numerical $\to$ boxplot is better than scatterplot.

5. \comm{geom\_point()} needs \comm{x} and \comm{y} aesthetics. No \comm{mapping} was provided.

## Solutions 2.2.1 (6--8)

6. \comm{na.rm = TRUE} silently removes NAs (default: \comm{FALSE}, which warns).

```{r}
#| eval: false
ggplot(penguins, aes(x = flipper_length_mm,
    y = body_mass_g)) +
  geom_point(na.rm = TRUE)
```

7. Add \comm{labs(caption = "Data come from the palmerpenguins package.")}

8. \comm{bill\_depth\_mm} mapped to \comm{color} at the **geom level** (so the smooth curve is uncolored):

```{r}
#| eval: false
ggplot(penguins, aes(x = flipper_length_mm,
    y = body_mass_g)) +
  geom_point(aes(color = bill_depth_mm)) +
  geom_smooth()
```

## Solutions 2.2.1 (9--10)

9. Three smooth curves (one per island, no confidence bands) with colored points.

10. **No** --- both plots look the same. The first specifies \comm{data} and \comm{mapping} globally; the second repeats them in each geom. Same result either way.

# Concise ggplot2 Calls

## Dropping Argument Names

Verbose:

```{r}
#| eval: false
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point()
```

Concise (first two arguments by position):

```{r}
#| eval: false
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```

With the pipe:

```{r}
#| eval: false
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```

# Visualizing Distributions

## Categorical Variable --- Bar Chart

```{r}
#| fig.height: 2.5
ggplot(penguins, aes(x = species)) +
  geom_bar()
```

## Reorder Bars by Frequency

```{r}
#| fig.height: 2.5
ggplot(penguins, aes(x = fct_infreq(species))) +
  geom_bar()
```

\comm{fct\_infreq()} reorders factor levels by decreasing frequency.

## Numerical Variable --- Histogram

```{r}
#| fig.height: 2.5
#| warning: false
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 200)
```

\comm{binwidth} controls bin width (in units of \comm{x}). Always try several values.

## Numerical Variable --- Density Plot

```{r}
#| fig.height: 2.5
ggplot(penguins, aes(x = body_mass_g)) +
  geom_density()
```

Smoothed version of a histogram --- good for seeing overall shape.

## Exercises 2.4.3

1. Bar plot of \comm{species} with \comm{species} on the $y$-axis. How is it different?
2. \comm{color = "red"} vs. \comm{fill = "red"} in \comm{geom\_bar()}. Which is more useful?
3. What does the \comm{bins} argument in \comm{geom\_histogram()} do?
4. Histogram of \comm{carat} in \comm{diamonds}. Which binwidth reveals the most interesting patterns?

## Solutions 2.4.3

1. Horizontal bars instead of vertical.
2. \comm{fill} changes the interior color; \comm{color} only the outline. \comm{fill} is more useful.
3. \comm{bins} sets the number of bins (default 30). Alternative to \comm{binwidth}.
4.

```{r}
#| fig.height: 2
ggplot(diamonds, aes(x = carat)) +
  geom_histogram(binwidth = 0.01)
```

Spikes at "round" values (0.5, 1.0, 1.5, 2.0).

# Visualizing Relationships

## Numerical + Categorical --- Boxplot

```{r}
#| fig.height: 2.5
#| warning: false
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot()
```

Box = IQR (25th--75th percentile); line = median; whiskers = 1.5$\times$IQR; dots = outliers.

## Numerical + Categorical --- Density

```{r}
#| fig.height: 2.5
#| warning: false
ggplot(penguins, aes(x = body_mass_g,
                     color = species, fill = species)) +
  geom_density(alpha = 0.5)
```

\comm{alpha = 0.5} adds transparency. We **map** variables to aesthetics; we **set** fixed values.

## Two Categorical Variables --- Stacked Bar

```{r}
#| fig.height: 2.5
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar()
```

## Relative Frequency Bar

```{r}
#| fig.height: 2.5
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(position = "fill") +
  labs(y = "proportion")
```

\comm{position = "fill"} scales each bar to 100\% for comparing proportions.

## Three or More Variables --- Extra Aesthetics

```{r}
#| fig.height: 2.5
#| warning: false
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = island))
```

Gets cluttered fast --- consider **facets** instead.

## Faceting

```{r}
#| fig.height: 2.3
#| warning: false
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = species)) +
  facet_wrap(~island)
```

\comm{facet\_wrap(\~var)} splits the plot into subplots by a categorical variable.

## Exercises 2.5.5 (1--4)

1. Which variables in \comm{mpg} are categorical? Numerical? How can you tell from \comm{mpg}?
2. Scatterplot of \comm{hwy} vs. \comm{displ}. Map a third variable to \comm{color}, then \comm{size}, then both, then \comm{shape}. Categorical vs. numerical differences?
3. What happens if you map a third variable to \comm{linewidth}?
4. What happens if you map the same variable to multiple aesthetics?

## Exercises 2.5.5 (5--7)

5. Scatterplot of \comm{bill\_depth\_mm} vs. \comm{bill\_length\_mm}, colored by \comm{species}. What does it reveal? What about faceting?
6. Why does this yield two separate legends? Fix it.

```{r}
#| eval: false
ggplot(penguins, aes(x = bill_length_mm,
    y = bill_depth_mm, color = species,
    shape = species)) +
  geom_point() + labs(color = "Species")
```

7. Create two stacked bar plots (island by species, species by island). Which question does each answer?

## Solutions 2.5.5 (1--4)

1. Categorical: \comm{manufacturer}, \comm{model}, \comm{trans}, \comm{drv}, \comm{fl}, \comm{class}. Numerical: \comm{displ}, \comm{year}, \comm{cyl}, \comm{cty}, \comm{hwy}. The tibble shows \comm{<chr>} vs. \comm{<dbl>}/\comm{<int>}.
2. Numerical variables produce a continuous color gradient or continuous size; categorical produce discrete colors/shapes. \comm{shape} cannot be used with numerical variables.
3. \comm{linewidth} has no effect on \comm{geom\_point()} --- it's for lines.
4. Redundant encoding; ggplot2 may combine legends.

## Solutions 2.5.5 (5--7)

5. Without color: negative trend (Simpson's paradox). With color: **positive** within each species. Faceting shows the same.

6. \comm{labs(color = "Species")} sets only the color legend label. Fix: also set \comm{labs(shape = "Species")} or use matching labels.

7. First (\comm{x = island, fill = species}): "What is the species composition of each island?" Second (\comm{x = species, fill = island}): "Which islands does each species inhabit?"

# Saving Plots

## `ggsave()`

```{r}
#| eval: false
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
ggsave(filename = "penguin-plot.png")
```

- Saves the **most recently displayed** plot by default
- Specify \comm{width} and \comm{height} for reproducibility
- File extension determines format: \comm{.png}, \comm{.pdf}, \comm{.svg}, etc.

## Exercises 2.6.1

1. Which plot is saved as \texttt{mpg-plot.png}? Why?

```{r}
#| eval: false
ggplot(mpg, aes(x = class)) + geom_bar()
ggplot(mpg, aes(x = cty, y = hwy)) + geom_point()
ggsave("mpg-plot.png")
```

2. Change the code to save as PDF instead. How can you find supported formats?

## Solutions 2.6.1

1. The **scatterplot** (\comm{cty} vs. \comm{hwy}) is saved, because \comm{ggsave()} defaults to the **last displayed plot**.

2. Change \comm{"mpg-plot.png"} to \comm{"mpg-plot.pdf"}. Run \comm{?ggsave} to see supported formats (or check the file extension).

# Common Problems

## Debugging ggplot2

Common mistakes:

- **Misplaced `+`**: must be at the **end** of the line, not the start

```{r}
#| eval: false
# WRONG:
ggplot(data = mpg)
+ geom_point(aes(x = displ, y = hwy))

# RIGHT:
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy))
```

\medskip

- **Missing aesthetic**: error says "stat requires ... aesthetic"
- **Unmatched quotes/parens**: console shows `+` prompt --- press **Escape**
- **Google the error message**: someone else has had the same problem

<!--
# Summary

## Key Takeaways

- ggplot2: data + aesthetics + geoms, built layer by layer
- \comm{aes()} **maps** variables to visual properties; fixed values are **set** outside \comm{aes()}
- Global vs. local aesthetics: \comm{ggplot(aes(...))} inherited by all layers; \comm{geom\_*(aes(...))} local only
- Distributions: \comm{geom\_bar()} (categorical), \comm{geom\_histogram()} / \comm{geom\_density()} (numerical)
- Relationships: \comm{geom\_boxplot()}, \comm{geom\_point()}, stacked bars
- Facets: \comm{facet\_wrap(\~var)} for subplots
- Save: \comm{ggsave("file.png", width, height)}
- Pipe: \comm{data |> ggplot(aes(...)) + geom\_*()}
-->
