---
title: "R4DS 14 -- Iteration & Base R"
---

```{r setup, include=FALSE}
source("_common.R")
library(tidyverse)
```

# Modifying Multiple Columns

## `across()` --- Apply to Many Columns

Instead of repeating the same operation per column:

```{r}
#| eval: false
df |> summarize(a = median(a), b = median(b),
                c = median(c), d = median(d))
```

Use \comm{across(.cols, .fns)}:

```{r}
#| eval: false
df |> summarize(across(a:d, median))
```

Common \comm{.cols} selectors:

| Selector | Matches |
|----------|---------|
| `a:d` | Columns a through d |
| `everything()` | All non-grouping columns |
| `where(is.numeric)` | All numeric columns |
| `starts_with("x")` | Columns starting with "x" |

## Anonymous Functions in `across()`

Pass the **function name**, not a call. For extra arguments, use an anonymous function:

```{r}
#| output.lines: 6
df_miss <- tibble(a = c(1, NA, 3, 4),
  b = c(NA, 5, 6, 7), c = c(7, 8, NA, 10))

# Without na.rm: NAs propagate
df_miss |> summarize(across(a:c, median))
```

```{r}
#| output.lines: 5
# Fix: anonymous function
df_miss |> summarize(
  across(a:c, \(x) median(x, na.rm = TRUE)))
```

`\(x)` is shorthand for `function(x)`.

## Multiple Functions and `.names`

Pass a **named list** of functions:

```{r}
#| output.lines: 5
df_miss |> summarize(across(a:c, list(
  med = \(x) median(x, na.rm = TRUE),
  n_miss = \(x) sum(is.na(x)))))
```

Control output names with \comm{.names} (glue syntax):

```{r}
#| eval: false
df_miss |> summarize(across(a:c, list(
  med = \(x) median(x, na.rm = TRUE),
  n_miss = \(x) sum(is.na(x))),
  .names = "{.fn}_{.col}"))
```

Placeholders: `{.col}` = column name, `{.fn}` = function name.

## `if_any()` and `if_all()`

Filter rows based on conditions across multiple columns:

```{r}
#| output.lines: 5
# Rows where ANY column has NA (OR logic)
df_miss |> filter(if_any(a:c, is.na))
```

```{r}
#| output.lines: 5
# Rows where ALL columns are non-NA (AND logic)
df_miss |> filter(if_all(a:c, \(x) !is.na(x)))
```

Use in \comm{mutate()} with \comm{coalesce()} to replace NAs:

```{r}
#| eval: false
df_miss |> mutate(across(a:c, \(x) coalesce(x, 0)))
```

## `across()` in Functions

Combine embracing with tidy-select:

```{r}
#| eval: false
summarize_means <- function(df,
    summary_vars = where(is.numeric)) {
  df |> summarize(
    across({{ summary_vars }},
      \(x) mean(x, na.rm = TRUE)),
    n = n(), .groups = "drop")
}

diamonds |> group_by(cut) |> summarize_means()
diamonds |> group_by(cut) |>
  summarize_means(c(carat, x:z))
```

Default \comm{where(is.numeric)} selects all numeric columns; user can override.

## Exercises (across)

1. Practice \comm{across()} skills:
   a. Compute unique values per column in \comm{palmerpenguins::penguins}
   b. Compute mean of every column in \comm{mtcars}
   c. Group \comm{diamonds} by \comm{cut}, \comm{clarity}, \comm{color}; count and compute mean of numeric columns

2. What happens with an unnamed list of functions in \comm{across()}?

3. Modify \comm{expand\_dates()} to auto-remove the original date columns after expansion.

4. Explain each step:

```{r}
#| eval: false
show_missing <- function(df, group_vars,
    summary_vars = everything()) {
  df |> group_by(pick({{ group_vars }})) |>
    summarize(across({{ summary_vars }},
      \(x) sum(is.na(x))), .groups = "drop") |>
    select(where(\(x) any(x > 0)))
}
```

## Solutions (across) --- 1

```{r}
#| output.lines: 5
#| message: false
# 1b. Mean of every column in mtcars
mtcars |>
  summarize(across(everything(), mean))
```

```{r}
#| output.lines: 5
#| message: false
# 1c. Grouped means of numeric columns
diamonds |> group_by(cut, clarity, color) |>
  summarize(n = n(),
    across(where(is.numeric), mean),
    .groups = "drop")
```

1a. Same pattern: `penguins |> summarize(across(everything(), n_distinct))`

## Solutions (across) --- 2--4

2. Unnamed functions get numeric suffixes: \comm{col\_1}, \comm{col\_2}, etc.

3. Add \comm{select(!where(is.Date))} at the end:

```{r}
#| eval: false
expand_dates <- function(df) {
  df |>
    mutate(across(where(is.Date),
      list(year = year, month = month,
           day = mday))) |>
    select(!where(is.Date))
}
```

4. Groups by \comm{group\_vars}, counts NAs per \comm{summary\_vars} column, then \comm{select(where(...))} keeps only columns where **any group has missing values**.

# Reading Multiple Files

## `map()` and `list_rbind()`

```{r}
#| eval: false
paths <- list.files("data/gapminder",
  pattern = "[.]xlsx$", full.names = TRUE)
```

\comm{map()} applies a function to each element, returning a **list**:

```{r}
#| eval: false
files <- map(paths, readxl::read_excel)
files[[1]]      # access first data frame
```

\comm{list\_rbind()} stacks all data frames into one:

```{r}
#| eval: false
paths |>
  map(readxl::read_excel) |>
  list_rbind()
```

Pass extra arguments with an anonymous function:

```{r}
#| eval: false
paths |>
  map(\(p) readxl::read_excel(p, n_max = 1)) |>
  list_rbind()
```

## Data in the Path

File names often contain data. Use \comm{set\_names()} + \comm{names\_to}:

```{r}
#| eval: false
paths |>
  set_names(basename) |>
  map(readxl::read_excel) |>
  list_rbind(names_to = "year") |>
  mutate(year = parse_number(year))
```

**Workflow:** list files $\to$ name them $\to$ read each $\to$ bind with name column $\to$ parse.

\medskip

Save processed data to avoid re-reading:

```{r}
#| eval: false
gapminder <- paths |>
  set_names(basename) |>
  map(readxl::read_excel) |>
  list_rbind(names_to = "year") |>
  mutate(year = parse_number(year))
write_csv(gapminder, "gapminder.csv")
```

## Handling Failures with `possibly()`

\comm{possibly()} wraps a function to return a default instead of erroring:

```{r}
#| eval: false
files <- paths |>
  map(possibly(
    \(p) readxl::read_excel(p), NULL))

data <- files |> list_rbind()

# Find which files failed
failed <- map_vec(files, is.null)
paths[failed]
```

Useful when some files are corrupt or have unexpected formats.

# Saving Multiple Outputs

## Writing Files with `walk()`

\comm{walk()} is like \comm{map()} but for **side effects** (writing, printing):

```{r}
#| output.lines: 5
by_clarity <- diamonds |> group_nest(clarity)
by_clarity
```

```{r}
#| eval: false
by_clarity <- by_clarity |>
  mutate(path = str_glue("diamonds-{clarity}.csv"))

# Write each subset to its own file
walk2(by_clarity$data, by_clarity$path, write_csv)
```

\comm{walk2()} maps over **two** inputs simultaneously.

## Saving Plots

```{r}
#| eval: false
carat_histogram <- function(df) {
  ggplot(df, aes(x = carat)) +
    geom_histogram(binwidth = 0.1)
}

by_clarity <- by_clarity |>
  mutate(
    plot = map(data, carat_histogram),
    path = str_glue("clarity-{clarity}.png"))

walk2(by_clarity$path, by_clarity$plot,
  \(path, plot) ggsave(path, plot,
    width = 6, height = 6))
```

Pattern: nest data $\to$ create plots with \comm{map()} $\to$ save with \comm{walk2()}.

# Base R Essentials

## Subsetting Vectors with `[`

Five ways to subset a vector:

```{r}
#| output.lines: 5
x <- c("one", "two", "three", "four", "five")
x[c(3, 2, 5)]           # positive integers
x[c(-1, -3, -5)]        # negative = exclude
```

```{r}
#| output.lines: 4
x <- c(10, 3, NA, 5, 8, 1, NA)
x[!is.na(x)]            # logical vector
x[x %% 2 == 0]          # NAs propagate!
```

```{r}
#| output.lines: 3
x <- c(abc = 1, def = 2, xyz = 5)
x[c("xyz", "def")]      # by name
```

Also: repeated indices (\comm{x[c(1,1,5)]}) and empty (\comm{x[]}) are valid.

## Subsetting Data Frames with `[`

`df[rows, cols]` --- rows and columns independently:

```{r}
#| output.lines: 5
df <- tibble(x = 1:3, y = c("a", "e", "f"),
             z = runif(3))
df[1, 2]                 # row 1, col 2
```

```{r}
#| output.lines: 5
df[df$x > 1, ]           # filter rows
```

\comm{df[, c("x", "y")]} selects columns. **Tibble vs. data.frame:** tibbles always return tibbles; \comm{data.frame} may drop to a vector.

## dplyr Equivalents

| dplyr | Base R |
|-------|--------|
| `filter(df, x > 1)` | `df[df$x > 1, ]` |
| `arrange(df, x)` | `df[order(df$x), ]` |
| `select(df, x, z)` | `df[, c("x", "z")]` |
| `mutate(df, z = x+y)` | `df$z <- df$x + df$y` |
| `pull(df, x)` | `df$x` or `df[["x"]]` |

\medskip

\comm{subset()} combines filter + select:

```{r}
#| eval: false
df |> subset(x > 1, c(y, z))
# same as: df |> filter(x > 1) |> select(y, z)
```

## Single Elements: `[[` and `$`

\comm{[[} extracts a **single element**; \comm{\$} is shorthand by name:

```{r}
#| output.lines: 5
tb <- tibble(x = 1:4, y = c(10, 4, 1, 21))
tb[["x"]]       # by name
tb$x             # shorthand
tb[[1]]          # by position
```

Create new columns with \comm{\$}:

```{r}
#| output.lines: 5
tb$z <- tb$x + tb$y
tb
```

**Tibbles are strict:** \comm{\$} warns on partial/missing names. \comm{data.frame} silently does partial matching.

## Lists: `[` vs. `[[`

```{r}
l <- list(a = 1:3, b = "a string",
          c = pi, d = list(-1, -5))
```

\comm{[} returns a **sub-list**; \comm{[[} extracts the **element itself**:

```{r}
#| output.lines: 6
str(l[1:2])     # sub-list of first 2
str(l[[1]])     # the vector itself
str(l[[4]])     # extracts the inner list
```

Same for data frames: \comm{df["x"]} = one-column data frame; \comm{df[["x"]]} = vector.

## Exercises (Base R)

1. Create functions that take a vector and return:
   a. Elements at even-numbered positions
   b. Every element except the last
   c. Only even values (no missing values)

2. Why is `x[-which(x > 0)]` not the same as `x[x <= 0]`?

3. What happens with \comm{[[} and an index larger than the vector length? With a non-existent name?

## Solutions (Base R)

1.

```{r}
#| output.lines: 5
even_pos <- function(x) x[seq(2, length(x), 2)]
drop_last <- function(x) x[-length(x)]
even_vals <- function(x) x[!is.na(x) & x %% 2 == 0]

even_pos(1:10)
drop_last(1:5)
even_vals(c(1, NA, 4, 7, 8))
```

2. With \comm{NA}: `x <= 0` returns \comm{NA} (keeps element as \comm{NA}), but \comm{which()} drops \comm{NA}s. Also, \comm{-which(x > 0)} on empty result gives \comm{-integer(0)} = select nothing.

3. \comm{[[} with out-of-bounds index: error. Non-existent name: \comm{NULL} for lists, error for atomic vectors.

# Apply Family and For Loops

## The Apply Family

| Function | Like purrr | Returns |
|----------|-----------|---------|
| `lapply(x, f)` | `map(x, f)` | Always a list |
| `sapply(x, f)` | --- | Simplified (unpredictable) |
| `vapply(x, f, type)` | `map_vec(x, f)` | Specified type (strict) |
| `tapply(x, grp, f)` | grouped summarize | Named vector |

```{r}
#| output.lines: 3
sapply(mtcars[1:4], is.numeric)
```

```{r}
#| output.lines: 3
tapply(diamonds$price, diamonds$cut, mean)
```

Prefer \comm{vapply()} over \comm{sapply()} --- it fails clearly on type mismatches.

## For Loops

```{r}
#| eval: false
for (element in vector) {
  # do something with element
}
```

Use for **side effects** (same purpose as \comm{walk()}):

```{r}
#| eval: false
# purrr
paths |> walk(append_file)

# base R equivalent
for (path in paths) {
  append_file(path)
}
```

For loops themselves aren't slow --- **growing vectors** inside them is.

## Pre-allocation and `seq_along()`

**Never grow** a vector in a loop --- pre-allocate:

```{r}
#| eval: false
# Good: pre-allocate, then fill
files <- vector("list", length(paths))
for (i in seq_along(paths)) {
  files[[i]] <- readxl::read_excel(paths[[i]])
}
do.call(rbind, files)  # combine list

# Bad: grows quadratically (O(n^2))
out <- NULL
for (path in paths) {
  out <- rbind(out, readxl::read_excel(path))
}
```

\comm{seq\_along()} generates indices; \comm{do.call(rbind, ...)} combines a list.

<!--
# Summary

## Key Takeaways

- **\comm{across(.cols, .fns)}:** apply functions to multiple columns; use \comm{where()}, \comm{everything()}, tidy-select
- **Anonymous functions:** `\(x) f(x, na.rm = TRUE)` for extra arguments
- **Multiple functions:** named list in \comm{.fns}; \comm{.names} controls output naming
- **\comm{if\_any()}/\comm{if\_all()}:** filter with OR/AND logic across columns
- **\comm{map()} + \comm{list\_rbind()}:** read multiple files $\to$ one data frame
- **\comm{set\_names()} + \comm{names\_to}:** extract metadata from file paths
- **\comm{possibly()}:** handle failures gracefully in \comm{map()}
- **\comm{walk()}/\comm{walk2()}:** like \comm{map()} but for side effects
- **Base R \comm{[}:** 5 subsetting modes (positive, negative, logical, character, empty)
- **\comm{[[} and \comm{\$}:** extract single elements; tibbles are stricter than data frames
- **Apply family:** \comm{lapply} $\approx$ \comm{map}, \comm{vapply} for strict types, \comm{tapply} for groups
- **For loops:** pre-allocate with \comm{vector()}, use \comm{seq\_along()}, never grow vectors
-->
