---
title: "R4DS 07 -- Communication"
---

```{r setup, include=FALSE}
source("_common.R")
library(tidyverse)
library(scales)
library(patchwork)
```

# Labels

## `labs()` --- Add Context to Plots

```{r}
#| fig.height: 2
#| message: false
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency decreases with engine size",
    subtitle = "Two seaters are an exception (light weight)",
    caption = "Data from fueleconomy.gov",
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    color = "Car type"
  )
```

## Best Practices for Labels

- **Title:** summarize the main **finding**, not the plot type
- **Subtitle:** additional context in smaller font
- **Caption:** data source, credits
- **Axis labels:** descriptive names with units
- **Legend title:** clear variable name via \comm{labs(color = ...)}

\medskip

Math labels: use \comm{quote()} instead of \comm{""}:

```{r}
#| eval: false
labs(x = quote(x[i]), y = quote(sum(x[i]^2, i == 1, n)))
```

## Exercises (Labels)

1. Create a plot of fuel economy data with customized \comm{title}, \comm{subtitle}, \comm{caption}, \comm{x}, \comm{y}, and \comm{color} labels.
2. Recreate: scatterplot of \comm{cty} vs. \comm{hwy}, colored and shaped by \comm{drv}, with combined legend label "Type of drive train".
3. Take an exploratory plot and add informative titles.

## Solutions (Labels)

1. (Open-ended --- use \comm{labs()} with all six arguments.)

2.

```{r}
#| fig.height: 2
ggplot(mpg, aes(x = cty, y = hwy, color = drv, shape = drv)) +
  geom_point() +
  labs(x = "City MPG", y = "Highway MPG",
       color = "Type of\ndrive train",
       shape = "Type of\ndrive train")
```

# Annotations

## Direct Labels with `geom_text()`

```{r}
#| fig.height: 2.2
#| message: false
#| warning: false
label_info <- mpg |>
  group_by(drv) |>
  arrange(desc(displ)) |>
  slice_head(n = 1) |>
  mutate(drive_type = case_when(
    drv == "f" ~ "front-wheel",
    drv == "r" ~ "rear-wheel",
    drv == "4" ~ "4-wheel"
  ))

ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point(alpha = 0.3) + geom_smooth(se = FALSE) +
  geom_text(data = label_info,
    aes(label = drive_type),
    fontface = "bold", size = 4, hjust = "right") +
  theme(legend.position = "none")
```

## Labeling Outliers

```{r}
#| fig.height: 2.2
#| warning: false
potential_outliers <- mpg |>
  filter(hwy > 40 | (hwy > 20 & displ > 5))

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_point(data = potential_outliers,
             color = "red", size = 3, shape = "circle open") +
  annotate(geom = "label", x = 4, y = 42,
           label = "Outliers", color = "red")
```

## Annotation Toolkit

| Function | Purpose |
|----------|---------|
| `geom_text()` / `geom_label()` | Label points from data |
| `annotate()` | Add single annotations (text, arrows, rectangles) |
| `geom_hline()` / `geom_vline()` | Reference lines |
| `geom_segment(arrow = ...)` | Arrows pointing to features |
| `geom_rect()` | Highlight rectangular regions |

\comm{annotate()} is for one-off elements; geoms are for data-driven labels.

## Exercises (Annotations)

1. Use \comm{geom\_text()} with infinite positions to place text at the four corners.
2. Use \comm{annotate()} to add a point geom in the middle of a plot.
3. How do \comm{geom\_text()} labels interact with faceting? How to label a single facet?
4. What arguments to \comm{geom\_label()} control the background box?
5. What are the four arguments to \comm{arrow()}?

## Solutions (Annotations)

1.
```{r}
#| eval: false
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() +
  geom_text(aes(x = x, y = y, label = label), data = tibble(
    x = c(-Inf, Inf, -Inf, Inf),
    y = c(-Inf, -Inf, Inf, Inf),
    label = c("BL", "BR", "TL", "TR")),
    hjust = c(0, 1, 0, 1), vjust = c(0, 0, 1, 1))
```

2. `annotate(geom = "point", x = 3.5, y = 25, size = 5, color = "red")`

3. Labels appear in all facets by default. To label one facet, pass a data frame with the faceting variable set to the desired level.

## Solutions (Annotations) --- cont.

4. \comm{label.padding}, \comm{label.r} (corner radius), \comm{label.size} (border width), \comm{fill}, \comm{color}.

5. \comm{angle}, \comm{length}, \comm{ends} ("last", "first", "both"), \comm{type} ("open", "closed").

# Scales

## Scale Naming Convention

```{r}
#| eval: false
# ggplot2 adds these defaults automatically:
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_color_discrete()
```

Pattern: \comm{scale\_} + aesthetic + \comm{\_} + type.

Override defaults to customize **breaks**, **labels**, or the entire scale.

## Custom Breaks and Labels

```{r}
#| fig.height: 2.5
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5))
```

## Formatting with scales Package

```{r}
#| fig.height: 2.5
ggplot(diamonds, aes(x = price, y = cut)) +
  geom_boxplot(alpha = 0.05) +
  scale_x_continuous(
    labels = label_dollar(scale = 1/1000, suffix = "K"),
    breaks = seq(1000, 19000, by = 6000)
  )
```

Also: \comm{label\_percent()}, \comm{label\_comma()}, \comm{label\_number()}.

## Color Scales

```{r}
#| fig.height: 2
#| layout-ncol: 2
#| fig-width: 4
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv))

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_color_brewer(palette = "Set1")
```

Left: default colors. Right: ColorBrewer palette + redundant shape (colorblind-friendly).

## Manual and Continuous Color Scales

```{r}
#| eval: false
# Manual: predefined value-to-color mapping
scale_color_manual(values = c(
  Republican = "#E81B23", Democratic = "#00AEF3"))

# Continuous: viridis (perceptually uniform, colorblind-safe)
scale_fill_viridis_c()   # continuous
scale_fill_viridis_d()   # discrete
scale_fill_viridis_b()   # binned
```

## Log Scales

```{r}
#| fig.height: 2.5
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_bin2d() +
  scale_x_log10() + scale_y_log10()
```

Transform in the scale (not \comm{aes()}) $\to$ axes labeled on original scale.

## Zooming: Three Approaches

```{r}
#| eval: false
# 1. Filter data (changes smooth curve!)
mpg |> filter(displ >= 5 & displ <= 6) |>
  ggplot(aes(x = displ, y = hwy)) + geom_smooth()

# 2. Set scale limits (drops data outside range)
scale_x_continuous(limits = c(5, 6))

# 3. coord_cartesian (just zooms, keeps all data)
coord_cartesian(xlim = c(5, 6), ylim = c(10, 25))
```

**Best practice:** use \comm{coord\_cartesian()} to zoom. Use \comm{limits} to match scales across plots.

## Exercises (Scales)

1. Why doesn't \comm{scale\_color\_gradient()} work on a \comm{geom\_hex()} plot?
2. What is the first argument to every scale? How does it compare to \comm{labs()}?
3. Customize the presidential terms plot: colors, breaks, labels, y-axis.
4. Use \comm{override.aes} to make a legend with \comm{alpha = 1/20} points easier to see.

## Solutions (Scales)

1. \comm{geom\_hex()} uses the \comm{fill} aesthetic, not \comm{color}. Need \comm{scale\_fill\_gradient()}.

2. The first argument is \comm{name} (the axis/legend title) --- same as \comm{labs()}.

3.

```{r}
#| eval: false
presidential |>
  mutate(id = 33 + row_number()) |>
  ggplot(aes(x = start, y = id, color = party)) +
  geom_point() + geom_segment(aes(xend = end, yend = id)) +
  scale_color_manual(values = c(
    Republican = "red", Democratic = "blue")) +
  scale_x_date(breaks = presidential$start,
               date_labels = "'%y")
```

4. Add \comm{guides(color = guide\_legend(override.aes = list(alpha = 1, size = 3)))}.

# Themes

## Built-in Themes

```{r}
#| fig.height: 2
#| message: false
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw()
```

## Theme Gallery

| Theme | Description |
|-------|-------------|
| `theme_gray()` | Default (gray background) |
| `theme_bw()` | White background, grid lines |
| `theme_minimal()` | Minimal, no background |
| `theme_classic()` | Axes only, no grid |
| `theme_light()` | Light axes and grid |
| `theme_dark()` | Dark background |
| `theme_void()` | Nothing --- just geoms |

More themes: **ggthemes** package.

## Customizing Theme Elements

```{r}
#| fig.height: 2
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  labs(title = "Engine size vs. fuel economy",
       caption = "Source: fueleconomy.gov") +
  theme(
    plot.title = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.caption = element_text(hjust = 0),
    legend.position = "bottom"
  )
```

## Exercises (Themes)

1. Pick a theme from **ggthemes** and apply it.
2. Make axis labels blue and bolded.

## Solutions (Themes)

1.

```{r}
#| eval: false
library(ggthemes)
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() + theme_economist()
```

2.

```{r}
#| eval: false
theme(axis.title = element_text(color = "blue", face = "bold"))
```

# Layout

## Combining Plots with patchwork

```{r}
#| fig.height: 2
p1 <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() + labs(title = "Plot 1")
p2 <- ggplot(mpg, aes(x = drv, y = hwy)) +
  geom_boxplot() + labs(title = "Plot 2")
p1 + p2
```

\comm{+} places plots side by side. Load **patchwork** to enable this.

## Complex Layouts

```{r}
#| fig.height: 2.5
p3 <- ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point() + labs(title = "Plot 3")
(p1 | p3) / p2
```

\comm{|} = side by side, \comm{/} = stack vertically, parentheses group.

## patchwork Features

```{r}
#| eval: false
# Shared legend and common title
(p1 + p2) +
  plot_annotation(
    title = "Combined plots",
    caption = "Source: mpg dataset"
  ) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

- \comm{plot\_annotation()}: shared title, subtitle, caption
- \comm{plot\_layout(guides = "collect")}: one combined legend
- \comm{\&}: apply theme to **all** subplots

## Exercises (Layout)

1. What happens if you omit parentheses in \comm{(p1 | p3) / p2}?

## Solutions (Layout)

1. Without parentheses, \comm{p1 | p3 / p2} evaluates \comm{p3 / p2} first (stacks them), then places \comm{p1} next to the stack. Parentheses control grouping, just like in math.

<!--
# Summary

## Key Takeaways

- **\comm{labs()}:** title (main finding), subtitle, caption, axis labels
- **Annotations:** \comm{geom\_text()}, \comm{annotate()}, \comm{geom\_segment(arrow)}
- **Scales:** \comm{scale\_*\_continuous()}, \comm{scale\_*\_discrete()}, \comm{scale\_color\_brewer()}
- **Breaks/labels:** customize tick marks and formatting (\comm{label\_dollar()}, \comm{label\_percent()})
- **Color:** ColorBrewer for categorical; viridis for continuous; manual for fixed mappings
- **Zooming:** \comm{coord\_cartesian()} (keeps data) vs. \comm{limits} (drops data)
- **Themes:** 8 built-in + \comm{theme()} for element-level control
- **Layout:** patchwork: \comm{+}, \comm{|}, \comm{/}, \comm{plot\_annotation()}, \comm{plot\_layout()}
-->
