---
title: "R4DS 04 -- Tidy Data, Import & Workflow"
---

```{r setup, include=FALSE}
source("_common.R")
library(tidyverse)
```

# Tidy Data

## What is Tidy Data?

> "Tidy datasets are all alike, but every messy dataset is messy in its own way." --- Hadley Wickham

Three rules:

1. Each **variable** is a column
2. Each **observation** is a row
3. Each **value** is a cell

\medskip

Why tidy? Consistent structure $\to$ tools just work (dplyr, ggplot2, etc.).

## Tidy vs. Messy

```{r}
#| output.lines: 8
table1
```

This is tidy: each row is a country-year, each column is a variable.

## Messy Example: Values in Column Names

```{r}
#| output.lines: 6
table2
```

Here \comm{type} mixes two variables (\comm{cases} and \comm{population}) into one column.

## Working with Tidy Data

```{r}
#| output.lines: 5
# Rate per 10,000
table1 |>
  mutate(rate = cases / population * 10000)
```

```{r}
#| output.lines: 4
# Total cases per year
table1 |>
  group_by(year) |>
  summarize(total_cases = sum(cases))
```

## Exercises (Tidy Data)

1. For each sample table (\comm{table1}, \comm{table2}, \comm{table3}), describe what each observation and column represents.

2. Sketch the process to calculate \comm{rate} for \comm{table2} and \comm{table3}: (a) extract cases per country-year, (b) extract population, (c) divide and multiply by 10,000, (d) store result.

## Solutions (Tidy Data)

1. \comm{table1}: each row = country-year; columns = country, year, cases, population. \comm{table2}: each row = country-year-type; \comm{count} mixes cases and population. \comm{table3}: each row = country-year; \comm{rate} column stores "cases/population" as a string.

2. For \comm{table2}: filter rows where \comm{type == "cases"}, filter where \comm{type == "population"}, join, divide. For \comm{table3}: use \comm{separate()} to split the \comm{rate} column by "/", convert to numbers, divide.

# Pivoting

## `pivot_longer()` --- Column Names to Values

The \comm{billboard} dataset has weeks as columns:

```{r}
#| output.lines: 5
billboard
```

## Pivoting Billboard Data

```{r}
#| output.lines: 7
billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  )
```

- \comm{cols}: which columns to pivot
- \comm{names\_to}: new variable from old column names
- \comm{values\_to}: new variable from cell values

## Cleaning After Pivot

```{r}
#| output.lines: 6
billboard_longer <- billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week", values_to = "rank",
    values_drop_na = TRUE
  ) |>
  mutate(week = parse_number(week))
billboard_longer
```

## Multiple Variables in Column Names

```{r}
#| output.lines: 5
who2
```

Columns like \comm{sp\_m\_014}: diagnosis / gender / age range.

## Splitting Column Names

```{r}
#| output.lines: 6
who2 |>
  pivot_longer(
    cols = !(country:year),
    names_to = c("diagnosis", "gender", "age"),
    names_sep = "_",
    values_to = "count"
  )
```

\comm{names\_sep} splits each column name into multiple variables.

## `pivot_wider()` --- Values to Column Names

```{r}
#| output.lines: 5
cms_patient_experience |>
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
```

\comm{pivot\_wider()} is the inverse of \comm{pivot\_longer()}: spreads rows into columns.

# Importing Data

## `read_csv()` --- Reading CSV Files

```{r}
#| eval: false
students <- read_csv("data/students.csv")
```

- First argument: path to file (or URL)
- Prints column names and types
- Returns a tibble

## Practical Advice

```{r}
#| eval: false
# Handle non-standard NA values
students <- read_csv("data/students.csv",
                     na = c("N/A", ""))

# Fix non-syntactic names
students |>
  rename(student_id = `Student ID`,
         full_name = `Full Name`)

# Or use janitor::clean_names() for all at once
students |> janitor::clean_names()
```

## Other Useful Arguments

```{r}
#| message: false
# Skip metadata lines
read_csv("metadata line 1
x,y,z
1,2,3", skip = 1)
```

```{r}
#| message: false
# No header row
read_csv("1,2,3
4,5,6", col_names = c("x", "y", "z"))
```

## Other File Types

| Function | Delimiter |
|----------|-----------|
| `read_csv()` | Comma `,` |
| `read_csv2()` | Semicolon `;` |
| `read_tsv()` | Tab |
| `read_delim()` | Any (auto-detect) |
| `read_fwf()` | Fixed width |

## Column Types

readr guesses types: logical $\to$ number $\to$ date $\to$ string.

Override with \comm{col\_types}:

```{r}
#| eval: false
read_csv("file.csv",
         col_types = list(x = col_double()))
```

Fix mismatched types with \comm{na} argument:

```{r}
#| message: false
read_csv("x\n10\n.\n20", na = ".")
```

## Reading Multiple Files

```{r}
#| eval: false
# Stack multiple CSVs into one data frame
sales_files <- list.files("data",
  pattern = "sales\\.csv$", full.names = TRUE)
read_csv(sales_files, id = "file")
```

\comm{id = "file"} adds a column identifying the source file.

## Writing Data

```{r}
#| eval: false
# CSV (loses type info)
write_csv(students, "students.csv")

# RDS (preserves exact R object)
write_rds(students, "students.rds")
read_rds("students.rds")

# Parquet (fast, cross-language)
library(arrow)
write_parquet(students, "students.parquet")
```

**Recommendation:** use RDS for caching R objects, parquet for sharing.

## Data Entry

```{r}
#| output.lines: 3
# By column
tibble(x = c(1, 2, 5), y = c("h", "m", "g"))
```

```{r}
#| output.lines: 3
# By row (easier to read)
tribble(
  ~x, ~y,
   1, "h",
   2, "m",
   5, "g"
)
```

## Exercises (Import)

1. What function reads files delimited with `|`?
2. What arguments do \comm{read\_csv()} and \comm{read\_tsv()} share (besides \comm{file}, \comm{skip}, \comm{comment})?
3. Most important arguments to \comm{read\_fwf()}?
4. How do you read: \comm{"x,y\\n1,'a,b'"}?
5. What's wrong with each?

```{r}
#| eval: false
read_csv("a,b\n1,2,3\n4,5,6")
read_csv("a,b,c\n1,2\n1,2,3,4")
read_csv("a,b\n\"1")
read_csv("a,b\n1,2\na,b")
read_csv("a;b\n1;3")
```

6. Practice with non-syntactic names (extract, plot, mutate, rename).

## Solutions (Import)

1. \comm{read\_delim(file, delim = "|")}

2. Nearly all: \comm{col\_names}, \comm{col\_types}, \comm{locale}, \comm{na}, \comm{trim\_ws}, \comm{n\_max}, \comm{guess\_max}, etc.

3. \comm{file} and \comm{col\_positions} (or \comm{fwf\_widths()} / \comm{fwf\_positions()}).

4. \comm{read\_csv("x,y\\n1,'a,b'", quote = "'")}

5. (a) 3 values but 2 columns --- 3rd value dropped. (b) Mismatched columns per row. (c) Only one value, one column missing. (d) \comm{a,b} in row 2 is character, forces column type to character. (e) Semicolons need \comm{read\_csv2()} or \comm{read\_delim()}.

6.
```{r}
#| eval: false
annoying <- tibble(`1` = 1:10, `2` = `1` * 2 + rnorm(10))
annoying$`1`                                    # extract
ggplot(annoying, aes(x = `1`, y = `2`)) + geom_point()
annoying |> mutate(`3` = `2` / `1`)
annoying |> rename(one = `1`, two = `2`)
```

# Code Style

## Naming Conventions

```{r}
#| eval: false
# Strive for:
short_flights <- flights |> filter(air_time < 60)

# Avoid:
SHORTFLIGHTS <- flights |> filter(air_time < 60)
```

- Use \comm{snake\_case} (lowercase + underscores)
- Prefer descriptive names over short ones
- Common prefix for related variables (autocomplete-friendly)

## Spacing Rules

```{r}
#| eval: false
# Strive for:
z <- (a + b)^2 / d
mean(x, na.rm = TRUE)

# Avoid:
z<-( a + b ) ^ 2/d
mean (x ,na.rm=TRUE)
```

Spaces around operators (\comm{+}, \comm{-}, \comm{==}, \comm{<-}) and after commas. No spaces inside parentheses.

## Pipe and ggplot2 Formatting

```{r}
#| eval: false
# Strive for:
flights |>
  filter(!is.na(arr_delay), !is.na(tailnum)) |>
  group_by(dest) |>
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

# ggplot2: treat + like |>
flights |>
  group_by(month) |>
  summarize(delay = mean(arr_delay, na.rm = TRUE)) |>
  ggplot(aes(x = month, y = delay)) +
  geom_point() +
  geom_line()
```

## Exercises (Style)

Restyle the following:

```{r}
#| eval: false
flights|>filter(dest=="IAH")|>group_by(year,month,
day)|>summarize(n=n(),delay=mean(arr_delay,
na.rm=TRUE))|>filter(n>10)

flights|>filter(carrier=="UA",dest%in%c("IAH",
"HOU"),sched_dep_time>0900,sched_arr_time<2000)|>
group_by(flight)|>summarize(delay=mean(arr_delay,
na.rm=TRUE),cancelled=sum(is.na(arr_delay)),
n=n())|>filter(n>10)
```

## Solutions (Style) --- 1

```{r}
#| eval: false
flights |>
  filter(dest == "IAH") |>
  group_by(year, month, day) |>
  summarize(n = n(),
            delay = mean(arr_delay, na.rm = TRUE)) |>
  filter(n > 10)
```

## Solutions (Style) --- 2

```{r}
#| eval: false
flights |>
  filter(
    carrier == "UA",
    dest %in% c("IAH", "HOU"),
    sched_dep_time > 0900,
    sched_arr_time < 2000
  ) |>
  group_by(flight) |>
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    cancelled = sum(is.na(arr_delay)),
    n = n()
  ) |>
  filter(n > 10)
```

# Workflow: Scripts & Help

## Scripts vs. Console

- **Console:** experiment, try things
- **Scripts** (\comm{.R} files): save your work, reproduce later
- Open new script: **Ctrl + Shift + N**
- Run current line: **Ctrl + Enter**
- Run entire script: **Ctrl + Shift + S**

\medskip

**Use sectioning comments** to organize:

```{r}
#| eval: false
# Load data ----
# Clean data ----
# Plot data ----
```

## RStudio Projects

- A project = a folder + a \comm{.Rproj} file
- Sets the **working directory** automatically
- All paths become **relative** to the project root
- **Never use** \comm{setwd()} or absolute paths
- Create via: File $\to$ New Project

## Getting Help

1. \comm{?function\_name} or \comm{help(function\_name)}
2. **Google** the error message
3. **Stack Overflow** and **Posit Community**
4. Create a **reprex** (reproducible example) with the \comm{reprex} package
5. Include: minimal data, minimal code, the exact error

<!--
# Summary

## Key Takeaways

- **Tidy data:** each variable a column, each observation a row, each value a cell
- **\comm{pivot\_longer()}:** column names $\to$ values (most common)
- **\comm{pivot\_wider()}:** values $\to$ column names
- **\comm{read\_csv()}:** main import tool; set \comm{na}, \comm{skip}, \comm{col\_types} as needed
- **\comm{write\_rds()} / parquet:** preserve types when saving
- **\comm{tibble()} / \comm{tribble()}:** create data frames by hand
- **Code style:** \comm{snake\_case}, spaces around operators, one pipe step per line
- **Projects:** use \comm{.Rproj}, relative paths, never \comm{setwd()}
-->
