---
title: "R4DS 09 -- Strings & Regular Expressions"
---

```{r setup, include=FALSE}
source("_common.R")
library(tidyverse)
```

# Creating Strings

## String Basics

Create strings with `"` (preferred) or `'`:

```{r}
string1 <- "This is a string"
string2 <- 'To include a "quote", use single quotes'
```

\medskip

**Escape sequences** with `\`:

```{r}
double_quote <- "\""
backslash <- "\\"
x <- c("one\ntwo", "one\ttwo", "\u00b5")
str_view(x)
```

## Raw Strings and Special Characters

**Raw strings** avoid escape headaches --- start with \comm{r"(} and end with \comm{)"}:

```{r}
tricky <- r"(She said "It's \"weird\"")"
str_view(tricky)
```

\medskip

Common special characters:

| Escape | Meaning |
|--------|---------|
| `\n` | Newline |
| `\t` | Tab |
| `\\` | Literal backslash |
| `\"` | Literal quote |
| `\uXXXX` | Unicode character |

## Exercises (Creating Strings)

1. Create strings containing:
   - `He said "That's amazing!"`
   - `\a\b\c\d`
   - `\\\\\\`

2. What does \comm{str\_view()} show for `"This\u00a0is\u00a0tricky"`? What is `\u00a0`?

## Solutions (Creating Strings)

1.

```{r}
str_view('He said "That\'s amazing!"')
str_view("\\a\\b\\c\\d")
str_view("\\\\\\\\\\\\")
```

2. `\u00a0` is a non-breaking space. Looks like a space but prevents line breaks. \comm{str\_view()} highlights it distinctly.

# String Operations

## `str_c()` --- Combining Strings

```{r}
str_c("x", "y", "z")
str_c("Hello ", c("John", "Susan"))
```

Propagates NAs (unlike \comm{paste0()}):

```{r}
#| output.lines: 6
df <- tibble(name = c("Flora", "David", "Terra", NA))
df |> mutate(greeting = str_c("Hi ", name, "!"))
```

## `str_glue()` --- Template Strings

Anything inside \comm{\{\}} is evaluated:

```{r}
#| output.lines: 6
df |> mutate(greeting = str_glue("Hi {name}!"))
```

Escape literal braces by doubling: \comm{\{\{} and \comm{\}\}}.

\medskip

\comm{str\_flatten()} collapses a vector into one string (pairs with \comm{summarize()}):

```{r}
str_flatten(c("x", "y", "z"), ", ")
str_flatten(c("x", "y", "z"), ", ", last = ", and ")
```

## Exercises (String Operations)

1. Compare \comm{paste0()} with \comm{str\_c()} for: \comm{str\_c("hi ", NA)} and \comm{str\_c(letters[1:2], letters[1:3])}.
2. What's the difference between \comm{paste()} and \comm{paste0()}? How to recreate \comm{paste()} with \comm{str\_c()}?
3. Convert: \comm{str\_c("The price of ", food, " is ", price)} to \comm{str\_glue()}.

## Solutions (String Operations)

1. \comm{str\_c("hi ", NA)} returns \comm{NA}; \comm{paste0()} returns \comm{"hi NA"}. \comm{str\_c(letters[1:2], letters[1:3])} gives an error (length mismatch); \comm{paste0()} recycles silently.

2. \comm{paste()} separates with a space by default; \comm{paste0()} uses no separator. Recreate: \comm{str\_c(..., sep = " ")}.

3. `str_glue("The price of {food} is {price}")`

# Extracting Data from Strings

## Separating into Rows

\comm{separate\_longer\_delim()} splits a string into rows by delimiter:

```{r}
#| output.lines: 8
df1 <- tibble(x = c("a,b,c", "d,e", "f"))
df1 |> separate_longer_delim(x, delim = ",")
```

\comm{separate\_longer\_position()} splits by character width:

```{r}
#| output.lines: 6
df2 <- tibble(x = c("1211", "131", "21"))
df2 |> separate_longer_position(x, width = 1)
```

## Separating into Columns

\comm{separate\_wider\_delim()} creates new columns:

```{r}
#| output.lines: 5
df3 <- tibble(x = c("a10.1.2022", "b10.2.2011",
                     "e15.1.2015"))
df3 |>
  separate_wider_delim(x, delim = ".",
    names = c("code", "edition", "year"))
```

\comm{separate\_wider\_position()} splits by fixed widths:

```{r}
#| output.lines: 5
df4 <- tibble(x = c("202215TX", "202122LA",
                     "202325CA"))
df4 |> separate_wider_position(x,
    widths = c(year = 4, age = 2, state = 2))
```

## Handling Mismatches

When pieces don't match expected count, use \comm{too\_few} and \comm{too\_many}:

```{r}
#| output.lines: 7
df <- tibble(a = c("1-1-1", "1-1-2", "1-3", "1"))
df |>
  separate_wider_delim(a, delim = "-",
    names = c("x", "y", "z"),
    too_few = "align_start")
```

Options: \comm{"debug"} (diagnose), \comm{"align\_start"} / \comm{"align\_end"} (fill NAs). For \comm{too\_many}: \comm{"drop"} or \comm{"merge"}.

# Working with Letters

## `str_length()` and `str_sub()`

```{r}
str_length(c("a", "R for data science", NA))
```

\comm{str\_sub()} extracts by position (negative = from end):

```{r}
x <- c("Apple", "Banana", "Pear")
str_sub(x, 1, 3)
str_sub(x, -3, -1)
```

Won't fail if string is too short:

```{r}
str_sub("a", 1, 5)
```

## Case Conversion and Locale

```{r}
str_to_upper("hello world")
str_to_lower("HELLO WORLD")
str_to_title("hello world")
```

\medskip

**Encoding:** UTF-8 is the modern standard. Use \comm{locale} argument if reading non-UTF-8 data:

```{r}
#| eval: false
read_csv(file, locale = locale(encoding = "Latin1"))
```

**Locale matters** for case and sorting in non-English text.

## Exercises (Letters)

1. Use \comm{str\_length()} and \comm{str\_sub()} to extract the middle letter of each name in \comm{c("Alice", "Bob", "Catherine")}. What if even length?
2. What does \comm{str\_sub("abc", 1, 5)} return? Why?

## Solutions (Letters)

1.

```{r}
x <- c("Alice", "Bob", "Catherine")
mid <- ceiling(str_length(x) / 2)
str_sub(x, mid, mid)
```

For even-length strings, this picks the left-of-center letter.

2. Returns \comm{"abc"} --- \comm{str\_sub()} silently returns as much as possible without error.

# Regular Expression Basics

## Pattern Matching with `str_view()`

Literal characters match exactly:

```{r}
#| output.lines: 4
str_view(fruit, "berry")
```

The dot \comm{.} matches any character:

```{r}
#| output.lines: 4
str_view(c("a", "ab", "ae", "bd", "ea"), "a.")
```

## Quantifiers

| Quantifier | Meaning |
|------------|---------|
| `?` | 0 or 1 (optional) |
| `+` | 1 or more |
| `*` | 0 or more |
| `{n}` | Exactly n |
| `{n,}` | At least n |
| `{n,m}` | Between n and m |

```{r}
str_view(c("a", "ab", "abb"), "ab?")
str_view(c("a", "ab", "abb"), "ab+")
```

## Character Classes and Alternation

\comm{[]} defines a set; \comm{[\^{}]} negates it:

```{r}
#| output.lines: 3
str_view(words, "[aeiou]x[aeiou]")
```

```{r}
#| output.lines: 3
str_view(words, "[^aeiou]y[^aeiou]")
```

**Alternation** with \comm{|}:

```{r}
#| output.lines: 4
str_view(fruit, "apple|melon|nut")
```

# Key Regex Functions

## `str_detect()` and `str_count()`

\comm{str\_detect()} returns logical --- pairs with \comm{filter()}:

```{r}
str_detect(c("apple", "banana", "pear"), "an")
```

```{r}
#| output.lines: 4
x <- c("apple", "banana", "pear")
str_count(x, "a")
```

\comm{str\_count()} with \comm{mutate()}:

```{r}
#| eval: false
df |> mutate(
  vowels = str_count(name, "[aeiouAEIOU]"))
```

## `str_replace()` and `str_remove()`

```{r}
x <- c("apple", "pear", "banana")
str_replace_all(x, "[aeiou]", "-")
```

```{r}
str_remove_all(x, "[aeiou]")
```

\comm{str\_replace()} replaces first match only; \comm{str\_replace\_all()} replaces all.

\comm{str\_remove()} / \comm{str\_remove\_all()} are shortcuts for replacing with \comm{""}.

## `separate_wider_regex()`

Extract data using regex patterns:

```{r}
#| output.lines: 6
df <- tribble(
  ~str,
  "<Sheryl>-F_34", "<Kisha>-F_45",
  "<Brandon>-N_33", "<Justin>-M_41",
)
df |> separate_wider_regex(str, patterns = c(
    "<", name = "[A-Za-z]+", ">-",
    gender = ".", "_", age = "[0-9]+"))
```

Named patterns become columns; unnamed patterns are consumed silently.

## Exercises (Key Functions)

1. Replace all forward slashes in \comm{"a/b/c/d/e"} with backslashes.
2. Implement a simple \comm{str\_to\_lower()} using \comm{str\_replace\_all()}.
3. Create a regex that matches US telephone numbers (e.g., 123-456-7890).

## Solutions (Key Functions)

1.

```{r}
str_replace_all("a/b/c/d/e", "/", "\\\\")
```

2.

```{r}
#| eval: false
str_replace_all(x,
  set_names(letters, LETTERS))
```

3. `"\\d{3}[- ]?\\d{3}[- ]?\\d{4}"` or `"\\(?\\d{3}\\)?[- ]?\\d{3}[- ]?\\d{4}"` (with optional parens).

# Pattern Details

## Escaping Metacharacters

To match literal \comm{.}, \comm{?}, \comm{+}, etc., escape with `\\`:

```{r}
str_view(c("abc", "a.c", "bef"), "a\\.c")
```

To match literal `\`:

```{r}
x <- "a\\b"
str_view(x, "\\\\")
```

Alternative: use \comm{fixed()} to disable regex entirely:

```{r}
str_view(c("", "a", "."), fixed("."))
```

## Anchors

\comm{\^{}} matches start, \comm{\$} matches end:

```{r}
#| output.lines: 3
str_view(fruit, "^a")
```

```{r}
#| output.lines: 3
str_view(fruit, "a$")
```

Match full string: \comm{\^{}apple\$}. Word boundary: `\\b`.

```{r}
x <- c("summary(x)", "summarize(df)", "sum(x)")
str_view(x, "\\bsum\\b")
```

## Shorthand Character Classes

| Shorthand | Matches | Inverse |
|-----------|---------|---------|
| `\d` | Digits | `\D` |
| `\s` | Whitespace | `\S` |
| `\w` | Word chars (letters, digits) | `\W` |

```{r}
x <- "abc 123 !@#"
str_view(x, "\\d+")
str_view(x, "\\w+")
```

Ranges: \comm{[a-z]}, \comm{[A-Z]}, \comm{[0-9]}.

## Grouping and Back References

Parentheses create **capturing groups**:

```{r}
#| output.lines: 3
# Repeated pair of letters
str_view(fruit, "(..)\\1")
```

Use in \comm{str\_replace()} to reorder:

```{r}
#| output.lines: 3
sentences |>
  str_replace("(\\w+) (\\w+) (\\w+)",
              "\\1 \\3 \\2") |>
  head(3)
```

Non-capturing group: \comm{(?:)}.

## Regex Flags

\comm{regex()} wraps pattern with options:

```{r}
bananas <- c("banana", "Banana", "BANANA")
str_view(bananas,
         regex("banana", ignore_case = TRUE))
```

Other flags: \comm{multiline} (\comm{\^{}}/\comm{\$} per line), \comm{dotall} (\comm{.} matches `\n`), \comm{comments} (allow whitespace/comments in pattern).

## Exercises (Pattern Details)

1. Create regexes to find words in \comm{stringr::words} that:
   - Start with "y"
   - End with "x"
   - Are exactly three letters long
   - Have seven letters or more
2. Describe what these match: `^.*$`, `\d{4}-\d{2}-\d{2}`, `(..)\1`.
3. Switch first and last letters of \comm{words}. Which are still valid words?

## Solutions (Pattern Details) --- 1

```{r}
#| output.lines: 3
str_view(words, "^y")       # start with y
```

```{r}
#| output.lines: 3
str_view(words, "x$")       # end with x
```

```{r}
#| output.lines: 3
str_view(words, "^...$")    # exactly 3 letters
```

```{r}
#| output.lines: 3
str_view(words, "^.{7,}$")  # 7+ letters
```

## Solutions (Pattern Details) --- 2--3

2. `^.*$` matches any entire string. `\d{4}-\d{2}-\d{2}` matches dates like 2024-03-31. `(..)\1` matches repeated two-character sequences (e.g., "anan" in banana).

3.

```{r}
#| output.lines: 5
swapped <- str_replace(words, "^(.)(.*)(.)",
                       "\\3\\2\\1")
intersect(swapped, words)
```

<!--
# Summary

## Key Takeaways

- **Creating:** `"..."`, escapes (`\n`, `\\`), raw strings `r"(...)"`
- **Combining:** \comm{str\_c()}, \comm{str\_glue("\{x\}")}, \comm{str\_flatten()}
- **Separating:** \comm{separate\_wider/longer\_delim()}, \comm{\_position()}, \comm{\_regex()}
- **Letters:** \comm{str\_length()}, \comm{str\_sub()}, \comm{str\_to\_lower/upper/title()}
- **Regex basics:** `.` (any), `?+*` (quantifiers), `[]` (class), `|` (alternation)
- **Functions:** \comm{str\_detect()}, \comm{str\_count()}, \comm{str\_replace\_all()}, \comm{str\_remove\_all()}
- **Escaping:** `\\.` for literal dot, `\\\\` for literal backslash, or use \comm{fixed()}
- **Anchors:** `^` start, `$` end, `\\b` word boundary
- **Shortcuts:** `\\d` digits, `\\s` whitespace, `\\w` word chars
- **Groups:** `()` capture, `\\1` back reference, `(?:)` non-capturing
-->
