---
title: "R4DS 01 -- Introduction & Workflow Basics"
---

```{r setup, include=FALSE}
source("_common.R")
```

# Introduction

## What is Data Science?

- Transform **raw data** into understanding, insight, and knowledge
- This course follows *R for Data Science* (2e) by Wickham, Çetinkaya-Rundel & Grolemund
- Tools: R + tidyverse ecosystem
- **80/20 rule**: the tools here cover ~80% of every project

## The Data Science Workflow

\begin{center}
\textbf{Import} $\;\rightarrow\;$ \textbf{Tidy} $\;\rightarrow\;$
\textbf{Transform} $\;\leftrightarrow\;$ \textbf{Visualize}
$\;\rightarrow\;$ \textbf{Communicate}
\end{center}

\medskip

- **Import**: load data from files, databases, APIs into R
- **Tidy**: each column = variable, each row = observation
- **Transform**: filter, create new variables, summarize
- **Visualize**: discover patterns, raise new questions
- **Communicate**: share results with others
- **Program**: the cross-cutting tool used in every step

## What This Course Covers (and Doesn't)

**Covers:**

- Data wrangling (import, tidy, transform)
- Visualization (ggplot2)
- Programming in R (functions, iteration)
- Communication (Quarto)

**Does not cover:**

- Modeling (see *Tidy Modeling with R*)
- Big data (see \comm{data.table} for 10--100 GB)
- Other languages (Python, Julia)

# Prerequisites

## R and RStudio

**R** --- the language

- Download from CRAN: \texttt{https://cloud.r-project.org}
- Recommend R $\geq$ 4.2.0

**RStudio** --- the IDE

- Download from: \texttt{https://posit.co/download/rstudio-desktop/}
- Console pane (left): type and run R code
- Output pane (right): plots, help, files

## The Tidyverse

An R **package** = collection of functions + data + documentation

Install the tidyverse (run once):

```{r}
#| eval: false
install.packages("tidyverse")
```

Load it (every session):

```{r}
library(tidyverse)
```

Core packages: \comm{ggplot2}, \comm{dplyr}, \comm{tidyr}, \comm{readr}, \comm{purrr}, \comm{tibble}, \comm{stringr}, \comm{forcats}, \comm{lubridate}

## Other Useful Packages

```{r}
#| eval: false
install.packages(
  c("arrow", "babynames", "curl", "duckdb",
    "gapminder", "ggrepel", "ggridges",
    "ggthemes", "hexbin", "janitor", "Lahman",
    "leaflet", "maps", "nycflights13",
    "openxlsx", "palmerpenguins", "repurrrsive",
    "tidymodels", "writexl")
)
```

If you see: \texttt{Error: there is no package called 'foo'} --- just run \comm{install.packages("foo")}

# Coding Basics

## Arithmetic

R works as a calculator:

```{r}
1 / 200 * 30
```

```{r}
(59 + 73 + 2) / 3
```

```{r}
sin(pi / 2)
```

## Assignment with `<-`

Create objects with the **assignment operator** `<-`:

```{r}
x <- 3 * 4
x
```

\medskip

- Read as: "x **gets** 3 times 4"
- RStudio shortcut: **Alt + -** (inserts \comm{<-} with spaces)
- The value is stored, not printed --- type the name to see it

## Vectors

Combine elements with \comm{c()}:

```{r}
primes <- c(2, 3, 5, 7, 11, 13)
primes
```

Arithmetic is **vectorized** (applied element-wise):

```{r}
primes * 2
```

```{r}
primes - 1
```

# Comments and Naming

## Comments

Everything after `#` is ignored by R:

```{r}
# create vector of primes
primes <- c(2, 3, 5, 7, 11, 13)

# multiply primes by 2
primes * 2
```

\medskip

**Best practice:** comment on *why*, not *what* or *how*

## Object Names

Rules:

- Must start with a letter
- Can contain: letters, numbers, `_`, `.`

Convention --- use **snake\_case**:

```{r}
#| eval: false
i_use_snake_case       # recommended
otherPeopleUseCamelCase
some.people.use.periods
```

## Case Sensitivity and Typos

R is **case-sensitive** and **exact**:

```{r}
r_rocks <- 2^3
```

```{r}
#| eval: false
r_rock
#> Error: object 'r_rock' not found
R_rocks
#> Error: object 'R_rocks' not found
```

\medskip

**The contract:** R does the tedious computation; you must be precise in your instructions.

# Calling Functions

## Function Syntax

General form:

```{r}
#| eval: false
function_name(argument1 = value1, argument2 = value2)
```

Example with \comm{seq()}:

```{r}
seq(from = 1, to = 10)
```

Named arguments can be shortened by position:

```{r}
seq(1, 10)
```

## RStudio Shortcuts

| Action | Shortcut |
|--------|----------|
| Assignment `<-` | Alt + - |
| Run current line | Ctrl + Enter |
| Tab completion | TAB |
| Command history | $\uparrow$ |
| Filter history | Ctrl + $\uparrow$ |
| Keyboard shortcut list | Alt + Shift + K |

\medskip

**Tip:** RStudio auto-completes function names, arguments, and file paths. Type a few characters and press TAB.

## Strings and Matching Pairs

```{r}
x <- "hello world"
x
```

\medskip

Quotation marks and parentheses must come in **matched pairs**.

If you see `+` in the console, R is waiting for more input:

```
> x <- "hello
+
```

Press **Escape** to abort, then fix the missing `"` or `)`.

# Exercises

## Exercise 1

Why does this code not work? Look carefully!

```{r}
#| error: true
my_variable <- 10
my_varıable
```

## Exercise 1 --- Solution

The second line uses a **Turkish dotless i** (\texttt{ı}) instead of a regular \texttt{i}.

\medskip

Correct code:

```{r}
my_variable <- 10
my_variable
```

**Lesson:** tiny typos matter. Train your eyes to spot differences.

## Exercise 2

Fix each line so it runs correctly:

```{r}
#| eval: false
libary(todyverse)

ggplot(dTA = mpg) +
  geom_point(maping = aes(x = displ y = hwy)) +
  geom_smooth(method = "lm)
```

## Exercise 2 --- Solution

```{r}
#| output.lines: 3
library(tidyverse)
```

```{r}
#| fig.height: 2.5
#| message: false
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Fixes: \comm{libary}$\to$\comm{library}, \comm{todyverse}$\to$\comm{tidyverse}, \comm{dTA}$\to$\comm{data}, \comm{maping}$\to$\comm{mapping}, missing comma after \comm{displ}, unclosed quote in \comm{"lm"}, \comm{aes()} moved to \comm{ggplot()} so all layers inherit it.

## Exercise 3

Press **Alt + Shift + K** (Windows/Linux) or **Option + Shift + K** (Mac).

\medskip

What happens?

\medskip

**Answer:** Opens the **Keyboard Shortcut Quick Reference** in RStudio. Same as: Help $\to$ Keyboard Shortcuts Help.

## Exercise 4

Which plot is saved as \texttt{mpg-plot.png}? Why?

```{r}
#| eval: false
my_bar_plot <- ggplot(mpg, aes(x = class)) +
  geom_bar()
my_scatter_plot <- ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
ggsave(filename = "mpg-plot.png", plot = my_bar_plot)
```

\medskip

**Answer:** The **bar plot** is saved, because \comm{ggsave()} uses the \comm{plot} argument explicitly. Without \comm{plot =}, it defaults to the last displayed plot.

<!--
# Summary

## Key Takeaways

- Data science workflow: Import $\to$ Tidy $\to$ Transform $\leftrightarrow$ Visualize $\to$ Communicate
- R + RStudio + tidyverse = core toolkit
- Assignment: \comm{x <- value}
- Vectors: \comm{c(1, 2, 3)} --- arithmetic is vectorized
- Functions: \comm{f(arg1, arg2)}
- Comments: \comm{\# why, not what}
- Names: \comm{snake\_case}, case-sensitive, no typos
- RStudio: TAB completion, Alt+- for \comm{<-}, Ctrl+Enter to run
-->
