---
title: "R4DS 10 -- Factors, Dates & Missing Values"
---

```{r setup, include=FALSE}
source("_common.R")
library(tidyverse)
library(nycflights13)
```

# Factor Basics

## Why Factors?

Factors handle **categorical variables** with a fixed set of possible values.

Two problems with plain strings:

```{r}
x1 <- c("Dec", "Apr", "Jan", "Mar")
sort(x1)  # alphabetical, not calendar order
```

\medskip

1. **No typo protection** --- any string is accepted
2. **No meaningful order** --- sorts alphabetically

## Creating Factors

Define valid **levels**, then create:

```{r}
#| output.lines: 6
month_levels <- c("Jan", "Feb", "Mar", "Apr",
  "May", "Jun", "Jul", "Aug",
  "Sep", "Oct", "Nov", "Dec")
y1 <- factor(x1, levels = month_levels)
y1
sort(y1)
```

Invalid values become \comm{NA} silently. Use \comm{fct()} for strict checking:

```{r}
#| error: true
#| output.lines: 3
fct(c("Dec", "Apr", "Jam"), levels = month_levels)
```

## The `gss_cat` Dataset

\comm{forcats::gss\_cat} --- sample from the General Social Survey:

```{r}
#| output.lines: 8
gss_cat
```

View levels with \comm{count()}:

```{r}
#| output.lines: 6
gss_cat |> count(race)
```

## Exercises (Factor Basics)

1. Explore the distribution of \comm{rincome} (reported income). What makes the default bar chart hard to understand? How to improve it?
2. What is the most common \comm{relig}? Most common \comm{partyid}?
3. Which \comm{relig} does \comm{denom} (denomination) apply to? How can you find out with a table? With a visualization?

## Solutions (Factor Basics)

1. Labels overlap and the order is counterintuitive (income categories are reversed). Improve: reorder levels or use \comm{coord\_flip()}.

2.

```{r}
#| output.lines: 3
gss_cat |> count(relig, sort = TRUE) |> head(1)
gss_cat |> count(partyid, sort = TRUE) |> head(1)
```

3. Mainly Protestant. Check: \comm{gss\_cat |> count(relig, denom)} or \comm{geom\_count(aes(x = relig, y = denom))}.

# Modifying Factor Order

## `fct_reorder()` --- Reorder by Another Variable

```{r}
#| fig.height: 2.5
relig_summary <- gss_cat |>
  group_by(relig) |>
  summarize(tvhours = mean(tvhours, na.rm = TRUE),
            n = n())

ggplot(relig_summary,
       aes(x = tvhours,
           y = fct_reorder(relig, tvhours))) +
  geom_point()
```

## `fct_relevel()` and `fct_reorder2()`

\comm{fct\_relevel()} moves specific levels to front (for principled orderings):

```{r}
#| eval: false
ggplot(rincome_summary,
  aes(x = age,
      y = fct_relevel(rincome, "Not applicable"))) +
  geom_point()
```

\medskip

\comm{fct\_reorder2(.f, .x, .y)} --- reorders by \comm{.y} at largest \comm{.x}. Legend aligns with line endpoints:

```{r}
#| eval: false
ggplot(by_age,
  aes(x = age, y = prop,
      color = fct_reorder2(marital, age, prop))) +
  geom_line() + labs(color = "marital")
```

## `fct_infreq()` and `fct_rev()`

Order by **frequency** --- simplest reordering (no extra variable):

```{r}
#| fig.height: 2.5
gss_cat |>
  mutate(marital = marital |>
           fct_infreq() |> fct_rev()) |>
  ggplot(aes(x = marital)) +
  geom_bar()
```

\comm{fct\_infreq()} orders most-to-least; \comm{fct\_rev()} reverses.

## Exercises (Factor Order)

1. There are suspiciously high numbers in \comm{tvhours}. Is the mean a good summary?
2. For each factor in \comm{gss\_cat}, identify whether the order of levels is arbitrary or principled.
3. Why did moving "Not applicable" to the front of levels move it to the **bottom** of the plot?

## Solutions (Factor Order)

1. No --- outliers skew the mean. Use median, or filter extremes first.

2. \comm{marital}, \comm{race}, \comm{relig}, \comm{denom}, \comm{partyid} are arbitrary. \comm{rincome} is principled (ordered income bands).

3. Because ggplot2 plots factor levels from bottom to top on the y-axis. The first level appears at the bottom.

# Modifying Factor Levels

## `fct_recode()` --- Rename Levels

New values on left, old on right:

```{r}
#| output.lines: 8
gss_cat |>
  mutate(partyid = fct_recode(partyid,
    "Republican, strong" = "Strong republican",
    "Republican, weak"   = "Not str republican",
    "Independent, near rep" = "Ind,near rep",
    "Independent, near dem" = "Ind,near dem",
    "Democrat, weak"  = "Not str democrat",
    "Democrat, strong" = "Strong democrat"
  )) |> count(partyid)
```

## `fct_collapse()` and `fct_lump_*()`

\comm{fct\_collapse()} groups multiple levels at once:

```{r}
#| output.lines: 7
gss_cat |>
  mutate(partyid = fct_collapse(partyid,
    "other" = c("No answer", "Don't know",
                "Other party"),
    "rep" = c("Strong republican",
              "Not str republican"),
    "ind" = c("Ind,near rep", "Independent",
              "Ind,near dem"),
    "dem" = c("Not str democrat",
              "Strong democrat")
  )) |> count(partyid)
```

## `fct_lump_*()` --- Lump Small Groups

```{r}
#| output.lines: 8
gss_cat |>
  mutate(relig = fct_lump_n(relig, n = 5)) |>
  count(relig, sort = TRUE)
```

Family: \comm{fct\_lump\_n()} (keep top n), \comm{fct\_lump\_min()} (min count), \comm{fct\_lump\_prop()} (min proportion), \comm{fct\_lump\_lowfreq()} (keep "Other" smallest).

## Exercises (Factor Levels)

1. How have proportions of people identifying as Democrat, Republican, and Independent changed over time?
2. How could you collapse \comm{rincome} into a small set of categories?
3. In the \comm{fct\_lump} example, why 9 groups (not 10) besides "Other"?

## Solutions (Factor Levels)

1.

```{r}
#| eval: false
gss_cat |>
  mutate(partyid = fct_collapse(partyid,
    "rep" = c("Strong republican",
              "Not str republican"),
    "dem" = c("Not str democrat",
              "Strong democrat"),
    "ind" = c("Ind,near rep", "Independent",
              "Ind,near dem"),
    "other" = c("No answer", "Don't know",
                "Other party"))) |>
  count(year, partyid) |>
  group_by(year) |> mutate(prop = n / sum(n)) |>
  ggplot(aes(year, prop, color = partyid)) +
  geom_line()
```

## Solutions (Factor Levels) --- cont.

2. Group adjacent income bands:

```{r}
#| eval: false
gss_cat |>
  mutate(rincome = fct_collapse(rincome,
    "Unknown" = c("No answer", "Don't know",
                  "Refused", "Not applicable"),
    "< $5000" = c("Lt $1000", "$1000 to 2999",
                  "$3000 to 3999",
                  "$4000 to 4999"),
    "$5000-9999" = c("$5000 to 5999",
      "$6000 to 6999", "$7000 to 7999",
      "$8000 to 9999"),
    "$10000+" = c("$10000 - 14999",
      "$15000 - 19999", "$20000 - 24999",
      "$25000 or more")))
```

3. "Other" counts as one of the groups internally. With \comm{n = 10}, you get 10 groups total: 9 named + "Other".

# Creating Dates and Times

## Three Types

| Type | Tibble prints | Example |
|------|---------------|---------|
| Date | `<date>` | `2024-03-31` |
| Time | `<time>` | `12:30:00` (use **hms** package) |
| Date-time | `<dttm>` | `2024-03-31 12:30:00` |

\medskip

Use the **simplest** type that works. Dates are simpler than date-times (no time zone issues).

```{r}
#| output.lines: 4
today()
now()
```

## From Strings: `ymd()`, `mdy()`, `dmy()`

Arrange "y", "m", "d" in the order they appear:

```{r}
#| output.lines: 6
ymd("2017-01-31")
mdy("January 31st, 2017")
dmy("31-Jan-2017")
```

Add time components with underscore:

```{r}
#| output.lines: 4
ymd_hms("2017-01-31 20:11:59")
mdy_hm("01/31/2017 08:01")
```

## From Components: `make_date()` / `make_datetime()`

When date parts are in separate columns:

```{r}
#| output.lines: 6
flights |>
  select(year, month, day, hour, minute) |>
  mutate(departure = make_datetime(
    year, month, day, hour, minute))
```

## Exercises (Creating Dates)

1. What happens if you parse an invalid date?

```{r}
#| eval: false
ymd(c("2010-10-10", "bananas"))
```

2. What does the \comm{tzone} argument to \comm{today()} do? Why is it important?
3. Parse each of these:

```{r}
#| eval: false
d1 <- "January 1, 2010"
d2 <- "2015-Mar-07"
d3 <- "06-Jun-2017"
d4 <- c("August 19 (2015)", "July 1 (2015)")
d5 <- "12/30/14"
```

## Solutions (Creating Dates)

1. Returns \comm{NA} with a warning for the invalid string.

2. \comm{tzone} sets the time zone used to determine the date. Important because "today" depends on your location --- it can be different dates in different time zones.

3.

```{r}
#| output.lines: 10
mdy("January 1, 2010")
ymd("2015-Mar-07")
dmy("06-Jun-2017")
mdy(c("August 19 (2015)", "July 1 (2015)"))
mdy("12/30/14")
```

# Date-Time Components

## Getting Components

Accessor functions extract parts of a date-time:

```{r}
#| output.lines: 8
datetime <- ymd_hms("2026-07-08 12:34:56")
year(datetime)
month(datetime)
mday(datetime)
yday(datetime)
wday(datetime)
```

\comm{label = TRUE} returns names: \comm{month(datetime, label = TRUE)}.

## Rounding Dates

\comm{floor\_date()}, \comm{round\_date()}, \comm{ceiling\_date()} round to a time unit:

```{r}
#| fig.height: 2.2
#| message: false
flights_dt <- flights |>
  filter(!is.na(dep_time), !is.na(arr_time)) |>
  mutate(dep_time = make_datetime(
    year, month, day,
    dep_time %/% 100, dep_time %% 100))

flights_dt |>
  count(week = floor_date(dep_time, "week")) |>
  ggplot(aes(x = week, y = n)) +
  geom_line() + geom_point()
```

# Time Spans

## Durations --- Exact Seconds

Subtracting dates gives a \comm{difftime}. Lubridate's **durations** always use seconds:

```{r}
#| output.lines: 8
dseconds(15)
dminutes(10)
dhours(c(12, 24))
ddays(0:3)
dyears(1)
```

Durations are exact: \comm{dyears(1)} = 365.25 days in seconds.

## Periods --- Human Units

**Periods** work with "human" time (days, months) without fixed second count:

```{r}
#| output.lines: 6
days(7)
months(1:6)
10 * (months(6) + days(1))
```

## Durations vs. Periods: DST

Durations add exact seconds; periods respect calendar rules:

```{r}
#| output.lines: 6
one_am <- ymd_hms("2026-03-08 01:00:00",
                   tz = "America/New_York")
one_am + ddays(1)   # 2am! (DST = 23-hr day)
one_am + days(1)    # 1am (as expected)
```

\medskip

Leap years:

```{r}
#| output.lines: 4
ymd("2024-01-01") + dyears(1)  # Dec 31!
ymd("2024-01-01") + years(1)   # Jan 1, 2025
```

## Intervals --- Duration with Start Point

\comm{start \%--\% end} creates an interval for precise calculations:

```{r}
#| output.lines: 6
y2023 <- ymd("2023-01-01") %--% ymd("2024-01-01")
y2024 <- ymd("2024-01-01") %--% ymd("2025-01-01")
y2023 / days(1)
y2024 / days(1)
```

**Rule of thumb:** physical time $\to$ duration, human time $\to$ period, how long in human units $\to$ interval.

## Exercises (Time Spans)

1. Explain \comm{days(overnight)} to a beginner. What key fact about R makes it work?
2. Create a vector of dates for the first day of every month in 2015.
3. Write a function that, given your birthday (as a date), returns your age in years.

## Solutions (Time Spans)

1. \comm{overnight} is logical (\comm{TRUE}/\comm{FALSE}). R coerces \comm{TRUE} to 1 and \comm{FALSE} to 0, so \comm{days(TRUE)} = 1 day, \comm{days(FALSE)} = 0 days.

2.

```{r}
#| output.lines: 3
ymd("2015-01-01") + months(0:11)
```

3.

```{r}
#| output.lines: 3
age <- function(bday) {
  (bday %--% today()) %/% years(1)
}
age(ymd("1990-05-15"))
```

## Time Zones

R uses IANA time zones: `"America/New_York"`, `"Europe/Paris"`, `"Pacific/Auckland"`.

\medskip

Two ways to change time zone:

| Function | What it does |
|----------|--------------|
| `with_tz()` | Same instant, different display |
| `force_tz()` | Changes the instant (fix mislabeled tz) |

```{r}
#| output.lines: 4
x <- ymd_hms("2024-06-01 12:00:00",
             tz = "America/New_York")
with_tz(x, "Europe/Paris")
```

# Explicit Missing Values

## `fill()` and `coalesce()`

\comm{fill()} carries the last observation forward:

```{r}
#| output.lines: 8
treatment <- tribble(
  ~person,           ~treatment, ~response,
  "Derrick Whitmore", 1,         7,
  NA,                 2,         10,
  NA,                 3,         NA,
  "Katherine Burke",  1,         4)
treatment |> fill(everything())
```

## `coalesce()` and `na_if()`

Replace \comm{NA} with a fixed value:

```{r}
x <- c(1, 4, 5, 7, NA)
coalesce(x, 0)
```

Replace a sentinel value with \comm{NA}:

```{r}
x <- c(1, 4, 5, 7, -99)
na_if(x, -99)
```

\medskip

**NaN** ("not a number") --- behaves like \comm{NA}. Arises from `0/0`, `Inf - Inf`, `sqrt(-1)`. Detect with \comm{is.nan()}.

# Implicit Missing Values

## Explicit vs. Implicit

```{r}
stocks <- tibble(
  year  = c(2020, 2020, 2020, 2020,
            2021, 2021, 2021),
  qtr   = c(1, 2, 3, 4, 2, 3, 4),
  price = c(1.88, 0.59, 0.35, NA,
            0.92, 0.17, 2.66))
```

- **Explicit**: 2020 Q4 price is \comm{NA} --- the presence of an absence
- **Implicit**: 2021 Q1 is missing entirely --- the absence of a presence

## `complete()` --- Make Implicit Explicit

```{r}
#| output.lines: 10
stocks |> complete(year, qtr)
```

Generates all combinations of variables, filling gaps with \comm{NA}.

Can supply ranges: \comm{complete(year = 2019:2021, qtr)}.

## `pivot_wider()` and Joins

\comm{pivot\_wider()} also reveals implicit missings:

```{r}
#| output.lines: 5
stocks |>
  pivot_wider(names_from = qtr,
              values_from = price)
```

\medskip

\comm{anti\_join()} finds rows in one table with no match in another --- useful for detecting missing data across datasets:

```{r}
#| eval: false
flights |> distinct(faa = dest) |>
  anti_join(airports)
```

## Factors and Empty Groups

Factor levels with zero observations are dropped by default:

```{r}
#| output.lines: 6
health <- tibble(
  name = c("Ikaia", "Oletta", "Leriah"),
  smoker = factor(c("no", "no", "no"),
                  levels = c("yes", "no")))
health |> count(smoker)
```

Keep all levels with \comm{.drop = FALSE}:

```{r}
#| output.lines: 5
health |> count(smoker, .drop = FALSE)
```

Same for \comm{group\_by(.drop = FALSE)} and ggplot2 \comm{scale\_x\_discrete(drop = FALSE)}.

## Exercises (Missing Values)

1. What happens to missing values in a histogram vs. a bar chart?
2. What do \comm{mean()} and \comm{sum()} return for a zero-length vector vs. a vector with \comm{NA}?
3. Can you find a relationship between \comm{carrier} and the rows missing from \comm{planes}?

## Solutions (Missing Values)

1. Histograms silently drop NAs (with a warning). Bar charts show NAs as a separate category.

2.

```{r}
#| output.lines: 6
mean(c(1, NA))        # NA (infectious)
mean(c(1, NA), na.rm = TRUE)  # 1
mean(numeric())       # NaN (0/0)
sum(numeric())        # 0
```

3. \comm{MQ} and \comm{AA} have many planes not in the \comm{planes} table --- they report tail numbers differently.

<!--
# Summary

## Key Takeaways

- **Factors:** \comm{factor()} / \comm{fct()} create; \comm{levels()} inspects
- **Reorder:** \comm{fct\_reorder()} by variable, \comm{fct\_infreq()} by count, \comm{fct\_relevel()} manually
- **Relabel:** \comm{fct\_recode()}, \comm{fct\_collapse()}, \comm{fct\_lump\_n()}
- **Create dates:** \comm{ymd()}, \comm{mdy()}, \comm{dmy()}; \comm{make\_date()}/\comm{make\_datetime()}
- **Components:** \comm{year()}, \comm{month()}, \comm{wday()}, \comm{hour()}, etc.
- **Rounding:** \comm{floor\_date()}, \comm{ceiling\_date()}, \comm{round\_date()}
- **Durations:** exact seconds (\comm{ddays()}, \comm{dyears()})
- **Periods:** human units (\comm{days()}, \comm{months()}, \comm{years()})
- **Intervals:** \comm{start \%--\% end} for precise spans
- **Time zones:** \comm{with\_tz()} (display) vs. \comm{force\_tz()} (fix)
- **Explicit NA:** \comm{fill()}, \comm{coalesce()}, \comm{na\_if()}
- **Implicit NA:** \comm{complete()}, \comm{pivot\_wider()}, \comm{anti\_join()}
- **Empty groups:** \comm{.drop = FALSE} in \comm{count()}/\comm{group\_by()}
-->
