---
title: "R4DS 06 -- Exploratory Data Analysis"
---

```{r setup, include=FALSE}
source("_common.R")
library(tidyverse)
```

# Introduction

## What is EDA?

> "Far better an approximate answer to the right question, which is often vague, than an exact answer to the wrong question." --- John Tukey

EDA is an **iterative cycle**:

1. Generate **questions** about your data
2. **Visualize**, transform, and model to search for answers
3. **Refine** your questions based on what you learn

\medskip

Two universal questions:

- What type of **variation** occurs within my variables?
- What type of **covariation** occurs between my variables?

# Variation

## Visualizing Distributions

```{r}
#| fig.height: 2.5
ggplot(diamonds, aes(x = carat)) +
  geom_histogram(binwidth = 0.5)
```

Right-skewed: most diamonds are small. Peaks at round values (0.5, 1, 1.5, 2).

## Zoom In: Smaller Binwidth

```{r}
#| fig.height: 2.5
smaller <- diamonds |> filter(carat < 3)

ggplot(smaller, aes(x = carat)) +
  geom_histogram(binwidth = 0.01)
```

Spikes at whole/half carats --- rounding by sellers. More diamonds just above each round number.

## Questions to Ask

When exploring a distribution, look for:

- **Typical values:** Which are most common? Why?
- **Rare values:** Which are rare? Does that match expectations?
- **Unusual patterns:** Clusters, spikes, gaps?
- **Subgroups:** Do clusters suggest distinct populations?

\medskip

Follow-up: How are observations within clusters similar? How do clusters differ?

## Unusual Values (Outliers)

```{r}
#| fig.height: 2
#| layout-ncol: 2
#| fig-width: 4
ggplot(diamonds, aes(x = y)) +
  geom_histogram(binwidth = 0.5)

ggplot(diamonds, aes(x = y)) +
  geom_histogram(binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 50))
```

\comm{coord\_cartesian()} zooms without dropping data. Reveals outliers at 0, ~30, ~60 mm.

## Identifying Outliers

```{r}
#| output.lines: 8
diamonds |>
  filter(y < 3 | y > 20) |>
  select(price, x, y, z) |>
  arrange(y)
```

Width of 0 mm = data entry error. 32 mm and 59 mm = implausible for these prices.

## Exercises (Variation)

1. Explore the distribution of \comm{x}, \comm{y}, and \comm{z} in \comm{diamonds}. What do you learn?
2. Explore the distribution of \comm{price}. Anything unusual? (Hint: try many binwidths.)
3. How many diamonds are 0.99 carat? How many are 1 carat? Why the difference?
4. Compare \comm{coord\_cartesian()} vs. \comm{xlim()}/\comm{ylim()} when zooming on a histogram.

## Solutions (Variation) --- 1--2

1.

```{r}
#| fig.height: 1.5
#| warning: false
ggplot(diamonds, aes(x = x)) +
  geom_histogram(binwidth = 0.1) +
  coord_cartesian(xlim = c(3, 11))
```

\comm{x} and \comm{y} are similar (length/width); \comm{z} is smaller (depth). All have outliers near 0 and at high extremes.

## Solutions (Variation) --- 2--4

2. No diamonds priced around \$1,500 --- a gap in the distribution. Visible with small binwidth.

3.

```{r}
#| output.lines: 4
diamonds |> filter(carat == 0.99) |> nrow()
diamonds |> filter(carat == 1) |> nrow()
```

23 vs. 1,558 --- sellers round up to 1 carat (higher perceived value).

4. \comm{coord\_cartesian()} zooms visually; \comm{xlim()}/\comm{ylim()} **drops data** outside the range (changes bin counts).

# Handling Unusual Values

## Replace, Don't Drop

```{r}
#| eval: false
# BAD: drops entire row
diamonds2 <- diamonds |> filter(between(y, 3, 20))

# GOOD: replace outliers with NA
diamonds2 <- diamonds |>
  mutate(y = if_else(y < 3 | y > 20, NA, y))
```

Don't drop rows --- one bad value doesn't invalidate the whole observation. Use \comm{if\_else()} to recode outliers as \comm{NA}.

## Working with Missing Values

```{r}
#| fig.height: 2.2
#| warning: false
#| message: false
nycflights13::flights |>
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) |>
  ggplot(aes(x = sched_dep_time)) +
  geom_freqpoly(aes(color = cancelled), binwidth = 1/4)
```

## Exercises (Missing Values)

1. What happens to missing values in a histogram? In a bar chart? Why the difference?
2. What does \comm{na.rm = TRUE} do in \comm{mean()} and \comm{sum()}?

## Solutions (Missing Values)

1. Histograms **silently drop** NAs (with a warning). Bar charts show NAs as a separate category. Histograms need numeric bins; NAs have no position. Bar charts treat NAs as another level.

2. \comm{na.rm = TRUE} removes NAs before computing. Without it, \comm{mean()} and \comm{sum()} return \comm{NA} if any input is missing.

# Covariation

## Categorical + Numerical: Frequency Polygon

```{r}
#| fig.height: 2.5
ggplot(diamonds, aes(x = price, y = after_stat(density))) +
  geom_freqpoly(aes(color = cut), binwidth = 500,
                linewidth = 0.75)
```

\comm{after\_stat(density)} standardizes area to 1. Surprise: Fair diamonds appear most expensive!

## Categorical + Numerical: Boxplot

```{r}
#| fig.height: 2.5
ggplot(diamonds, aes(x = cut, y = price)) +
  geom_boxplot()
```

Confirms: lower quality $\to$ higher median price. (Confounded by carat --- explained later.)

## Reordering with `fct_reorder()`

```{r}
#| fig.height: 2.5
ggplot(mpg, aes(x = hwy,
       y = fct_reorder(class, hwy, median))) +
  geom_boxplot()
```

\comm{fct\_reorder(class, hwy, median)} sorts categories by median \comm{hwy}. Swap axes for long labels.

## Exercises (Categorical + Numerical)

1. Improve the cancelled vs. non-cancelled flights visualization.
2. Which variable best predicts diamond price? How is it correlated with \comm{cut}? Why does this make lower-quality diamonds more expensive?
3. Try \comm{coord\_flip()} vs. swapping \comm{x}/\comm{y}.
4. Compare: \comm{geom\_violin()}, faceted \comm{geom\_histogram()}, colored \comm{geom\_freqpoly()}, colored \comm{geom\_density()}.

## Solutions (Cat. + Num.) --- 1--2

1. Use density or facet with \comm{scales = "free\_y"} to handle the count imbalance.

2. **Carat** predicts price best. Larger (lower-cut) diamonds tend to weigh more. When controlling for carat, better cut $\to$ higher price. Simpson's paradox: the confounding variable (carat) reverses the apparent relationship.

## Solutions (Cat. + Num.) --- 3--4

3. Same result visually. Swapping \comm{x}/\comm{y} is more explicit; \comm{coord\_flip()} is convenient for modifying existing code.

4. \comm{geom\_violin()}: shows full shape, compact. \comm{geom\_histogram()}: familiar, but faceting uses more space. \comm{geom\_freqpoly()}: good for overlaying, hard to read with many groups. \comm{geom\_density()}: smooth, but can hide features.

## Two Categorical Variables

```{r}
#| fig.height: 2.5
ggplot(diamonds, aes(x = cut, y = color)) +
  geom_count()
```

Point size $\propto$ count. Alternative: \comm{geom\_tile()} with \comm{fill = n}.

## Heatmap with `geom_tile()`

```{r}
#| fig.height: 2.5
diamonds |>
  count(color, cut) |>
  ggplot(aes(x = color, y = cut)) +
  geom_tile(aes(fill = n))
```

## Exercises (Two Categorical)

1. Rescale the count data to show distribution of \comm{cut} within \comm{color} (or vice versa).
2. Map \comm{color} to \comm{x} and \comm{cut} to \comm{fill} in a segmented bar chart. What insight?
3. Use \comm{geom\_tile()} with dplyr to show average departure delay by destination and month. What makes it hard to read?

## Solutions (Two Categorical)

1. Compute proportions within groups:

```{r}
#| eval: false
diamonds |>
  count(color, cut) |>
  group_by(color) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = color, y = cut, fill = prop)) +
  geom_tile()
```

2. Shows color composition per cut level. Compute with \comm{count(cut, color)}.

3. Too many destinations --- hard to read. Filter to busy destinations, or reorder by delay. Missing combos also create gaps.

## Two Numerical Variables

```{r}
#| fig.height: 2
#| layout-ncol: 2
#| fig-width: 4
#| warning: false
ggplot(smaller, aes(x = carat, y = price)) +
  geom_point(alpha = 1/100)

ggplot(smaller, aes(x = carat, y = price)) +
  geom_bin2d()
```

Left: \comm{alpha} reveals density. Right: \comm{geom\_bin2d()} bins into rectangles, colors by count.

## Binned Boxplots

```{r}
#| fig.height: 2.5
ggplot(smaller, aes(x = carat, y = price)) +
  geom_boxplot(aes(group = cut_width(carat, 0.1)))
```

\comm{cut\_width(carat, 0.1)} bins carat into 0.1-wide groups. Shows how price varies across carat ranges.

## Exercises (Two Numerical)

1. \comm{cut\_width()} vs. \comm{cut\_number()}: what to consider for frequency polygons?
2. Visualize distribution of \comm{carat} partitioned by \comm{price}.
3. How does price distribution of very large vs. small diamonds compare?
4. Combine techniques to visualize \comm{cut}, \comm{carat}, and \comm{price} together.

## Solutions (Two Numerical)

1. \comm{cut\_width()} creates equal-width bins (some may have few points). \comm{cut\_number()} creates equal-count bins (widths vary). For frequency polygons, \comm{cut\_number()} ensures each line has enough data.

2. `ggplot(diamonds, aes(x = price, y = carat)) + geom_boxplot(aes(group = cut_width(price, 2000)))`

3. Very large diamonds have high price variance. Fewer observations $\to$ less price competition $\to$ wider spread.

4. Example: `ggplot(diamonds, aes(x = carat, y = price, color = cut)) + geom_point(alpha = 0.1)` or facet by \comm{cut}.

# Patterns and Models

## Confounding: Cut vs. Price

Why are Fair diamonds more expensive? **Carat confounds the relationship.**

```{r}
#| fig.height: 2.5
ggplot(diamonds, aes(x = cut, y = carat)) +
  geom_boxplot()
```

Fair diamonds tend to be larger $\to$ higher carat $\to$ higher price.

## Removing Confounding with a Model

```{r}
#| fig.height: 2.2
#| message: false
mod <- lm(log(price) ~ log(carat), data = diamonds)
diamonds2 <- diamonds |>
  mutate(resid = exp(residuals(mod)))

ggplot(diamonds2, aes(x = cut, y = resid)) +
  geom_boxplot()
```

After removing carat effect: better cut $\to$ higher price (as expected).

<!--
# Summary

## Key Takeaways

- **EDA** is iterative: question $\to$ visualize $\to$ refine
- **Variation:** histograms, binwidth matters, look for clusters and outliers
- **Outliers:** use \comm{coord\_cartesian()} to zoom; replace with \comm{NA}, don't drop rows
- **Covariation (cat. + num.):** boxplots, \comm{geom\_freqpoly()}, \comm{fct\_reorder()}
- **Covariation (cat. + cat.):** \comm{geom\_count()}, \comm{geom\_tile()}
- **Covariation (num. + num.):** scatterplot + alpha, \comm{geom\_bin2d()}, binned boxplots
- **Confounding:** models remove strong effects to reveal subtler relationships
- Always ask: "Could this be misleading?"
-->
