---
title: "R4DS 03 -- Data Transformation"
---

```{r setup, include=FALSE}
source("_common.R")
```

# Introduction

## dplyr: The Data Transformation Toolbox

- **dplyr** provides verbs for the most common data manipulation tasks
- All verbs share a common pattern:
  1. First argument: a data frame
  2. Subsequent arguments: what to do (using column names, no quotes)
  3. Output: a new data frame
- Combine verbs with the **pipe** `|>`

## Prerequisites

```{r}
#| output.lines: 3
library(nycflights13)
library(tidyverse)
```

## The `flights` Data

```{r}
#| output.lines: 7
flights
```

- `r format(nrow(flights), big.mark = ",")` flights departing NYC in 2013
- Types: \comm{<int>} integer, \comm{<dbl>} double, \comm{<chr>} character, \comm{<dttm>} date-time

## `glimpse()` --- See All Variables

```{r}
#| output.lines: 12
glimpse(flights)
```

## dplyr Verbs at a Glance

| Category | Verb | Action |
|----------|------|--------|
| Rows | `filter()` | Keep rows that match conditions |
| Rows | `arrange()` | Reorder rows |
| Rows | `distinct()` | Remove duplicate rows |
| Columns | `mutate()` | Create/modify columns |
| Columns | `select()` | Pick columns |
| Columns | `rename()` | Rename columns |
| Columns | `relocate()` | Reorder columns |
| Groups | `group_by()` | Define groups |
| Groups | `summarize()` | Collapse groups to summaries |

# Rows

## `filter()` --- Keep Rows by Condition

```{r}
#| output.lines: 6
# Flights that departed more than 2 hours late
flights |>
  filter(dep_delay > 120)
```

## `filter()` --- Combining Conditions

```{r}
#| output.lines: 5
# Flights on January 1
flights |>
  filter(month == 1 & day == 1)
```

```{r}
#| output.lines: 5
# Flights in January or February (shortcut with %in%)
flights |>
  filter(month %in% c(1, 2))
```

## `filter()` --- Common Mistakes

\textbf{Use \comm{==} not \comm{=} for comparison:}

```{r}
#| error: true
flights |>
  filter(month = 1)
```

\textbf{Use \comm{\%in\%} or \comm{|}, not English "or":}

```{r}
#| eval: false
# WRONG: month == 1 | 2  (always TRUE!)
# RIGHT:
flights |> filter(month == 1 | month == 2)
flights |> filter(month %in% c(1, 2))
```

## `arrange()` --- Reorder Rows

```{r}
#| output.lines: 5
flights |>
  arrange(year, month, day, dep_time)
```

Use \comm{desc()} for descending order:

```{r}
#| output.lines: 5
flights |>
  arrange(desc(dep_delay))
```

## `distinct()` --- Unique Rows

```{r}
#| output.lines: 5
flights |>
  distinct(origin, dest)
```

Use \comm{count()} to also get frequencies:

```{r}
#| output.lines: 5
flights |>
  count(origin, dest, sort = TRUE)
```

## Exercises 3.2.5

1. Find all flights that:
   - Arrived $>$ 2 hours late
   - Flew to Houston (\comm{IAH} or \comm{HOU})
   - Operated by United, American, or Delta
   - Departed in summer (Jul--Sep)
   - Arrived $>$ 2 hours late but didn't leave late
   - Delayed $\geq$ 1 hour but made up $>$ 30 min in flight

## Exercises 3.2.5 (cont.)

2. Sort \comm{flights} to find longest departure delays. Find earliest morning flights.
3. Sort to find fastest flights. (Hint: use math in the function.)
4. Was there a flight on every day of 2013?
5. Which flights traveled farthest? Least distance?
6. Does order of \comm{filter()} and \comm{arrange()} matter? Why/why not?

## Solutions 3.2.5 (1)

```{r}
#| eval: false
# a) Arrived > 2 hours late
flights |> filter(arr_delay > 120)
# b) Flew to Houston
flights |> filter(dest %in% c("IAH", "HOU"))
# c) United, American, or Delta
flights |> filter(carrier %in% c("UA", "AA", "DL"))
# d) Summer
flights |> filter(month %in% c(7, 8, 9))
# e) Arrived >2h late, didn't leave late
flights |> filter(arr_delay > 120 & dep_delay <= 0)
# f) Delayed >=1h, made up >30 min
flights |> filter(dep_delay >= 60 &
                  dep_delay - arr_delay > 30)
```

## Solutions 3.2.5 (2--6)

2.
```{r}
#| eval: false
flights |> arrange(desc(dep_delay))   # longest delay
flights |> arrange(dep_time)          # earliest morning
```

3. `flights |> arrange(air_time)` or `flights |> arrange(distance / air_time)`

4. `flights |> distinct(month, day) |> nrow()` --- yes, 365 rows.

5. `flights |> arrange(desc(distance))` (farthest) and `flights |> arrange(distance)` (shortest).

6. **No** --- same rows and same order either way. But filtering first is more efficient (fewer rows to sort).

# Columns

## `mutate()` --- Create New Columns

```{r}
#| output.lines: 6
flights |>
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60,
    .before = 1
  )
```

\comm{.before = 1} places new columns at the front.

## `mutate()` --- `.keep` Argument

```{r}
#| output.lines: 6
flights |>
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .keep = "used"
  )
```

\comm{.keep = "used"} keeps only input and output columns.

## `select()` --- Pick Columns

```{r}
#| eval: false
flights |> select(year, month, day)       # by name
flights |> select(year:day)               # range
flights |> select(!year:day)              # exclude range
flights |> select(where(is.character))    # by type
```

**Helper functions:** \comm{starts\_with()}, \comm{ends\_with()}, \comm{contains()}, \comm{num\_range()}

Rename while selecting:

```{r}
#| output.lines: 5
flights |>
  select(tail_num = tailnum)
```

## `rename()` and `relocate()`

\comm{rename()} keeps all columns, just renames:

```{r}
#| eval: false
flights |> rename(tail_num = tailnum)
```

\comm{relocate()} moves columns:

```{r}
#| output.lines: 5
flights |>
  relocate(time_hour, air_time)
```

Use \comm{.before} / \comm{.after} to control position.

## Exercises 3.3.5

1. Compare \comm{dep\_time}, \comm{sched\_dep\_time}, \comm{dep\_delay}. How are they related?
2. Brainstorm ways to select \comm{dep\_time}, \comm{dep\_delay}, \comm{arr\_time}, \comm{arr\_delay}.
3. What if you specify the same variable multiple times in \comm{select()}?
4. What does \comm{any\_of()} do?
5. Does \comm{select(contains("TIME"))} surprise you?
6. Rename \comm{air\_time} to \comm{air\_time\_min} and move to front.
7. Why doesn't this work?

```{r}
#| eval: false
flights |> select(tailnum) |> arrange(arr_delay)
```

## Solutions 3.3.5 (1--4)

1. `dep_time = sched_dep_time + dep_delay` (in minutes, though `dep_time` is HHMM format).

2. Several ways:

```{r}
#| eval: false
flights |> select(dep_time, dep_delay, arr_time, arr_delay)
flights |> select(starts_with("dep"), starts_with("arr"))
flights |> select(matches("^(dep|arr)_(time|delay)$"))
```

3. Column appears only **once** --- duplicates are silently ignored.

4. \comm{any\_of()} selects columns from a character vector, ignoring names that don't exist (no error).

## Solutions 3.3.5 (5--7)

5. \comm{contains()} is **case-insensitive** by default. Change with \comm{contains("TIME", ignore.case = FALSE)}.

6.

```{r}
#| eval: false
flights |>
  rename(air_time_min = air_time) |>
  relocate(air_time_min)
```

7. After \comm{select(tailnum)}, only \comm{tailnum} remains --- \comm{arr\_delay} no longer exists.

# The Pipe

## Chaining Verbs with `|>`

```{r}
#| output.lines: 5
# Fastest flights to Houston
flights |>
  filter(dest == "IAH") |>
  mutate(speed = distance / air_time * 60) |>
  select(year:day, dep_time, carrier, flight, speed) |>
  arrange(desc(speed))
```

Read as: "Take flights, **then** filter, **then** mutate, **then** select, **then** arrange."

## Without the Pipe

Nesting (hard to read):

```{r}
#| eval: false
arrange(select(mutate(filter(flights,
  dest == "IAH"), speed = distance / air_time * 60),
  year:day, dep_time, carrier, flight, speed),
  desc(speed))
```

Intermediate objects (clutters environment):

```{r}
#| eval: false
flights1 <- filter(flights, dest == "IAH")
flights2 <- mutate(flights1, speed = distance / air_time * 60)
flights3 <- select(flights2, year:day, dep_time, carrier, flight, speed)
arrange(flights3, desc(speed))
```

The pipe is cleaner. Shortcut: **Ctrl + Shift + M**.

# Groups

## `group_by()` + `summarize()`

```{r}
#| output.lines: 8
flights |>
  group_by(month) |>
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n = n()
  )
```

- \comm{na.rm = TRUE} --- ignore missing values
- \comm{n()} --- count rows per group

## The `.by` Syntax (dplyr 1.1.0+)

Per-operation grouping --- no need for \comm{ungroup()}:

```{r}
#| output.lines: 6
flights |>
  summarize(
    delay = mean(dep_delay, na.rm = TRUE),
    n = n(),
    .by = c(origin, dest)
  )
```

\comm{.by} works with all verbs.

## `slice_*()` Functions

| Function | Action |
|----------|--------|
| `slice_head(n = 1)` | First row per group |
| `slice_tail(n = 1)` | Last row per group |
| `slice_min(x, n = 1)` | Row with smallest `x` |
| `slice_max(x, n = 1)` | Row with largest `x` |
| `slice_sample(n = 1)` | Random row per group |

```{r}
#| output.lines: 5
flights |>
  group_by(dest) |>
  slice_max(arr_delay, n = 1) |>
  relocate(dest)
```

## Grouping by Multiple Variables

```{r}
#| output.lines: 6
#| message: false
daily <- flights |> group_by(year, month, day)
daily |> summarize(n = n())
```

Each \comm{summarize()} peels off the last group. Use \comm{.groups = "drop"} to drop all.

## Exercises 3.5.7 (1--5)

1. Which carrier has the worst average delays?
2. Most delayed departure flight to each destination.
3. How do delays vary over the course of the day? (Plot it.)
4. What happens with negative \comm{n} in \comm{slice\_min()}?
5. Explain \comm{count()} in terms of dplyr verbs. What does \comm{sort} do?

## Exercise 3.5.7 (6)

Given:

```{r}
df <- tibble(
  x = 1:5,
  y = c("a", "b", "a", "a", "b"),
  z = c("K", "K", "L", "L", "K")
)
```

Predict the output of each, then check:

a. `df |> group_by(y)`
b. `df |> arrange(y)`
c. `df |> group_by(y) |> summarize(mean_x = mean(x))`
d. `df |> group_by(y, z) |> summarize(mean_x = mean(x))`
e. Same as (d) but with `.groups = "drop"`
f. Compare `summarize(mean_x = mean(x))` vs. `mutate(mean_x = mean(x))` after `group_by(y, z)`

## Solutions 3.5.7 (1--3)

1.

```{r}
#| output.lines: 5
flights |>
  group_by(carrier) |>
  summarize(avg_delay = mean(dep_delay, na.rm = TRUE)) |>
  arrange(desc(avg_delay))
```

Hard to disentangle airport vs. carrier effects (confounded).

## Solutions 3.5.7 (2--3)

2.

```{r}
#| eval: false
flights |>
  group_by(dest) |>
  slice_max(dep_delay, n = 1)
```

3.

```{r}
#| fig.height: 2.2
#| warning: false
flights |>
  group_by(hour) |>
  summarize(avg_delay = mean(dep_delay, na.rm = TRUE)) |>
  ggplot(aes(x = hour, y = avg_delay)) +
  geom_line() + geom_point()
```

Delays increase throughout the day.

## Solutions 3.5.7 (4--6)

4. Negative \comm{n} removes that many rows (keeps all except the $n$ smallest/largest).

5. `count(x)` is equivalent to `group_by(x) |> summarize(n = n())`. The \comm{sort = TRUE} argument arranges results in descending order of \comm{n}.

6.
   a. Same data, but grouped (header shows \comm{Groups: y [2]}).
   b. Rows reordered by \comm{y} (a's first, then b's). Not grouped.
   c. Two rows: mean of \comm{x} for \comm{y = "a"} (mean 3) and \comm{y = "b"} (mean 3.5).
   d. Three rows (one per y-z combo). Message: "grouped by y".
   e. Same three rows, but result is **ungrouped**.
   f. \comm{summarize} collapses to 3 rows; \comm{mutate} keeps all 5 rows, adding the group mean as a new column.

# Case Study

## Aggregates and Sample Size

```{r}
#| output.lines: 5
batters <- Lahman::Batting |>
  group_by(playerID) |>
  summarize(
    performance = sum(H, na.rm = TRUE) / sum(AB, na.rm = TRUE),
    n = sum(AB, na.rm = TRUE)
  )
batters
```

## Variation Decreases with Sample Size

```{r}
#| fig.height: 2.5
#| warning: false
batters |>
  filter(n > 100) |>
  ggplot(aes(x = n, y = performance)) +
  geom_point(alpha = 0.1) +
  geom_smooth(se = FALSE)
```

Always include \comm{n()} when aggregating --- don't draw conclusions from tiny groups.

<!--
# Summary

## Key Takeaways

- **Rows:** \comm{filter()} (keep), \comm{arrange()} (sort), \comm{distinct()} (unique)
- **Columns:** \comm{mutate()} (create), \comm{select()} (pick), \comm{rename()}, \comm{relocate()}
- **Groups:** \comm{group\_by() |> summarize()} or use \comm{.by}
- **Pipe** \comm{|>}: chain verbs, read left-to-right
- \comm{n()} counts rows per group; always pair with summaries
- \comm{na.rm = TRUE} for \comm{mean()}, \comm{sum()}, etc.
- Variation shrinks with sample size --- include counts in aggregates
-->
