---
title: "R4DS 12 -- Advanced Data Import"
---

```{r setup, include=FALSE}
source("_common.R")
library(tidyverse)
```

# Spreadsheets

## Reading Excel Files

\comm{readxl::read\_excel()} reads both `.xls` and `.xlsx`:

```{r}
#| eval: false
library(readxl)
students <- read_excel("students.xlsx",
  col_names = c("id", "name", "food",
                "meal_plan", "age"),
  skip = 1,
  na = c("", "N/A"),
  col_types = c("numeric", "text", "text",
                "text", "text"))
```

Key arguments:

| Argument | Purpose |
|----------|---------|
| `col_names` | Override column names |
| `skip` | Skip header rows |
| `na` | Strings to treat as `NA` |
| `col_types` | Force column types |
| `range` | Read specific cells (`"A5:F15"`) |

## Multiple Sheets

\comm{excel\_sheets()} lists all sheets; \comm{sheet} argument selects one:

```{r}
#| eval: false
excel_sheets("penguins.xlsx")

# Read and combine all sheets
penguins <- bind_rows(
  read_excel("penguins.xlsx",
    sheet = "Torgersen Island", na = "NA"),
  read_excel("penguins.xlsx",
    sheet = "Biscoe Island", na = "NA"),
  read_excel("penguins.xlsx",
    sheet = "Dream Island", na = "NA"))
```

## Writing Excel and Google Sheets

**Write Excel** with \comm{writexl::write\_xlsx()}:

```{r}
#| eval: false
library(writexl)
bake_sale <- tibble(
  item = c("brownie", "cupcake", "cookie"),
  quantity = c(10, 5, 8))
write_xlsx(bake_sale, path = "bake-sale.xlsx")
```

\medskip

**Google Sheets** with \comm{googlesheets4}:

```{r}
#| eval: false
library(googlesheets4)
gs4_deauth()  # for public sheets
read_sheet("sheet_url_or_id",
  col_types = "dcccc")  # shorthand types
```

## Exercises (Spreadsheets)

1. Read a messy Excel file and clean: handle merged cells with \comm{fill()}, filter out non-data rows.
2. What does \comm{janitor::clean\_names()} do? How does it improve imported column names?
3. What happens if you use \comm{read\_xls()} on an \comm{.xlsx} file?

## Solutions (Spreadsheets)

1. Read with \comm{read\_excel()}, then pipe to \comm{fill()} for merged cells and \comm{filter(!is.na(...))} to remove blank rows.

2. \comm{clean\_names()} converts to snake\_case, removes special characters, and handles duplicates:

```{r}
#| output.lines: 3
janitor::make_clean_names(
  c("First Name", "% Change", "1st Column"))
```

3. It will error or produce garbled results --- the formats are different (\comm{.xls} is binary, \comm{.xlsx} is XML-based).

# Databases

## Connecting with DBI

```{r}
#| eval: false
library(DBI)
library(duckdb)

# Connect (duckdb runs in-process, no server)
con <- dbConnect(duckdb())
dbWriteTable(con, "mpg", ggplot2::mpg)
dbWriteTable(con, "diamonds", ggplot2::diamonds)

# List tables and read
dbListTables(con)
dbReadTable(con, "mpg") |> as_tibble()
```

Other backends: \comm{RPostgres}, \comm{RMariaDB}, \comm{RSQLite}, \comm{odbc}.

## Lazy Queries with dbplyr

\comm{tbl()} creates a **lazy** reference --- no data is fetched until \comm{collect()}:

```{r}
#| eval: false
library(dbplyr)
diamonds_db <- tbl(con, "diamonds")

# Build query (lazy --- nothing executes yet)
big <- diamonds_db |>
  filter(price > 15000) |>
  select(carat:clarity, price)

big |> show_query()   # see generated SQL
big |> collect()      # execute and return tibble
```

\comm{show\_query()} is invaluable for learning SQL.

## dplyr $\to$ SQL Translation

| dplyr | SQL |
|-------|-----|
| `select()` | `SELECT` |
| `filter()` | `WHERE` |
| `arrange()` | `ORDER BY` |
| `group_by() + summarize()` | `GROUP BY` |
| `mutate()` | `SELECT expr AS col` |
| `rename()` | `AS` |
| `left_join()` | `LEFT JOIN ... ON` |

\medskip

```{r}
#| eval: false
flights |>
  filter(dest %in% c("IAH", "HOU")) |>
  show_query()
# SELECT * FROM flights WHERE dest IN ('IAH', 'HOU')
```

## Exercises (Databases)

1. What SQL does \comm{distinct()} translate to? What about \comm{head(n = 10)}?
2. Use dbplyr to recreate: find flights where \comm{dep\_delay < arr\_delay}, and compute speed (\comm{distance / air\_time * 60}).

## Solutions (Databases)

1. \comm{distinct()} $\to$ \comm{SELECT DISTINCT}. \comm{head(10)} $\to$ \comm{LIMIT 10}.

2.

```{r}
#| eval: false
# Flights where dep_delay < arr_delay
flights_db |>
  filter(dep_delay < arr_delay) |> collect()

# Speed calculation
flights_db |>
  mutate(speed = distance / air_time * 60) |>
  select(tailnum, speed) |> collect()
```

# Arrow and Parquet

## Why Parquet?

| Feature | CSV | Parquet |
|---------|-----|---------|
| Size | Large (text) | 50--60% smaller |
| Types | Guessed on read | Embedded metadata |
| Reading | Must read entire file | Column-selective |
| Speed | Slow for big data | Very fast |

Parquet is a **columnar** format --- reads only the columns you need.

## Reading and Writing Parquet

```{r}
#| eval: false
library(arrow)

# Read large CSV lazily
data <- open_dataset("checkouts.csv",
  format = "csv",
  col_types = schema(ISBN = string()))

# Convert to partitioned parquet
data |>
  group_by(CheckoutYear) |>
  write_dataset("checkouts_pq",
    format = "parquet")

# Query parquet (lazy until collect)
open_dataset("checkouts_pq") |>
  filter(CheckoutYear == 2021) |>
  group_by(CheckoutMonth) |>
  summarize(total = sum(Checkouts)) |>
  collect()
```

Partitioning by year means filtering by year only reads the relevant files.

## Exercises (Arrow)

1. Find the most popular book each year.
2. How have checkouts of physical books vs. ebooks changed over the last 10 years?

## Solutions (Arrow)

1.

```{r}
#| eval: false
open_dataset("checkouts_pq") |>
  filter(MaterialType == "BOOK") |>
  group_by(CheckoutYear, Title) |>
  summarize(n = sum(Checkouts)) |>
  slice_max(n, n = 1) |> collect()
```

2. Filter \comm{MaterialType \%in\% c("BOOK", "EBOOK")}, group by year and type, sum checkouts, plot as line chart.

# Rectangling

## List-Columns and Nested Data

Tibbles can hold **list-columns** --- columns where each cell is a list:

```{r}
#| output.lines: 5
df <- tibble(
  x = 1:2,
  y = list(list(a = 11, b = 12),
           list(a = 21, b = 22)))
df
```

Two key operations to flatten:

- \comm{unnest\_wider()}: named list $\to$ new **columns**
- \comm{unnest\_longer()}: unnamed list $\to$ new **rows**

## `unnest_wider()` --- Named Lists to Columns

```{r}
#| output.lines: 6
df |> unnest_wider(y)
```

Each **name** in the list becomes a column. Use when elements have consistent names.

\medskip

For nested lists, chain multiple unnest calls:

```{r}
#| eval: false
data |>
  unnest_wider(json) |>    # level 1
  unnest_wider(owner, names_sep = "_")  # level 2
```

\comm{names\_sep} prevents column name conflicts.

## `unnest_longer()` --- Unnamed Lists to Rows

```{r}
#| output.lines: 8
df2 <- tibble(
  x = 1:3,
  y = list(c(11, 12, 13), 21, c(31, 32)))
df2 |> unnest_longer(y)
```

Each **element** becomes a row. The parent row is duplicated.

## Working with JSON

JSON maps to R as: objects $\to$ named lists, arrays $\to$ unnamed lists.

```{r}
#| output.lines: 6
json <- '[
  {"name": "John", "age": 34},
  {"name": "Susan", "age": 27}]'

tibble(json = jsonlite::parse_json(json)) |>
  unnest_wider(json)
```

**Strategy:** start with \comm{unnest\_wider()}, inspect, then \comm{unnest\_longer()} for arrays, repeat.

## Exercises (Rectangling)

1. What happens if you use \comm{unnest\_wider()} on an unnamed list-column?
2. What happens if you use \comm{unnest\_longer()} on a named list-column?
3. Explain the difference between \comm{hoist()} and repeated \comm{unnest\_wider()} calls.

## Solutions (Rectangling)

1. It errors --- names are needed for column names. Use \comm{names\_repair} or provide names manually.

2. It creates rows from the values and adds an extra column with the names (the index column).

3. \comm{hoist()} extracts deeply nested components in one step using path specifications (e.g., \comm{c("bounds", "northeast", "lat")}). Repeated \comm{unnest\_wider()} flattens one level at a time. \comm{hoist()} is more concise for deep extraction.

# Web Scraping

## HTML Basics

```{r}
#| eval: false
library(rvest)
```

HTML is a tree of **elements**: `<tag attribute="value">content</tag>`

| Tag | Purpose |
|-----|---------|
| `<h1>` -- `<h6>` | Headings |
| `<p>` | Paragraphs |
| `<a href="...">` | Links |
| `<table>` | Tables |
| `<ul>`, `<li>` | Lists |
| `<div>`, `<span>` | Containers |

Attributes: \comm{id} (unique), \comm{class} (group), \comm{href} (link target).

## Extracting Data with rvest

```{r}
#| eval: false
html <- read_html("https://example.com")

# Find elements with CSS selectors
html |> html_elements("h2")          # by tag
html |> html_elements(".classname")  # by class
html |> html_elements("#myid")       # by id

# Extract content
html |> html_elements("h2") |> html_text2()
html |> html_elements("a") |>
  html_attr("href")

# Tables → tibbles automatically
html |> html_element("table") |> html_table()
```

## Nesting Pattern for Complex Pages

**Critical:** use \comm{html\_element()} (singular) inside a pipeline to maintain row alignment:

```{r}
#| eval: false
# Each <section> = one film
sections <- html |> html_elements("section")

tibble(
  title = sections |>
    html_element("h2") |> html_text2(),
  director = sections |>
    html_element(".director") |> html_text2()
)
```

\comm{html\_element()} returns \comm{NA} for missing matches (keeps alignment). \comm{html\_elements()} drops non-matches (breaks alignment).

## Ethics of Web Scraping

- **Public + non-personal + factual** = generally okay
- Always check **Terms of Service**
- Be careful with **personally identifiable information**
- Facts are not copyrighted; original content is
- Use the \comm{polite} package to rate-limit requests
- \comm{rvest} does **not** execute JavaScript --- use \comm{chromote} for dynamic sites

<!--
# Summary

## Key Takeaways

- **Excel:** \comm{read\_excel()} with \comm{col\_names}, \comm{skip}, \comm{na}, \comm{col\_types}; \comm{write\_xlsx()}
- **Multiple sheets:** \comm{excel\_sheets()} + \comm{bind\_rows()}
- **Databases:** \comm{DBI::dbConnect()} + \comm{tbl()} for lazy queries; \comm{collect()} to execute
- **dbplyr:** translates dplyr $\to$ SQL; \comm{show\_query()} reveals the SQL
- **Parquet:** columnar, typed, fast; \comm{open\_dataset()} + \comm{write\_dataset()}
- **Partitioning:** speeds queries by reading only relevant files
- **Rectangling:** \comm{unnest\_wider()} (named $\to$ cols), \comm{unnest\_longer()} (unnamed $\to$ rows)
- **JSON:** objects $\to$ named lists, arrays $\to$ unnamed lists; iterate unnest
- **Web scraping:** \comm{read\_html()} + \comm{html\_elements()} + \comm{html\_text2()}
- **Alignment:** use \comm{html\_element()} (singular) inside pipelines
- **Ethics:** respect ToS, avoid personal data, use \comm{polite} package
-->
