---
title: "R4DS 15 -- Quarto"
---

```{r setup, include=FALSE}
source("_common.R")
library(tidyverse)
```

# Quarto Basics

## What is Quarto?

**Quarto** is a unified authoring framework for data science:

- Combine **code**, **results**, and **prose** in one document
- Supports R, Python, Julia, and Observable JS
- A command-line tool (not an R package)

\medskip

Three uses:

1. **Communicate** to decision-makers (hide code, show results)
2. **Collaborate** with other data scientists (show code + reasoning)
3. **Environment** for doing data science (notebook for yourself)

\medskip

Pipeline: `.qmd` $\to$ **knitr** (executes R) $\to$ `.md` $\to$ **Pandoc** (converts) $\to$ output

## Document Structure

A `.qmd` file has three types of content:

**1. YAML header** (between `---`):

```yaml
title: "Diamond Sizes"
author: "Jane Doe"
date: today
format: html
```

**2. Code chunks** (between `` ``` ``):

```{r}
#| eval: false
smaller <- diamonds |> filter(carat <= 2.5)
```

**3. Text** with Markdown formatting: `**bold**`, `*italic*`, `` `code` ``, `# Heading`, `- bullet`, `[link](url)`.

## Code Chunks and Labels

Insert chunks with:

- **Keyboard:** Ctrl+Alt+I (Mac: Cmd+Option+I)
- **Button:** Insert $\to$ Code Chunk
- **Manual:** type `` ```{r} `` and `` ``` ``

\medskip

**Labels** identify chunks for navigation and cross-references:

```{r}
#| eval: false
#| label: simple-addition
1 + 1
```

Label rules: one word or words connected by **dashes**. Avoid `.`, `_`, and spaces. Figures must start with `fig-`, tables with `tbl-`.

## Chunk Options

Key options (set with `#|` inside chunks):

| Option | Effect |
|--------|--------|
| `eval: false` | Show code, don't run it |
| `include: false` | Run code, hide everything |
| `echo: false` | Run code, hide the code |
| `results: hide` | Run code, hide text output |
| `fig-show: hide` | Run code, hide plots |
| `message: false` | Suppress messages |
| `warning: false` | Suppress warnings |

## Global Options and Inline Code

**Global options** apply to all chunks via YAML:

```yaml
title: "My Report"
execute:
  echo: false
  warning: false
```

Override per-chunk with `#| echo: true`.

\medskip

**Inline code** embeds R results in text:

```markdown
We have data about `r nrow(diamonds)` diamonds.
```

Useful helper for readable numbers:

```{r}
#| output.lines: 3
comma <- function(x) format(x, digits = 2,
                            big.mark = ",")
comma(3452345)
```

## Exercises (Basics)

1. Create a new `.qmd` file. Render to HTML, PDF, and Word. How do the outputs differ?
2. Create a brief CV in Quarto: your name as title, headings for education and employment, bulleted lists, bold years.
3. Set `echo: false` globally. Write a report on diamond sizes for a non-R audience.
4. Use `scales::label_comma()` for readable inline numbers. Compute the percentage of diamonds $> 2.5$ carats.

## Solutions (Basics)

1. HTML is interactive (scrollable). PDF is fixed-layout (good for print). Word is editable. The `.qmd` input is identical --- only `format:` changes.

2. Use `**2020**` for bold years in a bulleted list under `## Education` and `## Employment`.

3. Set in YAML: `execute: echo: false`. Use `ggplot()` for visualizations, `knitr::kable()` for tables.

4.

```{r}
#| eval: false
# Inline: `r scales::label_comma()(nrow(diamonds))`
# Percentage:
pct <- round(sum(diamonds$carat > 2.5) /
  nrow(diamonds) * 100, 1)
# Inline: `r pct`% are larger than 2.5 carats
```

# Figures, Tables, and Caching

## Figure Sizing

Control with chunk options:

| Option | Effect |
|--------|--------|
| `fig-width` | Width in inches (default 7) |
| `fig-height` | Height in inches |
| `fig-asp` | Aspect ratio (height/width) |
| `out-width` | Scaling in output (e.g., `"70%"`) |
| `layout-ncol` | Multi-figure layout |

\medskip

**Recommendations:**

- Set `fig-width` between 4--8 inches
- Use `fig-asp: 0.618` (golden ratio) for aesthetics
- `out-width: "70%"` for a good default size in documents
- Cross-reference figures with `fig-` label prefix + `fig-cap`

## Tables with `knitr::kable()`

Default prints are basic. Use \comm{knitr::kable()} for polished tables:

```{r}
knitr::kable(mtcars[1:4, 1:4])
```

Cross-reference with `tbl-` label prefix and `tbl-cap`:

```{r}
#| eval: false
#| label: tbl-example
#| tbl-cap: "First 5 cars"
knitr::kable(mtcars[1:5, ])
```

Other packages: \comm{gt}, \comm{kableExtra}, \comm{huxtable}, \comm{flextable}.

## Caching

Expensive computations can be cached:

```{r}
#| eval: false
#| label: raw-data
#| cache: true
rawdata <- read_csv("a_very_large_file.csv")
```

Track dependencies with \comm{dependson}:

```{r}
#| eval: false
#| label: processed-data
#| cache: true
#| dependson: "raw-data"
processed <- rawdata |>
  filter(!is.na(important_var))
```

Invalidate when source file changes:

```yaml
cache.extra: !expr file.mtime("large_file.csv")
```

Or enable document-wide: `execute: cache: true` in YAML.

## Exercises (Figures/Tables/Caching)

1. Try each figure option one at a time: `fig-width: 10`, `fig-height: 3`, `out-width: "100%"`, `out-width: "20%"`. How does each change the output?

2. Display \comm{diamonds[1:5, ]} with \comm{knitr::kable()}. Add a `tbl-` label and `tbl-cap`. Cross-reference it in text.

3. Set up cached chunks: `d` depends on `c` and `b`, both depend on `a`. Print \comm{lubridate::now()} in each. Verify caching behavior.

## Solutions (Figures/Tables/Caching)

1. `fig-width: 10`: wider figure, elements appear smaller. `fig-height: 3`: short, wide figure. `out-width: "100%"`: fills page width. `out-width: "20%"`: tiny figure.

2.

```{r}
#| eval: false
#| label: tbl-diamonds
#| tbl-cap: "First 5 diamonds"
knitr::kable(diamonds[1:5, ])
```

Reference in text: `@tbl-diamonds`.

3. Use `dependson: "a"` for b and c, `dependson: c("b", "c")` for d. Only `a` re-runs when its code changes; downstream chunks cascade.

# Configuration

## Self-Contained Documents and Parameters

**Self-contained HTML** embeds all images, CSS, JS:

```yaml
format:
  html:
    embed-resources: true
```

\medskip

**Parameters** make documents reusable:

```yaml
params:
  start: !expr lubridate::ymd("2015-01-01")
  snapshot: !expr lubridate::ymd_hms("2015-01-01 12:30:00")
```

Access with `params$start` in R code. Render with different values programmatically.

## Citations and Bibliography

Add a `.bib` file and CSL style:

```yaml
bibliography: references.bib
csl: apa.csl
```

\medskip

Citation syntax in text:

| Syntax | Output |
|--------|--------|
| `[@smith04]` | (Smith 2004) |
| `@smith04` | Smith (2004) |
| `[@smith04; @doe99]` | (Smith 2004; Doe 1999) |
| `[-@smith04]` | (2004) --- suppress author |

Bibliography is auto-inserted at document end.

## Troubleshooting

Common issues and fixes:

1. **Duplicated chunk labels** --- every label must be unique
2. **Working directory** --- Quarto renders from the `.qmd` file's directory
3. **Object not found** --- code runs top-to-bottom during render; test with "Restart R and Run All Chunks"
4. **Hard-to-debug errors** --- add `error: true` to show the error in output
5. **Package conflicts** --- use `pkg::fun()` calls or \comm{conflicted} package

# Output Formats

## Document Formats

Set output in YAML:

| Format | YAML | Output |
|--------|------|--------|
| HTML | `html` | Web page |
| PDF | `pdf` | Fixed-layout (needs LaTeX) |
| Word | `docx` | Editable document |
| Markdown | `gfm` | GitHub-flavored markdown |

\medskip

Multiple formats from one file:

```yaml
format:
  html:
    toc: true
  pdf: default
  docx: default
```

Render all: `quarto render file.qmd --to all`

## Presentations

Quarto creates slides with header-based divisions:

| Format | YAML | Technology |
|--------|------|-----------|
| RevealJS | `revealjs` | HTML/JS |
| PowerPoint | `pptx` | MS Office |
| Beamer | `beamer` | LaTeX/PDF |

\medskip

- `#` creates section title slides
- `##` creates content slides
- `---` creates untitled slides
- Speaker notes go in `:::{.notes}` divs

These slides are built with Quarto + Beamer!

## Interactivity

**htmlwidgets** --- client-side (no R server needed):

```{r}
#| eval: false
library(leaflet)
leaflet() |>
  setView(174.764, -36.877, zoom = 16) |>
  addTiles() |>
  addMarkers(174.764, -36.877,
    popup = "Maungawhau")
```

Packages: \comm{leaflet} (maps), \comm{DT} (tables), \comm{plotly} (interactive plots).

\medskip

**Shiny** --- server-side (needs R running): add `server: shiny` to YAML.

## Websites and Books

Multi-document projects use \comm{\_quarto.yml}:

```yaml
project:
  type: website

website:
  title: "My Site"
  navbar:
    left:
      - href: index.qmd
        text: Home
      - about.qmd
```

- **Websites:** linked pages with navigation bars
- **Books:** ordered chapters; supports HTML + PDF + EPUB
- Render all with `quarto render` from project root

<!--
# Summary

## Key Takeaways

- **Pipeline:** `.qmd` $\to$ knitr $\to$ `.md` $\to$ Pandoc $\to$ output
- **Three parts:** YAML header, code chunks, Markdown text
- **Chunk options:** `eval`, `echo`, `include`, `message`, `warning`
- **Global options:** `execute:` in YAML; override per-chunk
- **Inline code:** embed R results directly in text
- **Figure sizing:** `fig-width`, `fig-height`, `fig-asp`, `out-width`
- **Tables:** \comm{knitr::kable()} or \comm{gt::gt()} for polished output
- **Caching:** `cache: true` + `dependson` for expensive computations
- **Self-contained:** `embed-resources: true` for portable HTML
- **Parameters:** `params:` in YAML for reusable documents
- **Citations:** `.bib` file + `[@key]` syntax
- **Formats:** HTML, PDF, Word, RevealJS, Beamer, PowerPoint
- **Projects:** \comm{\_quarto.yml} for websites and books
-->
