---
title: "R4DS 11 -- Joins"
---

```{r setup, include=FALSE}
source("_common.R")
library(tidyverse)
library(nycflights13)
```

# Keys

## Primary and Foreign Keys

A **primary key** uniquely identifies each row. A **foreign key** links to another table's primary key.

\medskip

| Table | Primary key | Identifies |
|-------|-------------|------------|
| `airlines` | `carrier` | Airline |
| `airports` | `faa` | Airport |
| `planes` | `tailnum` | Plane |
| `weather` | `origin`, `time_hour` | Weather obs. |

\medskip

Foreign keys in \comm{flights}: \comm{carrier} $\to$ \comm{airlines}, \comm{tailnum} $\to$ \comm{planes}, \comm{origin}/\comm{dest} $\to$ \comm{airports\$faa}, (\comm{origin}, \comm{time\_hour}) $\to$ \comm{weather}.

## Checking Primary Keys

Verify uniqueness with \comm{count()} + \comm{filter()}:

```{r}
#| output.lines: 5
planes |>
  count(tailnum) |> filter(n > 1)
```

Also check for missing values:

```{r}
#| output.lines: 5
planes |> filter(is.na(tailnum))
```

**Surrogate keys:** when no natural key exists, add one with \comm{row\_number()}:

```{r}
#| eval: false
flights |> mutate(id = row_number(), .before = 1)
```

## Exercises (Keys)

1. What is the relationship between \comm{weather} and \comm{airports}?
2. If \comm{weather} contained records for all US airports, what additional connection would it make to \comm{flights}?
3. The \comm{year}, \comm{month}, \comm{day}, \comm{hour}, and \comm{origin} variables almost form a compound key for \comm{weather}, but one hour has duplicates. What's special about it?
4. How would you represent "special days" (e.g., Christmas) as a data frame? What would be its primary key?

## Solutions (Keys)

1. \comm{weather\$origin} is a foreign key to \comm{airports\$faa}. Only the three NYC airports appear.

2. \comm{weather} could also connect to \comm{flights\$dest}, providing weather at the destination.

3. The duplicate hour is likely the fall-back DST hour (November, when clocks go back --- the same hour occurs twice).

4. A tibble with \comm{month} and \comm{day} as compound primary key. Connects to \comm{flights} via those columns.

# Mutating Joins

## `left_join()` --- The Most Common Join

A **mutating join** adds variables from one table to another by matching keys.

```{r}
#| output.lines: 8
flights2 <- flights |>
  select(year, time_hour, origin, dest,
         tailnum, carrier)

flights2 |> left_join(airlines)
```

\comm{left\_join()} keeps **all rows** in \comm{x}, fills \comm{NA} for non-matches.

## `left_join()` --- Adding Weather

```{r}
#| output.lines: 8
flights2 |>
  left_join(weather |>
    select(origin, time_hour, temp, wind_speed))
```

Non-matches produce \comm{NA} for the new columns.

## `left_join()` --- Adding Plane Info

```{r}
#| output.lines: 8
flights2 |>
  left_join(planes |>
    select(tailnum, type, engines, seats))
```

Many planes have \comm{NA} for \comm{type}/\comm{seats} --- those tail numbers are missing from the \comm{planes} table.

## Specifying Join Keys with `join_by()`

Default: join on **all shared column names** (natural join). This can go wrong:

```{r}
#| output.lines: 4
#| warning: false
# year means different things in flights vs planes!
flights2 |> left_join(planes) |>
  select(tailnum, type) |> head(3)
```

Fix: specify the key explicitly:

```{r}
#| eval: false
flights2 |> left_join(planes, join_by(tailnum))
```

For **different column names**, use \comm{==}:

```{r}
#| eval: false
flights2 |>
  left_join(airports, join_by(dest == faa))
```

## Four Types of Mutating Joins

Simple example:

```{r}
x <- tribble(~key, ~val_x, 1, "x1",
             2, "x2", 3, "x3")
y <- tribble(~key, ~val_y, 1, "y1",
             2, "y2", 4, "y3")
```

| Join | Keeps rows from | Unmatched rows |
|------|----------------|----------------|
| `left_join(x, y)` | All of `x` | `y` gets `NA` |
| `right_join(x, y)` | All of `y` | `x` gets `NA` |
| `full_join(x, y)` | All of `x` and `y` | Both get `NA` |
| `inner_join(x, y)` | Only matches | Dropped |

## `inner_join()` and `left_join()`

```{r}
#| output.lines: 6
x |> inner_join(y, join_by(key))
```

```{r}
#| output.lines: 7
x |> left_join(y, join_by(key))
```

## `full_join()`

```{r}
#| output.lines: 8
x |> full_join(y, join_by(key))
```

Key 3 exists only in \comm{x} $\to$ \comm{val\_y} is \comm{NA}. Key 4 exists only in \comm{y} $\to$ \comm{val\_x} is \comm{NA}.

## Filtering Joins: `semi_join()` and `anti_join()`

**Filtering joins** filter rows --- they never add columns.

\comm{semi\_join(x, y)}: keep rows in \comm{x} that **have** a match in \comm{y}:

```{r}
#| output.lines: 6
airports |>
  semi_join(flights2, join_by(faa == origin))
```

\comm{anti\_join(x, y)}: keep rows in \comm{x} that have **no** match in \comm{y}:

```{r}
#| output.lines: 6
flights2 |>
  anti_join(airports, join_by(dest == faa)) |>
  distinct(dest)
```

## Row Matching

Three outcomes for a row in \comm{x}:

- **0 matches** --- dropped (inner) or filled with \comm{NA} (left/right/full)
- **1 match** --- preserved as-is
- **>1 matches** --- **duplicated** once per match

\medskip

**Many-to-many** joins cause row explosion:

```{r}
#| output.lines: 7
#| warning: false
df1 <- tibble(key = c(1, 2, 2),
              val_x = c("x1", "x2", "x3"))
df2 <- tibble(key = c(1, 2, 2),
              val_y = c("y1", "y2", "y3"))
df1 |> inner_join(df2, join_by(key))
```

## Exercises (Basic Joins)

1. Find the 48 hours with the worst delays over the year. Cross-reference with \comm{weather}.
2. Find all flights to the top 10 most popular destinations.
3. Does every departing flight have corresponding weather data?
4. What do tail numbers missing from \comm{planes} have in common? (Hint: one variable explains ~90\%.)
5. Add origin **and** destination airport latitude/longitude to \comm{flights}.
6. Compute average delay by destination, then plot on a US map.

## Solutions (Basic Joins) --- 1--3

1.

```{r}
#| eval: false
flights |>
  mutate(hour = dep_time %/% 100) |>
  group_by(year, month, day, hour) |>
  summarize(avg_delay = mean(dep_delay,
    na.rm = TRUE)) |>
  ungroup() |> slice_max(avg_delay, n = 48) |>
  left_join(weather)
```

2.

```{r}
#| eval: false
top_dest <- flights |>
  count(dest, sort = TRUE) |> head(10)
flights |> semi_join(top_dest, join_by(dest))
```

3. No --- \comm{flights |> anti\_join(weather)} reveals some flights without matching weather.

## Solutions (Basic Joins) --- 4--6

4. Mostly \comm{carrier == "MQ"} or \comm{"AA"} --- they report tail numbers differently.

5.

```{r}
#| eval: false
flights |>
  left_join(airports |> select(faa, lat, lon),
    join_by(origin == faa)) |>
  rename(origin_lat = lat, origin_lon = lon) |>
  left_join(airports |> select(faa, lat, lon),
    join_by(dest == faa)) |>
  rename(dest_lat = lat, dest_lon = lon)
```

Rename **after** the join to avoid ambiguity.

6.

```{r}
#| eval: false
flights |>
  group_by(dest) |>
  summarize(delay = mean(arr_delay, na.rm = TRUE)) |>
  inner_join(airports, join_by(dest == faa)) |>
  ggplot(aes(lon, lat, color = delay)) +
  borders("state") + geom_point() +
  coord_quickmap()
```

# Non-Equi Joins

## Cross Joins

\comm{cross\_join()} matches every row in \comm{x} with every row in \comm{y} (Cartesian product):

```{r}
#| output.lines: 8
df <- tibble(name = c("John", "Simon",
                      "Tracy", "Max"))
df |> cross_join(df)
```

Useful for generating all permutations. Output has $n_x \times n_y$ rows.

## Inequality Joins

Use `<`, `<=`, `>`, `>=` in \comm{join\_by()} instead of `==`:

```{r}
#| output.lines: 8
df <- tibble(id = 1:4,
  name = c("John", "Simon", "Tracy", "Max"))
df |> inner_join(df, join_by(id < id))
```

Generates all **combinations** (not permutations) --- avoids duplicates.

## Rolling Joins

Match the **closest** row satisfying an inequality with \comm{closest()}:

```{r}
#| output.lines: 8
parties <- tibble(
  q = 1:4,
  party = ymd(c("2022-01-10", "2022-04-04",
                "2022-07-11", "2022-10-03")))

set.seed(123)
employees <- tibble(
  name = c("Alice", "Bob", "Carol", "Dan"),
  birthday = ymd(c("2022-03-15", "2022-06-20",
                    "2022-01-05", "2022-11-01")))

employees |>
  left_join(parties,
    join_by(closest(birthday >= party)))
```

## Overlap Joins

Three helpers for working with date ranges:

| Helper | Condition |
|--------|-----------|
| `between(x, y_lower, y_upper)` | `x >= y_lower & x <= y_upper` |
| `within(xl, xu, yl, yu)` | `xl >= yl & xu <= yu` |
| `overlaps(xl, xu, yl, yu)` | `xl <= yu & xu >= yl` |

```{r}
#| eval: false
parties <- parties |>
  mutate(start = ymd(c("2022-01-01",
    "2022-04-04", "2022-07-11", "2022-10-03")),
    end = ymd(c("2022-04-03", "2022-07-10",
      "2022-10-02", "2022-12-31")))

employees |>
  inner_join(parties,
    join_by(between(birthday, start, end)))
```

## Exercises (Non-Equi Joins)

1. Explain the difference in output between these two:

```{r}
#| eval: false
x |> full_join(y, join_by(key == key))
x |> full_join(y, join_by(key == key),
               keep = TRUE)
```

2. When checking party period overlaps, why is \comm{q < q} needed in the \comm{join\_by()}? What happens without it?

## Solutions (Non-Equi Joins)

1. Without \comm{keep = TRUE}, the key columns are coalesced into one (showing the non-\comm{NA} value). With \comm{keep = TRUE}, both \comm{key.x} and \comm{key.y} are shown separately.

2. Without \comm{q < q}, every interval overlaps with itself, giving spurious matches. The inequality ensures each pair is compared only once and excludes self-matches.

<!--
# Summary

## Key Takeaways

- **Primary key:** uniquely identifies rows; **foreign key:** links tables
- **Surrogate key:** \comm{row\_number()} when no natural key exists
- **Mutating joins** add columns:
  - \comm{left\_join()}: keeps all rows in \comm{x} (most common)
  - \comm{inner\_join()}: keeps only matches
  - \comm{right\_join()}: keeps all rows in \comm{y}
  - \comm{full\_join()}: keeps all rows in both
- **Filtering joins** filter rows:
  - \comm{semi\_join()}: keeps \comm{x} rows with a match in \comm{y}
  - \comm{anti\_join()}: keeps \comm{x} rows without a match in \comm{y}
- **Join keys:** \comm{join\_by(x == y)} for different names; natural join by default
- **Row matching:** 0 matches $\to$ \comm{NA}; 1 match $\to$ kept; >1 match $\to$ duplicated
- **Non-equi joins:** \comm{cross\_join()}, inequality (\comm{<}, \comm{>=}), \comm{closest()}, \comm{between()}/\comm{overlaps()}
-->
