---
title: "R Fundamentals"
author: ""
date: ""
format:
  beamer:
    theme: Madrid
    colortheme: beaver
    fontsize: 9pt
    aspectratio: 169
    section-titles: true
    toc: true
    toc-depth: 2
    navigation: empty
    pdf-engine: xelatex
    header-includes: |
      \usepackage{booktabs}
      \usepackage{dsfont}
      \setbeamertemplate{caption}[numbered]{}
      \setbeamertemplate{footline}[text line]{%
        \parbox{\linewidth}{\vspace*{-8pt}\hfill\insertframenumber\,/\,\inserttotalframenumber~~~~~~~~~}}
      \setbeamertemplate{navigation symbols}{}
      \setbeamercolor{item projected}{fg=white}
      \setbeamercolor{frametitle}{fg=black}
      \setbeamercolor{block title}{fg=white}
      \definecolor{foo}{rgb}{.2,.2,.7}
      \AtBeginSection[]{
        \begin{frame}
        \vfill
        \centering
        \begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{section page}
          \usebeamerfont{title}{\color{foo}\insertsectionhead}\par
        \end{beamercolorbox}
        \vfill
        \end{frame}
      }
      \AtBeginSubsection{}
      \setbeamertemplate{headline}{}
      \newcommand{\comm}[1]{\textcolor[rgb]{0.737,0.353,0.396}{{\tt #1}}}
      \usepackage{fancyvrb}
      \definecolor{errorred}{rgb}{0.80,0.10,0.10}
      \definecolor{warnbrown}{rgb}{0.70,0.40,0.05}
      \DefineVerbatimEnvironment{ErrorVerbatim}{Verbatim}{formatcom=\color{errorred},fontsize=\small}
      \DefineVerbatimEnvironment{WarnVerbatim}{Verbatim}{formatcom=\color{warnbrown},fontsize=\small}
      %% ── tighten spacing around code blocks and output ──
      \setlength{\parskip}{0pt}
      \setlength{\OuterFrameSep}{0pt}
      \usepackage{etoolbox}
      \makeatletter
      \preto{\@verbatim}{\topsep=2pt \partopsep=0pt}
      \makeatother
      \AtBeginEnvironment{Shaded}{\vspace{0pt}}
      \AfterEndEnvironment{Shaded}{\vspace{0pt}}
      \AtBeginEnvironment{verbatim}{\vspace{0pt}}
      \AfterEndEnvironment{verbatim}{\vspace{0pt}}
execute:
  echo: true
  warning: false
knitr:
  opts_chunk:
    comment: '#'
    fig.align: 'center'
---

```{r}
#| include: false
knitr::knit_hooks$set(error = function(x, options) {
  x <- gsub('\n$', '', x)
  paste0('\n\n\\begin{ErrorVerbatim}\n', x, '\n\\end{ErrorVerbatim}\n\n')
})
knitr::knit_hooks$set(warning = function(x, options) {
  x <- gsub('\n$', '', x)
  paste0('\n\n\\begin{WarnVerbatim}\n', x, '\n\\end{WarnVerbatim}\n\n')
})
```

# First Encounter

### R as a Calculator

- R as a calculator --- show the numerical result of $1 + 2\times\frac{3}{4} - 5 + 6 \sin\frac{\pi}{2} - \sqrt[8]{7}$

```{r}
1 + 2 * (3 / 4) - 5 + 6 * sin(pi / 2) - 7^(1/8)
```

\vfill

- Print the content of the built-in dataset `Nile`

```{r}
Nile
```

---

### A First Glance at Plots

:::::: columns
::: column
- Histogram of `Nile`

```{r, out.width="90%"}
hist(Nile)
```
:::
::: column
- Time series plot of `Nile`

```{r, out.width="90%"}
plot(Nile)
```
:::
::::::

---

### A First Glance at Vectors

```{r}
Nile[2]          # the 2nd item
Nile[2:5]        # subsequence from 2nd to 5th
x <- c(10, 20, -1, 3)   # create a vector with c()
length(x); mean(x)
```

```{r}
x > 5                # logical comparison
sum(x > 5)           # count TRUEs
```

# Vectors

## Atomic Vectors

### Vector Types and Scalars

- Flavors: **atomic** (all elements same type), **generic** (lists, elements can differ)
- Primary atomic types: `logical`, `integer`, `double`, `character`; integer + double = numeric
- Scalars are simply vectors of length 1

```{r}
v_logical   <- c(TRUE, FALSE)
v_integer   <- c(1L, 6L, 10L)
v_double    <- c(1, 2.5, 4.5)
v_character <- c('these', 'are', 'characters')
typeof(v_logical); typeof(v_integer); typeof(v_double); typeof(v_character)
```

```{r}
# concatenation of atomic vectors yields flat atomic vectors
(v <- c(c(1, 2), c(3, 4)))
```

---

### `NA` and `NULL`

- `NA` (Not Applicable): value **exists** but is **unknown**
- `NULL`: value **does not exist**

```{r}
NA > 5; 20 * NA; !NA             # NA is infectious!
NA ^ 0; NA | TRUE; NA & FALSE    # some exceptions
```

```{r}
x <- c(NA, 5, NA, 10)
x == NA          # WRONG way to test (always NA)
is.na(x)         # CORRECT way to test
```

```{r}
u <- NULL; length(u)   # 0
v <- NA;   length(v)   # 1
```

---

### Coercion Hierarchy

- Automatic coercion: **character > double > integer > logical**

```{r}
typeof(c('a', 1))      # character wins
c(1, FALSE)             # logical -> double
c(TRUE, 1L)             # logical -> integer
```

```{r}
# explicit coercion with as.*()
x <- c(TRUE, FALSE, TRUE, TRUE); as.numeric(x)
as.integer(c('1', '1.5', 'a'))   # 'a' -> NA with warning
```

---

### Type Testing and Coercion Functions

\begin{table}[!htbp]
\centering\small
\begin{tabular}{ll}
\toprule
Testing & Coercion \\
\midrule
\comm{is.logical(x)} & \comm{as.logical(x)} \\
\comm{is.integer(x)} & \comm{as.integer(x)} \\
\comm{is.double(x)} & \comm{as.double(x)} \\
\comm{is.numeric(x)} & \comm{as.numeric(x)} \\
\comm{is.character(x)} & \comm{as.character(x)} \\
\comm{is.na(x)}, \comm{is.null(x)} & --- \\
\comm{is.vector(x)} & \comm{as.vector(x)} \\
\comm{is.list(x)} & \comm{as.list(x)} \\
\comm{is.data.frame(x)} & \comm{as.data.frame(x)} \\
\comm{is.matrix(x)} & \comm{as.matrix(x)} \\
\comm{is.factor(x)} & \comm{as.factor(x)} \\
\bottomrule
\end{tabular}
\end{table}

```{r}
x <- c(3, 5, 7, 8)
is.integer(x); is.numeric(x)
as.integer(5.7); as.integer(-5.7)
```

## Creating Vectors

### Sequences --- `:` and `seq()`

```{r}
5:10; 5:1
i <- 5; 1:i-1; 1:(i-1)   # note operator precedence!
```

```{r}
seq(from = 11, to = 30, by = 3)
seq(from = 1.1, to = 2, length = 10)
```

```{r}
# seq_along and seq_len: safer alternatives for loops
x <- c(9, 5, 17); seq_along(x)
x <- NULL; seq_along(x)   # integer(0), not 1:0 !
seq_len(5)
```

---

### Repetition: `rep()`

```{r}
rep(3, 5)
rep(c(10, 15, 3), 4)
rep(c(10, 15, 3), each = 4)   # interleave with 'each'
```

```{r}
rep(1:4, 1:4)                  # variable repetition counts
rep(c('C', 'G', 'H'), c(4, 5, 2))
```

---

### Other Generators: `sample()`, `sequence()`

```{r}
set.seed(42)
sample(1:6, 10, replace = TRUE)     # dice rolls
sample(letters[1:20], 5)            # no replacement
```

```{r}
sequence(c(6, 7, 4), from = c(8, 1, 9), by = c(-1, 1, -2))
```

## Indexing and Selection

### Numeric Indexing

```{r}
fib <- c(0, 1, 1, 2, 3, 5, 8, 13, 21, 34, 55, 89, 144)
fib[1]; fib[5]; fib[2:5]; fib[c(2, 4, 8, 1)]
```

```{r}
# 'minus' means 'exclude'; no mixing of positive/negative!
fib[-3]; fib[-(5:9)]
```

```{r}
# negative indexing to build new vectors
x <- c(10, 20, -1, 3, 5, 6, 30, 9)
(y <- x[-2])                  # remove 2nd element
(z <- x[c(-2, -3, -6)])       # remove 2nd, 3rd, 6th
(w <- c(x, y, z))             # combine vectors
```

---

### Logical Indexing

```{r}
fib[fib < 10]                           # elements < 10
fib[fib %% 2 == 0 | fib %% 3 == 0]     # multiples of 2 or 3
fib[c(TRUE, FALSE)]                      # odd-indexed (recycling!)
```

```{r}
v <- c(2, 8, 1, 7, -11, 6, 20, 3, 12, 35, 21, 3, 5, 7, -2)
v[v > median(v)]                             # above median
v[v < quantile(v, 0.05) | v > quantile(v, 0.95)]  # extreme 5%
v[abs(v - mean(v)) > sd(v)]                  # beyond 1 SD
```

---

### `which()`, `which.min()`, `which.max()`

```{r}
which(Nile > 1200)            # positions where TRUE
Nile[Nile > 1200]             # the values themselves
```

```{r}
x <- c(5, 3, 8, 1, 9, 2)
which.min(x); which.max(x)   # index of first min/max
```

---

### Removing `NA` from Vectors

```{r}
v <- c(21, 2, 3, NA, 5, NULL, NA, 32, 55)
v[!is.na(v)]                  # NULL is already dropped by c()
```

```{r}
na.omit(v)                    # alternative; returns with attr
```

## Vector Properties

### Recycling Rule

- In operations involving two vectors of different lengths, R **repeats** the shorter one:

```{r}
c(2, 4, 3) + c(6, 10, -1, 3, 5)
```

```{r}
# useful applications
x <- 1:5
x + 10           # add a constant (scalar = length-1 vector)
x * c(1, -1)     # alternate signs
```

---

### `all()`, `any()`, and Vector Equality

```{r}
x <- 1:100; any(x > 200); all(x > 0)
```

```{r}
# == is element-wise; don't use for whole-vector equality!
x <- 1:3; y0 <- c(1, 2, 3); y1 <- c(1, 3, 4)
x == y0; x == y1
```

---

### `identical()` and `all.equal()`

```{r}
x <- 1:3; y0 <- c(1, 2, 3)
identical(x, y0)          # FALSE: integer vs double!
typeof(x); typeof(y0)
```

```{r}
all.equal(x, y0)          # TRUE (with tolerance)
all.equal(x, c(1, 3, 4)) # shows difference
```

---

### Named Vectors

```{r}
# 3 ways to name a vector
x <- c(a = 1, b = 2, c = 3)
x <- 1:3; names(x) <- c('a', 'b', 'c')
x <- setNames(1:3, c('a', 'b', 'c'))
```

```{r}
# access by name
x['b']; x[c('a', 'c')]
```

---

### Attributes and `structure()`

```{r}
a <- 1:3; attr(a, 'x') <- 'ab'; attr(a, 'y') <- 4:6
str(attributes(a))
a <- structure(1:3, x = 'ab', y = 4:6)
str(attributes(a))
```

---

### Factors

- Factors represent **categorical variables** with a fixed set of levels.

:::::: columns
::: column

```{r}
(x <- factor(c('a', 'b', 'b', 'a')))
typeof(x); levels(x)
```

:::
::: column

```{r}
# ordered factors
(size <- factor(c('S','M','L','M','S'),
    levels=c('S','M','L'), ordered=TRUE))
size > 'S'
```

:::
::::::

---

### Selecting Even/Odd Indexed Elements

```{r}
x <- 1:10
x[x %% 2 == 0]                   # even values
x[seq_along(x) %% 2 == 0]        # even-indexed elements
x[c(FALSE, TRUE)]                 # same, using recycling
```

---

### Appending and Inserting into Vectors

```{r, eval=FALSE}
v <- c(1, 2, 3); it <- c(6, 7, 8)
c(v, it)                       # concatenate
v[length(v) + 1] <- 42; v     # auto-extend (NAs fill gaps)
```

```{r}
append(1:10, 99, after = 5)   # insert 99 after position 5
append(1:10, 99, after = 0)   # insert at the beginning
```

# Operators

### Operator Table

\begin{table}[!htbp]
\centering\small
\begin{tabular}{ll}
\toprule
Operators & Definition \\
\midrule
\comm{+ - * /} & plus, minus, times, divide\\
\comm{\%/\% \%\% \textasciicircum} & integer quotient, modulo, power \\
\comm{> >= < <=} & greater, greater-or-equal, less, less-or-equal\\
\comm{== !=} & equals, not equals\\
\comm{! \& |} & not, and (elementwise), or (elementwise)\\
\comm{<- ->} & assignment\\
\comm{\$} & list/data.frame element access\\
\comm{:} & sequence creation\\
\comm{\textasciitilde} & model formulae\\
\midrule
\comm{\%*\%} & matrix product \\
\comm{\%in\%} & test matching \\
\comm{\&\& ||} & short-circuit and, or (scalar only)\\
\comm{::} & namespace access \\
\comm{xor(a, b)} & elementwise exclusive OR \\
\comm{<<-} & global / parent-env assignment \\
\comm{|>} & native pipe (R 4.1+) \\
\bottomrule
\end{tabular}
\end{table}

---

### Operator Precedence (High to Low)

\begin{table}[!htbp]
\centering\small
\begin{tabular}{ll}
\toprule
Precedence & Operators \\
\midrule
1 (highest) & \comm{::}, \comm{:::} \\
2 & \comm{\$}, \comm{@} \\
3 & \comm{[}, \comm{[[} \\
4 & \comm{\textasciicircum} (right-associative) \\
5 & unary \comm{-}, \comm{+} \\
6 & \comm{:} \\
7 & \comm{\%any\%} (including \comm{\%*\%}, \comm{\%/\%}, \comm{\%\%}, \comm{\%in\%}) \\
8 & \comm{*}, \comm{/} \\
9 & \comm{+}, \comm{-} \\
10 & \comm{<}, \comm{>}, \comm{<=}, \comm{>=}, \comm{==}, \comm{!=} \\
11 & \comm{!} \\
12 & \comm{\&}, \comm{\&\&} \\
13 & \comm{|}, \comm{||} \\
14 & \comm{|>} \\
15 & \comm{\textasciitilde} \\
16 (lowest) & \comm{<-}, \comm{<<-}, \comm{->}, \comm{->>} \\
\bottomrule
\end{tabular}
\end{table}

---

### `%in%` and `match()`

```{r}
x <- c(1, 5, 3, 8, 10)
x %in% c(3, 5, 7)      # which elements of x are in the set?
```

```{r}
# match() returns the position of first match (NA if not found)
match(c(3, 5, 99), x)
```

```{r}
# practical: filter data frame rows by a set of values
tg <- ToothGrowth
tg[tg$dose %in% c(0.5, 2), ][1:6, ]
```

# Common Functions

### Math Functions

\begin{table}[!htbp]
\centering\small
\begin{tabular}{ll}
\toprule
Function & Definition \\
\midrule
\comm{abs(x)} & absolute value \\
\comm{floor(x)}, \comm{ceiling(x)}, \comm{trunc(x)} & rounding toward $-\infty$, $+\infty$, 0\\
\comm{round(x, digits=0)} & round (banker's rounding)\\
\comm{signif(x, digits=6)} & significant digits\\
\comm{sqrt(x)}, \comm{exp(x)} & $\sqrt{x}$, $e^x$ \\
\comm{log(x)}, \comm{log10(x)}, \comm{log2(x)} & natural, base-10, base-2 log\\
\comm{factorial(x)}, \comm{choose(n, m)} & $x!$, $\binom{n}{m}$ \\
\comm{sin, cos, tan, asin, acos, atan} & trig functions\\
\comm{beta(a, b)}, \comm{gamma(x)} & special functions\\
\bottomrule
\end{tabular}
\end{table}

---

### Examples: Math Functions

```{r}
abs(-1); floor(5.7); ceiling(-5.7); trunc(-5.7)
round(5.7); round(5.5); round(2.5)  # banker's rounding!
signif(12345678, 4)
```

```{r}
sin(pi / 2); cos(pi / 2)  # cos not exactly 0 (floating point)
119 %/% 13; 119 %% 13
```

---

### Statistical Functions

\begin{table}[!htbp]
\centering\small
\begin{tabular}{ll}
\toprule
Function & Definition \\
\midrule
\comm{max(x)}, \comm{min(x)}, \comm{range(x)} & max, min, c(min, max)\\
\comm{sum(x)}, \comm{prod(x)} & sum, product\\
\comm{mean(x)}, \comm{median(x)} & mean, median\\
\comm{quantile(x, p)} & quantile at probability \comm{p}\\
\comm{var(x)}, \comm{sd(x)} & sample variance, std deviation\\
\comm{cor(x, y)}, \comm{cov(x, y)} & correlation, covariance\\
\comm{sort(x)}, \comm{rev(x)} & sorted copy, reversed copy\\
\comm{rank(x)}, \comm{order(x)} & ranks, permutation to sort\\
\midrule
\comm{cumsum}, \comm{cumprod}, \comm{cummax}, \comm{cummin} & cumulative operations\\
\comm{diff(x)} & lagged differences\\
\comm{pmax(x,y)}, \comm{pmin(x,y)} & parallel (elementwise) max/min\\
\comm{table(x)} & frequency table\\
\comm{colMeans}, \comm{colSums} & column means/sums\\
\comm{rowMeans}, \comm{rowSums} & row means/sums\\
\bottomrule
\end{tabular}
\end{table}

---

### Floating Point Surprises

```{r}
0.1 + 0.2 == 0.3              # FALSE!
all.equal(0.1 + 0.2, 0.3)     # TRUE (with tolerance)
isTRUE(all.equal(0.1 + 0.2, 0.3))  # safe for if()
```

```{r}
# trunc vs floor for negative numbers
trunc(-3.7); floor(-3.7)     # toward 0 vs toward -Inf
```

# Data Frames

## Basics

### What Is a Data Frame?

- A rectangular table: one row per observation, one column per variable.
- **Technically**: a named list of equal-length vectors (we'll revisit this in Lists).

```{r}
tg <- ToothGrowth    # built-in data; used throughout
head(tg, 4)
```

```{r}
# Each column is a vector; use $ to extract
tg$len[1:5]
```

---

### Data Frame Inspection

```{r}
str(tg)
```

```{r}
nrow(tg); ncol(tg); dim(tg); names(tg)
```

---

### `summary()` --- Quick Statistics

```{r}
summary(tg)
```

---

### Indexing: `[row, col]`

```{r}
tg[3, 1]              # row 3, column 1
tg$len[3]             # equivalent via $
(z <- tg[2:4, c(1, 3)])   # rows 2-4, columns 1 & 3
```

```{r}
head(tg[, c(1, 3)], 3)   # all rows, columns 1 & 3
head(tg[, -2], 3)         # all rows, remove column 2
```

---

### Creating Data Frames

:::::: columns
::: column

```{r}
age <- c(55, 58, 45)
name <- c('Alice', 'Bill', 'Cathy')
(d <- data.frame(age, name))
```

- Vectors of different lengths are not allowed!

```{r, error=TRUE}
blood <- c('O', 'B', 'A', 'AB')
data.frame(age, name, blood)
```

:::
::: column

```{r, eval=FALSE}
# with custom column names
data.frame(p1 = pred1, p2 = pred2)

# from a named list of vectors
l <- list(p1 = pred1, p2 = pred2)
as.data.frame(l)
```

```{r, eval=FALSE}
# from row data via rbind
r1 <- data.frame(a=1, b=2, c="X")
r2 <- data.frame(a=3, b=4, c="Y")
rbind(r1, r2)
```

:::
::::::

## Extracting and Filtering

### Row Filtering with Logical Conditions

```{r}
tg_vc <- tg[tg$supp == 'VC', ]
tg_oj <- tg[tg$supp == 'OJ', ]
mean(tg_vc$len); mean(tg_oj$len)
```

```{r}
# AND condition
tg[tg$supp == 'OJ' & tg$len < 8.8, ]
```

```{r}
# OR condition
head(tg[tg$len > 28 | tg$dose == 1.0, ], 6)
```

- Note the comma (`,`) and the use of `==`, not `=`!

---

### `subset()` --- Cleaner Filtering

```{r}
# equivalent to tg[tg$supp == 'OJ' & tg$len > 20, ]
subset(tg, supp == 'OJ' & len > 20)
```

```{r}
# select specific columns
subset(tg, dose == 2, select = c(len, supp))
```

---

### `with()` and `within()`

```{r}
# with() evaluates an expression in the data frame environment
with(tg, mean(len[supp == 'VC']))
```

```{r}
# within() modifies and returns the data frame
tg2 <- within(tg, {
  len_cm  <- len / 10
  log_len <- log(len)
})
head(tg2, 3)
```

## Modifying Data Frames

### Adding, Removing, Renaming Columns

```{r}
d <- data.frame(x = 1:3, y = c(10, 20, 30))
d$z <- d$x + d$y              # add column with $
d[['w']] <- c('a', 'b', 'c') # add column with [[]]
d
```

```{r}
d$z <- NULL                   # remove column
names(d)[names(d) == 'w'] <- 'label'   # rename
d
```

---

### Sorting Data Frames

```{r}
d <- data.frame(name = c('Zara', 'Amy', 'Mike'),
                score = c(88, 95, 72))
d[order(d$score), ]                        # ascending
d[order(d$score, decreasing = TRUE), ]     # descending
```

```{r}
# sort by multiple columns
d2 <- data.frame(g = c('A','B','A','B'), x = c(3,1,2,4))
d2[order(d2$g, d2$x), ]
```

---

### Handling NAs

```{r}
df <- data.frame(x = c(1, NA, 3, 4, 5), y = c(1, 2, NA, 4, 5))
na.omit(df)                      # remove rows with any NA
```

```{r}
complete.cases(df)                # logical: which rows complete?
df[complete.cases(df), ]
```

```{r, eval=FALSE}
df$x[is.na(df$x)] <- 0           # replace specific column NAs
df[is.na(df)] <- 0               # replace all NAs
```

---

### Combining Data Frames: `cbind`, `rbind`

```{r}
# column-bind (side by side) --- same number of rows
df1 <- data.frame(a = c(1, 2)); df2 <- data.frame(b = c(7, 8))
cbind(df1, df2)
```

```{r}
# row-bind (stack) --- same column names
df1 <- data.frame(x = c("a", "a"), y = c(5, 6))
df2 <- data.frame(x = c("b", "b"), y = c(9, 10))
rbind(df1, df2)
```

## Split--Apply--Combine

### The Split--Apply--Combine Paradigm

\begin{table}[!htbp]
\centering\small
\begin{tabular}{lll}
\toprule
Step & Base R tool & What it does \\
\midrule
Split & \comm{split()} & partition data by factor $\to$ list \\
Apply & \comm{lapply/sapply/tapply/\ldots} & run function on each piece \\
Combine & \comm{unsplit/rbind/do.call/\ldots} & reassemble results \\
\midrule
All-in-one & \comm{aggregate()} & split + apply + combine in one call \\
Horizontal combine & \comm{merge()} & join two tables by key \\
\bottomrule
\end{tabular}
\end{table}

---

### `split()` --- Partition Data

```{r}
scores <- c(88, 92, 78, 95, 85, 70, 91, 83)
group  <- c('A','B','A','B','A','B','A','B')
(sg <- split(scores, group))
sapply(sg, mean)         # per-group mean
```

```{r}
# split a data frame --- each element is a complete data frame
tg_split <- split(tg, tg$supp)
head(tg_split$OJ, 3)
```

---

### `split()` --- Multi-Factor and `unsplit()`

```{r}
tg_split2 <- split(tg, list(tg$supp, tg$dose))
names(tg_split2)
```

```{r}
# unsplit: put pieces back in their original positions
sg <- split(1:8, rep(c('A','B'), 4))
sg$A <- sg$A * 10; sg$B <- sg$B * (-1)
unsplit(sg, rep(c('A','B'), 4))
```

---

### The `split` + `lapply` + `rbind` Pipeline

```{r}
# add a z-score column within each group
standardize <- function(df) {
  df$len_z <- (df$len - mean(df$len)) / sd(df$len)
  df
}
parts <- split(tg, tg$supp)
parts <- lapply(parts, standardize)
tg2 <- do.call(rbind, parts)
head(tg2[order(as.numeric(rownames(tg2))), ], 4)
```

---

### `aggregate()` --- Group-wise Summaries

```{r}
# formula syntax: response ~ grouping
aggregate(len ~ supp, data = tg, FUN = mean)
aggregate(len ~ supp + dose, data = tg, FUN = mean)
```

---

### `aggregate()` --- Custom Summary and Multiple Responses

```{r}
# custom function: result is a matrix column!
aggregate(len ~ supp, data = tg,
          FUN = function(x) c(mean = mean(x), sd = sd(x)))
```

```{r}
# multiple response columns with cbind()
aggregate(cbind(mpg, hp, wt) ~ cyl, data = mtcars, FUN = mean)
```

---

### `aggregate()` --- Expand Matrix Column and `.` Shorthand

```{r}
# expand matrix column into proper data frame
res <- aggregate(len ~ supp, data = tg,
                 FUN = function(x) c(m = mean(x), s = sd(x)))
cbind(res[1], as.data.frame(res$len))
```

```{r}
# . means "all other columns"
aggregate(. ~ cyl, data = mtcars[, c('mpg','hp','wt','cyl')],
          FUN = mean)
```

---

### `merge()` --- Joining Data Frames

```{r}
customers <- data.frame(id = 1:4,
  name = c('Alice', 'Bob', 'Cathy', 'Dan'))
orders <- data.frame(cust_id = c(2, 3, 3, 1, 5),
  product = c('pen','book','ink','paper','tape'),
  amount = c(12, 35, 8, 20, 5))
# inner join (default): only matched rows
merge(customers, orders, by.x = 'id', by.y = 'cust_id')
```

---

### `merge()` --- Join Types

```{r}
# left join: keep all customers (NA for no match)
merge(customers, orders, by.x='id', by.y='cust_id', all.x=TRUE)
```

\vfill

\begin{table}[!htbp]
\centering\small
\begin{tabular}{lll}
\toprule
SQL & \texttt{merge()} arguments & Keeps \\
\midrule
\texttt{INNER JOIN} & default (\comm{all=FALSE}) & matched rows only \\
\texttt{LEFT JOIN} & \comm{all.x = TRUE} & all rows in \comm{x} \\
\texttt{RIGHT JOIN} & \comm{all.y = TRUE} & all rows in \comm{y} \\
\texttt{FULL OUTER JOIN} & \comm{all = TRUE} & all rows in both \\
\bottomrule
\end{tabular}
\end{table}

---

### `merge()` --- Multi-Column Keys and Suffixes

```{r}
# multi-column key
sales <- data.frame(year=c(2023,2023,2024), qtr=c(1,2,1), rev=c(100,120,110))
targets <- data.frame(year=c(2023,2023,2024), qtr=c(1,2,1), target=c(95,115,105))
merge(sales, targets, by = c('year', 'qtr'))
```

```{r}
# duplicate column names get suffixes
df1 <- data.frame(id = 1:3, value = c(10, 20, 30))
df2 <- data.frame(id = 1:3, value = c(100, 200, 300))
merge(df1, df2, by = 'id', suffixes = c('_old', '_new'))
```

# Lists

## List Basics

### What Is a List?

- A **generic vector**: elements can be of **different types and lengths**.

```{r}
vn <- c(1, 1.2, 2.3, 3.4, 4.5)
vb <- c(TRUE, TRUE, FALSE)
vc <- c('limestone', 'marl', 'oolite', 'CaCO3')
```

```{r, error=TRUE}
data.frame(vn, vb, vc)   # ERROR: different lengths
```

---

### Creating Lists

```{r}
# unnamed list
l <- list(3.14, 'Moe', c(1, 3, 5, 2), mean)
str(l)
```

```{r}
# named list + incremental construction
l <- list(max = 5, right = 0.6, mean = 20)
l <- list(); l$left <- 0.1; l$right <- 0.52; l$mid <- 0.33
l
```

---

### `[[` vs `[` --- The Most Important Distinction

:::::: columns
::: column

```{r}
l <- list(a = 1:3, b = 'hello',
          c = TRUE)
l[[1]]     # the element itself
l[1]       # a sub-list
```

```{r}
class(l[[1]]); class(l[1])
```

:::
::: column

```{r}
l$a   # same as l[['a']]
l$b
```

\vfill

**Analogy**: `[[` reaches into the box and pulls out the item; `[` selects boxes (still wrapped).

:::
::::::

---

### Modifying, Removing, Concatenating

```{r}
l <- list(x = 1, y = 2, z = 3)
l$x <- 100; l[['y']] <- 200; l$w <- 'new'   # modify / add
l[c('x', 'w')] <- NULL; l                     # remove
```

```{r}
l1 <- list(a = 1, b = 2); l2 <- list(c = 3, d = 4)
c(l1, l2)    # concatenate lists
```

```{r}
# nested lists
nested <- list(student = list(name = 'Alice', age = 20),
               scores  = list(math = 95, english = 88))
nested$student$name; nested[['scores']][['math']]
```

---

### `unlist()` --- Flatten a List

```{r}
iq <- list(100, 120, 140, 112, 105)
mean(unlist(iq))       # must unlist before mean()
```

```{r}
l <- list(a = 1:3, b = 4:5)
unlist(l)                        # names preserved with prefix
unlist(l, use.names = FALSE)     # drop names
```

---

### `split()` --- Split a Vector into a List

```{r}
(mtl <- split(mtcars$mpg, mtcars$cyl))
mtl$`6`; mtl[[2]]   # access by name or position
```

- `tapply(x, f, FUN)` is equivalent to `sapply(split(x, f), FUN)`

## List--Data Frame Connection

### A Data Frame Is a List

```{r}
df <- data.frame(x = 1:3, y = c('a', 'b', 'c'))
is.list(df); typeof(df)
```

```{r}
df$x; df[['x']]     # extracts column vector (like [[)
df['x']              # returns 1-column data.frame (like [)
```

```{r}
# split() a data frame -> list of data frames
str(split(tg, tg$supp))
```

---

### Conversions Between Data Structures

\begin{table}[!htbp]
\centering\small
\begin{tabular}{llll}
\toprule
From & To & How & Note\\
\midrule
vector & list & \comm{as.list(v)} & Don't use \comm{list(v)}!\\
vector & matrix & \comm{matrix(v, n, m)} & \\
vector & data.frame & \comm{as.data.frame(v)} & \\
list & vector & \comm{unlist(l)} & Not \comm{as.vector}\\
list & data.frame & \comm{as.data.frame(l)} & Equal-length only\\
matrix & vector & \comm{as.vector(m)} & \\
matrix & data.frame & \comm{as.data.frame(m)} & \\
data.frame & list & \comm{as.list(df)} & Removes class attr\\
data.frame & matrix & \comm{as.matrix(df)} & Coercion!\\
\bottomrule
\end{tabular}
\end{table}

# Matrices

### Creating Matrices

```{r}
v <- 1:6; matrix(v, nrow = 2, ncol = 3)
```

```{r}
# named rows and columns
m <- matrix(c(1, .56, .38, .56, 1, .44, .38, .44, 1), 3, 3)
colnames(m) <- rownames(m) <- c('AAPL', 'MSFT', 'GOOG')
m; m['MSFT', 'GOOG']
```

---

### Matrix Operations

```{r}
A <- matrix(1:4, 2, 2); B <- matrix(5:8, 2, 2)
A * B        # element-wise
A %*% B      # matrix multiplication
t(A); det(A); solve(A)   # transpose, determinant, inverse
```

```{r}
# row/column selection: drop=FALSE preserves matrix class
A[1, ]; A[1, , drop = FALSE]
```

---

### Matrix Functions Summary

```{r, eval=FALSE}
nrow(m); ncol(m); dim(m)
diag(m)                    # diagonal elements
diag(3)                    # 3x3 identity matrix
crossprod(A, B)            # t(A) %*% B (faster)
tcrossprod(A, B)           # A %*% t(B)
outer(1:3, 1:3, '+')      # outer product
eigen(A)                   # eigenvalue decomposition
svd(A)                     # SVD
qr(A)                      # QR decomposition
```

# Programming

## Functions

### Defining Functions

```{r, eval=FALSE}
function_name <- function(arg1, arg2 = default_value, ...) {
    # code
    return(some_object)
}
```

```{r}
# functions can read global variables
y_squared <- function() { return(y^2) }
y <- 2; y_squared()
```

```{r}
# but local assignment doesn't change globals
add_to_y <- function(n) { y <- y + n }
y <- 1; add_to_y(2); y      # y is still 1
```

```{r}
# <<- modifies globals (use sparingly!)
add_to_y_global <- function(n) { y <<- y + n }
y <- 1; add_to_y_global(2); y   # y is now 3
```

---

### `return()` and Implicit Return

```{r}
avg <- function(x) { return(sum(x) / length(x)) }
avg(c(2, 3, 6))
avg(c(TRUE, FALSE, TRUE))   # TRUE=1, FALSE=0
```

```{r}
# implicit return: the last evaluated expression is returned
avg_bad <- function(x) { r <- sum(x) / length(x) }
avg_ok  <- function(x) { sum(x) / length(x) }
avg_bad(1:10)   # silent! (assignment returns invisibly)
avg_ok(1:10)    # returns result
```

---

### Default Arguments and Named Arguments

```{r}
power_n <- function(x, n = 2) { return(x^n) }
power_n(3)              # default n = 2
power_n(3, 3)           # supplied n = 3
power_n(x = 5, n = 2)  # named arguments
power_n(n = 2, x = 5)  # order doesn't matter with names
```

---

### Functions as Arguments and `...`

```{r}
apply_to_first2 <- function(x, func, ...) {
    func(x[1:2], ...)
}
x <- c(NA, 2, 3)
apply_to_first2(x, sum, na.rm = TRUE)
```

---

### Anonymous Functions and `\(x)` Shorthand

```{r}
sapply(1:5, function(x) x^2 + 1)
sapply(1:5, \(x) x^2 + 1)   # R 4.1+ shorthand
```

```{r}
# useful in apply family
l <- list(a = 1:5, b = 6:10)
lapply(l, \(v) c(mean = mean(v), sd = sd(v)))
```

## Closures and Lexical Scoping

### Lexical Scoping Rule

When R encounters a free variable inside a function, it searches the **environment where the function was defined** (not where it was called):

```{r}
x <- 10
f <- function() x          # f is born in global env
g <- function() {
  x <- 999
  f()                       # calls f, but f ignores this x
}
g()                         # 10, not 999
```

\vfill

- **Lexical** (static): determined by where the code is **written**
- R, Python, JavaScript all use lexical scoping

---

### What Is a Closure?

> **Closure** = a function **bundled together** with the environment in which it was created.

When function A returns function B, B retains a **live reference** to A's local environment --- even after A has finished executing.

```{r}
power_factory <- function(n) {
  function(x) x^n             # captures n
}
square <- power_factory(2)
cube   <- power_factory(3)
square(5); cube(5)
```

```{r}
# each call creates a new, independent environment
environment(square)$n; environment(cube)$n
```

---

### Inspecting Function Internals

R does not distinguish "closure" as a special type. Every function stores `formals`, `body`, and `environment`:

```{r}
f <- function(x) x + 1
typeof(f)          # "closure"
formals(f); body(f); environment(f)
```

---

### `<-` vs `<<-` and Mutable State

- `<-` assigns **only** in the current (local) environment
- `<<-` walks **up** the enclosing environment chain to modify an existing variable

```{r}
make_counter <- function() {
  n <- 0
  list(
    increment = function() { n <<- n + 1; n },
    get       = function() n,
    reset     = function() { n <<- 0 }
  )
}
ctr <- make_counter()
ctr$increment(); ctr$increment(); ctr$get()
ctr$reset(); ctr$get()
```

---

### Independent Instances and `local()`

```{r}
ctr1 <- make_counter(); ctr2 <- make_counter()
ctr1$increment(); ctr1$increment(); ctr2$increment()
cat("ctr1:", ctr1$get(), " ctr2:", ctr2$get(), "\n")
```

```{r}
# local() creates a single-instance closure
counter <- local({
  n <- 0
  list(
    increment = function() { n <<- n + 1; n },
    get       = function() n,
    reset     = function() { n <<- 0 })
})
counter$increment(); counter$increment(); counter$get()
```

---

### Classic Pitfall: Loop Capture

```{r}
funs <- list()
for (i in 1:5) funs[[i]] <- function() i
sapply(funs, \(f) f())  # all 5!
```

All five functions share the **same** environment; after the loop, `i = 5`.

```{r}
# Fix 1: force() defeats lazy evaluation
make_fun <- function(val) { force(val); function() val }
funs <- list()
for (i in 1:5) funs[[i]] <- make_fun(i)
sapply(funs, \(f) f())
```

```{r}
# Fix 2 (recommended): lapply creates separate environments
funs <- lapply(1:5, \(i) function() i)
sapply(funs, \(f) f())
```

---

### Base R Closures: `ecdf`, `approxfun`, `Negate`

```{r}
x <- c(1, 3, 3, 5, 7, 9); Fn <- ecdf(x)
Fn(4); Fn(7)        # P(X<=4)=3/6; P(X<=7)=5/6
typeof(Fn)           # "closure"
```

```{r}
xp <- 1:5; yp <- c(2, 5, 3, 8, 6)
linear_interp <- approxfun(xp, yp)
linear_interp(2.5)   # returns a closure over the knot data
```

```{r}
is_even <- function(x) x %% 2 == 0
is_odd  <- Negate(is_even)          # closure combinator
Filter(is_odd, 1:10)
```

---

### Decorator and Iterator Patterns

```{r}
with_logging <- function(f, fname = deparse(substitute(f))) {
  call_count <- 0L
  function(...) {
    call_count <<- call_count + 1L
    cat(sprintf("[%s] call #%d\n", fname, call_count))
    f(...)
  }
}
logged_mean <- with_logging(mean)
logged_mean(1:10); logged_mean(c(3, NA, 7), na.rm = TRUE)
```

---

### Moving Average and Fibonacci Iterator

```{r}
make_ema <- function(alpha = 0.3) {
  value <- NULL
  function(x) {
    if (is.null(value)) { value <<- x }
    else { value <<- alpha * x + (1 - alpha) * value }
    value
  }
}
ema <- make_ema(0.2)
sapply(c(100, 102, 101, 105, 103, 107, 106), ema)
```

```{r}
make_fib_iterator <- function() {
  a <- 0; b <- 1
  list(next_val = function() {
    val <- a; new <- a + b; a <<- b; b <<- new; val
  }, reset = function() { a <<- 0; b <<- 1 })
}
fib_it <- make_fib_iterator()
replicate(10, fib_it$next_val())
```

## Control Flow

### `if` / `else`

```{r}
reciprocal <- function(x) {
  if (is.na(x)) {
    cat("Error! Division by NA.\n")
  } else if (is.infinite(x)) {
    cat("Error! Division by infinity.\n")
  } else if (x == 0) {
    cat("Error! Division by zero.\n")
  } else { 1 / x }
}
reciprocal(2); reciprocal(0); reciprocal(NA)
```

---

### `ifelse()` --- Vectorized Conditional

```{r}
x <- c(-1, 2, -100, 20, 50)
ifelse(x > 0, 'positive', 'negative')
```

```{r}
# nested ifelse
v <- c(20, 18, 45, 33, 25, 9, 17, 55, 69, 81, 44, 32)
ifelse(v >= 20 & v < 40, 'young adult',
ifelse(v >= 40 & v < 60, 'middle age', 'other'))
```

---

### `switch()`

```{r}
day_type <- function(day) {
  switch(day,
    Monday=, Tuesday=, Wednesday=,
    Thursday=, Friday= "Weekday",
    Saturday=, Sunday= "Weekend",
    "Unknown")
}
day_type("Monday"); day_type("Sunday"); day_type("Holiday")
```

---

### `&&` vs `&` (Short-Circuit Evaluation)

```{r}
# first, ensure 'aa' does not exist from earlier chunks
if (exists("aa")) rm(aa)
```

```{r}
# && is short-circuited: stops at first FALSE
if (exists("aa") && aa > 0) {
    cat("aa exists and is positive.\n")
} else {
    cat("aa doesn't exist or is negative.\n")
}
```

```{r, error=TRUE}
# & evaluates BOTH sides -- error because aa doesn't exist!
if (exists("aa") & aa > 0) {
    cat("aa exists and is positive.\n")
} else {
    cat("aa doesn't exist or is negative.\n")
}
```

- **Rule**: Use `&&` / `||` in `if` (scalar); use `&` / `|` for vectors.

### `if (FALSE) { }` --- Code Block Toggle

```{r, eval=FALSE}
# wrap example/test code that shouldn't run
if (FALSE) {
  # these lines are "commented out" but remain valid R code
  book(date = "2026/04/02", headless = FALSE)
  cancel(order_id = "01901879")
}
```

- Unlike `#` comments, the code inside is still checked by R's parser

## Loops

### `for` Loop

```{r}
# sum 1 to 10000
ans <- 0
for (i in 1:10000) { ans <- ans + i }
ans
```

```{r}
# sum of min(2^k, k^4) for k = 1..100
ans <- 0
for (k in 1:100) { ans <- ans + min(2^k, k^4) }
ans
```

```{r}
# vectorized equivalent (much faster)
v <- 1:100; sum(pmin(2^v, v^4))
```

---

### `while`, `repeat`, `break`, `next`

```{r}
# while: find smallest n such that sum(1:n) > 1000
n <- 0; s <- 0
while (s <= 1000) { n <- n + 1; s <- s + n }
cat("n =", n, ", sum =", s, "\n")
```

```{r}
# repeat + break
n <- 1
repeat { if (n^2 > 100) break; n <- n + 1 }
cat("First n with n^2 > 100:", n, "\n")
```

```{r}
# next: skip iterations
for (i in 1:10) {
  if (i %% 3 == 0) next    # skip multiples of 3
  cat(i, '')
}
```

---

### Vectorization vs Loops: Speed Comparison

```{r}
my_sum <- function(x) {
    ans <- 0
    for (i in seq_along(x)) { ans <- ans + x[i] }
    ans
}
v <- 1:1000000; system.time(my_sum(v)); system.time(sum(v))
```

```{r}
# pre-allocate vs grow-in-loop vs vectorized
n <- 1e5
system.time({v <- c(); for (i in 1:n) v <- c(v, i^2)})
system.time({v <- numeric(n); for (i in 1:n) v[i] <- i^2})
system.time({v <- (1:n)^2})
```

## Pipe Operator

### Native Pipe `|>` (R 4.1+)

```{r}
# without pipe: nested calls are hard to read
round(mean(abs(rnorm(100))), 3)
```

```{r}
# with pipe: read left to right
set.seed(1)
rnorm(100) |> abs() |> mean() |> round(3)
```

```{r}
# pipe with data frame operations
mtcars |> subset(cyl == 6) |> head(3)
```

```{r, eval=FALSE}
# use _ as placeholder for non-first argument (R 4.2+)
mtcars |> lm(mpg ~ wt, data = _) |> summary()
```

# The Apply Family

### `apply()` --- Over Matrix / Data Frame Margins

```{r}
m <- matrix(1:12, nrow = 3,
            dimnames = list(paste0('r',1:3), paste0('c',1:4)))
m
apply(m, 1, sum)     # MARGIN=1: rows
apply(m, 2, sum)     # MARGIN=2: columns
```

```{r}
# when FUN returns a vector, result is k x n matrix
apply(m, 2, range)
apply(m, 2, quantile, probs = c(0.25, 0.75))
```

---

### `apply()` on Data Frames --- Beware of Coercion!

```{r}
df <- data.frame(x = 1:3, y = c('a','b','c'), z = 4:6)
apply(df, 2, class)       # all "character"!
```

- `apply()` converts data frames to matrices first --- **all columns become character** if any is character. Use `sapply`/`lapply` for mixed types.

```{r}
# faster alternatives for row/column sums/means:
colSums(m); rowSums(m); colMeans(m); rowMeans(m)
```

---

### `lapply()` and `sapply()`

```{r}
x <- list(a = 1:5, b = rnorm(10), c = c(TRUE, FALSE, TRUE))
lapply(x, mean)      # always returns a list
sapply(x, mean)      # simplifies to named vector
```

```{r}
# sapply can be unpredictable; vapply is safer
vapply(x, mean, numeric(1))
```

---

### `sapply()` Is Unpredictable --- Use `vapply()` in Production

```{r}
# when all lengths equal: matrix. When they differ: list!
sapply(list(a = 1:3, b = 4:6), identity)
sapply(list(a = 1:3, b = 4:5), identity)
```

```{r}
# vapply: explicit type template; error if mismatch
vapply(iris[,1:4], median, numeric(1))
```

---

### `tapply()` --- Apply by Groups

```{r}
# mean mpg by cylinder count
tapply(mtcars$mpg, mtcars$cyl, mean)
```

```{r}
# two-way grouping -> matrix
tapply(mtcars$mpg, list(mtcars$cyl, mtcars$am), mean)
```

- `tapply(x, f, FUN)` $\equiv$ `sapply(split(x, f), FUN)`

---

### `mapply()` and `Map()`

```{r}
# apply to corresponding elements of multiple vectors
mapply(function(x, y) x + y, 1:5, 10:14)
mapply(rep, 1:3, 3:1)   # variable-length results
```

```{r}
# Map: always returns a list (no simplification)
Map('+', 1:3, 10:12)
```

```{r}
# MoreArgs: constant arguments not varied across elements
mapply(function(x, y, suffix) paste0(x, y, suffix),
       x = c('Alice', 'Bob'), y = c(25, 30),
       MoreArgs = list(suffix = '!'))
```

---

### `rapply()` --- Recursive Apply

```{r}
nested <- list(a = 1:3,
  b = list(c = 4:6, d = list(e = 7:9, f = 'text')))
rapply(nested, sum, classes = 'integer')         # unlist
rapply(nested, sum, classes = 'integer', how = 'list')
rapply(nested, \(x) x * 10, classes = 'integer', how = 'replace')
```

---

### `Reduce()` --- Sequential Accumulation

```{r}
Reduce('+', 1:5)                          # 15
Reduce('+', 1:5, accumulate = TRUE)       # running total
```

```{r}
# merge multiple data frames
dfs <- list(data.frame(id=1:3, a=10:12),
            data.frame(id=2:4, b=20:22),
            data.frame(id=1:3, c=30:32))
Reduce(\(x, y) merge(x, y, by='id', all=TRUE), dfs)
```

---

### `Filter()`, `Find()`, `Position()`, `Negate()`

```{r}
x <- list(1:3, "hello", 4:6, TRUE, NULL)
Filter(is.numeric, x)          # keep numeric elements
Find(is.character, x)          # first character element
Position(is.character, x)      # its position
```

```{r}
Filter(\(x) x %% 15 == 0, 1:50)   # multiples of 15
```

```{r}
# Find/Position with direction
Find(\(x) x > 5, c(2, 8, 3, 9, 1), right = TRUE)
Position(\(x) x > 5, c(2, 8, 3, 9, 1), right = TRUE)
```

---

### Decision Table: Choosing the Right Tool

\begin{table}[!htbp]
\centering\footnotesize
\begin{tabular}{lllp{3cm}}
\toprule
Function & Input & Output & Use case \\
\midrule
\comm{apply} & matrix/array & vector/matrix & row/column ops \\
\comm{lapply} & vector/list & list & general iteration \\
\comm{sapply} & vector/list & simplified & interactive only \\
\comm{vapply} & vector/list & typed output & production code \\
\comm{tapply} & vector + factor & array & group summaries \\
\comm{mapply/Map} & multiple vectors & simplified/list & parallel iteration \\
\comm{rapply} & nested list & various & recursive ops \\
\comm{Reduce} & vector/list & single value & sequential fold \\
\comm{Filter} & vector/list & subset & keep matching \\
\comm{Find/Position} & vector/list & one element & first match \\
\comm{Negate} & function & function & invert predicate \\
\bottomrule
\end{tabular}
\end{table}

---

### Common Equivalences

```{r, eval=FALSE}
tapply(x, f, FUN)         <==>  sapply(split(x, f), FUN)
sapply(X, FUN)            <==>  unlist(lapply(X, FUN))   # scalar FUN
mapply(FUN, x, y)         <==>  sapply(seq_along(x),
                                  \(i) FUN(x[i], y[i]))
Map(FUN, x, y)            <==>  mapply(FUN, x, y, SIMPLIFY = FALSE)
Reduce('+', 1:5)          <==>  ((((1+2)+3)+4)+5)
Filter(f, x)              <==>  x[sapply(x, f)]
aggregate(y~g, d, mean)   <==>  tapply(d$y, d$g, mean)  # approx
```

# String Operations

### Basic String Functions

```{r}
nchar('hello world')
toupper('hello'); tolower('HELLO')
trimws('  hello  '); trimws('  hello  ', which = 'left')
```

```{r}
substr('abcdefgh', 3, 5)
startsWith('hello world', 'hello')
endsWith('hello world', 'world')
chartr('abc', 'ABC', 'a big cat')
```

---

### `paste()` and `paste0()`

```{r}
paste('a', 'b', 'c', sep = '-')
paste0('item', 1:5)
```

```{r}
(labels <- paste(c('a', 'b'), 1:6, sep = '-'))
```

```{r}
# collapse: join multiple strings into one
paste('item', 1:5, sep = '_', collapse = ', ')
```

---

### `sprintf()` --- Formatted Strings

```{r}
# %d: integer; %.2f: 2 decimal places; %s: string
sprintf('Name: %s, Age: %d, Score: %.2f', 'Alice', 25, 95.678)
```

```{r}
sprintf('%05d', 42)         # zero-padded
sprintf('%.2f%%', 12.345)   # percentage
sprintf('%-10s|', 'left')   # left-align
```

---

### `cat()` --- Print to Console

```{r}
v <- c(20, 30, 3 * pi)
cat("The content of v is:", v, "...\n")
```

```{r}
for (i in 50 * (0:2)) {
  cat(sprintf('exp(sin(%3d)) = %.6f', i, exp(sin(i))), '\n')
}
```

---

### `strsplit()` --- Split Strings

```{r}
strsplit('a.b.c.d', split = '\\.')
strsplit('hello world foo', split = ' ')
strsplit(c('a-b', 'c-d-e'), split = '-')
strsplit('hello', split = '')   # every character
```

---

### `grep()`, `grepl()` --- Pattern Matching

```{r}
x <- c('apple', 'banana', 'cherry', 'apricot', 'blueberry')
grep('an', x)                                  # indices
grep('an', x, value = TRUE)                   # values
grepl('an', x)                                  # logical
grep('APPLE', x, ignore.case = TRUE, value = TRUE)
```

---

### `sub()`, `gsub()` --- Pattern Replacement

```{r}
x <- 'Hello World Hello'
sub('Hello', 'Hi', x)      # replace FIRST match
gsub('Hello', 'Hi', x)     # replace ALL matches
```

```{r}
gsub('[0-9]', '', 'abc123def456')      # remove all digits
gsub(' +', ' ', 'too   many    spaces')  # compress spaces
```

---

### `regexpr()`, `gregexpr()`, `regmatches()`

```{r}
x <- 'Phone: 02-1234-5678, Fax: 04-8765-4321'
m <- gregexpr('[0-9]+-[0-9]+-[0-9]+', x)
regmatches(x, m)    # extract all matches
```

# Regular Expressions

### What Are Regular Expressions?

- Patterns for matching text. Used in `grep`, `grepl`, `sub`, `gsub`, `regexpr`, `strsplit`, etc.
- In R strings, backslash must be **doubled**: write `\\d` for the regex `\d`.

---

### Basic Metacharacters

\begin{table}[!htbp]
\centering\small
\begin{tabular}{ll}
\toprule
Pattern & Meaning \\
\midrule
\comm{.} & any single character (except newline) \\
\comm{\textasciicircum} & start of string \\
\comm{\$} & end of string \\
\comm{*} & 0 or more of preceding \\
\comm{+} & 1 or more of preceding \\
\comm{?} & 0 or 1 of preceding \\
\comm{\{n\}}, \comm{\{n,m\}} & exactly \comm{n}; between \comm{n} and \comm{m} \\
\comm{|} & alternation (or) \\
\comm{()} & grouping and capture \\
\comm{[]} & character class \\
\comm{\textbackslash} & escape next character \\
\bottomrule
\end{tabular}
\end{table}

---

### Character Classes

\begin{table}[!htbp]
\centering\small
\begin{tabular}{ll}
\toprule
Class & Meaning \\
\midrule
\comm{[abc]} & matches a, b, or c \\
\comm{[\textasciicircum abc]} & matches anything except a, b, c \\
\comm{[a-z]}, \comm{[A-Z]}, \comm{[0-9]} & ranges \\
\comm{\textbackslash d} / \comm{\textbackslash D} & digit / non-digit \\
\comm{\textbackslash w} / \comm{\textbackslash W} & word char / non-word char \\
\comm{\textbackslash s} / \comm{\textbackslash S} & whitespace / non-whitespace \\
\comm{\textbackslash b} & word boundary \\
\bottomrule
\end{tabular}
\end{table}

```{r}
x <- c('hello123', 'world', '456', 'test!', 'a1b2')
grep('[0-9]', x, value = TRUE)       # contains a digit
grep('^[0-9]+$', x, value = TRUE)    # all digits
```

---

### Anchors, Quantifiers, Alternation

```{r}
x <- c('cat', 'concatenate', 'scat', 'catalog')
grep('^cat', x, value = TRUE)     # starts with 'cat'
grep('cat$', x, value = TRUE)     # ends with 'cat'
```

```{r}
x <- c('cat', 'dog', 'catfish', 'dogfish', 'bird')
grep('cat|dog', x, value = TRUE)       # cat OR dog
grep('^(cat|dog)$', x, value = TRUE)   # exactly cat or dog
```

---

### Capture Groups and Backreferences

```{r}
# swap first and last name
x <- c('John Smith', 'Jane Doe', 'Bob Lee')
sub('(\\w+) (\\w+)', '\\2, \\1', x)
```

```{r}
# extract parts of a date
dates <- c('2024-01-15', '2023-12-31')
sub('(\\d{4})-(\\d{2})-(\\d{2})', 'Year:\\1, Month:\\2', dates)
```

---

### Greedy vs Lazy Matching

```{r}
x <- '<b>bold</b> and <i>italic</i>'
regmatches(x, gregexpr('<.*>', x))     # greedy
regmatches(x, gregexpr('<.*?>', x))    # lazy (non-greedy)
```

---

### Lookahead and Lookbehind (Perl)

```{r}
x <- c('$100', '200', '$300', '400')
regmatches(x, gregexpr('(?<=\\$)\\d+', x, perl = TRUE))
```

```{r}
# negative lookahead: files NOT ending in .txt
x <- c('file.txt', 'image.png', 'doc.txt', 'photo.jpg')
grep('\\.(?!txt)', x, value = TRUE, perl = TRUE)
```

---

### `regexec()` --- Full Capture Group Extraction

```{r}
x <- c('2024-01-15 14:30', '2023-12-31 09:00')
pattern <- '(\\d{4})-(\\d{2})-(\\d{2}) (\\d{2}):(\\d{2})'
m <- regexec(pattern, x)
regmatches(x, m)
```

---

### Practical Regex Examples

```{r}
# validate email (simplified)
emails <- c('user@example.com', 'bad@@email',
            'test@domain.org', 'no_at_sign')
grep('^[\\w.]+@[\\w.]+\\.[a-z]{2,}$', emails,
     value = TRUE, perl = TRUE)
```

```{r}
# clean column names: replace non-alphanumeric with _
col_names <- c('First Name', 'Last.Name', 'Age (years)')
gsub('[^a-zA-Z0-9]', '_', col_names)
```

# Graphics

### Scatter Plots

```{r, out.width="50%"}
plot(cars$speed, cars$dist, xlab = 'Speed', ylab = 'Distance',
     main = 'cars dataset', pch = 19, col = 'steelblue')
abline(lm(dist ~ speed, data = cars), col = 'red', lwd = 2)
```

---

### Essential Plot Parameters

```{r, out.width="45%"}
x <- seq(0, 2*pi, length.out = 100)
plot(x, sin(x), type = 'l', col = 'blue', lwd = 2,
     main = 'Trig Functions', xlab = 'x', ylab = 'y')
lines(x, cos(x), col = 'red', lwd = 2, lty = 2)
legend('topright', legend = c('sin', 'cos'),
       col = c('blue', 'red'), lty = c(1, 2), lwd = 2)
```

---

### Plot Types: `type` Argument

```{r, out.width="60%", fig.height=4}
par(mfrow = c(2, 3), mar = c(2,2,2,1))
y <- c(3,1,4,1,5,9,2,6)
for (tp in c('p','l','b','o','h','s')) {
  plot(y, type = tp, main = paste("type =", tp))
}
```

---

### Multi-panel Plots with `par(mfrow)`

```{r, out.width="55%", fig.height=4}
par(mfrow = c(1, 3), mar = c(4,4,2,1))
hist(rnorm(1000), main = 'Histogram', col = 'lightblue')
boxplot(len ~ supp, data = tg, main = 'Boxplot',
        col = c('orange', 'cyan'))
barplot(table(mtcars$cyl), main = 'Barplot',
        col = c('red', 'green', 'blue'))
```

---

### Adding Elements to Plots

```{r, out.width="45%"}
plot(1:10, (1:10)^2, type = 'b', pch = 19)
abline(h = 50, col = 'red', lty = 2)
abline(v = 7, col = 'blue', lty = 3)
text(5, 80, 'annotation', col = 'darkgreen', cex = 1.2)
points(c(3, 6), c(9, 36), pch = 17, col = 'red', cex = 2)
```

---

### Color by Factor and Saving Plots

```{r, out.width="45%"}
plot(tg$dose, tg$len, col = as.factor(tg$supp),
     pch = 19, xlab = 'Dose', ylab = 'Tooth Length',
     main = 'ToothGrowth by Supplement')
legend('bottomright', legend = levels(as.factor(tg$supp)),
       col = 1:2, pch = 19)
```

```{r, eval=FALSE}
png('my_plot.png', width = 800, height = 600)
plot(1:10, main = 'My Plot'); dev.off()
pdf('my_plot.pdf', width = 8, height = 6)
plot(1:10, main = 'My Plot'); dev.off()
```

# Date and Time

### Date Objects

```{r}
today <- Sys.Date(); today; class(today)
```

```{r}
as.Date('2024-01-15')
as.Date('15/01/2024', format = '%d/%m/%Y')
as.Date('Jan 15, 2024', format = '%b %d, %Y')
```

```{r}
today + 30
as.Date('2024-12-31') - as.Date('2024-01-01')
```

---

### Date Formatting

```{r}
d <- as.Date('2024-07-15')
format(d, '%Y'); format(d, '%B'); format(d, '%A')
format(d, '%d %b %Y')
weekdays(d); months(d); quarters(d)
```

---

### Date-Time: `POSIXct` and `POSIXlt`

```{r}
now <- Sys.time(); now; class(now)
as.POSIXct('2024-01-15 14:30:00', tz = 'Asia/Taipei')
```

```{r}
# POSIXlt: list-based access to components
lt <- as.POSIXlt(now)
lt$hour; lt$min; lt$wday    # 0=Sunday
```

---

### `difftime`, Sequences, and Format Codes

```{r}
d1 <- as.Date('2024-01-01'); d2 <- as.Date('2024-12-31')
difftime(d2, d1, units = 'weeks')
seq(as.Date('2024-01-01'), as.Date('2024-06-01'), by = 'month')
```

\begin{table}[!htbp]
\centering\small
\begin{tabular}{llll}
\toprule
Code & Meaning & Code & Meaning \\
\midrule
\comm{\%Y} & 4-digit year & \comm{\%H} & hour (00--23) \\
\comm{\%m} & month (01--12) & \comm{\%M} & minute (00--59) \\
\comm{\%d} & day (01--31) & \comm{\%S} & second (00--59) \\
\comm{\%B}/\comm{\%b} & month name/abbrev & \comm{\%A}/\comm{\%a} & weekday name/abbrev \\
\bottomrule
\end{tabular}
\end{table}

---

### `strftime()` and `strptime()`

```{r}
strftime(Sys.time(), '%Y-%m-%d %H:%M:%S %Z')
strptime('15-Jan-2024 14:30', format = '%d-%b-%Y %H:%M')
```

# File I/O

### Reading and Writing Text Files

```{r, eval=FALSE}
df <- read.csv('data.csv')
df <- read.csv('data.csv', header = TRUE, stringsAsFactors = FALSE)
df <- read.delim('data.tsv')               # tab-separated
df <- read.table('data.txt', header = TRUE, sep = '|')
```

```{r, eval=FALSE}
write.csv(df, 'output.csv', row.names = FALSE)
write.table(df, 'output.tsv', sep = '\t',
            row.names = FALSE, quote = FALSE)
```

---

### Reading and Writing R Objects

```{r, eval=FALSE}
# save multiple R objects (binary, preserves types)
save(x, y, file = 'my_data.RData')
load('my_data.RData')

# save/load a single object
saveRDS(df, 'my_df.rds')
df <- readRDS('my_df.rds')
```

---

### `readLines`, `writeLines`, `scan`

```{r, eval=FALSE}
lines <- readLines('file.txt')
lines <- readLines('file.txt', n = 10)     # first 10 lines
writeLines(c('line 1', 'line 2'), 'out.txt')
nums  <- scan('numbers.txt')               # numeric
words <- scan('words.txt', what = '')       # character
```

---

### File System Functions

```{r, eval=FALSE}
getwd(); setwd('/path/to/dir')
list.files(); list.files(pattern = '\\.csv$')
file.exists('myfile.csv'); dir.exists('mydir')
file.info('myfile.csv')
file.copy('src.csv', 'dst.csv')
file.rename('old.csv', 'new.csv')
file.remove('temp.csv')
dir.create('new_folder')
normalizePath('relative/path')       # full absolute path
```

---

### Binary I/O and Downloads

```{r}
# practical: inspect raw type
x <- charToRaw("ABC"); x; typeof(x)
rawToChar(x); as.integer(x)     # see the byte values
```

```{r, eval=FALSE}
# write/read raw bytes
raw_data <- charToRaw("Hello, world!")
writeBin(raw_data, 'hello.bin')
dat <- readBin('hello.bin', what = 'raw', n = 100)
rawToChar(dat)
```

```{r, eval=FALSE}
# download files from the internet
download.file("https://example.com/data.csv", "data.csv")
unzip("archive.zip", exdir = "output_dir")
```

---

### `readline()` and `interactive()`

```{r, eval=FALSE}
name <- readline(prompt = "Enter your name: ")
cat("Hello,", name, "\n")
n <- as.numeric(readline("Enter a number: "))
```

```{r, eval=FALSE}
if (interactive()) {
  cat("Running interactively\n")
} else {
  cat("Running in batch mode\n")
}
```

# Error Handling

### `tryCatch()` --- Structured Error Handling

```{r}
# note: tryCatch catches the FIRST condition raised
result <- tryCatch({
  log('a')      # error: non-numeric argument
}, warning = function(w) {
  cat("Warning:", conditionMessage(w), "\n"); NA
}, error = function(e) {
  cat("Error:", conditionMessage(e), "\n"); NA
}, finally = {
  cat("(cleanup code runs always)\n")
})
result
```

---

### `try()` --- Simple Error Handling

```{r}
result <- try(log('a'), silent = TRUE)
inherits(result, 'try-error')
```

```{r}
# useful in loops: skip failing inputs
inputs <- list(1, 'a', 4, NULL, 9)
results <- list()
for (i in seq_along(inputs)) {
  results[[i]] <- try(sqrt(inputs[[i]]), silent = TRUE)
}
Filter(\(r) !inherits(r, 'try-error'), results)
```

---

### Common Error-Handling Idioms

```{r}
# silently return NULL on failure
safe_parse <- function(x) {
  tryCatch(as.numeric(x), warning = function(w) NULL,
                          error   = function(e) NULL)
}
safe_parse("3.14"); safe_parse("abc"); safe_parse(NULL)
```

```{r}
# wrap risky operations with a default value
safe_log <- function(x, default = NA) {
  tryCatch(log(x), error = function(e) default)
}
sapply(list(10, -1, 'a', NULL), safe_log)
```

---

### `stop()`, `warning()`, `message()`

```{r, error=TRUE}
safe_div <- function(a, b) {
  if (b == 0) stop("Division by zero!")
  a / b
}
safe_div(10, 2); safe_div(10, 0)
```

```{r}
check_val <- function(x) {
  if (x < 0) warning("Negative value!")
  sqrt(abs(x))
}
check_val(-4)
```

---

### `stopifnot()` --- Assertions

```{r, error=TRUE}
my_mean <- function(x) {
  stopifnot(is.numeric(x), length(x) > 0)
  sum(x) / length(x)
}
my_mean(1:5)
my_mean('hello')
```

# Useful Utilities

### `do.call()`

```{r}
dfs <- list(
  data.frame(x = 1:2, y = 3:4),
  data.frame(x = 5:6, y = 7:8),
  data.frame(x = 9:10, y = 11:12))
do.call(rbind, dfs)
```

```{r}
args <- list(1:10, na.rm = TRUE)
do.call(mean, args)
```

---

### `Vectorize()` --- Scalar to Vector

```{r}
f <- function(x, y) {
  if (x > y) x - y else x + y
}
f_vec <- Vectorize(f)
f_vec(1:3, 2)
```

---

### `table()` and `prop.table()`

```{r}
x <- c('A','B','A','C','B','A','C','C','B','A')
table(x); prop.table(table(x))
```

```{r}
y <- c(1,1,2,1,2,2,1,2,1,2)
addmargins(table(x, y))
```

---

### `match.arg()` and `missing()`

```{r}
my_plot <- function(type = c('scatter', 'line', 'bar')) {
  type <- match.arg(type)
  cat("Plotting:", type, "\n")
}
my_plot()         # default: first option
my_plot('l')      # partial matching
```

```{r}
f <- function(x, y) {
  cat("nargs:", nargs(), "\n")
  if (missing(y)) cat("y is missing\n")
}
f(1); f(1, 2)
```

---

### `invisible()` --- Silent Return

```{r}
# invisible: returns value without printing
f <- function(x) invisible(x^2)
f(5)                 # nothing printed
y_val <- f(5); y_val # but value is captured
```

```{r}
# common pattern: side-effect functions return invisible(NULL)
my_print <- function(x) {
  cat("Value:", x, "\n")
  invisible(NULL)
}
result <- my_print(42); is.null(result)
```

---

### `on.exit()` --- Guaranteed Cleanup

```{r, eval=FALSE}
safe_read <- function(path) {
  con <- file(path, 'r')
  on.exit(close(con))       # guaranteed cleanup
  readLines(con)
}
```

```{r, eval=FALSE}
# add = TRUE: stack multiple on.exit calls
process <- function(path) {
  tmp <- tempfile()
  on.exit(unlink(tmp), add = TRUE)      # cleanup 1
  con <- file(path, 'r')
  on.exit(close(con), add = TRUE)       # cleanup 2
}
```

---

### Recursive Functions and `Recall()`

```{r}
my_factorial <- function(n) {
  if (n <= 1) return(1)
  n * Recall(n - 1)
}
my_factorial(5)
```

```{r}
fib <- function(n) {
  if (n <= 2) return(1)
  Recall(n - 1) + Recall(n - 2)
}
sapply(1:10, fib)
```

---

### Local Helper Functions

```{r}
search_with_retry <- function(query, max_tries = 3) {
  is_valid <- function(result) {
    !is.null(result) && length(result) > 0
  }
  for (i in seq_len(max_tries)) {
    result <- tryCatch(grep(query, letters, value = TRUE),
                       error = function(e) NULL)
    if (is_valid(result)) return(result)
    Sys.sleep(0.1)
  }
  NULL
}
search_with_retry("^[a-c]")
```

---

### Naming Conventions and Dot-Prefix

```{r}
# names starting with . are "hidden" (not shown by ls())
e <- new.env()
e$.internal <- function(x) x + 1
e$public    <- function(x) x * 2
ls(e)                        # .internal not shown
ls(e, all.names = TRUE)      # now it's shown
```

- Convention: `.name` signals "internal/private"
- Many base R internals use this: `.Machine`, `.Platform`, `.GlobalEnv`

---

### `Sys.sleep()`, `Sys.info()`, `Sys.time()`

```{r, eval=FALSE}
Sys.sleep(1.5)              # pause execution for 1.5 seconds
Sys.time()                  # current date-time (POSIXct)
Sys.info()                  # OS name, machine, user, etc.
.Platform$OS.type           # "unix" or "windows"
.Machine$double.eps         # machine epsilon
```

---

### Benchmarking: `system.time()` and `proc.time()`

```{r}
system.time({
  x <- matrix(rnorm(1e6), 1000, 1000)
  y_mat <- crossprod(x)
})
```

```{r}
pt <- proc.time()
x <- sum(1:1e7)
proc.time() - pt
```

---

### `system2()` --- Run External Commands

```{r, eval=FALSE}
system2("echo", args = c("hello", "world"))
out <- system2("date", stdout = TRUE); cat(out, "\n")
system2("sleep", args = "5", wait = FALSE)     # background
```

```{r, eval=FALSE}
# cross-platform branching
if (.Platform$OS.type == "windows") {
  system2("taskkill", args = c("/F", "/IM", "myapp.exe"),
          stdout = NULL, stderr = NULL)
} else {
  system2("pkill", args = "myapp",
          stdout = NULL, stderr = NULL)
}
```

---

### `Encoding()` and `iconv()`

```{r, eval=FALSE}
Encoding(x)
iconv(x, from = 'UTF-8', to = 'BIG5')
iconv(x, from = 'BIG5', to = 'UTF-8')
Sys.setlocale('LC_ALL', 'zh_TW.UTF-8')
```

# Advanced Topics

### S3 Object System Basics

```{r}
person <- list(name = 'Alice', age = 30)
class(person) <- 'Person'
print.Person <- function(x, ...) {
  cat(sprintf("Person: %s (age %d)\n", x$name, x$age))
}
person
```

---

### Formula Objects

```{r}
f <- y ~ x1 + x2
class(f); typeof(f); all.vars(f)
```

```{r, eval=FALSE}
y ~ x                 # simple regression
y ~ x1 + x2           # multiple regression
y ~ x1 * x2           # with interaction (= x1 + x2 + x1:x2)
y ~ .                  # all other columns
y ~ poly(x, 3)        # polynomial
y ~ log(x)            # transformed
```

---

### Linear Regression Example

```{r, out.width="40%"}
model <- lm(dist ~ speed, data = cars)
model$coefficients
plot(cars$speed, cars$dist, pch = 19)
abline(model, col = 'red', lwd = 2)
```

---

### Random Number Generation

```{r}
set.seed(42)
runif(5); rnorm(5, mean = 0, sd = 1)
rbinom(10, size = 1, prob = 0.5)
rpois(10, lambda = 3)
rexp(5, rate = 1)
```

---

### `options()` and `getOption()`

```{r, eval=FALSE}
options(digits = 15)           # more decimal places
options(scipen = 999)          # avoid scientific notation
getOption('digits')
options(warn = 2)              # warnings as errors
```

---

### Package Management

```{r, eval=FALSE}
install.packages('dplyr')
library(dplyr)
require(dplyr)                  # returns TRUE/FALSE
requireNamespace('dplyr', quietly = TRUE)  # no attach
installed.packages()[, 'Package']
update.packages()
remove.packages('dplyr')
```

---

### Help System and Useful One-Liners

```{r, eval=FALSE}
?prod                            # help on a function
example(cumsum)                  # run examples
help.search('multivariate normal')
help(mgcv::rmvn)                # package::function help
```

```{r, eval=FALSE}
rm(list = ls())                 # clear workspace
suppressWarnings(as.integer(c('1', 'a', '3')))
sort(table(x), decreasing = TRUE)   # frequency count
source('my_script.R')
args <- commandArgs(trailingOnly = TRUE)
```
