---
title: "R Fundamentals"
---

```{r setup, include=FALSE}
source("_common.R")
```

# Vectors

## Vector Types and Scalars

- Flavors: **atomic** (all elements same type), **generic** (lists, elements can differ)
- Primary atomic types: `logical`, `integer`, `double`, `character`; integer + double = numeric
- Scalars are simply vectors of length 1

```{r}
#| output.lines: 3
v_logical   <- c(TRUE, FALSE)
v_integer   <- c(1L, 6L, 10L)
v_double    <- c(1, 2.5, 4.5)
v_character <- c('these', 'are', 'characters')
typeof(v_logical); typeof(v_integer); typeof(v_double); typeof(v_character)
```

```{r}
#| output.lines: 3
# concatenation of atomic vectors yields flat atomic vectors
(v <- c(c(1, 2), c(3, 4)))
```

---

## `NA` and `NULL`

- `NA` (Not Applicable): value **exists** but is **unknown**
- `NULL`: value **does not exist**

```{r}
#| output.lines: 3
NA > 5; 20 * NA; !NA             # NA is infectious!
NA ^ 0; NA | TRUE; NA & FALSE    # some exceptions
```

```{r}
#| output.lines: 3
x <- c(NA, 5, NA, 10)
x == NA          # WRONG way to test (always NA)
is.na(x)         # CORRECT way to test
```

```{r}
#| output.lines: 3
u <- NULL; length(u)   # 0
v <- NA;   length(v)   # 1
```

---

## Coercion Hierarchy

- Automatic coercion: **character > double > integer > logical**

```{r}
#| output.lines: 3
typeof(c('a', 1))      # character wins
c(1, FALSE)             # logical -> double
c(TRUE, 1L)             # logical -> integer
```

```{r}
#| output.lines: 3
# explicit coercion with as.*()
x <- c(TRUE, FALSE, TRUE, TRUE); as.numeric(x)
as.integer(c('1', '1.5', 'a'))   # 'a' -> NA with warning
```

---

## Type Testing and Coercion Functions

\begin{table}[!htbp]
\centering\small
\begin{tabular}{ll}
\toprule
Testing & Coercion \\
\midrule
\comm{is.logical(x)} & \comm{as.logical(x)} \\
\comm{is.integer(x)} & \comm{as.integer(x)} \\
\comm{is.double(x)} & \comm{as.double(x)} \\
\comm{is.numeric(x)} & \comm{as.numeric(x)} \\
\comm{is.character(x)} & \comm{as.character(x)} \\
\comm{is.na(x)}, \comm{is.null(x)} & --- \\
\comm{is.vector(x)} & \comm{as.vector(x)} \\
\comm{is.list(x)} & \comm{as.list(x)} \\
\comm{is.data.frame(x)} & \comm{as.data.frame(x)} \\
\comm{is.matrix(x)} & \comm{as.matrix(x)} \\
\comm{is.factor(x)} & \comm{as.factor(x)} \\
\bottomrule
\end{tabular}
\end{table}


## Type Testing and Coercion Functions Cont'd

```{r}
#| output.lines: 2
x <- c(3, 5, 7, 8)
is.integer(x); is.numeric(x)
as.integer(5.7); as.integer(-5.7)
```

## Sequences --- `:` and `seq()`

```{r}
#| output.lines: 3
5:10; 5:1
i <- 5; 1:i-1; 1:(i-1)   # note operator precedence!
```

```{r}
#| output.lines: 3
seq(from = 11, to = 30, by = 3)
seq(from = 1.1, to = 2, length = 10)
```

```{r}
#| output.lines: 3
# seq_along and seq_len: safer alternatives for loops
x <- c(9, 5, 17); seq_along(x)
x <- NULL; seq_along(x)   # integer(0), not 1:0 !
seq_len(5)
```

---

## Repetition: `rep()`

```{r}
#| output.lines: 3
rep(3, 5)
rep(c(10, 15, 3), 4)
rep(c(10, 15, 3), each = 4)   # interleave with 'each'
```

```{r}
#| output.lines: 3
rep(1:4, 1:4)                  # variable repetition counts
rep(c('C', 'G', 'H'), c(4, 5, 2))
```

---

## Other Generators: `sample()`, `sequence()`

```{r}
#| output.lines: 3
set.seed(42)
sample(1:6, 10, replace = TRUE)     # dice rolls
sample(letters[1:20], 5)            # no replacement
```

```{r}
#| output.lines: 3
sequence(c(6, 7, 4), from = c(8, 1, 9), by = c(-1, 1, -2))
```

## Numeric Indexing

```{r}
#| output.lines: 3
fib <- c(0, 1, 1, 2, 3, 5, 8, 13, 21, 34, 55, 89, 144)
fib[1]; fib[5]; fib[2:5]; fib[c(2, 4, 8, 1)]
```

```{r}
#| output.lines: 3
# 'minus' means 'exclude'; no mixing of positive/negative!
fib[-3]; fib[-(5:9)]
```

## Numeric Indexing Cont'd


```{r}
#| output.lines: 3
# negative indexing to build new vectors
x <- c(10, 20, -1, 3, 5, 6, 30, 9)
(y <- x[-2])                  # remove 2nd element
(z <- x[c(-2, -3, -6)])       # remove 2nd, 3rd, 6th
(w <- c(x, y, z))             # combine vectors
```

---

## Logical Indexing

```{r}
#| output.lines: 3
fib[fib < 10]                           # elements < 10
fib[fib %% 2 == 0 | fib %% 3 == 0]     # multiples of 2 or 3
fib[c(TRUE, FALSE)]                      # odd-indexed (recycling!)
```

## Logical Indexing Cont'd


```{r}
#| output.lines: 3
v <- c(2, 8, 1, 7, -11, 6, 20, 3, 12, 35, 21, 3, 5, 7, -2)
v[v > median(v)]                             # above median
v[v < quantile(v, 0.05) | v > quantile(v, 0.95)]  # extreme 5%
v[abs(v - mean(v)) > sd(v)]                  # beyond 1 SD
```

---

## `which()`, `which.min()`, `which.max()`

```{r}
#| output.lines: 3
which(Nile > 1200)            # positions where TRUE
Nile[Nile > 1200]             # the values themselves
```

```{r}
#| output.lines: 3
x <- c(5, 3, 8, 1, 9, 2)
which.min(x); which.max(x)   # index of first min/max
```

---

## Removing `NA` from Vectors

```{r}
#| output.lines: 3
v <- c(21, 2, 3, NA, 5, NULL, NA, 32, 55)
v[!is.na(v)]                  # NULL is already dropped by c()
```

```{r}
#| output.lines: 3
na.omit(v)                    # alternative; returns with attr
```

## Recycling Rule

- In operations involving two vectors of different lengths, R **repeats** the shorter one:

```{r}
#| output.lines: 3
c(2, 4, 3) + c(6, 10, -1, 3, 5)
```

```{r}
#| output.lines: 3
# useful applications
x <- 1:5
x + 10           # add a constant (scalar = length-1 vector)
x * c(1, -1)     # alternate signs
```

---

## `all()`, `any()`, and Vector Equality

```{r}
#| output.lines: 3
x <- 1:100; any(x > 200); all(x > 0)
```

```{r}
#| output.lines: 3
# == is element-wise; don't use for whole-vector equality!
x <- 1:3; y0 <- c(1, 2, 3); y1 <- c(1, 3, 4)
x == y0; x == y1
```

---

## `identical()` and `all.equal()`

```{r}
#| output.lines: 3
x <- 1:3; y0 <- c(1, 2, 3)
identical(x, y0)          # FALSE: integer vs double!
typeof(x); typeof(y0)
```

```{r}
#| output.lines: 3
all.equal(x, y0)          # TRUE (with tolerance)
all.equal(x, c(1, 3, 4)) # shows difference
```

---

## Named Vectors

```{r}
#| output.lines: 3
# 3 ways to name a vector
x <- c(a = 1, b = 2, c = 3)
x <- 1:3; names(x) <- c('a', 'b', 'c')
x <- setNames(1:3, c('a', 'b', 'c'))
```

```{r}
#| output.lines: 3
# access by name
x['b']; x[c('a', 'c')]
```

---

## Factors

- Factors represent **categorical variables** with a fixed set of levels.

:::::: columns
::: column

```{r}
#| output.lines: 3
(x <- factor(c('a', 'b', 'b', 'a')))
typeof(x); levels(x)
```

:::
::: column

```{r}
#| output.lines: 3
# ordered factors
(size <- factor(c('S','M','L','M','S'),
    levels=c('S','M','L'), ordered=TRUE))
size > 'S'
```

:::
::::::

## Ex 1: Logical Indexing

Let `v <- c(14, -3, 22, 7, -8, 0, 31, 5, -1, 18)`.

- Extract all elements strictly between 5 and 25
- Replace every negative element with `0` in a copy `w`
- How many elements are negative or exceed 20?
- Find indices where $|v_i|$ exceeds its rank


## Sol 1

```{r}
#| output.lines: 3
v <- c(14, -3, 22, 7, -8, 0, 31, 5, -1, 18)
v[v > 5 & v < 25]
w <- v; w[w < 0] <- 0; w
sum(v < 0 | v > 20)
which(abs(v) > rank(v))
```


## Ex 2: Named Vectors

Create `stock <- c(AAPL=189.5, MSFT=415.2, GOOG=172.3, AMZN=185.7)`.

- Access `GOOG` by name; add `NVDA = 875.4`
- Compute percentage change from base 150; round to 1 dp
- Return stocks above mean price, sorted descending
- Rename all tickers to lowercase


## Sol 2

```{r}
#| output.lines: 3
stock <- c(AAPL=189.5, MSFT=415.2, GOOG=172.3, AMZN=185.7)
stock["GOOG"]; stock["NVDA"] <- 875.4
round((stock - 150) / 150 * 100, 1)
sort(stock[stock > mean(stock)], decreasing=TRUE)
names(stock) <- tolower(names(stock)); names(stock)
```


## Ex 3: `which`, `which.min`, `which.max`

Let `x <- c(3, 7, 2, 9, 1, 5, 8, 4, 6, 10)`.

- Indices of elements > 6
- Index of min and max
- Second-largest value (no `sort`)
- Local maxima: elements greater than both immediate neighbours (ignore endpoints)


## Sol 3

```{r}
#| output.lines: 3
x <- c(3, 7, 2, 9, 1, 5, 8, 4, 6, 10)
which(x > 6)
which.min(x); which.max(x)
max(x[x != max(x)])
n <- length(x); idx <- 2:(n-1)
x[idx][x[idx] > x[idx-1] & x[idx] > x[idx+1]]
```


## Ex 4: Cumulative and Parallel Functions

Let `x <- c(2,5,3,8,1,7,4)` and `y <- c(6,2,9,1,4,3,8)`.

- Running maximum of `x`
- Running sum and product
- Element-wise min and max
- Verify $\sum\text{pmin}(x,y) + \sum\text{pmax}(x,y) = \sum x + \sum y$


## Sol 4

```{r}
#| output.lines: 3
x <- c(2,5,3,8,1,7,4); y <- c(6,2,9,1,4,3,8)
cummax(x)
cumsum(x); cumprod(x)
pmin(x,y); pmax(x,y)
sum(pmin(x,y)) + sum(pmax(x,y)) == sum(x) + sum(y)
```


## Ex 5: Set Operations

Let `a <- c(1,3,5,7,9,3,5)` and `b <- c(2,3,5,6,7,2)`.

- Unique elements of `a`
- Positions of duplicated values in `a`
- $a \cup b$, $a \cap b$, $a \setminus b$
- `match(c(3,5,4), b)` — explain the `NA`


## Sol 5

```{r}
#| output.lines: 3
a <- c(1,3,5,7,9,3,5); b <- c(2,3,5,6,7,2)
unique(a)
which(duplicated(a))
union(a,b); intersect(a,b); setdiff(a,b)
match(c(3,5,4), b)  # 4 not found -> NA
```


## Ex 6: Recycling Rule

- What does `c(1,2,3,4,5,6) + c(10,20)` return? Why?
- Let `m <- matrix(1:12, nrow=3)`. What does `m + c(100,200,300)` do?
- Extract odd-indexed and even-indexed elements of `1:12` using `c(TRUE,FALSE)`
- Generate `1 0 1 0 1 0 1 0 1 0` using `rep`


## Sol 6

```{r}
#| output.lines: 3
c(1,2,3,4,5,6) + c(10,20)                         # recycled
m <- matrix(1:12, nrow=3)
m + c(100,200,300)    # adds per column (col-major)
x <- 1:12; x[c(T,F)]; x[c(F,T)]
rep(c(1,0), 5)
```


## Ex 7: Sorting and Ranking

Let `x <- c(40,15,30,10,50,25)`.

- Sort ascending and descending
- Find `order(x)` and verify `x[order(x)] == sort(x)`
- What rank does `30` receive?
- Sort `c("banana","apple","cherry")` alphabetically


## Sol 7

```{r}
#| output.lines: 3
x <- c(40,15,30,10,50,25)
sort(x); sort(x, decreasing=TRUE)
ord <- order(x); all(x[ord] == sort(x))
rank(x)  # 30 -> rank 4
sort(c("banana","apple","cherry"))
```


## Ex 8: `NA`, `NULL`, `NaN`, `Inf`

- Given `x <- c(3, NA, -1, NaN, Inf, -Inf, 0)`, apply `is.na`, `is.nan`, `is.finite`, `is.infinite`. Why is `is.na(NaN)` TRUE?
- Remove non-finite values and compute mean
- Show `c(1, NULL, 3)` vs `c(1, NA, 3)` lengths
- Compute `NA^0`, `NA | TRUE`, `NA & FALSE`


## Sol 8

```{r}
#| output.lines: 3
x <- c(3, NA, -1, NaN, Inf, -Inf, 0)
is.na(x); is.nan(x); is.finite(x)
mean(x[is.finite(x)])
length(c(1, NULL, 3)); length(c(1, NA, 3))        # 2 vs 3
NA^0; NA | TRUE; NA & FALSE
```


## Ex 9: Type Coercion

- What does `c(TRUE, 1L, 2.5, "x")` produce? What type?
- Coerce `c("3","1.5","NA","7")` to numeric. What happens to `"NA"`?
- Predict `as.integer(c(3.9, -2.7, TRUE, "5"))`
- Why does `1 == "1"` return `TRUE`?


## Sol 9

```{r}
#| output.lines: 3
x <- c(TRUE, 1L, 2.5, "x"); x; typeof(x)         # character
as.numeric(c("3","1.5","NA","7"))                  # NA + warning
as.integer(c(3.9, -2.7, TRUE, "5"))               # 3,-2,1,5
1 == "1"  # 1 coerced to "1", then string compare
```


## Ex 10: `seq` and `seq_along`

- Generate `1.0 2.5 4.0 5.5 7.0 8.5 10.0` using `seq`
- Show `1:length(x)` is unsafe when `x <- NULL`; fix with `seq_along`
- 6 equally-spaced values from 0 to $2\pi$
- What does `seq_along(integer(0))` return?


## Sol 10

```{r}
#| output.lines: 3
seq(1, 10, by=1.5)
x <- NULL; 1:length(x)                            # c(1,0) WRONG
seq_along(x)                                       # integer(0) SAFE
seq(0, 2*pi, length.out=6)
seq_along(integer(0))                              # integer(0)
```


## Ex 11: `rep` Patterns

Using `rep`, generate:

- `1 1 2 2 3 3 4 4 5 5`
- `1 2 3 1 2 3 1 2 3 1 2 3`
- `"A" "A" "A" "B" "B" "C"`
- Alternating `TRUE FALSE` of length 20


## Sol 11

```{r}
#| output.lines: 3
rep(1:5, each=2)
rep(1:3, times=4)
rep(c("A","B","C"), c(3,2,1))
rep(c(TRUE,FALSE), 10)
```


## Ex 12: `%in%` and `match`

`haystack <- c(3,1,4,1,5,9,2,6,5,3)`; `needles <- c(5,7,2)`.

- Which needles are present?
- First occurrence position of each
- Difference between `match("a", c("b","a","a"))` and `which(c("b","a","a") == "a")`
- Filter `mtcars` rows where `cyl %in% c(4,6)`


## Sol 12

```{r}
#| output.lines: 3
haystack <- c(3,1,4,1,5,9,2,6,5,3); needles <- c(5,7,2)
needles %in% haystack
match(needles, haystack)                           # NA for 7
match("a", c("b","a","a"))                         # 2 (first)
which(c("b","a","a") == "a")                       # c(2,3) (all)
nrow(mtcars[mtcars$cyl %in% c(4,6), ])
```


## Ex 13: Vector Arithmetic

- Compute $H_{100} = \sum_{k=1}^{100}\frac{1}{k}$ in one line
- Compute $\sum_{k=0}^{50}\frac{(-1)^k}{k!}$. Compare with `exp(-1)`
- For `x <- seq(0.1,2,by=0.1)`, compute $e^x\sin x - \ln x$, rounded to 4 dp
- Compute `diff(cumsum(1:10))` and explain


## Sol 13

```{r}
#| output.lines: 3
sum(1 / 1:100)                                     # H_100
k <- 0:50; sum((-1)^k / factorial(k)); exp(-1)
x <- seq(0.1, 2, by=0.1)
round(exp(x)*sin(x) - log(x), 4)
diff(cumsum(1:10))  # recovers 2:10
```


## Ex 14: Attributes and `structure`

- Create `a <- 1:4`; attach attributes `"unit"="kg"` and `"source"="lab"`. Print
- Recreate using `structure()` in one call
- `x <- 1:3` vs `y <- c(1,2,3)`: are they `identical`? `all.equal`? Why?
- When to use `isTRUE(all.equal())` in an `if`?


## Sol 14

```{r}
#| output.lines: 3
a <- 1:4; attr(a,"unit") <- "kg"; attr(a,"source") <- "lab"
attributes(a)
b <- structure(1:4, unit="kg", source="lab")
x <- 1:3; y <- c(1,2,3)
identical(x, y); all.equal(x, y)                   # F; T
isTRUE(all.equal(0.1+0.2, 0.3))                   # TRUE
```


## Ex 15: Factors

- Create factor `f <- factor(c("low","med","high","med","low"))`. What are the levels?
- Create an ordered factor with levels `low < med < high`. Test `f > "low"`
- Use `table(f)` to count occurrences
- What does `as.integer(f)` return and why?


## Sol 15

```{r}
#| output.lines: 3
f <- factor(c("low","med","high","med","low"))
levels(f)                                          # alphabetical
fo <- factor(f, levels=c("low","med","high"), ordered=TRUE)
fo > "low"
table(f)
as.integer(f)  # underlying integer codes
```


## Ex 16: Floating Point

- Is `0.1 + 0.2 == 0.3` TRUE or FALSE? Why?
- Use `all.equal` to compare correctly
- What is `trunc(-3.7)` vs `floor(-3.7)`?
- Demonstrate banker's rounding: `round(0.5)`, `round(1.5)`, `round(2.5)`


## Sol 16

```{r}
#| output.lines: 3
0.1 + 0.2 == 0.3  # FALSE (floating point)
all.equal(0.1 + 0.2, 0.3)                         # TRUE
trunc(-3.7); floor(-3.7)  # -3 vs -4
round(0.5); round(1.5); round(2.5)  # 0; 2; 2
```


## Ex 17: Appending and Inserting

- Start with `v <- 1:5`. Append `c(10,11)` to the end
- Insert `99` after position 3 using `append`
- What happens if you assign `v[10] <- 42` when `length(v)` is 7?
- Remove the 3rd element without using negative indexing (use logical)


## Sol 17

```{r}
#| output.lines: 3
v <- 1:5; v <- c(v, c(10,11)); v
append(v, 99, after=3)
v[10] <- 42; v   # NAs fill gaps
v[seq_along(v) != 3]
```


## Ex 18: `sample` and `sequence`

- Simulate 20 dice rolls with `sample`
- Use `sample` to shuffle `letters[1:10]`
- Using `sequence`, generate `1 2 3 4  1 2 3  1 2  1`
- Generate `5 4 3 2 1  6 5 4 3 2 1` using `sequence`


## Sol 18

```{r}
#| output.lines: 3
set.seed(42)
sample(1:6, 20, replace=TRUE)
sample(letters[1:10])
sequence(4:1)
sequence(5:6, from=5:6, by=-1)
```


# Operators

## Operator Table

\begin{table}[!htbp]
\centering\small
\begin{tabular}{ll}
\toprule
Operators & Definition \\
\midrule
\comm{+ - * /} & plus, minus, times, divide\\
\comm{\%/\% \%\% \textasciicircum} & integer quotient, modulo, power \\
\comm{> >= < <=} & greater, greater-or-equal, less, less-or-equal\\
\comm{== !=} & equals, not equals\\
\comm{! \& |} & not, and (elementwise), or (elementwise)\\
\comm{<- ->} & assignment\\
\comm{\$} & list/data.frame element access\\
\bottomrule
\end{tabular}
\end{table}

## Operator Table Cont'd

\begin{table}[!htbp]
\centering\small
\begin{tabular}{ll}
\toprule
Operators & Definition \\
\midrule
\comm{:} & sequence creation\\
\comm{\textasciitilde} & model formulae\\
\midrule
\comm{\%*\%} & matrix product \\
\comm{\%in\%} & test matching \\
\comm{\&\& ||} & short-circuit and, or (scalar only)\\
\comm{::} & namespace access \\
\comm{xor(a, b)} & elementwise exclusive OR \\
\comm{<<-} & global / parent-env assignment \\
\comm{|>} & native pipe (R 4.1+) \\
\bottomrule
\end{tabular}
\end{table}

---

## Operator Precedence (High to Low)

\begin{table}[!htbp]
\centering\small
\begin{tabular}{ll}
\toprule
Precedence & Operators \\
\midrule
1 (highest) & \comm{::}, \comm{:::} \\
2 & \comm{\$}, \comm{@} \\
3 & \comm{[}, \comm{[[} \\
4 & \comm{\textasciicircum} (right-associative) \\
5 & unary \comm{-}, \comm{+} \\
6 & \comm{:} \\
\bottomrule
\end{tabular}
\end{table}

## Operator Precedence (High to Low) Cont'd

\begin{table}[!htbp]
\centering\small
\begin{tabular}{ll}
\toprule
Precedence & Operators \\
\midrule
7 & \comm{\%any\%} (including \comm{\%*\%}, \comm{\%/\%}, \comm{\%\%}, \comm{\%in\%}) \\
8 & \comm{*}, \comm{/} \\
9 & \comm{+}, \comm{-} \\
10 & \comm{<}, \comm{>}, \comm{<=}, \comm{>=}, \comm{==}, \comm{!=} \\
11 & \comm{!} \\
12 & \comm{\&}, \comm{\&\&} \\
13 & \comm{|}, \comm{||} \\
14 & \comm{|>} \\
15 & \comm{\textasciitilde} \\
16 (lowest) & \comm{<-}, \comm{<<-}, \comm{->}, \comm{->>} \\
\bottomrule
\end{tabular}
\end{table}

---

## `%in%` and `match()`

```{r}
#| output.lines: 3
x <- c(1, 5, 3, 8, 10)
x %in% c(3, 5, 7)      # which elements of x are in the set?
```

```{r}
#| output.lines: 3
# match() returns the position of first match (NA if not found)
match(c(3, 5, 99), x)
```

## `%in%` and `match()` Cont'd


```{r}
#| output.lines: 3
# practical: filter data frame rows by a set of values
tg <- ToothGrowth
tg[tg$dose %in% c(0.5, 2), ][1:6, ]
```

# Common Functions

## Math Functions

\begin{table}[!htbp]
\centering\small
\begin{tabular}{ll}
\toprule
Function & Definition \\
\midrule
\comm{abs(x)} & absolute value \\
\comm{floor(x)}, \comm{ceiling(x)}, \comm{trunc(x)} & rounding toward $-\infty$, $+\infty$, 0\\
\comm{round(x, digits=0)} & round (banker's rounding)\\
\comm{signif(x, digits=6)} & significant digits\\
\comm{sqrt(x)}, \comm{exp(x)} & $\sqrt{x}$, $e^x$ \\
\comm{log(x)}, \comm{log10(x)}, \comm{log2(x)} & natural, base-10, base-2 log\\
\comm{factorial(x)}, \comm{choose(n, m)} & $x!$, $\binom{n}{m}$ \\
\comm{sin, cos, tan, asin, acos, atan} & trig functions\\
\comm{beta(a, b)}, \comm{gamma(x)} & special functions\\
\bottomrule
\end{tabular}
\end{table}

---

## Examples: Math Functions

```{r}
#| output.lines: 2
abs(-1); floor(5.7); ceiling(-5.7); trunc(-5.7)
round(5.7); round(5.5); round(2.5)  # banker's rounding!
signif(12345678, 4)
```

```{r}
#| output.lines: 2
sin(pi / 2); cos(pi / 2)  # cos not exactly 0 (floating point)
119 %/% 13; 119 %% 13
```

---

## Statistical Functions

\begin{table}[!htbp]
\centering\small
\begin{tabular}{ll}
\toprule
Function & Definition \\
\midrule
\comm{max(x)}, \comm{min(x)}, \comm{range(x)} & max, min, c(min, max)\\
\comm{sum(x)}, \comm{prod(x)} & sum, product\\
\comm{mean(x)}, \comm{median(x)} & mean, median\\
\comm{quantile(x, p)} & quantile at probability \comm{p}\\
\comm{var(x)}, \comm{sd(x)} & sample variance, std deviation\\
\comm{cor(x, y)}, \comm{cov(x, y)} & correlation, covariance\\
\comm{sort(x)}, \comm{rev(x)} & sorted copy, reversed copy\\
\comm{rank(x)}, \comm{order(x)} & ranks, permutation to sort\\
\midrule
\comm{cumsum}, \comm{cumprod}, \comm{cummax}, \comm{cummin} & cumulative operations\\
\comm{diff(x)} & lagged differences\\
\comm{pmax(x,y)}, \comm{pmin(x,y)} & parallel (elementwise) max/min\\
\comm{table(x)} & frequency table\\
\comm{colMeans}, \comm{colSums} & column means/sums\\
\comm{rowMeans}, \comm{rowSums} & row means/sums\\
\bottomrule
\end{tabular}
\end{table}

---

## Floating Point Surprises

```{r}
#| output.lines: 3
0.1 + 0.2 == 0.3              # FALSE!
all.equal(0.1 + 0.2, 0.3)     # TRUE (with tolerance)
isTRUE(all.equal(0.1 + 0.2, 0.3))  # safe for if()
```

```{r}
#| output.lines: 3
# trunc vs floor for negative numbers
trunc(-3.7); floor(-3.7)     # toward 0 vs toward -Inf
```

## Ex 19: Operator Precedence

- Without running, predict: `2 + 3 * 4`, `2^3^2`, `1:3 + 1`, `1:(3+1)`
- Why does `-2^2` give `-4` instead of `4`?
- Predict: `TRUE | FALSE & FALSE`
- Write `x <- 5; x %% 3 == 2 & x > 3` and trace operator precedence


## Sol 19

```{r}
#| output.lines: 3
2 + 3 * 4     # 14: * before +
2^3^2          # 512: ^ right-associative = 2^9
1:3 + 1        # 2 3 4: : before +
1:(3+1)        # 1 2 3 4
-2^2           # -4: ^ before unary -
TRUE | FALSE & FALSE  # TRUE: & before |
x <- 5; x %% 3 == 2 & x > 3  # TRUE
```


## Ex 20: Math Functions

- Compute `floor`, `ceiling`, `trunc`, `round` of `-5.7`
- Compute $\binom{10}{3}$ and verify with `factorial`
- Compute `119 %/% 13` and `119 %% 13`. Verify `119 == 13 * (119 %/% 13) + 119 %% 13`
- `signif(12345678, 4)`


## Sol 20

```{r}
#| output.lines: 3
floor(-5.7); ceiling(-5.7); trunc(-5.7); round(-5.7)
choose(10, 3); factorial(10)/(factorial(3)*factorial(7))
```


## Sol 20 Cont'd

```{r}
#| output.lines: 3
119 %/% 13; 119 %% 13
119 == 13 * (119 %/% 13) + 119 %% 13              # TRUE
signif(12345678, 4)
```

## Ex 21: Statistical Summary

`v <- c(4, 7, 2, 9, 1, 5, 8, 3, 6, 10)`.

- mean, median, var, sd, range
- `quantile(v, c(0.25, 0.75))` and IQR
- Compute z-scores: $(v - \bar v)/s$. How many exceed $|z| > 1$?
- Use `diff(v)` to find the largest jump


## Sol 21

```{r}
#| output.lines: 3
v <- c(4, 7, 2, 9, 1, 5, 8, 3, 6, 10)
mean(v); median(v); var(v); sd(v); range(v)
```


## Sol 21 Cont'd

```{r, echo=FALSE}
v <- c(4, 7, 2, 9, 1, 5, 8, 3, 6, 10)
```

```{r}
#| output.lines: 3
quantile(v, c(0.25, 0.75)); IQR(v)
z <- (v - mean(v)) / sd(v); sum(abs(z) > 1)
max(abs(diff(v)))                                  # largest jump
```

## Ex 22: `table` and `prop.table`

`x <- c("A","B","A","C","B","A","C","C","B","A")`.

- Frequency table of `x`
- Proportion table
- Create `y <- c(1,1,2,1,2,2,1,2,1,2)`. Cross-tabulate `x` and `y`; add margins
- Which category of `x` is most frequent?


## Sol 22

```{r}
#| output.lines: 3
x <- c("A","B","A","C","B","A","C","C","B","A")
table(x)
prop.table(table(x))
```


## Sol 22 Cont'd

```{r, echo=FALSE}
x <- c("A","B","A","C","B","A","C","C","B","A")
```

```{r}
#| output.lines: 3
y <- c(1,1,2,1,2,2,1,2,1,2)
addmargins(table(x, y))
names(which.max(table(x)))                         # "A"
```

## Ex 23: Harmonic Series Convergence

- Compute $H_n = \sum_{k=1}^{n}\frac{1}{k}$ for $n \in \{10, 100, 1000, 10000\}$
- Compare each $H_n$ with $\ln(n) + \gamma$ where $\gamma \approx 0.5772$
- Plot $H_n - \ln n$ for $n = 1, \ldots, 200$


## Sol 23

```{r, out.width="45%"}
#| output.lines: 1
ns <- c(10, 100, 1000, 10000)
Hn <- sapply(ns, function(n) sum(1/1:n))
data.frame(n=ns, Hn=round(Hn,4),
           approx=round(log(ns)+0.5772,4))
x <- 1:200; y <- cumsum(1/x) - log(x)
plot(x, y, type="l", ylab=expression(H[n]-ln(n)))
abline(h=0.5772, col="red", lty=2)
```


## Ex 24: `pmin`/`pmax` Application

- Clip vector `v <- c(-5,3,12,-2,8,25)` to range $[0, 10]$ using `pmin` and `pmax`
- Given two portfolios `a <- c(5,8,3,9)` and `b <- c(7,2,6,4)`, compute the "best of both" portfolio
- Compute $\sum_{k=1}^{100}\min(2^k, k^4)$ without loops


## Sol 24

```{r}
#| output.lines: 3
v <- c(-5, 3, 12, -2, 8, 25)
pmin(pmax(v, 0), 10)                               # clip to [0,10]
a <- c(5,8,3,9); b <- c(7,2,6,4)
pmax(a, b)
k <- 1:100; sum(pmin(2^k, k^4))
```


## Ex 25: Cumulative Functions

`x <- c(3, -1, 4, -1, 5, -9, 2, 6)`.

- Compute `cumsum(x)`. At which index does it first become negative?
- Maximum drawdown: $\max_i(\text{cummax}_i - \text{cumsum}_i)$
- Compute `cumprod(1 + c(0.05, -0.02, 0.03, -0.01))` (portfolio growth)


## Sol 25

```{r}
#| output.lines: 3
x <- c(3, -1, 4, -1, 5, -9, 2, 6)
cs <- cumsum(x); cs
which(cs < 0)[1]                                   # first negative
cm <- cummax(cs); max(cm - cs)                     # max drawdown
cumprod(1 + c(0.05, -0.02, 0.03, -0.01))          # growth
```


## Ex 26: `order` and Data Sorting

`d <- data.frame(name=c("Zara","Amy","Mike","Bob"), score=c(88,95,72,95), age=c(22,20,25,21))`.

- Sort by score ascending
- Sort by score descending, ties broken by age ascending
- Find the name of the person with the highest score (youngest if tied)


## Sol 26

```{r}
#| output.lines: 3
d <- data.frame(name=c("Zara","Amy","Mike","Bob"),
                score=c(88,95,72,95), age=c(22,20,25,21))
d[order(d$score), ]
d[order(-d$score, d$age), ]
d[order(-d$score, d$age), ]$name[1]                # Bob
```


# Data Frames

## What Is a Data Frame?

- A rectangular table: one row per observation, one column per variable.
- **Technically**: a named list of equal-length vectors (we'll revisit this in Lists).

```{r}
#| output.lines: 3
tg <- ToothGrowth    # built-in data; used throughout
head(tg, 4)
```

```{r}
#| output.lines: 3
# Each column is a vector; use $ to extract
tg$len[1:5]
```

---

## Data Frame Inspection

```{r}
#| output.lines: 3
str(tg)
```

```{r}
#| output.lines: 3
nrow(tg); ncol(tg); dim(tg); names(tg)
```

---

## `summary()` --- Quick Statistics

```{r}
#| output.lines: 3
summary(tg)
```

---

## Indexing: `[row, col]`

```{r}
#| output.lines: 3
tg[3, 1]              # row 3, column 1
tg$len[3]             # equivalent via $
(z <- tg[2:4, c(1, 3)])   # rows 2-4, columns 1 & 3
```

## Indexing: `[row, col]` Cont'd


```{r}
#| output.lines: 3
head(tg[, c(1, 3)], 3)   # all rows, columns 1 & 3
head(tg[, -2], 3)         # all rows, remove column 2
```

---

## Creating Data Frames

:::::: columns
::: column

```{r}
#| output.lines: 3
age <- c(55, 58, 45)
name <- c('Alice', 'Bill', 'Cathy')
(d <- data.frame(age, name))
```

- Vectors of different lengths are not allowed!

```{r, error=TRUE}
#| output.lines: 3
blood <- c('O', 'B', 'A', 'AB')
data.frame(age, name, blood)
```

:::
::: column

```{r, eval=FALSE}
# with custom column names
data.frame(p1 = pred1, p2 = pred2)

# from a named list of vectors
l <- list(p1 = pred1, p2 = pred2)
as.data.frame(l)
```

```{r, eval=FALSE}
# from row data via rbind
r1 <- data.frame(a=1, b=2, c="X")
r2 <- data.frame(a=3, b=4, c="Y")
rbind(r1, r2)
```

:::
::::::

## Row Filtering with Logical Conditions

```{r}
#| output.lines: 3
tg_vc <- tg[tg$supp == 'VC', ]
tg_oj <- tg[tg$supp == 'OJ', ]
mean(tg_vc$len); mean(tg_oj$len)
```

## Row Filtering with Logical Conditions Cont'd


```{r}
#| output.lines: 3
# AND condition
tg[tg$supp == 'OJ' & tg$len < 8.8, ]
```

```{r}
#| output.lines: 3
# OR condition
head(tg[tg$len > 28 | tg$dose == 1.0, ], 6)
```

- Note the comma (`,`) and the use of `==`, not `=`!

---

## `subset()` --- Cleaner Filtering

```{r}
#| output.lines: 3
# equivalent to tg[tg$supp == 'OJ' & tg$len > 20, ]
subset(tg, supp == 'OJ' & len > 20)
```

```{r}
#| output.lines: 3
# select specific columns
subset(tg, dose == 2, select = c(len, supp))
```

---

## `with()` and `within()`

```{r}
#| output.lines: 3
# with() evaluates an expression in the data frame environment
with(tg, mean(len[supp == 'VC']))
```

```{r}
#| output.lines: 3
# within() modifies and returns the data frame
tg2 <- within(tg, {
  len_cm  <- len / 10
  log_len <- log(len)
})
head(tg2, 3)
```

## Adding, Removing, Renaming Columns

```{r}
#| output.lines: 3
d <- data.frame(x = 1:3, y = c(10, 20, 30))
d$z <- d$x + d$y              # add column with $
d[['w']] <- c('a', 'b', 'c') # add column with [[]]
d
```

```{r}
#| output.lines: 3
d$z <- NULL                   # remove column
names(d)[names(d) == 'w'] <- 'label'   # rename
d
```

---

## Sorting Data Frames

```{r}
#| output.lines: 3
d <- data.frame(name = c('Zara', 'Amy', 'Mike'),
                score = c(88, 95, 72))
d[order(d$score), ]                        # ascending
d[order(d$score, decreasing = TRUE), ]     # descending
```

## Sorting Data Frames Cont'd


```{r}
#| output.lines: 3
# sort by multiple columns
d2 <- data.frame(g = c('A','B','A','B'), x = c(3,1,2,4))
d2[order(d2$g, d2$x), ]
```

---

## Handling NAs

```{r}
#| output.lines: 3
df <- data.frame(x = c(1, NA, 3, 4, 5), y = c(1, 2, NA, 4, 5))
na.omit(df)                      # remove rows with any NA
```

## Handling NAs Cont'd


```{r}
#| output.lines: 3
complete.cases(df)                # logical: which rows complete?
df[complete.cases(df), ]
```

```{r, eval=FALSE}
df$x[is.na(df$x)] <- 0           # replace specific column NAs
df[is.na(df)] <- 0               # replace all NAs
```

---

## Combining Data Frames: `cbind`, `rbind`

```{r}
#| output.lines: 3
# column-bind (side by side) --- same number of rows
df1 <- data.frame(a = c(1, 2)); df2 <- data.frame(b = c(7, 8))
cbind(df1, df2)
```

```{r}
#| output.lines: 3
# row-bind (stack) --- same column names
df1 <- data.frame(x = c("a", "a"), y = c(5, 6))
df2 <- data.frame(x = c("b", "b"), y = c(9, 10))
rbind(df1, df2)
```

## The Split--Apply--Combine Paradigm

\begin{table}[!htbp]
\centering\small
\begin{tabular}{lll}
\toprule
Step & Base R tool & What it does \\
\midrule
Split & \comm{split()} & partition data by factor $\to$ list \\
Apply & \comm{lapply/sapply/tapply/\ldots} & run function on each piece \\
Combine & \comm{unsplit/rbind/do.call/\ldots} & reassemble results \\
\midrule
All-in-one & \comm{aggregate()} & split + apply + combine in one call \\
Horizontal combine & \comm{merge()} & join two tables by key \\
\bottomrule
\end{tabular}
\end{table}

---

## `split()` --- Partition Data

```{r}
#| output.lines: 3
scores <- c(88, 92, 78, 95, 85, 70, 91, 83)
group  <- c('A','B','A','B','A','B','A','B')
(sg <- split(scores, group))
sapply(sg, mean)         # per-group mean
```

## `split()` --- Partition Data Cont'd


```{r}
#| output.lines: 3
# split a data frame --- each element is a complete data frame
tg_split <- split(tg, tg$supp)
head(tg_split$OJ, 3)
```

---

## `split()` --- Multi-Factor and `unsplit()`

```{r}
#| output.lines: 3
tg_split2 <- split(tg, list(tg$supp, tg$dose))
names(tg_split2)
```

```{r}
#| output.lines: 3
# unsplit: put pieces back in their original positions
sg <- split(1:8, rep(c('A','B'), 4))
sg$A <- sg$A * 10; sg$B <- sg$B * (-1)
unsplit(sg, rep(c('A','B'), 4))
```

---

## The `split` + `lapply` + `rbind` Pipeline

```{r}
#| output.lines: 3
# add a z-score column within each group
standardize <- function(df) {
  df$len_z <- (df$len - mean(df$len)) / sd(df$len)
  df
}
parts <- split(tg, tg$supp)
parts <- lapply(parts, standardize)
tg2 <- do.call(rbind, parts)
head(tg2[order(as.numeric(rownames(tg2))), ], 4)
```

---

## `aggregate()` --- Group-wise Summaries

```{r}
#| output.lines: 3
# formula syntax: response ~ grouping
aggregate(len ~ supp, data = tg, FUN = mean)
aggregate(len ~ supp + dose, data = tg, FUN = mean)
```

---

## `aggregate()` --- Custom Summary and Multiple Responses

```{r}
#| output.lines: 3
# custom function: result is a matrix column!
aggregate(len ~ supp, data = tg,
          FUN = function(x) c(mean = mean(x), sd = sd(x)))
```

```{r}
#| output.lines: 3
# multiple response columns with cbind()
aggregate(cbind(mpg, hp, wt) ~ cyl, data = mtcars, FUN = mean)
```

---

## `aggregate()` --- Expand Matrix Column and `.` Shorthand

```{r}
#| output.lines: 3
# expand matrix column into proper data frame
res <- aggregate(len ~ supp, data = tg,
                 FUN = function(x) c(m = mean(x), s = sd(x)))
cbind(res[1], as.data.frame(res$len))
```

```{r}
#| output.lines: 3
# . means "all other columns"
aggregate(. ~ cyl, data = mtcars[, c('mpg','hp','wt','cyl')],
          FUN = mean)
```

---

## `merge()` --- Joining Data Frames

```{r}
#| output.lines: 3
customers <- data.frame(id = 1:4,
  name = c('Alice', 'Bob', 'Cathy', 'Dan'))
orders <- data.frame(cust_id = c(2, 3, 3, 1, 5),
  product = c('pen','book','ink','paper','tape'),
  amount = c(12, 35, 8, 20, 5))
# inner join (default): only matched rows
merge(customers, orders, by.x = 'id', by.y = 'cust_id')
```

---

## `merge()` --- Join Types

```{r}
#| output.lines: 3
# left join: keep all customers (NA for no match)
merge(customers, orders, by.x='id', by.y='cust_id', all.x=TRUE)
```

\vfill

\begin{table}[!htbp]
\centering\small
\begin{tabular}{lll}
\toprule
SQL & \texttt{merge()} arguments & Keeps \\
\midrule
\texttt{INNER JOIN} & default (\comm{all=FALSE}) & matched rows only \\
\texttt{LEFT JOIN} & \comm{all.x = TRUE} & all rows in \comm{x} \\
\texttt{RIGHT JOIN} & \comm{all.y = TRUE} & all rows in \comm{y} \\
\texttt{FULL OUTER JOIN} & \comm{all = TRUE} & all rows in both \\
\bottomrule
\end{tabular}
\end{table}

---

## `merge()` --- Multi-Column Keys and Suffixes

```{r}
#| output.lines: 3
# multi-column key
sales <- data.frame(year=c(2023,2023,2024), qtr=c(1,2,1), rev=c(100,120,110))
targets <- data.frame(year=c(2023,2023,2024), qtr=c(1,2,1), target=c(95,115,105))
merge(sales, targets, by = c('year', 'qtr'))
```

```{r}
#| output.lines: 3
# duplicate column names get suffixes
df1 <- data.frame(id = 1:3, value = c(10, 20, 30))
df2 <- data.frame(id = 1:3, value = c(100, 200, 300))
merge(df1, df2, by = 'id', suffixes = c('_old', '_new'))
```

## Ex 27: Creating and Inspecting

- Create a data frame with columns `name` (Alice, Bob, Cathy), `age` (25, 30, 28), `score` (88.5, 92.3, 76.8)
- Show `str`, `dim`, `names`, `summary`
- What class is `df$name`?
- Access `score` of row 2 by name and by position


## Sol 27

```{r}
#| output.lines: 3
df <- data.frame(name=c("Alice","Bob","Cathy"),
                 age=c(25,30,28), score=c(88.5,92.3,76.8))
str(df); dim(df); names(df)
class(df$name)
df$score[2]; df[2, "score"]
```


## Ex 28: Row Filtering

Using `tg <- ToothGrowth`:

- Rows where `supp == "OJ"` and `dose >= 1`
- Mean tooth length for VC vs OJ
- Rows where `len` is above its overall median


## Sol 28

```{r}
#| output.lines: 3
tg <- ToothGrowth
head(tg[tg$supp == "OJ" & tg$dose >= 1, ], 3)
tapply(tg$len, tg$supp, mean)
head(tg[tg$len > median(tg$len), ], 3)
```


## Ex 29: Adding and Removing Columns

Start with `df <- data.frame(x=1:5, y=c(10,20,30,40,50))`.

- Add column `z = x * y`
- Add column `label` using `paste0("item", 1:5)`
- Remove column `z`
- Rename `y` to `value`


## Sol 29

```{r}
#| output.lines: 3
df <- data.frame(x=1:5, y=c(10,20,30,40,50))
df$z <- df$x * df$y
df$label <- paste0("item", 1:5)
df$z <- NULL
names(df)[names(df)=="y"] <- "value"; df
```


## Ex 30: `with()` and `within()`

Using `tg <- ToothGrowth`:

- Use `with()` to compute `mean(len[dose == 2])`
- Use `within()` to add `log_len <- log(len)` and `len_centered <- len - mean(len)`


## Sol 30

```{r}
#| output.lines: 3
tg <- ToothGrowth
with(tg, mean(len[dose == 2]))
tg2 <- within(tg, {
  log_len <- log(len)
  len_centered <- len - mean(len)
})
head(tg2, 3)
```


## Ex 31: Handling NAs

`df <- data.frame(x=c(1,NA,3,4,NA), y=c(NA,2,3,NA,5), z=c(1,2,3,4,5))`.

- Which rows are complete?
- Remove all incomplete rows
- Replace NAs in column `x` with the mean of `x` (ignoring NAs)
- Count NAs per column


## Sol 31

```{r}
#| output.lines: 3
df <- data.frame(x=c(1,NA,3,4,NA), y=c(NA,2,3,NA,5), z=1:5)
complete.cases(df)
df[complete.cases(df), ]
df$x[is.na(df$x)] <- mean(df$x, na.rm=TRUE); df
colSums(is.na(df))
```


## Ex 32: `merge()`

`students <- data.frame(id=1:4, name=c("A","B","C","D"))`.
`grades <- data.frame(sid=c(1,2,2,4), course=c("Math","Math","Eng","Eng"), grade=c(85,90,78,92))`.

- Inner join on id
- Left join (keep all students)
- How many rows in each result? Why?


## Sol 32

```{r}
#| output.lines: 3
students <- data.frame(id=1:4, name=c("A","B","C","D"))
grades <- data.frame(sid=c(1,2,2,4),
  course=c("Math","Math","Eng","Eng"), grade=c(85,90,78,92))
merge(students, grades, by.x="id", by.y="sid")     # inner
merge(students, grades, by.x="id", by.y="sid",
      all.x=TRUE)                                   # left
# inner: 4 rows; left: 5 rows (C has NA)
```


## Ex 33: `aggregate()`

Using `mtcars`:

- Mean `mpg` by `cyl`
- Mean and sd of `mpg` by `cyl` and `am`
- Max `hp` by `cyl`


## Sol 33

```{r}
#| output.lines: 3
aggregate(mpg ~ cyl, data=mtcars, FUN=mean)
```


## Sol 33 Cont'd

```{r}
#| output.lines: 3
aggregate(mpg ~ cyl + am, data=mtcars,
  FUN=function(x) c(m=mean(x), s=sd(x)))
aggregate(hp ~ cyl, data=mtcars, FUN=max)
```

## Ex 34: `cbind` and `rbind`

- Create two 1-column data frames and `cbind` them
- Create two data frames with same columns and `rbind`
- Use `do.call(rbind, list_of_dfs)` with 3 data frames


## Sol 34

```{r}
#| output.lines: 2
d1 <- data.frame(a=1:3); d2 <- data.frame(b=4:6)
cbind(d1, d2)
e1 <- data.frame(x="a", y=1); e2 <- data.frame(x="b", y=2)
rbind(e1, e2)
dfs <- list(data.frame(x=1,y=2), data.frame(x=3,y=4),
            data.frame(x=5,y=6))
do.call(rbind, dfs)
```


## Ex 35: Data Frame as Named List

- Show `is.list(mtcars)` is TRUE. What does `typeof(mtcars)` return?
- Access the `mpg` column using `[[` notation
- What is the difference between `mtcars["mpg"]` and `mtcars[["mpg"]]`?
- Convert `mtcars[1:3, 1:3]` to a list and back


## Sol 35

```{r}
#| output.lines: 3
is.list(mtcars); typeof(mtcars)                    # TRUE; "list"
head(mtcars[["mpg"]])
class(mtcars["mpg"]); class(mtcars[["mpg"]])       # df vs numeric
l <- as.list(mtcars[1:3, 1:3]); as.data.frame(l)
```


## Ex 36: `split()` on Data Frames

- Split `iris` by `Species`. What type is the result?
- Compute mean `Sepal.Length` per species using `split` + `sapply`
- Recombine with `do.call(rbind, ...)`


## Sol 36

```{r}
#| output.lines: 3
sp <- split(iris, iris$Species)
class(sp); names(sp)                               # list
```


## Sol 36 Cont'd

```{r, echo=FALSE}
sp <- split(iris, iris$Species)
```

```{r}
#| output.lines: 3
sapply(sp, function(d) mean(d$Sepal.Length))
head(do.call(rbind, sp), 4)
```

## Ex 37: `subset()` with `select`

Using `mtcars`:

- Select only `mpg`, `cyl`, `hp` for cars with `mpg > 25`
- Exclude `drat` and `wt` columns
- Why can you write `subset(mtcars, mpg > 25)` without quoting `mpg`?


## Sol 37

```{r}
#| output.lines: 3
subset(mtcars, mpg > 25, select=c(mpg, cyl, hp))
head(subset(mtcars, select=-c(drat, wt)), 3)
# subset uses non-standard evaluation: it captures
# the expression and evaluates it within the data frame
```


## Ex 38: Sorting with Multiple Keys

`df <- data.frame(dept=c("A","B","A","B","A"), name=c("Zoe","Amy","Bob","Cal","Amy"), salary=c(50,60,70,55,65))`.

- Sort by dept ascending, then salary descending
- Find the highest-paid person in each department (using `order` + `!duplicated`)


## Sol 38

```{r}
#| output.lines: 3
df <- data.frame(dept=c("A","B","A","B","A"),
  name=c("Zoe","Amy","Bob","Cal","Amy"), salary=c(50,60,70,55,65))
df[order(df$dept, -df$salary), ]
sorted <- df[order(df$dept, -df$salary), ]
sorted[!duplicated(sorted$dept), ]
```


## Ex 39: Column-wise Operations

- Use `colMeans` on `mtcars[,1:4]`
- Use `sapply` to find `max` of each column
- Standardize (z-score) all numeric columns using `scale()`. Compare with manual `sweep`


## Sol 39

```{r}
#| output.lines: 3
colMeans(mtcars[,1:4])
sapply(mtcars[,1:4], max)
head(scale(mtcars[,1:2]), 3)
```


## Sol 39 Cont'd

```{r}
#| output.lines: 3
m <- as.matrix(mtcars[,1:2])
head(sweep(sweep(m, 2, colMeans(m)), 2, apply(m,2,sd), "/"), 3)
```

## Ex 40: Building a Gradebook

- Create `grades <- data.frame(name=c("A","B","C","D","E"), hw=c(85,92,78,95,88), exam=c(90,85,92,88,76))`
- Add `final = 0.4*hw + 0.6*exam`
- Add `letter` using nested `ifelse` (A/B/C/F)
- Sort by `final` descending


## Sol 40

```{r}
#| output.lines: 3
grades <- data.frame(name=c("A","B","C","D","E"),
  hw=c(85,92,78,95,88), exam=c(90,85,92,88,76))
grades$final <- 0.4*grades$hw + 0.6*grades$exam
grades$letter <- ifelse(grades$final>=90,"A",
  ifelse(grades$final>=80,"B",
  ifelse(grades$final>=70,"C","F")))
grades[order(-grades$final), ]
```


## Ex 41: Multi-Factor Split and Unsplit

Using `ToothGrowth`:

- Split by both `supp` and `dose`; how many groups?
- Compute mean `len` per group using `sapply`
- Add a z-score column within each group using `split` + `lapply` + `do.call(rbind,...)`
- Use `unsplit` to restore original ordering after modifying split pieces


## Sol 41

```{r}
#| output.lines: 3
tg <- ToothGrowth
sp <- split(tg, list(tg$supp, tg$dose))
length(sp)                                           # 6
sapply(sp, \(d) mean(d$len))
```


## Sol 41 Cont'd

```{r, echo=FALSE}
tg <- ToothGrowth
sp <- split(tg, list(tg$supp, tg$dose))
```

```{r}
#| output.lines: 3
parts <- lapply(sp, \(d) {
  d$len_z <- (d$len - mean(d$len)) / sd(d$len); d })
tg2 <- do.call(rbind, parts)
head(tg2[order(as.numeric(rownames(tg2))), ], 4)
```

```{r}
#| output.lines: 3
# unsplit example
sg <- split(1:8, rep(c('A','B'), 4))
sg$A <- sg$A * 10; sg$B <- sg$B * (-1)
unsplit(sg, rep(c('A','B'), 4))
```

## Ex 42: `aggregate()` --- Matrix Column Gotcha

Using `ToothGrowth`:

- Use `aggregate` with `FUN = function(x) c(m=mean(x), s=sd(x))`. What type is the result column?
- Expand the matrix column into a proper flat data frame
- Use `cbind(mpg, hp, wt) ~ cyl` to summarize multiple columns of `mtcars`
- Use `.` shorthand: `aggregate(. ~ cyl, ...)` on a subset of `mtcars`


## Sol 42

```{r}
#| output.lines: 3
tg <- ToothGrowth
res <- aggregate(len ~ supp, data = tg,
  FUN = function(x) c(m=mean(x), s=sd(x)))
str(res)                                             # matrix col!
cbind(res[1], as.data.frame(res$len))                # flatten
```


## Sol 42 Cont'd

```{r}
#| output.lines: 3
aggregate(cbind(mpg, hp, wt) ~ cyl,
          data = mtcars, FUN = mean)
aggregate(. ~ cyl,
  data = mtcars[, c('mpg','hp','wt','cyl')], FUN = mean)
```

## Ex 43: `merge()` --- All Join Types

`cust <- data.frame(id=1:4, name=c("A","B","C","D"))`.
`ord <- data.frame(cid=c(2,3,3,5), prod=c("pen","book","ink","tape"))`.

- Inner join. Which rows are dropped and why?
- Left join. What appears for customer D?
- Right join. What appears for `cid=5`?
- Full outer join


## Sol 43

```{r}
#| output.lines: 3
cust <- data.frame(id=1:4, name=c("A","B","C","D"))
ord <- data.frame(cid=c(2,3,3,5),
                  prod=c("pen","book","ink","tape"))
merge(cust, ord, by.x="id", by.y="cid")              # inner
merge(cust, ord, by.x="id", by.y="cid", all.x=TRUE)  # left
```


## Sol 43 Cont'd

```{r, echo=FALSE}
cust <- data.frame(id=1:4, name=c("A","B","C","D"))
ord <- data.frame(cid=c(2,3,3,5),
                  prod=c("pen","book","ink","tape"))
```

```{r}
#| output.lines: 3
merge(cust, ord, by.x="id", by.y="cid", all.y=TRUE)  # right
merge(cust, ord, by.x="id", by.y="cid", all=TRUE)    # full
```

## Ex 44: Multi-Table Merge with `Reduce`

- Create `products(pid, name)`, `orders(oid, pid, qty)`, `prices(pid, unit_price)`
- Inner-join all three; compute `total = qty * unit_price`
- Add `regions(pid, region)`. Use `Reduce` to merge all four


## Sol 44

```{r}
#| output.lines: 1
products <- data.frame(pid=1:5,
  name=c('Pen','Book','Ink','Paper','Tape'))
orders <- data.frame(oid=101:106,
  pid=c(2,1,3,2,5,1), qty=c(3,10,2,5,1,8))
prices <- data.frame(pid=1:5,
  unit_price=c(1.5, 12, 8, 3, 2.5))
m <- merge(merge(products, orders, by='pid'),
           prices, by='pid')
m$total <- m$qty * m$unit_price; m[, c('oid','name','total')]

regions <- data.frame(pid=1:5, region=c('N','S','N','S','N'))
Reduce(\(x, y) merge(x, y, by='pid', all=TRUE),
       list(products, orders, prices, regions))
```


## Ex 45: `vapply` vs `sapply` Safety

- `sapply(list(a=1:3, b=4:6), identity)` returns a matrix. `sapply(list(a=1:3, b=4:5), identity)` returns a list. Demonstrate
- Show that `vapply(list(a=1:3, b=4:6), identity, numeric(1))` throws an error
- Use `vapply` correctly with `FUN.VALUE = numeric(3)` for same-length vectors
- Use `vapply(iris[,1:4], median, numeric(1))` to safely compute medians


## Sol 45

```{r}
#| output.lines: 3
sapply(list(a=1:3, b=4:6), identity)                 # matrix
sapply(list(a=1:3, b=4:5), identity)                 # list!
```

```{r, error=TRUE}
#| output.lines: 3
vapply(list(a=1:3, b=4:6), identity, numeric(1))    # error!
vapply(list(a=1:3, b=4:6), identity, numeric(3))    # OK
vapply(iris[,1:4], median, numeric(1))               # safe
```


## Ex 46: `rapply()` --- Recursive Apply

`nested <- list(a=1:3, b=list(c=4:6, d=list(e=7:9, f="text")))`.

- Use `rapply` with `sum`, `classes="integer"` to sum only integer leaves
- Same with `how="list"` to preserve structure
- Use `how="replace"` to multiply all integer leaves by 10
- What happens to the `"text"` element in each case?


## Sol 46

```{r}
#| output.lines: 3
nested <- list(a=1:3,
  b=list(c=4:6, d=list(e=7:9, f="text")))
rapply(nested, sum, classes="integer")               # unlist
rapply(nested, sum, classes="integer", how="list")   # structure
rapply(nested, \(x) x*10, classes="integer",
       how="replace")                                # replace
# "text" is left unchanged (not class "integer")
```


## Ex 47: `Reduce` --- Running Max Drawdown

\texttt{r <- c(0.02, -0.01, 0.03, -0.04, 0.01, -0.02, 0.05)}.

- Use `Reduce` with `accumulate=TRUE` to compute cumulative wealth $W_t = W_{t-1}(1+r_t)$, starting from $W_0=1$
- Compute running max of cumulative wealth using `Reduce` and `max`
- Compute drawdown: $\text{DD}_t = 1 - W_t / \max_{s \le t} W_s$
- Verify with vectorized `1 - cumprod(1+r) / cummax(cumprod(1+r))`


## Sol 47

```{r}
#| output.lines: 3
r <- c(0.02, -0.01, 0.03, -0.04, 0.01, -0.02, 0.05)
W <- Reduce(\(w, ri) w*(1+ri), r, init=1,
            accumulate=TRUE)[-1]
Wmax <- Reduce(\(mx, w) max(mx, w), W,
               accumulate=TRUE)
DD <- 1 - W / Wmax; round(DD, 6)
DD2 <- 1 - cumprod(1+r) / cummax(cumprod(1+r))
all.equal(DD, DD2)                                   # TRUE
```


## Ex 48: `Filter` Pipeline with `|>`

`words <- c("apple","Banana","cherry","Date","123","fig","GRAPE","hi")`.

- Use `Filter` to keep only all-lowercase words
- Use `Filter` + `Negate` to remove strings containing digits
- Find first word with $>5$ characters using `Find`; find its position with `Position`
- Chain with `|>`: filter lowercase $\to$ filter length $>3$ $\to$ sort


## Sol 48

```{r}
#| output.lines: 3
words <- c("apple","Banana","cherry","Date","123",
           "fig","GRAPE","hi")
Filter(\(w) w == tolower(w), words)
Filter(Negate(\(w) grepl("[0-9]", w)), words)
Find(\(w) nchar(w) > 5, words)
Position(\(w) nchar(w) > 5, words)

words |>
  Filter(f = \(w) w == tolower(w)) |>
  Filter(f = \(w) nchar(w) > 3) |>
  sort()
```


# Lists

## What Is a List?

- A **generic vector**: elements can be of **different types and lengths**.

```{r}
#| output.lines: 3
vn <- c(1, 1.2, 2.3, 3.4, 4.5)
vb <- c(TRUE, TRUE, FALSE)
vc <- c('limestone', 'marl', 'oolite', 'CaCO3')
```

```{r, error=TRUE}
#| output.lines: 3
data.frame(vn, vb, vc)   # ERROR: different lengths
```

---

## Creating Lists

```{r}
#| output.lines: 3
# unnamed list
l <- list(3.14, 'Moe', c(1, 3, 5, 2), mean)
str(l)
```

```{r}
#| output.lines: 3
# named list + incremental construction
l <- list(max = 5, right = 0.6, mean = 20)
l <- list(); l$left <- 0.1; l$right <- 0.52; l$mid <- 0.33
l
```

---

## `[[` vs `[` --- The Most Important Distinction

:::::: columns
::: column

```{r}
#| output.lines: 3
l <- list(a = 1:3, b = 'hello',
          c = TRUE)
l[[1]]     # the element itself
l[1]       # a sub-list
```

```{r}
#| output.lines: 3
class(l[[1]]); class(l[1])
```

:::
::: column

```{r}
#| output.lines: 3
l$a   # same as l[['a']]
l$b
```

\vfill

**Analogy**: `[[` reaches into the box and pulls out the item; `[` selects boxes (still wrapped).

:::
::::::

---

## Modifying, Removing, Concatenating

```{r}
#| output.lines: 3
l <- list(x = 1, y = 2, z = 3)
l$x <- 100; l[['y']] <- 200; l$w <- 'new'   # modify / add
l[c('x', 'w')] <- NULL; l                     # remove
```

```{r}
#| output.lines: 3
l1 <- list(a = 1, b = 2); l2 <- list(c = 3, d = 4)
c(l1, l2)    # concatenate lists
```

## Modifying, Removing, Concatenating Cont'd


```{r}
#| output.lines: 3
# nested lists
nested <- list(student = list(name = 'Alice', age = 20),
               scores  = list(math = 95, english = 88))
nested$student$name; nested[['scores']][['math']]
```

---

## `unlist()` --- Flatten a List

```{r}
#| output.lines: 3
iq <- list(100, 120, 140, 112, 105)
mean(unlist(iq))       # must unlist before mean()
```

```{r}
#| output.lines: 3
l <- list(a = 1:3, b = 4:5)
unlist(l)                        # names preserved with prefix
unlist(l, use.names = FALSE)     # drop names
```

---

## `split()` --- Split a Vector into a List

```{r}
#| output.lines: 3
(mtl <- split(mtcars$mpg, mtcars$cyl))
mtl$`6`; mtl[[2]]   # access by name or position
```

- `tapply(x, f, FUN)` is equivalent to `sapply(split(x, f), FUN)`

## A Data Frame Is a List

```{r}
#| output.lines: 3
df <- data.frame(x = 1:3, y = c('a', 'b', 'c'))
is.list(df); typeof(df)
```

```{r}
#| output.lines: 3
df$x; df[['x']]     # extracts column vector (like [[)
df['x']              # returns 1-column data.frame (like [)
```

```{r, eval=FALSE}
# split() a data frame -> list of data frames
split(tg, tg$supp)  # returns list of data frames
```

---

## Conversions Between Data Structures

\begin{table}[!htbp]
\centering\small
\begin{tabular}{llll}
\toprule
From & To & How & Note\\
\midrule
vector & list & \comm{as.list(v)} & Don't use \comm{list(v)}!\\
vector & matrix & \comm{matrix(v, n, m)} & \\
vector & data.frame & \comm{as.data.frame(v)} & \\
list & vector & \comm{unlist(l)} & Not \comm{as.vector}\\
list & data.frame & \comm{as.data.frame(l)} & Equal-length only\\
matrix & vector & \comm{as.vector(m)} & \\
matrix & data.frame & \comm{as.data.frame(m)} & \\
data.frame & list & \comm{as.list(df)} & Removes class attr\\
data.frame & matrix & \comm{as.matrix(df)} & Coercion!\\
\bottomrule
\end{tabular}
\end{table}

## Ex 49: Creating and Accessing Lists

- Create `l <- list(name="Alice", scores=c(85,92,78), passed=TRUE)`
- Access `scores` with `$`, `[[]]`, and `[]`. Show the different classes
- Add element `grade = "B"`. Remove `passed`
- What is `length(l)` before and after?


## Sol 49

```{r}
#| output.lines: 3
l <- list(name="Alice", scores=c(85,92,78), passed=TRUE)
l$scores; l[["scores"]]; l["scores"]
class(l$scores); class(l["scores"])                # vector vs list
l$grade <- "B"; l$passed <- NULL
length(l)  # now 3
```


## Ex 50: `unlist()` Behavior

- `l <- list(a=1:3, b=4:5)`. What does `unlist(l)` produce? Explain the names
- `unlist(l, use.names=FALSE)`
- `l2 <- list(x="hello", y=1:3)`. What type is `unlist(l2)`? Why?
- Mean of a list of numbers: `list(10, 20, 30)`


## Sol 50

```{r}
#| output.lines: 3
l <- list(a=1:3, b=4:5)
unlist(l)                # named: a1 a2 a3 b1 b2
unlist(l, use.names=FALSE)
unlist(list(x="hello", y=1:3))  # all character
mean(unlist(list(10, 20, 30)))                     # 20
```


## Ex 51: Nested Lists

- Create `student <- list(info=list(name="Bob", age=20), grades=list(math=90, eng=85))`
- Access Bob's math grade
- Add `grades$sci = 92`
- Use `str()` to visualize the structure


## Sol 51

```{r}
#| output.lines: 3
student <- list(info=list(name="Bob", age=20),
                grades=list(math=90, eng=85))
student$grades$math                                # 90
student$grades$sci <- 92
str(student)
```


## Ex 52: `split()` and Recombine

- `split(mtcars$mpg, mtcars$cyl)` — what type?
- Compute `mean` per group using `sapply`
- Same using `tapply` in one line
- `split(1:10, rep(c("odd","even"), 5))`


## Sol 52

```{r}
#| output.lines: 3
sp <- split(mtcars$mpg, mtcars$cyl); class(sp)    # list
sapply(sp, mean)
tapply(mtcars$mpg, mtcars$cyl, mean)
split(1:10, rep(c("odd","even"), 5))
```


## Ex 53: List Concatenation

- Concatenate `list(a=1)` and `list(b=2)` with `c()`
- What happens with `c(list(a=1), 5)`?
- Build a list incrementally in a loop: for i=1:5, append `i^2`


## Sol 53

```{r}
#| output.lines: 3
c(list(a=1), list(b=2))
c(list(a=1), 5)                                    # 5 becomes element
l <- list()
for (i in 1:5) l[[i]] <- i^2; l
```


## Ex 54: Conversion Table

- Vector to list: `as.list(1:3)` vs `list(1:3)`. Explain difference
- List to data frame: `as.data.frame(list(a=1:3, b=4:6))`
- Data frame to matrix: what happens with `as.matrix(data.frame(x=1:3, y=c("a","b","c")))`?
- List to vector: why use `unlist` not `as.vector`?


## Sol 54

```{r}
#| output.lines: 3
as.list(1:3)    # 3 elements
list(1:3)       # 1 element (the vector)
as.data.frame(list(a=1:3, b=4:6))
as.matrix(data.frame(x=1:3, y=c("a","b","c")))    # all char!
unlist(list(a=1, b=2))
```


## Ex 55: Named List as Lookup Table

- Create a lookup: `grade_points <- list(A=4.0, B=3.0, C=2.0, D=1.0, F=0)`
- Look up GPA for grades `c("A","B","F","A","C")` using `sapply` + `[[`
- Compute mean GPA


## Sol 55

```{r}
#| output.lines: 3
grade_points <- list(A=4.0, B=3.0, C=2.0, D=1.0, F=0)
g <- c("A","B","F","A","C")
gpa <- sapply(g, function(x) grade_points[[x]])
gpa
mean(gpa)                                          # 2.6
```


## Ex 56: `lapply`/`sapply` on Lists

`l <- list(a=1:10, b=rnorm(20), c=c(TRUE,FALSE,TRUE,TRUE))`.

- Compute `mean` of each element with `lapply` and `sapply`
- Compute `length` of each
- Apply custom function: range (max - min)


## Sol 56

```{r}
#| output.lines: 3
set.seed(1)
l <- list(a=1:10, b=rnorm(20), c=c(TRUE,FALSE,TRUE,TRUE))
lapply(l, mean); sapply(l, mean)
sapply(l, length)
sapply(l, function(x) diff(range(x)))
```


## Ex 57: `Filter`, `Find`, `Position`

`l <- list(1:3, "hello", 4:6, TRUE, NULL, c(1.5, 2.5))`.

- Keep only numeric elements
- Find first character element
- Position of first character element
- Filter integers from `1:30` that are divisible by both 3 and 7


## Sol 57

```{r}
#| output.lines: 3
l <- list(1:3, "hello", 4:6, TRUE, NULL, c(1.5, 2.5))
Filter(is.numeric, l)
Find(is.character, l)                              # "hello"
Position(is.character, l)                          # 2
Filter(\(x) x%%3==0 & x%%7==0, 1:30)              # 21
```


## Ex 58: `Reduce()`

- Compute `1+2+3+4+5` using `Reduce("+", 1:5)`. Show accumulated steps
- Find intersection of 3 vectors via `Reduce(intersect, ...)`
- Merge 3 data frames pairwise using `Reduce`


## Sol 58

```{r}
#| output.lines: 3
Reduce("+", 1:5)                                   # 15
Reduce("+", 1:5, accumulate=TRUE)                  # steps
vecs <- list(c(1,2,3,4), c(2,3,4,5), c(3,4,5,6))
Reduce(intersect, vecs)                            # 3 4
dfs <- list(data.frame(id=1:3, a=10:12),
            data.frame(id=2:4, b=20:22))
Reduce(\(x,y) merge(x,y,by="id",all=TRUE), dfs)
```


## Ex 59: List Modification Patterns

- Given `l <- list(x=1, y=2, z=3)`, remove `x` and `z` in one operation
- Replace all numeric elements with their squares using `lapply`
- Deep copy a nested list and modify without affecting original


## Sol 59

```{r}
#| output.lines: 3
l <- list(x=1, y=2, z=3)
l[c("x","z")] <- NULL; l                          # only y
l2 <- list(a=3, b="hi", c=4)
lapply(l2, function(x) if(is.numeric(x)) x^2 else x)
orig <- list(inner=list(val=10))
copy <- orig; copy$inner$val <- 99
orig$inner$val  # still 10 (R copies on modify)
```


## Ex 60: `rapply()` --- Recursive Apply

`l <- list(a=1:3, b=list(c=4:6, d=list(e=7:9)))`.

- Apply `sum` to all leaf elements using `rapply(l, sum)`
- Use `rapply(l, sum, how="unlist")` vs `how="list"`
- Double all values with `rapply(l, function(x) x*2, how="replace")`


## Sol 60

```{r}
#| output.lines: 3
l <- list(a=1:3, b=list(c=4:6, d=list(e=7:9)))
rapply(l, sum)
rapply(l, sum, how="unlist")                       # named vector
rapply(l, sum, how="list")                         # list structure
str(rapply(l, \(x) x*2, how="replace"))            # doubled
```


# Matrices

## Creating Matrices

```{r}
#| output.lines: 3
v <- 1:6; matrix(v, nrow = 2, ncol = 3)
```

```{r}
#| output.lines: 3
# named rows and columns
m <- matrix(c(1, .56, .38, .56, 1, .44, .38, .44, 1), 3, 3)
colnames(m) <- rownames(m) <- c('AAPL', 'MSFT', 'GOOG')
m; m['MSFT', 'GOOG']
```

---

## Matrix Operations

```{r}
#| output.lines: 3
A <- matrix(1:4, 2, 2); B <- matrix(5:8, 2, 2)
A * B        # element-wise
A %*% B      # matrix multiplication
t(A); det(A); solve(A)   # transpose, determinant, inverse
```

## Matrix Operations Cont'd


```{r}
#| output.lines: 3
# row/column selection: drop=FALSE preserves matrix class
A[1, ]; A[1, , drop = FALSE]
```

---

## Matrix Functions Summary

```{r, eval=FALSE}
nrow(m); ncol(m); dim(m)
diag(m)                    # diagonal elements
diag(3)                    # 3x3 identity matrix
crossprod(A, B)            # t(A) %*% B (faster)
tcrossprod(A, B)           # A %*% t(B)
outer(1:3, 1:3, '+')      # outer product
```

## Ex 61: Creating Matrices

- Create 3×4 matrix filled column-wise from `1:12`
- Create same matrix filled row-wise
- Name rows `r1,r2,r3` and columns `c1,...,c4`. Access `m["r2","c3"]`
- Create a 3×3 identity matrix with `diag`


## Sol 61

```{r}
#| output.lines: 3
matrix(1:12, nrow=3)
matrix(1:12, nrow=3, byrow=TRUE)
m <- matrix(1:12, 3, 4)
rownames(m) <- paste0("r",1:3); colnames(m) <- paste0("c",1:4)
m["r2","c3"]                                       # 8
diag(3)
```


## Ex 62: Matrix Arithmetic

`A <- matrix(1:4, 2, 2); B <- matrix(5:8, 2, 2)`.

- Element-wise `*` vs matrix `%*%`
- Compute `t(A)`, `det(A)`, `solve(A)`
- Verify `A %*% solve(A)` is the identity
- `crossprod(A,B)` vs `t(A) %*% B`


## Sol 62

```{r}
#| output.lines: 2
A <- matrix(1:4, 2, 2); B <- matrix(5:8, 2, 2)
A * B; A %*% B
t(A); det(A); solve(A)
round(A %*% solve(A))                              # identity
all.equal(crossprod(A,B), t(A) %*% B)             # TRUE
```


## Ex 63: `apply()` on Matrices

`m <- matrix(1:12, nrow=3)`.

- Row sums and column sums via `apply`
- Per-column range
- Standardize each column (subtract mean, divide by sd)
- Compare `apply(m,2,sum)` speed with `colSums(m)`


## Sol 63

```{r}
#| output.lines: 3
m <- matrix(1:12, nrow=3)
apply(m, 1, sum); apply(m, 2, sum)
apply(m, 2, range)
apply(m, 2, \(x) (x-mean(x))/sd(x))
all.equal(apply(m,2,sum), colSums(m))              # TRUE
```


## Ex 64: `outer()` Product

- Multiplication table 1..5 × 1..5 using `outer`
- Compute $f(x,y)=x^2+y^2$ for `x=1:4`, `y=1:3`
- All pairwise `paste` of `c("a","b")` and `c("1","2","3")`


## Sol 64

```{r}
#| output.lines: 3
outer(1:5, 1:5, "*")
outer(1:4, 1:3, \(x,y) x^2 + y^2)
outer(c("a","b"), c("1","2","3"), paste0)
```


## Ex 65: Matrix Row/Column Selection

`m <- matrix(1:12, 3, 4)`.

- Extract row 2 as a vector vs as a 1-row matrix (`drop`)
- Extract column 3 both ways
- Select submatrix: rows 1-2, columns 2-3
- Replace all elements > 8 with `NA`


## Sol 65

```{r}
#| output.lines: 3
m <- matrix(1:12, 3, 4)
m[2,]; m[2,,drop=FALSE]
m[,3]; m[,3,drop=FALSE]
m[1:2, 2:3]
m[m > 8] <- NA; m
```


## Ex 66: Linear System

Solve $2x + 3y = 7, \; 4x - y = 1$.

- Set up as `A %*% x = b` and solve with `solve(A, b)`
- Verify by multiplying back
- Compute eigenvalues of `A`


## Sol 66

```{r}
#| output.lines: 3
A <- matrix(c(2,4,3,-1), 2, 2); b <- c(7,1)
x <- solve(A, b); x
A %*% x  # should be c(7,1)
eigen(A)$values
```


# Programming

## Defining Functions

```{r, eval=FALSE}
function_name <- function(arg1, arg2 = default_value, ...) {
    # code
    return(some_object)
}
```

```{r}
#| output.lines: 3
# functions can read global variables
y_squared <- function() { return(y^2) }
y <- 2; y_squared()
```

```{r}
#| output.lines: 3
# but local assignment doesn't change globals
add_to_y <- function(n) { y <- y + n }
y <- 1; add_to_y(2); y      # y is still 1
```

```{r}
#| output.lines: 3
# <<- modifies globals (use sparingly!)
add_to_y_global <- function(n) { y <<- y + n }
y <- 1; add_to_y_global(2); y   # y is now 3
```

---

## `return()` and Implicit Return

```{r}
#| output.lines: 3
avg <- function(x) { return(sum(x) / length(x)) }
avg(c(2, 3, 6))
avg(c(TRUE, FALSE, TRUE))   # TRUE=1, FALSE=0
```

```{r}
#| output.lines: 3
# implicit return: the last evaluated expression is returned
avg_bad <- function(x) { r <- sum(x) / length(x) }
avg_ok  <- function(x) { sum(x) / length(x) }
avg_bad(1:10)   # silent! (assignment returns invisibly)
avg_ok(1:10)    # returns result
```

---

## Default Arguments and Named Arguments

```{r}
#| output.lines: 3
power_n <- function(x, n = 2) { return(x^n) }
power_n(3)              # default n = 2
power_n(3, 3)           # supplied n = 3
power_n(x = 5, n = 2)  # named arguments
power_n(n = 2, x = 5)  # order doesn't matter with names
```

---

## Functions as Arguments and `...`

```{r}
#| output.lines: 3
apply_to_first2 <- function(x, func, ...) {
    func(x[1:2], ...)
}
x <- c(NA, 2, 3)
apply_to_first2(x, sum, na.rm = TRUE)
```

---

## Anonymous Functions and `\(x)` Shorthand

```{r}
#| output.lines: 3
sapply(1:5, function(x) x^2 + 1)
sapply(1:5, \(x) x^2 + 1)   # R 4.1+ shorthand
```

```{r}
#| output.lines: 3
# useful in apply family
l <- list(a = 1:5, b = 6:10)
lapply(l, \(v) c(mean = mean(v), sd = sd(v)))
```

## Lexical Scoping Rule

When R encounters a free variable inside a function, it searches the **environment where the function was defined** (not where it was called):

```{r}
#| output.lines: 3
x <- 10
f <- function() x          # f is born in global env
g <- function() {
  x <- 999
  f()                       # calls f, but f ignores this x
}
g()                         # 10, not 999
```

\vfill

- **Lexical** (static): determined by where the code is **written**
- R, Python, JavaScript all use lexical scoping

---

## What Is a Closure?

> **Closure** = a function **bundled together** with the environment in which it was created.

When function A returns function B, B retains a **live reference** to A's local environment --- even after A has finished executing.

```{r}
#| output.lines: 3
power_factory <- function(n) {
  function(x) x^n             # captures n
}
square <- power_factory(2)
cube   <- power_factory(3)
square(5); cube(5)
```

```{r}
#| output.lines: 3
# each call creates a new, independent environment
environment(square)$n; environment(cube)$n
```

---

## Inspecting Function Internals

R does not distinguish "closure" as a special type. Every function stores `formals`, `body`, and `environment`:

```{r}
#| output.lines: 3
f <- function(x) x + 1
typeof(f)          # "closure"
formals(f); body(f); environment(f)
```

---

## `<-` vs `<<-` and Mutable State

- `<-` assigns **only** in the current (local) environment
- `<<-` walks **up** the enclosing environment chain to modify an existing variable

```{r, eval=FALSE}
make_counter <- function() {
  n <- 0
  list(
    increment = function() { n <<- n + 1; n },
    get       = function() n,
    reset     = function() { n <<- 0 }
  )
}
```


## `<-` vs `<<-` and Mutable State Cont'd

```{r}
#| output.lines: 2
make_counter <- function() {
  n <- 0
  list(
    increment = function() { n <<- n + 1; n },
    get       = function() n,
    reset     = function() { n <<- 0 }
  )
}
ctr <- make_counter()
ctr$increment(); ctr$increment(); ctr$get()
ctr$reset(); ctr$get()
```

---

## Independent Instances and `local()`

```{r}
#| output.lines: 3
ctr1 <- make_counter(); ctr2 <- make_counter()
ctr1$increment(); ctr1$increment(); ctr2$increment()
cat("ctr1:", ctr1$get(), " ctr2:", ctr2$get(), "\n")
```

```{r}
#| output.lines: 3
# local() creates a single-instance closure
counter <- local({
  n <- 0
  list(
    increment = function() { n <<- n + 1; n },
    get       = function() n,
    reset     = function() { n <<- 0 })
})
counter$increment(); counter$increment(); counter$get()
```

---

## Classic Pitfall: Loop Capture

```{r}
#| output.lines: 3
funs <- list()
for (i in 1:5) funs[[i]] <- function() i
sapply(funs, \(f) f())  # all 5!
```

All five functions share the **same** environment; after the loop, `i = 5`.

```{r}
#| output.lines: 3
# Fix 1: force() defeats lazy evaluation
make_fun <- function(val) { force(val); function() val }
funs <- list()
for (i in 1:5) funs[[i]] <- make_fun(i)
sapply(funs, \(f) f())
```

```{r}
#| output.lines: 3
# Fix 2 (recommended): lapply creates separate environments
funs <- lapply(1:5, \(i) function() i)
sapply(funs, \(f) f())
```

---

## Base R Closures: `ecdf`, `approxfun`, `Negate`

```{r}
#| output.lines: 3
x <- c(1, 3, 3, 5, 7, 9); Fn <- ecdf(x)
Fn(4); Fn(7)        # P(X<=4)=3/6; P(X<=7)=5/6
typeof(Fn)           # "closure"
```

```{r}
#| output.lines: 3
xp <- 1:5; yp <- c(2, 5, 3, 8, 6)
linear_interp <- approxfun(xp, yp)
linear_interp(2.5)   # returns a closure over the knot data
```

```{r}
#| output.lines: 3
is_even <- function(x) x %% 2 == 0
is_odd  <- Negate(is_even)          # closure combinator
Filter(is_odd, 1:10)
```

## `if` / `else`

```{r}
#| output.lines: 3
reciprocal <- function(x) {
  if (is.na(x)) {
    cat("Error! Division by NA.\n")
  } else if (is.infinite(x)) {
    cat("Error! Division by infinity.\n")
  } else if (x == 0) {
    cat("Error! Division by zero.\n")
  } else { 1 / x }
}
reciprocal(2); reciprocal(0); reciprocal(NA)
```

---

## `ifelse()` --- Vectorized Conditional

```{r}
#| output.lines: 3
x <- c(-1, 2, -100, 20, 50)
ifelse(x > 0, 'positive', 'negative')
```

```{r}
#| output.lines: 3
# nested ifelse
v <- c(20, 18, 45, 33, 25, 9, 17, 55, 69, 81, 44, 32)
ifelse(v >= 20 & v < 40, 'young adult',
ifelse(v >= 40 & v < 60, 'middle age', 'other'))
```

---

## `switch()`

```{r}
#| output.lines: 3
day_type <- function(day) {
  switch(day,
    Monday=, Tuesday=, Wednesday=,
    Thursday=, Friday= "Weekday",
    Saturday=, Sunday= "Weekend",
    "Unknown")
}
day_type("Monday"); day_type("Sunday"); day_type("Holiday")
```

---

## `&&` vs `&` (Short-Circuit Evaluation)

```{r}
#| output.lines: 2
# first, ensure 'aa' does not exist from earlier chunks
if (exists("aa")) rm(aa)
```

```{r}
#| output.lines: 2
# && is short-circuited: stops at first FALSE
if (exists("aa") && aa > 0) {
    cat("aa exists and is positive.\n")
} else {
    cat("aa doesn't exist or is negative.\n")
}
```



## `&&` vs `&` (Short-Circuit Evaluation) Cont'd

```{r, error=TRUE}
#| output.lines: 2
# & evaluates BOTH sides -- error because aa doesn't exist!
if (exists("aa") & aa > 0) {
    cat("aa exists and is positive.\n")
} else {
    cat("aa doesn't exist or is negative.\n")
}
```

- **Rule**: Use `&&` / `||` in `if` (scalar); use `&` / `|` for vectors.

## `if (FALSE) { }` --- Code Block Toggle

```{r, eval=FALSE}
# wrap example/test code that shouldn't run
if (FALSE) {
  # these lines are "commented out" but remain valid R code
  book(date = "2026/04/02", headless = FALSE)
  cancel(order_id = "01901879")
}
```

- Unlike `#` comments, the code inside is still checked by R's parser

## `for` Loop

```{r}
#| output.lines: 3
# sum 1 to 10000
ans <- 0
for (i in 1:10000) { ans <- ans + i }
ans
```

```{r}
#| output.lines: 3
# sum of min(2^k, k^4) for k = 1..100
ans <- 0
for (k in 1:100) { ans <- ans + min(2^k, k^4) }
ans
```

```{r}
#| output.lines: 3
# vectorized equivalent (much faster)
v <- 1:100; sum(pmin(2^v, v^4))
```

---

## `while`, `repeat`, `break`, `next`

```{r}
#| output.lines: 3
# while: find smallest n such that sum(1:n) > 1000
n <- 0; s <- 0
while (s <= 1000) { n <- n + 1; s <- s + n }
cat("n =", n, ", sum =", s, "\n")
```

```{r}
#| output.lines: 3
# repeat + break
n <- 1
repeat { if (n^2 > 100) break; n <- n + 1 }
cat("First n with n^2 > 100:", n, "\n")
```

```{r}
#| output.lines: 3
# next: skip iterations
for (i in 1:10) {
  if (i %% 3 == 0) next    # skip multiples of 3
  cat(i, '')
}
```

---

## Vectorization vs Loops: Speed Comparison

```{r}
#| output.lines: 3
my_sum <- function(x) {
    ans <- 0
    for (i in seq_along(x)) { ans <- ans + x[i] }
    ans
}
v <- 1:1000000; system.time(my_sum(v)); system.time(sum(v))
```

## Vectorization vs Loops: Speed Comparison Cont'd


```{r}
#| output.lines: 3
# pre-allocate vs grow-in-loop vs vectorized
n <- 1e5
system.time({v <- c(); for (i in 1:n) v <- c(v, i^2)})
system.time({v <- numeric(n); for (i in 1:n) v[i] <- i^2})
system.time({v <- (1:n)^2})
```

## Native Pipe `|>` (R 4.1+)

```{r}
#| output.lines: 3
# without pipe: nested calls are hard to read
round(mean(abs(rnorm(100))), 3)
```

```{r}
#| output.lines: 3
# with pipe: read left to right
set.seed(1)
rnorm(100) |> abs() |> mean() |> round(3)
```

## Native Pipe `|>` (R 4.1+) Cont'd


```{r}
#| output.lines: 3
# pipe with data frame operations
mtcars |> subset(cyl == 6) |> head(3)
```

```{r, eval=FALSE}
# use _ as placeholder for non-first argument (R 4.2+)
mtcars |> lm(mpg ~ wt, data = _) |> summary()
```

## Ex 67: Default Arguments

- Write `power(x, n=2)` that computes `x^n`. Test with 1 and 2 arguments
- Write `greet(name, greeting="Hello")` that returns `"Hello, Alice!"`
- What happens if you call `power(n=3)` without `x`?


## Sol 67

```{r, error=TRUE}
#| output.lines: 3
power <- function(x, n=2) x^n
power(3); power(3, 3); power(x=5, n=2)
greet <- function(name, greeting="Hello")
  paste0(greeting, ", ", name, "!")
greet("Alice"); greet("Bob", "Hi")
power(n=3)  # error: x is missing
```


## Ex 68: Functions as Arguments

- Write `apply_fn(x, fn)` that applies `fn` to `x`. Test with `sqrt`, `log`, `abs`
- Extend to `apply_fn(x, fn, ...)` supporting extra arguments. Test: `apply_fn(c(1,NA,3), mean, na.rm=TRUE)`


## Sol 68

```{r}
#| output.lines: 3
apply_fn <- function(x, fn, ...) fn(x, ...)
apply_fn(c(4,9,16), sqrt)
apply_fn(c(1,NA,3), mean, na.rm=TRUE)             # 2
```


## Ex 69: Implicit Return

- Write two versions of average: one with explicit `return()`, one without
- Write `bad_avg` that assigns to a local variable as last line. Show it returns silently
- Capture the invisible result with `x <- bad_avg(1:10)`


## Sol 69

```{r}
#| output.lines: 3
avg_explicit <- function(x) { return(sum(x)/length(x)) }
avg_implicit <- function(x) { sum(x)/length(x) }
avg_explicit(1:10); avg_implicit(1:10)             # same
bad_avg <- function(x) { result <- sum(x)/length(x) }
bad_avg(1:10)  # no output!
x <- bad_avg(1:10); x  # but value is captured
```


## Ex 70: Anonymous Functions

- Use `sapply(1:5, function(x) x^2 + 1)`
- Rewrite with `\(x)` shorthand
- Use anonymous function in `Filter` to keep even numbers from `1:20`
- Nested: `sapply(1:3, \(i) sapply(1:3, \(j) i*j))`


## Sol 70

```{r}
#| output.lines: 3
sapply(1:5, function(x) x^2+1)
sapply(1:5, \(x) x^2+1)
Filter(\(x) x%%2==0, 1:20)
sapply(1:3, \(i) sapply(1:3, \(j) i*j))            # 3x3 matrix
```


## Ex 71: Closures

- Write `make_adder(n)` that returns a function adding `n`. Test `add5 <- make_adder(5); add5(3)`
- Write `make_counter()` returning a list with `increment`, `get`, `reset` functions using `<<-`
- Verify the counter works across multiple calls


## Sol 71

```{r}
#| output.lines: 3
make_adder <- function(n) function(x) x + n
add5 <- make_adder(5); add5(3)                     # 8
make_counter <- function() {
  n <- 0
  list(inc = function() {n <<- n+1; n},
       get = function() n,
       reset = function() {n <<- 0})
}
ctr <- make_counter()
ctr$inc(); ctr$inc(); ctr$get()                    # 2
ctr$reset(); ctr$get()                             # 0
```


## Ex 72: Dot-Prefix Naming

- Define `.helper <- function(x) x+1` and `public <- function(x) .helper(x)*2`
- Show `.helper` doesn't appear in `ls()` but does in `ls(all.names=TRUE)`
- Name 3 built-in R objects starting with `.`


## Sol 72

```{r}
#| output.lines: 3
e <- new.env()
e$.helper <- function(x) x+1
e$public  <- function(x) e$.helper(x)*2
e$public(5)  # 12
ls(e); ls(e, all.names=TRUE)
# .Machine, .Platform, .GlobalEnv
```


## Ex 73: `if (FALSE)` Idiom

- Wrap code in `if (FALSE) \{ ... \`} and show it doesn't execute
- Change to `if (TRUE)` and show it runs
- Why is this better than commenting with `#` for multi-line blocks?


## Sol 73

```{r}
#| output.lines: 3
if (FALSE) {
  cat("This won't print\n")
  x <- 999
}
exists("x")  # from previous chunks, may be TRUE
if (TRUE) { cat("This prints\n") }
# Unlike #, the code inside is still parsed by R,
# so syntax errors are caught immediately.
```


## Ex 74: Local Helper Functions

- Write `outer_fn(x)` that defines inner helper `validate(v)` checking `is.numeric(v) && length(v) > 0`, then uses it
- Show that `validate` is not accessible outside `outer_fn`
- Why is this pattern useful?


## Sol 74

```{r, error=TRUE}
#| output.lines: 3
outer_fn <- function(x) {
  validate <- function(v) is.numeric(v) && length(v) > 0
  if (!validate(x)) stop("invalid input")
  mean(x)
}
outer_fn(1:10)                                     # 5.5
outer_fn("abc")                                    # error
# validate is local, not in global env
# keeps global namespace clean; helper has access to
#     outer function's variables via lexical scoping
```


## Ex 75: `if`/`else` and `ifelse`

- Write scalar `classify(x)` returning "pos", "neg", or "zero" using `if/else`
- Vectorize: classify `c(-3,0,5,2,-1)` using `ifelse`
- Nested `ifelse` to map scores to grades: $\geq 90$ A, $\geq 80$ B, $\geq 70$ C, else F


## Sol 75

```{r}
#| output.lines: 3
classify <- function(x) {
  if (x > 0) "pos" else if (x < 0) "neg" else "zero"
}
classify(5); classify(-1); classify(0)
ifelse(c(-3,0,5,2,-1)>0, "pos",
  ifelse(c(-3,0,5,2,-1)<0, "neg", "zero"))
s <- c(95,82,71,65,90)
ifelse(s>=90,"A",ifelse(s>=80,"B",ifelse(s>=70,"C","F")))
```


## Ex 76: `switch()`

- Write `calc(op, a, b)` using `switch(op, "+"=a+b, "-"=a-b, "*"=a*b, "/"=a/b, stop("unknown"))`
- Test with all 4 operators and an invalid one
- Write `day_type(day)` classifying weekdays vs weekends


## Sol 76

```{r, error=TRUE}
#| output.lines: 3
calc <- function(op, a, b)
  switch(op, "+"=a+b, "-"=a-b, "*"=a*b, "/"=a/b,
         stop("unknown op"))
calc("+",3,4); calc("*",5,6); calc("/",10,3)
calc("^",2,3)  # error
day_type <- function(d) switch(d, Monday=,Tuesday=,
  Wednesday=,Thursday=,Friday="Weekday",
  Saturday=,Sunday="Weekend","Unknown")
day_type("Monday"); day_type("Sunday")
```


## Ex 77: `&&`/`||` vs `&`/`|`

- Show `&&` short-circuits: `FALSE && stop("boom")` doesn't error
- Show `&` doesn't short-circuit: `c(FALSE,TRUE) & c(TRUE,TRUE)` is vectorized
- Write safe null-check: `if (!is.null(x) && x > 0)` and explain why `&&` is required


## Sol 77

```{r}
#| output.lines: 3
FALSE && stop("boom")  # no error: short-circuit
c(FALSE,TRUE) & c(TRUE,TRUE)                      # vectorized
x <- NULL
if (!is.null(x) && x > 0) cat("pos") else cat("safe")
# && stops after !is.null(x) is FALSE; & would try x>0
```


## Ex 78: `for` Loop --- FizzBuzz

- Classic FizzBuzz 1-20: print "Fizz" if divisible by 3, "Buzz" if by 5, "FizzBuzz" if both, else the number
- Vectorized version using `ifelse` (no loop)
- Compare with `paste0` trick


## Sol 78

```{r}
#| output.lines: 3
for (i in 1:20) {
  if (i%%15==0) cat("FizzBuzz ")
  else if (i%%3==0) cat("Fizz ")
  else if (i%%5==0) cat("Buzz ")
  else cat(i, "")
}
cat("\n")
x <- 1:20
ifelse(x%%15==0,"FizzBuzz",ifelse(x%%3==0,"Fizz",
  ifelse(x%%5==0,"Buzz",as.character(x))))
```


## Ex 79: `while` Loop --- Newton's Method

Implement Newton's method for $\sqrt{a}$: $x_{n+1} = \frac{1}{2}(x_n + a/x_n)$.

- Write `my_sqrt(a, tol=1e-10)` using `while`
- Test on 2, 9, 100
- Count iterations needed for `a=2`


## Sol 79

```{r, eval=FALSE}
my_sqrt <- function(a, tol=1e-10) {
  x <- a; n <- 0
  while (TRUE) {
    x_new <- 0.5 * (x + a/x); n <- n + 1
    if (abs(x_new - x) < tol) break
    x <- x_new
  }
  list(root=x_new, iterations=n)
}
```


## Sol 79 Cont'd

```{r, include=FALSE}
my_sqrt <- function(a, tol=1e-10) {
  x <- a; n <- 0
  while (TRUE) {
    x_new <- 0.5 * (x + a/x); n <- n + 1
    if (abs(x_new - x) < tol) break
    x <- x_new
  }
  list(root=x_new, iterations=n)
}
```

```{r}
#| output.lines: 3
my_sqrt(2); my_sqrt(9)$root; my_sqrt(100)$root
my_sqrt(2)$iterations
```


## Ex 80: `repeat`/`break` --- Collatz

Collatz: if $n$ even, $n/2$; if odd, $3n+1$. Conjecture: always reaches 1.

- Write `collatz(n)` returning the sequence using `repeat`/`break`
- Find the starting number 1--100 with the longest sequence
- Plot the sequence for `n=27`


## Sol 80

```{r, eval=FALSE}
collatz <- function(n) {
  seq <- n
  repeat { if (n==1) break; n <- if(n%%2==0) n/2 else 3*n+1
           seq <- c(seq, n) }
  seq
}
lengths <- sapply(1:100, \(n) length(collatz(n)))
which.max(lengths)  # starting number
```


## Sol 80 Cont'd

```{r, include=FALSE}
collatz <- function(n) {
  seq <- n
  repeat { if (n==1) break; n <- if(n%%2==0) n/2 else 3*n+1
           seq <- c(seq, n) }
  seq
}
```

```{r, out.width="35%"}
#| output.lines: 2
lengths <- sapply(1:100, \(n) length(collatz(n)))
which.max(lengths)  # starting number
plot(collatz(27), type="l", main="Collatz(27)")
```


## Ex 81: `next` --- Filtered Sum

- Sum integers 1-100 that are NOT multiples of 7, using `for`+`next`
- Same with vectorized filtering (no loop)
- Sum only prime numbers 1-50. Write `is_prime(n)` helper


## Sol 81

```{r}
#| output.lines: 3
s <- 0; for(i in 1:100) {if(i%%7==0) next; s<-s+i}; s
sum((1:100)[(1:100)%%7 != 0])                      # same
is_prime <- function(n) {
  if (n < 2) return(FALSE)
  if (n == 2) return(TRUE)
  all(n %% 2:floor(sqrt(n)) != 0)
}
sum(Filter(is_prime, 1:50))
```


## Ex 82: Vectorization Speed

- Write loop version of `sum(x^2)` for `x <- 1:1e5`
- Compare `system.time` with vectorized `sum(x^2)`
- Pre-allocate vs grow-in-loop: time `v <- c(); for(...) v <- c(v,i)` vs `v <- numeric(n); for(...) v[i] <- i`


## Sol 82

```{r}
#| output.lines: 3
n <- 1e5; x <- 1:n
loop_sum <- function(x) {s<-0; for(v in x) s<-s+v^2; s}
system.time(loop_sum(x))                           # slow
system.time(sum(x^2))                              # fast
system.time({v<-c(); for(i in 1:n) v<-c(v,i)})    # very slow
system.time({v<-numeric(n); for(i in 1:n) v[i]<-i}) # faster
```


## Ex 83: Sieve of Eratosthenes

- Implement `sieve(n)` returning all primes up to `n` using a logical vector and nested loops
- Test with `sieve(50)`
- How many primes below 1000?


## Sol 83

```{r}
#| output.lines: 3
sieve <- function(n) {
  is_p <- rep(TRUE, n); is_p[1] <- FALSE
  for (i in 2:floor(sqrt(n)))
    if (is_p[i]) is_p[seq(i*2, n, i)] <- FALSE
  which(is_p)
}
sieve(50)
length(sieve(1000))                                # 168
```


## Ex 84: Recursive Factorial and Fibonacci

- Write `my_fact(n)` using `Recall`
- Write `fib(n)` using `Recall`
- Compute `sapply(1:10, fib)`
- Why is naive recursive `fib` slow for large `n`?


## Sol 84

```{r}
#| output.lines: 3
my_fact <- function(n) {
  if (n <= 1) return(1); n * Recall(n-1)
}
my_fact(5); my_fact(10)
fib <- function(n) {
  if (n <= 2) return(1); Recall(n-1) + Recall(n-2)
}
sapply(1:10, fib)
# exponential time: each call branches into 2
```


## Ex 85: `match.arg` and `missing`

- Write `summary_stat(x, type=c("mean","median","sd"))` using `match.arg`. Test partial matching
- Write `f(x, y)` that checks `missing(y)` and uses default `y=0`
- Show `nargs()` inside a function


## Sol 85

```{r}
#| output.lines: 3
summary_stat <- function(x, type=c("mean","median","sd")) {
  type <- match.arg(type)
  switch(type, mean=mean(x), median=median(x), sd=sd(x))
}
summary_stat(1:10, "mea")                          # mean
f <- function(x, y) {
  if (missing(y)) y <- 0; x + y
}
f(5); f(5, 3)
g <- function(a,b,c) nargs(); g(1); g(1,2,3)
```


## Ex 86: `invisible` and Side Effects

- Write `my_print(x)` that `cat`s and returns `invisible(x)`. Show it doesn't print the return
- Capture the invisible return
- Write `log_and_return(x)` that prints a message and returns `invisible(NULL)`


## Sol 86

```{r}
#| output.lines: 3
my_print <- function(x) { cat("Value:", x, "\n"); invisible(x) }
my_print(42)          # prints message, no [1] 42
val <- my_print(42); val                           # captured
log_return <- function(x) {
  cat("Processed:", x, "\n"); invisible(NULL)
}
r <- log_return(5); is.null(r)                     # TRUE
```


## Ex 87: `on.exit` with `add=TRUE`

- Write function that sets `options(digits=3)`, uses `on.exit` to restore
- Write function with two `on.exit(add=TRUE)` calls. Show both run
- Show cleanup runs even on error


## Sol 87

```{r, eval=FALSE}
show_digits <- function() {
  old <- options(digits=3)
  on.exit(options(old))
  cat("Inside:", getOption("digits"), "\n")
}
```


## Sol 87 Cont'd

```{r, include=FALSE}
show_digits <- function() {
  old <- options(digits=3)
  on.exit(options(old))
  cat("Inside:", getOption("digits"), "\n")
}
show_digits(); cat("Outside:", getOption("digits"), "\n")
two_exits <- function() {
  on.exit(cat("exit1\n"), add=TRUE)
  on.exit(cat("exit2\n"), add=TRUE)
  cat("body\n")
}
```

```{r, eval=FALSE}
show_digits(); cat("Outside:", getOption("digits"), "\n")
two_exits <- function() {
  on.exit(cat("exit1\n"), add=TRUE)
  on.exit(cat("exit2\n"), add=TRUE)
  cat("body\n")
}
two_exits()
tryCatch({
  f <- function() { on.exit(cat("cleanup!\n")); stop("err") }
  f()
}, error = function(e) cat("caught:", e$message, "\n"))
```

```{r, echo=FALSE}
#| output.lines: 1
two_exits()
tryCatch({
  f <- function() { on.exit(cat("cleanup!\n")); stop("err") }
  f()
}, error = function(e) cat("caught:", e$message, "\n"))
```

## Ex 88: `Vectorize()`

- Write scalar `classify(x)` using `if/else` (not vectorized)
- Use `Vectorize` to create vector version. Test on `c(-2, 0, 5)`
- Why is `Vectorize` just a wrapper around `mapply`?


## Sol 88

```{r}
#| output.lines: 3
classify <- function(x) {
  if (x > 0) "pos" else if (x < 0) "neg" else "zero"
}
classify_v <- Vectorize(classify)
classify_v(c(-2, 0, 5))
# Vectorize calls mapply internally, applying the
# scalar function element-wise. Not truly vectorized (still loops).
```


## Ex 89: Native Pipe `|>`

- Rewrite `round(mean(abs(c(-3,4,-1,2))), 2)` using `|>`
- Pipe `mtcars` through `subset` then `head`
- Use `_` placeholder: `mtcars |> lm(mpg ~ wt, data = _)`


## Sol 89

```{r}
#| output.lines: 3
c(-3,4,-1,2) |> abs() |> mean() |> round(2)
mtcars |> subset(cyl == 4) |> head(3)
mtcars |> lm(mpg ~ wt, data = _) |>
  coef()
```


## Ex 90: `do.call`

- Use `do.call(paste, c(as.list(c("a","b","c")), sep="-"))`
- Use `do.call(rbind, list_of_dfs)` to stack 3 data frames
- Build a call to `seq` dynamically from a list of arguments


## Sol 90

```{r}
#| output.lines: 3
do.call(paste, c(as.list(c("a","b","c")), sep="-"))
dfs <- list(data.frame(x=1,y=2), data.frame(x=3,y=4))
do.call(rbind, dfs)
do.call(seq, list(from=1, to=10, by=2))
```


## Ex 91: `tryCatch` and `try`

- Use `tryCatch(log("a"), error=function(e) NA)`
- Loop over `list(1,"a",4)`, wrapping each in `try`. Keep only successes
- Write `safe_div(a,b)` that `stop`s on `b==0` and test with `tryCatch`


## Sol 91

```{r}
#| output.lines: 3
tryCatch(log("a"), error=function(e) NA)
inputs <- list(1, "a", 4)
results <- lapply(inputs, \(x) try(sqrt(x), silent=TRUE))
Filter(\(r) !inherits(r,"try-error"), results)
safe_div <- function(a,b) {if(b==0) stop("div0!"); a/b}
tryCatch(safe_div(10,0), error=function(e) Inf)
```


## Ex 92: `stopifnot` and `warning`

- Write `my_mean(x)` with `stopifnot(is.numeric(x), length(x)>0)`
- Write `safe_sqrt(x)` that issues `warning` for negative input and returns `sqrt(abs(x))`
- Use `suppressWarnings` to silence it


## Sol 92

```{r, error=TRUE}
#| output.lines: 3
my_mean <- function(x) {
  stopifnot(is.numeric(x), length(x) > 0)
  sum(x)/length(x)
}
my_mean(1:5); my_mean("abc")
safe_sqrt <- function(x) {
  if (any(x < 0)) warning("Negative values!")
  sqrt(abs(x))
}
safe_sqrt(c(4, -9, 16))
suppressWarnings(safe_sqrt(-1))
```


## Ex 93: Power Factory

- Write `power_factory(n)` that returns a function computing $x^n$
- Create `square` and `cube` from it; test on 5
- Inspect the captured `n` with `environment(square)$n`
- Verify `square` and `cube` have different environments


## Sol 93

```{r}
#| output.lines: 3
power_factory <- function(n) function(x) x^n
square <- power_factory(2); cube <- power_factory(3)
square(5); cube(5)
environment(square)$n; environment(cube)$n          # 2; 3
identical(environment(square), environment(cube))   # FALSE
```


## Ex 94: Closure Counter

Write `make_counter()` returning `list(increment, get, reset)`.

- `increment()` adds 1 to internal `n` and returns it
- `get()` returns current `n`; `reset()` sets `n` back to 0
- Create two independent counters; show they don't share state
- Rewrite using `local()` for a single-instance counter


## Sol 94

```{r}
#| output.lines: 3
make_counter <- function() {
  n <- 0
  list(increment = function() { n <<- n + 1; n },
       get = function() n,
       reset = function() { n <<- 0 })
}
c1 <- make_counter(); c2 <- make_counter()
c1$increment(); c1$increment(); c2$increment()
cat("c1:", c1$get(), " c2:", c2$get(), "\n")
c1$reset(); c1$get()
```

```{r}
#| output.lines: 3
# local() single-instance
counter <- local({
  n <- 0
  list(increment = function() { n <<- n + 1; n },
       get = function() n)
})
counter$increment(); counter$increment(); counter$get()
```


## Ex 95: Loop Capture Bug

- Create 5 functions in a `for` loop: `funs[[i]] <- function() i`. Call all 5. What happens?
- Fix using `force()` inside a wrapper function
- Fix using `lapply`
- Explain *why* the bug occurs (lazy evaluation + shared environment)


## Sol 95

```{r, eval=FALSE}
funs <- list()
for (i in 1:5) funs[[i]] <- function() i
sapply(funs, \(f) f())                              # all 5!
# Bug: all closures share the same env; i=5 after loop
```


## Sol 95 Cont'd

```{r, include=FALSE}
funs <- list()
for (i in 1:5) funs[[i]] <- function() i
sapply(funs, \(f) f())                              # all 5!
# Bug: all closures share the same env; i=5 after loop
```

```{r}
#| output.lines: 3
make_fun <- function(val) { force(val); function() val }
funs <- list()
for (i in 1:5) funs[[i]] <- make_fun(i)
sapply(funs, \(f) f())

funs <- lapply(1:5, \(i) function() i)
sapply(funs, \(f) f())
# Lazy eval: function() i captures a promise to i,
# not its value. By call time, loop finished, i=5.
```

## Ex 96: Memoized Fibonacci

Write `make_memo_fib()` returning `list(fib, cache_size, clear_cache)`.

- Use a named list as cache; `fib(n)` checks cache before computing
- Time `fib(35)` first call vs second call
- How large is the cache after `fib(35)`?
- Clear cache and verify `fib(10)` still works


## Sol 96

```{r, eval=FALSE}
make_memo_fib <- function() {
  cache <- list()
  fib <- function(n) {
    key <- as.character(n)
    if (!is.null(cache[[key]])) return(cache[[key]])
    result <- if (n <= 2) 1 else fib(n-1) + fib(n-2)
    cache[[key]] <<- result; result
  }
  list(fib = fib,
       cache_size = function() length(cache),
       clear_cache = function() { cache <<- list() })
}
```


## Sol 96 Cont'd

```{r, include=FALSE}
make_memo_fib <- function() {
  cache <- list()
  fib <- function(n) {
    key <- as.character(n)
    if (!is.null(cache[[key]])) return(cache[[key]])
    result <- if (n <= 2) 1 else fib(n-1) + fib(n-2)
    cache[[key]] <<- result; result
  }
  list(fib = fib,
       cache_size = function() length(cache),
       clear_cache = function() { cache <<- list() })
}
```

```{r, eval=FALSE}
mf <- make_memo_fib()
system.time(mf$fib(35))                             # first
system.time(mf$fib(35))                             # second: instant
mf$cache_size()                                      # 35
mf$clear_cache(); mf$fib(10)
```

```{r, echo=FALSE}
#| output.lines: 3
mf <- make_memo_fib()
system.time(mf$fib(35))                             # first
system.time(mf$fib(35))                             # second: instant
mf$cache_size()                                      # 35
mf$clear_cache(); mf$fib(10)
```

## Ex 97: Decorator --- Logging Wrapper

Write `with_logging(f)` that wraps `f` to print call count on each invocation.

- Return a closure that increments a counter and calls `f(...)`
- Test: `logged_mean <- with_logging(mean)`. Call it 3 times
- Extend to also print elapsed time per call


## Sol 97

```{r, eval=FALSE}
with_logging <- function(f, fname = deparse(substitute(f))) {
  count <- 0L
  function(...) {
    count <<- count + 1L
    t0 <- proc.time()["elapsed"]
    cat(sprintf("[%s] call #%d\n", fname, count))
    result <- f(...)
    cat(sprintf("  elapsed: %.4fs\n",
        proc.time()["elapsed"] - t0))
    result
  }
}
```


## Sol 97 Cont'd

```{r, include=FALSE}
with_logging <- function(f, fname = deparse(substitute(f))) {
  count <- 0L
  function(...) {
    count <<- count + 1L
    t0 <- proc.time()["elapsed"]
    cat(sprintf("[%s] call #%d\n", fname, count))
    result <- f(...)
    cat(sprintf("  elapsed: %.4fs\n",
        proc.time()["elapsed"] - t0))
    result
  }
}
```

```{r}
#| output.lines: 3
logged_mean <- with_logging(mean)
logged_mean(1:100); logged_mean(rnorm(1e5))
```

## Ex 98: Bank Account with Transaction Log

Write `make_account(owner, balance)` returning `list(deposit, withdraw, statement, balance)`.

- `deposit(amount)` adds and logs. `withdraw(amount)` subtracts; `stop()` if insufficient
- `statement()` returns a data frame of all transactions
- Create two accounts; verify independence


## Sol 98

```{r, eval=FALSE}
make_account <- function(owner, init = 0) {
  bal <- init
  log <- data.frame(time=Sys.time()[0], type=character(0),
                    amount=numeric(0), balance=numeric(0))
  record <- function(type, amount) {
    log <<- rbind(log, data.frame(
      time=Sys.time(), type=type, amount=amount, balance=bal))
  }
  list(
    deposit = function(amt) {
      stopifnot(amt > 0); bal <<- bal + amt
      record("deposit", amt); invisible(bal)
```


## Sol 98 Cont'd

```{r, include=FALSE}
make_account <- function(owner, init = 0) {
  bal <- init
  log <- data.frame(time=Sys.time()[0], type=character(0),
                    amount=numeric(0), balance=numeric(0))
  record <- function(type, amount) {
    log <<- rbind(log, data.frame(
      time=Sys.time(), type=type, amount=amount, balance=bal))
  }
  list(
    deposit = function(amt) {
      stopifnot(amt > 0); bal <<- bal + amt
      record("deposit", amt); invisible(bal)
    },
    withdraw = function(amt) {
      stopifnot(amt > 0)
      if (amt > bal) stop("Insufficient funds")
      bal <<- bal - amt; record("withdraw", amt); invisible(bal)
    },
    balance = function() bal,
    statement = function() log)
}
```

```{r, eval=FALSE}
    balance = function() bal,
    statement = function() log)
}
alice <- make_account("Alice", 1000)
bob   <- make_account("Bob", 500)
alice$deposit(200); alice$withdraw(50); bob$deposit(100)
alice$balance(); bob$balance()
alice$statement()
```

```{r, echo=FALSE}
#| output.lines: 2
alice <- make_account("Alice", 1000)
bob   <- make_account("Bob", 500)
alice$deposit(200); alice$withdraw(50); bob$deposit(100)
alice$balance(); bob$balance()
alice$statement()
```

## Ex 99: Moving Average Filter

Write `make_ema(alpha)` returning a function that computes an exponential moving average: $\text{EMA}_t = \alpha x_t + (1-\alpha)\text{EMA}_{t-1}$.

- First call initializes EMA to the input value
- Test with prices `c(100, 102, 101, 105, 103, 107, 106)` and $\alpha=0.2$
- Create two EMAs with different $\alpha$; verify independence


## Sol 99

```{r}
#| output.lines: 3
make_ema <- function(alpha = 0.3) {
  value <- NULL
  function(x) {
    if (is.null(value)) value <<- x
    else value <<- alpha * x + (1 - alpha) * value
    value
  }
}
ema <- make_ema(0.2)
sapply(c(100, 102, 101, 105, 103, 107, 106), ema)
ema_fast <- make_ema(0.5); ema_slow <- make_ema(0.1)
sapply(c(100, 110, 90), ema_fast)
sapply(c(100, 110, 90), ema_slow)
```


## Ex 100: Fibonacci Iterator

Write `make_fib_iterator()` returning `list(next_val, reset)`.

- `next_val()` returns the next Fibonacci number on each call
- Use `replicate(12, fib_it$next_val())` to get first 12 values
- Reset and verify the sequence restarts


## Sol 100

```{r}
#| output.lines: 3
make_fib_iterator <- function() {
  a <- 0; b <- 1
  list(
    next_val = function() {
      val <- a; new <- a + b; a <<- b; b <<- new; val
    },
    reset = function() { a <<- 0; b <<- 1 })
}
fib_it <- make_fib_iterator()
replicate(12, fib_it$next_val())
fib_it$reset(); replicate(5, fib_it$next_val())
```


## Ex 101: Closure-Based ODE Solver (Euler Method)

Write `make_ode_solver(f, y0, dt)` where $f(t,y)$ defines $dy/dt$.

- `step()` advances one time step; `run(n)` returns trajectory as a data frame
- Solve $dy/dt = -0.5y$, $y(0)=10$, $dt=0.1$ for 100 steps
- Plot Euler vs exact $y=10e^{-0.5t}$


## Sol 101

```{r, out.width="40%", eval=FALSE}
make_ode_solver <- function(f, y0, dt = 0.01) {
  t <- 0; y <- y0
  list(
    run = function(n) {
      traj <- matrix(NA, n+1, 2, dimnames=list(NULL, c("t","y")))
      traj[1, ] <- c(t, y)
      for (i in seq_len(n)) {
        y <<- y + f(t, y)*dt; t <<- t + dt
        traj[i+1, ] <- c(t, y)
      }
      as.data.frame(traj)
    },
    state = function() c(t=t, y=y),
    reset = function(y0_new=y0) { t <<- 0; y <<- y0_new })
}
```


## Sol 101 Cont'd

```{r, include=FALSE}
make_ode_solver <- function(f, y0, dt = 0.01) {
  t <- 0; y <- y0
  list(
    run = function(n) {
      traj <- matrix(NA, n+1, 2, dimnames=list(NULL, c("t","y")))
      traj[1, ] <- c(t, y)
      for (i in seq_len(n)) {
        y <<- y + f(t, y)*dt; t <<- t + dt
        traj[i+1, ] <- c(t, y)
      }
      as.data.frame(traj)
    },
    state = function() c(t=t, y=y),
    reset = function(y0_new=y0) { t <<- 0; y <<- y0_new })
}
```

```{r, eval=FALSE}
solver <- make_ode_solver(\(t, y) -0.5*y, y0=10, dt=0.1)
traj <- solver$run(100)
plot(traj$t, traj$y, type='l', main='Euler vs Exact')
curve(10*exp(-0.5*x), add=TRUE, col='red', lty=2)
```

```{r, echo=FALSE, out.width="30%"}
solver <- make_ode_solver(\(t, y) -0.5*y, y0=10, dt=0.1)
traj <- solver$run(100)
plot(traj$t, traj$y, type='l', lwd=2, col='steelblue',
     xlab='t', ylab='y', main='Euler vs Exact')
curve(10*exp(-0.5*x), add=TRUE, col='red', lwd=2, lty=2)
legend('topright', c('Euler','Exact'),
       col=c('steelblue','red'), lty=1:2, lwd=2)
```

## Ex 102: Rate Limiter

Write `make_rate_limiter(f, max_calls, period)` wrapping `f` so at most `max_calls` invocations are allowed per `period`-second window.

- If limit exceeded, return `NULL` with a warning
- Track timestamps in a closure variable
- Test: limiter allowing 3 calls per 2 seconds; call 5 times rapidly


## Sol 102

```{r, eval=FALSE}
make_rate_limiter <- function(f, max_calls=3, period=2) {
  timestamps <- numeric(0)
  function(...) {
    now <- as.numeric(Sys.time())
    timestamps <<- timestamps[timestamps > now - period]
    if (length(timestamps) >= max_calls) {
      warning("Rate limit exceeded"); return(NULL)
    }
    timestamps <<- c(timestamps, now)
    f(...)
  }
}
```


## Sol 102 Cont'd

```{r, include=FALSE}
make_rate_limiter <- function(f, max_calls=3, period=2) {
  timestamps <- numeric(0)
  function(...) {
    now <- as.numeric(Sys.time())
    timestamps <<- timestamps[timestamps > now - period]
    if (length(timestamps) >= max_calls) {
      warning("Rate limit exceeded"); return(NULL)
    }
    timestamps <<- c(timestamps, now)
    f(...)
  }
}
```

```{r, eval=FALSE}
limited_sqrt <- make_rate_limiter(sqrt, 3, 2)
for (i in 1:5) { r <- limited_sqrt(i*10)
  cat(sprintf("Call %d: %s\n", i,
      if (is.null(r)) "BLOCKED" else round(r, 4))) }
```

```{r, echo=FALSE}
#| output.lines: 3
limited_sqrt <- make_rate_limiter(sqrt, 3, 2)
for (i in 1:5) { r <- limited_sqrt(i*10)
  cat(sprintf("Call %d: %s\n", i,
      if (is.null(r)) "BLOCKED" else round(r, 4))) }
```

## Ex 103: Portfolio Analytics Factory

Write `make_portfolio(weights, tickers)` returning:

- `returns(price_matrix)` --- weighted portfolio returns
- `sharpe(price_matrix, rf)` --- annualized Sharpe ratio
- `drawdown(price_matrix)` --- maximum drawdown
- Test with simulated 252-day prices for 3 assets


## Sol 103 (Function)

```{r, eval=FALSE}

make_portfolio <- function(w, tickers) {
  w <- w / sum(w); names(w) <- tickers
  list(
    returns = function(P) {
      r <- apply(P, 2, \(x) diff(x)/head(x,-1)); drop(r %*% w)
    },
    sharpe = function(P, rf=0) {
      r <- apply(P, 2, \(x) diff(x)/head(x,-1))
```


## Sol 103 (Function) Cont'd

```{r, eval=FALSE}
      pr <- drop(r %*% w)
      (mean(pr) - rf/252) / sd(pr) * sqrt(252)
    },
    drawdown = function(P) {
      r <- apply(P, 2, \(x) diff(x)/head(x,-1))
      cum <- cumprod(1 + drop(r %*% w))
      max(1 - cum / cummax(cum))
    })
}
```

## Sol 103 (Test)

```{r, include=FALSE}
make_portfolio <- function(w, tickers) {
  w <- w / sum(w); names(w) <- tickers
  list(
    returns = function(P) {
      r <- apply(P, 2, \(x) diff(x)/head(x,-1)); drop(r %*% w)
    },
    sharpe = function(P, rf=0) {
      r <- apply(P, 2, \(x) diff(x)/head(x,-1))
      pr <- drop(r %*% w)
      (mean(pr) - rf/252) / sd(pr) * sqrt(252)
    },
    drawdown = function(P) {
      r <- apply(P, 2, \(x) diff(x)/head(x,-1))
      cum <- cumprod(1 + drop(r %*% w))
      max(1 - cum / cummax(cum))
    })
}
```

```{r, out.width="40%"}
#| output.lines: 3
set.seed(42); P <- matrix(NA, 253, 3); P[1,] <- c(100,50,200)
for (i in 2:253) P[i,] <- P[i-1,]*exp(rnorm(3, 3e-4,
                                        c(.015,.02,.01)))
pf <- make_portfolio(c(.4,.3,.3), c("A","G","M"))
cat("Sharpe:", round(pf$sharpe(P), 3), "\n")
cat("MaxDD:", round(pf$drawdown(P)*100, 2), "%\n")
r <- pf$returns(P)
plot(cumprod(1+r), type='l', lwd=2, col='steelblue',
     main='Portfolio Growth', ylab='Growth of $1')
abline(h=1, lty=2, col='grey50')
```


## Ex 104: `by()` and Full EDA Report

Using `mtcars`:

- Use `by(mtcars, mtcars$cyl, \(d) cor(d$mpg, d$wt))` to compute per-cylinder correlation
- Mean, sd, min, max of `mpg` by `cyl` in one `aggregate` call; expand the matrix column
- Merge with group counts from `aggregate(..., FUN=length)`
- Sort by mean `mpg` descending


## Sol 104

```{r}
#| output.lines: 3
by(mtcars, mtcars$cyl, \(d) cor(d$mpg, d$wt))
res <- aggregate(mpg ~ cyl, data=mtcars,
  FUN=\(x) c(mean=mean(x), sd=sd(x), min=min(x), max=max(x)))
flat <- cbind(res[1], as.data.frame(res$mpg))

counts <- aggregate(mpg ~ cyl, data=mtcars, FUN=length)
names(counts)[2] <- 'n'
flat <- merge(flat, counts, by='cyl')
flat[order(-flat$mean), ]
```


# The Apply Family

## `apply()` --- Over Matrix / Data Frame Margins

```{r}
#| output.lines: 3
m <- matrix(1:12, nrow = 3,
            dimnames = list(paste0('r',1:3), paste0('c',1:4)))
m
apply(m, 1, sum)     # MARGIN=1: rows
apply(m, 2, sum)     # MARGIN=2: columns
```

---

## `apply()` --- When FUN Returns a Vector

```{r}
#| output.lines: 3
# when FUN returns a vector, result is k x n matrix
apply(m, 2, range)
apply(m, 2, quantile, probs = c(0.25, 0.75))
```

---

## `apply()` on Data Frames --- Beware of Coercion!

```{r}
#| output.lines: 3
df <- data.frame(x = 1:3, y = c('a','b','c'), z = 4:6)
apply(df, 2, class)       # all "character"!
```

- `apply()` converts data frames to matrices first --- **all columns become character** if any is character. Use `sapply`/`lapply` for mixed types.

```{r, eval=FALSE}
# faster alternatives for row/column sums/means:
colSums(m); rowSums(m); colMeans(m); rowMeans(m)
```

---

## `lapply()` and `sapply()`

```{r}
#| output.lines: 3
x <- list(a = 1:5, b = rnorm(10), c = c(TRUE, FALSE, TRUE))
lapply(x, mean)      # always returns a list
sapply(x, mean)      # simplifies to named vector
```

```{r}
#| output.lines: 3
# sapply can be unpredictable; vapply is safer
vapply(x, mean, numeric(1))
```

---

## `sapply()` Is Unpredictable --- Use `vapply()` in Production

```{r}
#| output.lines: 3
# when all lengths equal: matrix. When they differ: list!
sapply(list(a = 1:3, b = 4:6), identity)
sapply(list(a = 1:3, b = 4:5), identity)
```

```{r}
#| output.lines: 3
# vapply: explicit type template; error if mismatch
vapply(iris[,1:4], median, numeric(1))
```

---

## `tapply()` --- Apply by Groups

```{r}
#| output.lines: 3
# mean mpg by cylinder count
tapply(mtcars$mpg, mtcars$cyl, mean)
```

```{r}
#| output.lines: 3
# two-way grouping -> matrix
tapply(mtcars$mpg, list(mtcars$cyl, mtcars$am), mean)
```

- `tapply(x, f, FUN)` $\equiv$ `sapply(split(x, f), FUN)`

---

## `mapply()` and `Map()`

```{r}
#| output.lines: 3
# apply to corresponding elements of multiple vectors
mapply(function(x, y) x + y, 1:5, 10:14)
mapply(rep, 1:3, 3:1)   # variable-length results
```

```{r}
#| output.lines: 3
# Map: always returns a list (no simplification)
Map('+', 1:3, 10:12)
```

## `mapply()` and `Map()` Cont'd


```{r}
#| output.lines: 3
# MoreArgs: constant arguments not varied across elements
mapply(function(x, y, suffix) paste0(x, y, suffix),
       x = c('Alice', 'Bob'), y = c(25, 30),
       MoreArgs = list(suffix = '!'))
```

---

## `rapply()` --- Recursive Apply

```{r}
#| output.lines: 2
nested <- list(a = 1:3,
  b = list(c = 4:6, d = list(e = 7:9, f = 'text')))
rapply(nested, sum, classes = 'integer')         # unlist
rapply(nested, sum, classes = 'integer', how = 'list')
rapply(nested, \(x) x * 10, classes = 'integer', how = 'replace')
```

---

## `Reduce()` --- Sequential Accumulation

```{r}
#| output.lines: 3
Reduce('+', 1:5)                          # 15
Reduce('+', 1:5, accumulate = TRUE)       # running total
```

```{r}
#| output.lines: 3
# merge multiple data frames
dfs <- list(data.frame(id=1:3, a=10:12),
            data.frame(id=2:4, b=20:22),
            data.frame(id=1:3, c=30:32))
Reduce(\(x, y) merge(x, y, by='id', all=TRUE), dfs)
```

---

## `Filter()`, `Find()`, `Position()`, `Negate()`

```{r}
#| output.lines: 3
x <- list(1:3, "hello", 4:6, TRUE, NULL)
Filter(is.numeric, x)          # keep numeric elements
Find(is.character, x)          # first character element
Position(is.character, x)      # its position
```

```{r}
#| output.lines: 3
Filter(\(x) x %% 15 == 0, 1:50)   # multiples of 15
```

```{r}
#| output.lines: 3
# Find/Position with direction
Find(\(x) x > 5, c(2, 8, 3, 9, 1), right = TRUE)
Position(\(x) x > 5, c(2, 8, 3, 9, 1), right = TRUE)
```

---

## Decision Table: Choosing the Right Tool

\begin{table}[!htbp]
\centering\footnotesize
\begin{tabular}{lllp{3cm}}
\toprule
Function & Input & Output & Use case \\
\midrule
\comm{apply} & matrix/array & vector/matrix & row/column ops \\
\comm{lapply} & vector/list & list & general iteration \\
\comm{sapply} & vector/list & simplified & interactive only \\
\comm{vapply} & vector/list & typed output & production code \\
\comm{tapply} & vector + factor & array & group summaries \\
\comm{mapply/Map} & multiple vectors & simplified/list & parallel iteration \\
\comm{rapply} & nested list & various & recursive ops \\
\comm{Reduce} & vector/list & single value & sequential fold \\
\comm{Filter} & vector/list & subset & keep matching \\
\comm{Find/Position} & vector/list & one element & first match \\
\comm{Negate} & function & function & invert predicate \\
\bottomrule
\end{tabular}
\end{table}

---

## Common Equivalences

```{r, eval=FALSE}
tapply(x, f, FUN)         <==>  sapply(split(x, f), FUN)
sapply(X, FUN)            <==>  unlist(lapply(X, FUN))   # scalar FUN
mapply(FUN, x, y)         <==>  sapply(seq_along(x),
                                  \(i) FUN(x[i], y[i]))
Map(FUN, x, y)            <==>  mapply(FUN, x, y, SIMPLIFY = FALSE)
Reduce('+', 1:5)          <==>  ((((1+2)+3)+4)+5)
Filter(f, x)              <==>  x[sapply(x, f)]
aggregate(y~g, d, mean)   <==>  tapply(d$y, d$g, mean)  # approx
```

## Ex 105: `apply` --- Row and Column

`m <- matrix(c(3,1,4,1,5,9,2,6,5,3,5,8), nrow=3)`.

- Column means via `apply`
- Row ranges
- Apply custom function returning a named vector per column
- Apply on a data frame's numeric columns


## Sol 105

```{r}
#| output.lines: 3
m <- matrix(c(3,1,4,1,5,9,2,6,5,3,5,8), nrow=3)
apply(m, 2, mean)
apply(m, 1, range)                                 # 2x3 matrix
apply(m, 2, \(x) c(mu=mean(x), sd=sd(x)))
apply(mtcars[1:3, 1:4], 2, mean)
```


## Ex 106: `lapply` and `sapply`

`l <- list(a=1:10, b=seq(0,1,by=0.1), c=c(T,F,T,T,F))`.

- `sapply(l, length)`
- `sapply(l, range)` — what shape?
- `vapply(l, mean, numeric(1))`
- Why prefer `vapply` over `sapply`?


## Sol 106

```{r}
#| output.lines: 3
l <- list(a=1:10, b=seq(0,1,by=0.1), c=c(T,F,T,T,F))
sapply(l, length)                                  # named vector
sapply(l, range)                                   # 2x3 matrix
vapply(l, mean, numeric(1))                        # named vector
# vapply guarantees return type; sapply may return list
# unexpectedly if elements have different lengths
```


## Ex 107: `tapply`

- Mean `mpg` by `cyl` in `mtcars`
- Two-way: mean `mpg` by `cyl` × `am`
- Use `tapply` on `ToothGrowth`: mean `len` by `supp` and `dose`


## Sol 107

```{r}
#| output.lines: 3
tapply(mtcars$mpg, mtcars$cyl, mean)
tapply(mtcars$mpg, list(mtcars$cyl, mtcars$am), mean)
tapply(ToothGrowth$len,
  list(ToothGrowth$supp, ToothGrowth$dose), mean)
```


## Ex 108: `mapply` and `Map`

- `mapply(function(x,y) x+y, 1:5, 10:14)`
- `mapply(rep, 1:4, 4:1)`
- `Map("+", 1:3, 10:12)` — what type?
- Use `Map` to apply `paste` element-wise to two vectors


## Sol 108

```{r}
#| output.lines: 3
mapply(function(x,y) x+y, 1:5, 10:14)
mapply(rep, 1:4, 4:1)                              # list
Map("+", 1:3, 10:12)                               # list
Map(paste, c("a","b","c"), 1:3, sep="-")
```


## Ex 109: `Reduce` Applications

- Factorial via `Reduce("*", 1:6)`
- Cumulative products via `accumulate=TRUE`
- Find the common elements of 4 vectors using `Reduce(intersect, ...)`


## Sol 109

```{r}
#| output.lines: 3
Reduce("*", 1:6)                                   # 720
Reduce("*", 1:6, accumulate=TRUE)
Reduce(intersect, list(1:10, 3:12, 5:15, 7:20))   # 7:10
```


## Ex 110: Functional Pipeline

Process `x <- c(-5, 3, -1, 8, -2, 7)`:

- Filter positive → square → sum. Do with nested calls
- Same with pipe `|>`
- Same with `Reduce` composing 3 functions


## Sol 110

```{r}
#| output.lines: 3
x <- c(-5, 3, -1, 8, -2, 7)
sum(x[x>0]^2)                                     # nested
x[x>0] |> (\(v) v^2)() |> sum()                    # pipe
fns <- list(\(v) v[v>0], \(v) v^2, sum)
Reduce(\(v,f) f(v), fns, init=x)
```


## Ex 111: `tapply` + `aggregate` Pipeline

Using `iris`:

- Mean `Sepal.Length` by `Species` with `tapply`
- Same with `aggregate`
- Mean of all numeric columns by `Species` with `aggregate(. ~ Species, ...)`


## Sol 111

```{r}
#| output.lines: 3
tapply(iris$Sepal.Length, iris$Species, mean)
aggregate(Sepal.Length ~ Species, data=iris, mean)
aggregate(. ~ Species, data=iris, mean)
```


## Ex 112: Memoization via Closure

- Write `make_memo(fn)` that returns a memoized version storing results in a list environment
- Memoize `fib`
- Show second call is instant


## Sol 112

```{r}
#| output.lines: 3
make_memo <- function(fn) {
  cache <- list()
  function(...) {
    key <- paste(..., sep=",")
    if (is.null(cache[[key]])) cache[[key]] <<- fn(...)
    cache[[key]]
  }
}
slow_fn <- function(x) { Sys.sleep(0.1); x^2 }
fast_fn <- make_memo(slow_fn)
system.time(fast_fn(5))                            # slow
system.time(fast_fn(5))                            # instant
```


## Ex 113: `lapply` with Index

- Given `l <- list(a=1:3, b=4:6)`, use `lapply(seq_along(l), ...)` to create `"a: 1 2 3"`, `"b: 4 5 6"`
- Same with `Map` and `names(l)`
- Same with `mapply`


## Sol 113

```{r}
#| output.lines: 2
l <- list(a=1:3, b=4:6)
lapply(seq_along(l), \(i)
  paste0(names(l)[i], ": ", paste(l[[i]], collapse=" ")))
Map(\(nm, v) paste0(nm, ": ", paste(v,collapse=" ")),
    names(l), l)
mapply(\(nm,v) paste0(nm,": ",paste(v,collapse=" ")),
       names(l), l)
```


## Ex 114: Nested `lapply`

- Generate a 3×3 multiplication table using `lapply(1:3, \(i) sapply(1:3, \(j) i*j))`
- Convert to a matrix with `do.call(rbind, ...)`
- Compare with `outer(1:3, 1:3)`


## Sol 114

```{r}
#| output.lines: 3
l <- lapply(1:3, \(i) sapply(1:3, \(j) i*j))
do.call(rbind, l)
outer(1:3, 1:3)                                    # same
```


## Ex 115: `Map` for Parallel Processing

`prices <- list(c(100,102,98), c(50,55,52), c(200,210,205))`.
`names_v <- c("AAPL", "GOOG", "MSFT")`.

- Compute return `diff(p)/p[-length(p)]` for each using `Map`
- Compute mean return per stock


## Sol 115

```{r}
#| output.lines: 3
prices <- list(c(100,102,98), c(50,55,52), c(200,210,205))
names_v <- c("AAPL","GOOG","MSFT")
returns <- Map(\(p) diff(p)/p[-length(p)], prices)
names(returns) <- names_v
sapply(returns, mean)
```


## Ex 116: `Filter` + `Reduce` Combo

- From `1:100`, filter multiples of 3, then sum with `Reduce`
- From a list of vectors, keep only those with length > 3, then find their common intersection
- Compare with direct `sum(Filter(...))`


## Sol 116

```{r}
#| output.lines: 3
Reduce("+", Filter(\(x) x%%3==0, 1:100))
vecs <- list(1:10, 3:8, 2:15, 5:12, 4:20)
long <- Filter(\(v) length(v)>3, vecs)
Reduce(intersect, long)                            # 5:8
sum(Filter(\(x) x%%3==0, 1:100))                    # same as (a)
```


# String Operations

## Basic String Functions

```{r}
#| output.lines: 3
nchar('hello world')
toupper('hello'); tolower('HELLO')
trimws('  hello  '); trimws('  hello  ', which = 'left')
```

## Basic String Functions Cont'd


```{r}
#| output.lines: 3
substr('abcdefgh', 3, 5)
startsWith('hello world', 'hello')
endsWith('hello world', 'world')
chartr('abc', 'ABC', 'a big cat')
```

---

## `paste()` and `paste0()`

```{r}
#| output.lines: 3
paste('a', 'b', 'c', sep = '-')
paste0('item', 1:5)
```

```{r}
#| output.lines: 3
(labels <- paste(c('a', 'b'), 1:6, sep = '-'))
```

```{r}
#| output.lines: 3
# collapse: join multiple strings into one
paste('item', 1:5, sep = '_', collapse = ', ')
```

---

## `sprintf()` --- Formatted Strings

```{r}
#| output.lines: 3
# %d: integer; %.2f: 2 decimal places; %s: string
sprintf('Name: %s, Age: %d, Score: %.2f', 'Alice', 25, 95.678)
```

```{r}
#| output.lines: 3
sprintf('%05d', 42)         # zero-padded
sprintf('%.2f%%', 12.345)   # percentage
sprintf('%-10s|', 'left')   # left-align
```

---

## `cat()` --- Print to Console

```{r}
#| output.lines: 3
v <- c(20, 30, 3 * pi)
cat("The content of v is:", v, "...\n")
```

```{r}
#| output.lines: 3
for (i in 50 * (0:2)) {
  cat(sprintf('exp(sin(%3d)) = %.6f', i, exp(sin(i))), '\n')
}
```

---

## `strsplit()` --- Split Strings

```{r}
#| output.lines: 2
strsplit('a.b.c.d', split = '\\.')
strsplit('hello world foo', split = ' ')
strsplit(c('a-b', 'c-d-e'), split = '-')
strsplit('hello', split = '')   # every character
```

---

## `grep()`, `grepl()` --- Pattern Matching

```{r}
#| output.lines: 3
x <- c('apple', 'banana', 'cherry', 'apricot', 'blueberry')
grep('an', x)                                  # indices
grep('an', x, value = TRUE)                   # values
grepl('an', x)                                  # logical
grep('APPLE', x, ignore.case = TRUE, value = TRUE)
```

---

## `sub()`, `gsub()` --- Pattern Replacement

```{r}
#| output.lines: 3
x <- 'Hello World Hello'
sub('Hello', 'Hi', x)      # replace FIRST match
gsub('Hello', 'Hi', x)     # replace ALL matches
```

```{r}
#| output.lines: 3
gsub('[0-9]', '', 'abc123def456')      # remove all digits
gsub(' +', ' ', 'too   many    spaces')  # compress spaces
```

---

## `regexpr()`, `gregexpr()`, `regmatches()`

```{r}
#| output.lines: 3
x <- 'Phone: 02-1234-5678, Fax: 04-8765-4321'
m <- gregexpr('[0-9]+-[0-9]+-[0-9]+', x)
regmatches(x, m)    # extract all matches
```

## Ex 117: Basic String Functions

`x <- c("  Hello World  ", "R Programming", "data SCIENCE")`.

- Trim whitespace; count chars of each
- Convert to uppercase, lowercase
- `startsWith` / `endsWith` tests
- Extract first 4 chars using `substr`


## Sol 117

```{r}
#| output.lines: 3
x <- c("  Hello World  ","R Programming","data SCIENCE")
xt <- trimws(x); nchar(xt)
toupper(xt); tolower(xt)
startsWith(xt, "Hello"); endsWith(xt, "SCIENCE")
substr(xt, 1, 4)
```


## Ex 118: `paste` and `sprintf`

- `paste("item", 1:5, sep="_")`
- Same with `collapse=", "`
- `sprintf("Name: %s, Score: %.1f", "Alice", 95.678)`
- Zero-pad: `sprintf("%03d", 1:5)`


## Sol 118

```{r}
#| output.lines: 3
paste("item", 1:5, sep="_")
paste("item", 1:5, sep="_", collapse=", ")
sprintf("Name: %s, Score: %.1f", "Alice", 95.678)
sprintf("%03d", 1:5)
```


## Ex 119: `strsplit`

- Split `"a.b.c.d"` by `"."` (note the escape!)
- Split `"hello world foo"` by space
- Split each character of `"ABCDE"`
- Split `c("a-b", "c-d-e")` — what shape is the result?


## Sol 119

```{r}
#| output.lines: 3
strsplit("a.b.c.d", "\\.")
strsplit("hello world foo", " ")
strsplit("ABCDE", "")
strsplit(c("a-b","c-d-e"), "-")                    # list of 2
```


# Regular Expressions

## What Are Regular Expressions?

- Patterns for matching text. Used in `grep`, `grepl`, `sub`, `gsub`, `regexpr`, `strsplit`, etc.
- In R strings, backslash must be **doubled**: write `\\d` for the regex `\d`.

---

## Basic Metacharacters

\begin{table}[!htbp]
\centering\small
\begin{tabular}{ll}
\toprule
Pattern & Meaning \\
\midrule
\comm{.} & any single character (except newline) \\
\comm{\textasciicircum} & start of string \\
\comm{\$} & end of string \\
\comm{*} & 0 or more of preceding \\
\comm{+} & 1 or more of preceding \\
\comm{?} & 0 or 1 of preceding \\
\comm{\{n\}}, \comm{\{n,m\}} & exactly \comm{n}; between \comm{n} and \comm{m} \\
\comm{|} & alternation (or) \\
\comm{()} & grouping and capture \\
\comm{[]} & character class \\
\comm{\textbackslash} & escape next character \\
\bottomrule
\end{tabular}
\end{table}

---

## Character Classes

\begin{table}[!htbp]
\centering\small
\begin{tabular}{ll}
\toprule
Class & Meaning \\
\midrule
\comm{[abc]} & matches a, b, or c \\
\comm{[\textasciicircum abc]} & matches anything except a, b, c \\
\comm{[a-z]}, \comm{[A-Z]}, \comm{[0-9]} & ranges \\
\comm{\textbackslash d} / \comm{\textbackslash D} & digit / non-digit \\
\comm{\textbackslash w} / \comm{\textbackslash W} & word char / non-word char \\
\comm{\textbackslash s} / \comm{\textbackslash S} & whitespace / non-whitespace \\
\comm{\textbackslash b} & word boundary \\
\bottomrule
\end{tabular}
\end{table}

```{r}
#| output.lines: 3
x <- c('hello123', 'world', '456', 'test!', 'a1b2')
grep('[0-9]', x, value = TRUE)       # contains a digit
grep('^[0-9]+$', x, value = TRUE)    # all digits
```

---

## Anchors, Quantifiers, Alternation

```{r}
#| output.lines: 3
x <- c('cat', 'concatenate', 'scat', 'catalog')
grep('^cat', x, value = TRUE)     # starts with 'cat'
grep('cat$', x, value = TRUE)     # ends with 'cat'
```

```{r}
#| output.lines: 3
x <- c('cat', 'dog', 'catfish', 'dogfish', 'bird')
grep('cat|dog', x, value = TRUE)       # cat OR dog
grep('^(cat|dog)$', x, value = TRUE)   # exactly cat or dog
```

---

## Capture Groups and Backreferences

```{r}
#| output.lines: 3
# swap first and last name
x <- c('John Smith', 'Jane Doe', 'Bob Lee')
sub('(\\w+) (\\w+)', '\\2, \\1', x)
```

```{r}
#| output.lines: 3
# extract parts of a date
dates <- c('2024-01-15', '2023-12-31')
sub('(\\d{4})-(\\d{2})-(\\d{2})', 'Year:\\1, Month:\\2', dates)
```

---

## Greedy vs Lazy Matching

```{r}
#| output.lines: 3
x <- '<b>bold</b> and <i>italic</i>'
regmatches(x, gregexpr('<.*>', x))     # greedy
regmatches(x, gregexpr('<.*?>', x))    # lazy (non-greedy)
```

---

## Lookahead and Lookbehind (Perl)

```{r}
#| output.lines: 3
x <- c('$100', '200', '$300', '400')
regmatches(x, gregexpr('(?<=\\$)\\d+', x, perl = TRUE))
```

```{r}
#| output.lines: 3
# negative lookahead: files NOT ending in .txt
x <- c('file.txt', 'image.png', 'doc.txt', 'photo.jpg')
grep('\\.(?!txt)', x, value = TRUE, perl = TRUE)
```

---

## `regexec()` --- Full Capture Group Extraction

```{r}
#| output.lines: 3
x <- c('2024-01-15 14:30', '2023-12-31 09:00')
pattern <- '(\\d{4})-(\\d{2})-(\\d{2}) (\\d{2}):(\\d{2})'
m <- regexec(pattern, x)
regmatches(x, m)
```

---

## Practical Regex Examples

```{r}
#| output.lines: 3
# validate email (simplified)
emails <- c('user@example.com', 'bad@@email',
            'test@domain.org', 'no_at_sign')
grep('^[\\w.]+@[\\w.]+\\.[a-z]{2,}$', emails,
     value = TRUE, perl = TRUE)
```

```{r}
#| output.lines: 3
# clean column names: replace non-alphanumeric with _
col_names <- c('First Name', 'Last.Name', 'Age (years)')
gsub('[^a-zA-Z0-9]', '_', col_names)
```

## Ex 120: `grep` and `grepl`

`w <- c("statistics","computation","algorithm","distribution","estimation")`.

- Words containing `"tion"`
- Words starting with vowel
- Words of $\geq$ 12 characters
- `grepl` to create logical mask


## Sol 120

```{r}
#| output.lines: 3
w <- c("statistics","computation","algorithm",
       "distribution","estimation")
grep("tion", w, value=TRUE)
grep("^[aeiou]", w, value=TRUE)
w[nchar(w) >= 12]
grepl("tion", w)
```


## Ex 121: `sub` and `gsub`

- Replace first space with `"_"` in `"hello world foo"`
- Replace all spaces
- Remove all digits from `"abc123def456"`
- Compress multiple spaces to one


## Sol 121

```{r}
#| output.lines: 3
sub(" ", "_", "hello world foo")                   # first only
gsub(" ", "_", "hello world foo")                  # all
gsub("[0-9]", "", "abc123def456")
gsub(" +", " ", "too   many    spaces")
```


## Ex 122: Capture Groups

- Swap first/last name: `sub("(\\w+) (\\w+)", "\\2, \\1", "John Smith")`
- Extract domain from email: `sub(".*@", "", "user@example.com")`
- Extract year from dates using capture group


## Sol 122

```{r}
#| output.lines: 3
sub("(\\w+) (\\w+)", "\\2, \\1",
    c("John Smith","Jane Doe"))
sub(".*@", "", c("user@example.com","a@b.org"))
sub("(\\d{4})-\\d{2}-\\d{2}", "\\1",
    c("2024-01-15","2023-12-31"))                  # year
```


## Ex 123: `regexpr` and `gregexpr`

`x <- "Phone: 02-1234-5678, Fax: 04-8765-4321"`.

- Extract all phone numbers with `gregexpr` + `regmatches`
- Extract first match only with `regexpr`
- Extract all words with `\\b\\w+\\b`


## Sol 123

```{r}
#| output.lines: 3
x <- "Phone: 02-1234-5678, Fax: 04-8765-4321"
m <- gregexpr("[0-9]+-[0-9]+-[0-9]+", x)
regmatches(x, m)
m1 <- regexpr("[0-9]+-[0-9]+-[0-9]+", x)
regmatches(x, m1)
regmatches(x, gregexpr("\\b\\w+\\b", x))
```


## Ex 124: Greedy vs Lazy

`x <- "<b>bold</b> and <i>italic</i>"`.

- `gregexpr("<.*>", x)` + `regmatches` — what matches? (greedy)
- `gregexpr("<.*?>", x)` — what matches? (lazy)
- Explain the difference


## Sol 124

```{r}
#| output.lines: 3
x <- "<b>bold</b> and <i>italic</i>"
regmatches(x, gregexpr("<.*>", x))                 # one big match
regmatches(x, gregexpr("<.*?>", x))                # 4 tags
# greedy: .* matches as much as possible (all the way
#     to last >); lazy: .*? matches as little as possible
```


## Ex 125: Anchors and Quantifiers

`x <- c("cat","concatenate","scat","catalog")`.

- Starts with `cat`
- Ends with `cat`
- Exactly `cat`
- `c("a","ab","abb","abbb")`: match `ab+` vs `ab*` vs `ab?`


## Sol 125

```{r}
#| output.lines: 3
x <- c("cat","concatenate","scat","catalog")
grep("^cat", x, value=TRUE)
grep("cat$", x, value=TRUE)
grep("^cat$", x, value=TRUE)
y <- c("a","ab","abb","abbb")
grep("ab+", y, value=TRUE); grep("ab*", y, value=TRUE)
```


## Ex 126: POSIX Classes

`x <- c("Hello!", "123", "abc", "A1!")`.

- Contains punctuation
- All digits
- Contains only alphabetic
- All alphanumeric


## Sol 126

```{r}
#| output.lines: 3
x <- c("Hello!", "123", "abc", "A1!")
grep("[[:punct:]]", x, value=TRUE)
grep("^[[:digit:]]+$", x, value=TRUE)
grep("^[[:alpha:]]+$", x, value=TRUE)
grep("^[[:alnum:]]+$", x, value=TRUE)
```


## Ex 127: Lookahead/Lookbehind

- Extract numbers preceded by `$`: `c("$100","200","$300")`
- Find files NOT ending in `.txt`: `c("a.txt","b.png","c.txt","d.jpg")`
- Explain why `perl=TRUE` is needed


## Sol 127

```{r}
#| output.lines: 3
x <- c("$100","200","$300")
regmatches(x, gregexpr("(?<=\\$)\\d+", x, perl=TRUE))
y <- c("a.txt","b.png","c.txt","d.jpg")
grep("\\.(?!txt$)", y, value=TRUE, perl=TRUE)
# lookahead/lookbehind are Perl-compatible regex
# extensions not available in base (POSIX) regex
```


## Ex 128: `regexec` --- Structured Extraction

`dates <- c("2024-01-15 14:30", "2023-12-31 09:00")`.

- Write pattern to capture year, month, day, hour, minute separately
- Extract with `regexec` + `regmatches`
- Build a data frame from the results


## Sol 128

```{r}
#| output.lines: 3
dates <- c("2024-01-15 14:30", "2023-12-31 09:00")
pat <- "(\\d{4})-(\\d{2})-(\\d{2}) (\\d{2}):(\\d{2})"
m <- regexec(pat, dates)
parts <- regmatches(dates, m)
do.call(rbind, lapply(parts, \(p)
  setNames(p[-1], c("Y","M","D","h","m"))))
```


## Ex 129: camelCase ↔ snake_case

- Write `camel_to_snake(s)` converting `"camelCase"` → `"camel_case"`
- Write `snake_to_camel(s)` converting `"snake_case"` → `"snakeCase"`
- Test both on `"myHTTPRequest"` / `"my_http_request"`


## Sol 129

```{r}
#| output.lines: 3
camel_to_snake <- function(s) tolower(gsub("([A-Z])","_\\1",s))
camel_to_snake("camelCase")
camel_to_snake("myHTTPRequest")
snake_to_camel <- function(s) {
  parts <- strsplit(s, "_")[[1]]
  paste0(parts[1], paste0(toupper(substring(parts[-1],1,1)),
         substring(parts[-1],2), collapse=""))
}
snake_to_camel("my_http_request")
```


## Ex 130: Clean Column Names

`cols <- c("First Name", "Last.Name", "Age (years)", "Income$")`.

- Replace all non-alphanumeric chars with `_`
- Remove leading/trailing `_`
- Collapse consecutive `_` to single `_`
- Convert to lowercase


## Sol 130

```{r}
#| output.lines: 3
cols <- c("First Name","Last.Name","Age (years)","Income$")
x <- gsub("[^a-zA-Z0-9]", "_", cols)
x <- gsub("^_+|_+$", "", x)
x <- gsub("_+", "_", x)
tolower(x)
```


# Date and Time

## Date Objects

```{r}
#| output.lines: 3
today <- Sys.Date(); today; class(today)
```

```{r}
#| output.lines: 3
as.Date('2024-01-15')
as.Date('15/01/2024', format = '%d/%m/%Y')
as.Date('Jan 15, 2024', format = '%b %d, %Y')
```

```{r}
#| output.lines: 3
today + 30
as.Date('2024-12-31') - as.Date('2024-01-01')
```

---

## Date Formatting

```{r}
#| output.lines: 3
d <- as.Date('2024-07-15')
format(d, '%Y'); format(d, '%B'); format(d, '%A')
format(d, '%d %b %Y')
weekdays(d); months(d); quarters(d)
```

---

## Date-Time: `POSIXct` and `POSIXlt`

```{r}
#| output.lines: 3
now <- Sys.time(); now; class(now)
as.POSIXct('2024-01-15 14:30:00', tz = 'Asia/Taipei')
```

```{r}
#| output.lines: 3
# POSIXlt: list-based access to components
lt <- as.POSIXlt(now)
lt$hour; lt$min; lt$wday    # 0=Sunday
```

---

## `difftime`, Sequences, and Format Codes

```{r}
#| output.lines: 3
d1 <- as.Date('2024-01-01'); d2 <- as.Date('2024-12-31')
difftime(d2, d1, units = 'weeks')
seq(as.Date('2024-01-01'), as.Date('2024-06-01'), by = 'month')
```

\begin{table}[!htbp]
\centering\small
\begin{tabular}{llll}
\toprule
Code & Meaning & Code & Meaning \\
\midrule
\comm{\%Y} & 4-digit year & \comm{\%H} & hour (00--23) \\
\comm{\%m} & month (01--12) & \comm{\%M} & minute (00--59) \\
\comm{\%d} & day (01--31) & \comm{\%S} & second (00--59) \\
\comm{\%B}/\comm{\%b} & month name/abbrev & \comm{\%A}/\comm{\%a} & weekday name/abbrev \\
\bottomrule
\end{tabular}
\end{table}

---

## `strftime()` and `strptime()`

```{r}
#| output.lines: 3
strftime(Sys.time(), '%Y-%m-%d %H:%M:%S %Z')
strptime('15-Jan-2024 14:30', format = '%d-%b-%Y %H:%M')
```

## Ex 131: Date Arithmetic

- Create today's date with `Sys.Date()`. Add 30 days
- Days between `2024-01-01` and `2024-12-31`
- Generate monthly sequence Jan--Jun 2024
- Extract weekday name and month name from a date


## Sol 131

```{r}
#| output.lines: 3
today <- Sys.Date(); today + 30
as.Date("2024-12-31") - as.Date("2024-01-01")     # 365
seq(as.Date("2024-01-01"), as.Date("2024-06-01"),
    by="month")
d <- as.Date("2024-07-15")
weekdays(d); months(d); format(d, "%A %B %d, %Y")
```


## Ex 132: `POSIXct` and Timezones

- Create `as.POSIXct("2024-01-15 14:30:00", tz="Asia/Taipei")`
- Extract hour, minute via `POSIXlt`
- Difference between two timestamps in hours
- Format with `strftime`


## Sol 132

```{r}
#| output.lines: 3
t1 <- as.POSIXct("2024-01-15 14:30:00", tz="Asia/Taipei")
lt <- as.POSIXlt(t1); lt$hour; lt$min
t2 <- as.POSIXct("2024-01-15 18:45:00", tz="Asia/Taipei")
difftime(t2, t1, units="hours")
strftime(t1, "%Y/%m/%d %H:%M %Z")
```


# File I/O

## Reading and Writing Text Files

```{r, eval=FALSE}
df <- read.csv('data.csv')
df <- read.csv('data.csv', header = TRUE, stringsAsFactors = FALSE)
df <- read.delim('data.tsv')               # tab-separated
df <- read.table('data.txt', header = TRUE, sep = '|')
```

```{r, eval=FALSE}
write.csv(df, 'output.csv', row.names = FALSE)
write.table(df, 'output.tsv', sep = '\t',
            row.names = FALSE, quote = FALSE)
```

---

## Reading and Writing R Objects

```{r, eval=FALSE}
# save multiple R objects (binary, preserves types)
save(x, y, file = 'my_data.RData')
load('my_data.RData')

# save/load a single object
saveRDS(df, 'my_df.rds')
df <- readRDS('my_df.rds')
```

---

## `readLines`, `writeLines`, `scan`

```{r, eval=FALSE}
lines <- readLines('file.txt')
lines <- readLines('file.txt', n = 10)     # first 10 lines
writeLines(c('line 1', 'line 2'), 'out.txt')
nums  <- scan('numbers.txt')               # numeric
words <- scan('words.txt', what = '')       # character
```

---

## `readline()` and `interactive()`

```{r, eval=FALSE}
name <- readline(prompt = "Enter your name: ")
cat("Hello,", name, "\n")
n <- as.numeric(readline("Enter a number: "))
```

```{r, eval=FALSE}
if (interactive()) {
  cat("Running interactively\n")
} else {
  cat("Running in batch mode\n")
}
```

## Ex 133: Binary I/O

- `charToRaw("Hello")` — what type?
- Convert back with `rawToChar`
- `as.integer(charToRaw("ABC"))` — what are the byte values?
- Write raw bytes to a temp file with `writeBin` and read back


## Sol 133

```{r}
#| output.lines: 3
r <- charToRaw("Hello"); r; typeof(r)             # raw
rawToChar(r)                                       # "Hello"
as.integer(charToRaw("ABC"))                       # 65 66 67
tmp <- tempfile()
writeBin(charToRaw("test123"), tmp)
rawToChar(readBin(tmp, "raw", 20))
```


## Ex 134: `readline` and `interactive`

- Write a function that uses `readline` to ask for user's name (use `eval=FALSE`)
- Use `interactive()` to branch: interactive → readline, batch → default
- Why does `readline` always return character?


## Sol 134

```{r, eval=FALSE}
#| output.lines: 3
ask_name <- function() {
  if (interactive()) {
    name <- readline("Your name: ")
  } else {
    name <- "default_user"
  }
  cat("Hello,", name, "\n")
}
# readline reads raw text from console; R has no way
# to know the intended type, so it always returns character
```


# Error Handling

## `tryCatch()` --- Structured Error Handling

```{r}
#| output.lines: 3
# note: tryCatch catches the FIRST condition raised
result <- tryCatch({
  log('a')      # error: non-numeric argument
}, warning = function(w) {
  cat("Warning:", conditionMessage(w), "\n"); NA
}, error = function(e) {
  cat("Error:", conditionMessage(e), "\n"); NA
}, finally = {
  cat("(cleanup code runs always)\n")
})
result
```

---

## `try()` --- Simple Error Handling

```{r}
#| output.lines: 3
result <- try(log('a'), silent = TRUE)
inherits(result, 'try-error')
```

```{r}
#| output.lines: 3
# useful in loops: skip failing inputs
inputs <- list(1, 'a', 4, NULL, 9)
results <- list()
for (i in seq_along(inputs)) {
  results[[i]] <- try(sqrt(inputs[[i]]), silent = TRUE)
}
Filter(\(r) !inherits(r, 'try-error'), results)
```

---

## Common Error-Handling Idioms

```{r}
#| output.lines: 3
# silently return NULL on failure
safe_parse <- function(x) {
  tryCatch(as.numeric(x), warning = function(w) NULL,
                          error   = function(e) NULL)
}
safe_parse("3.14"); safe_parse("abc"); safe_parse(NULL)
```

```{r}
#| output.lines: 3
# wrap risky operations with a default value
safe_log <- function(x, default = NA) {
  tryCatch(log(x), error = function(e) default)
}
sapply(list(10, -1, 'a', NULL), safe_log)
```

---

## `stop()`, `warning()`, `message()`

```{r, error=TRUE}
#| output.lines: 3
safe_div <- function(a, b) {
  if (b == 0) stop("Division by zero!")
  a / b
}
safe_div(10, 2); safe_div(10, 0)
```

```{r}
#| output.lines: 3
check_val <- function(x) {
  if (x < 0) warning("Negative value!")
  sqrt(abs(x))
}
check_val(-4)
```

---

## `stopifnot()` --- Assertions

```{r, error=TRUE}
#| output.lines: 3
my_mean <- function(x) {
  stopifnot(is.numeric(x), length(x) > 0)
  sum(x) / length(x)
}
my_mean(1:5)
my_mean('hello')
```

## Ex 135: `system2`

- Run `system2("echo", args=c("hello","world"))`
- Capture output with `stdout=TRUE`
- Run in background with `wait=FALSE`
- Cross-platform check with `.Platform$OS.type`


## Sol 135

```{r, eval=FALSE}
#| output.lines: 3
system2("echo", args=c("hello","world"))
out <- system2("date", stdout=TRUE); out
system2("sleep", args="1", wait=FALSE)
if (.Platform$OS.type == "unix") {
  system2("uname", args="-a", stdout=TRUE)
}
```


# Useful Utilities

## `do.call()`

```{r}
#| output.lines: 3
dfs <- list(
  data.frame(x = 1:2, y = 3:4),
  data.frame(x = 5:6, y = 7:8),
  data.frame(x = 9:10, y = 11:12))
do.call(rbind, dfs)
```

```{r}
#| output.lines: 3
args <- list(1:10, na.rm = TRUE)
do.call(mean, args)
```

---

## `Vectorize()` --- Scalar to Vector

```{r}
#| output.lines: 3
f <- function(x, y) {
  if (x > y) x - y else x + y
}
f_vec <- Vectorize(f)
f_vec(1:3, 2)
```

---

## `invisible()` --- Silent Return

```{r}
#| output.lines: 3
# invisible: returns value without printing
f <- function(x) invisible(x^2)
f(5)                 # nothing printed
y_val <- f(5); y_val # but value is captured
```

```{r}
#| output.lines: 3
# common pattern: side-effect functions return invisible(NULL)
my_print <- function(x) {
  cat("Value:", x, "\n")
  invisible(NULL)
}
result <- my_print(42); is.null(result)
```

---

## `on.exit()` --- Guaranteed Cleanup

```{r, eval=FALSE}
safe_read <- function(path) {
  con <- file(path, 'r')
  on.exit(close(con))       # guaranteed cleanup
  readLines(con)
}
```

```{r, eval=FALSE}
# add = TRUE: stack multiple on.exit calls
process <- function(path) {
  tmp <- tempfile()
  on.exit(unlink(tmp), add = TRUE)      # cleanup 1
  con <- file(path, 'r')
  on.exit(close(con), add = TRUE)       # cleanup 2
}
```

---

## Benchmarking: `system.time()` and `proc.time()`

```{r}
#| output.lines: 3
system.time({
  x <- matrix(rnorm(1e6), 1000, 1000)
  y_mat <- crossprod(x)
})
```

```{r}
#| output.lines: 3
pt <- proc.time()
x <- sum(1:1e7)
proc.time() - pt
```

---

## `system2()` --- Run External Commands

```{r, eval=FALSE}
system2("echo", args = c("hello", "world"))
out <- system2("date", stdout = TRUE); cat(out, "\n")
system2("sleep", args = "5", wait = FALSE)     # background
```

```{r, eval=FALSE}
# cross-platform branching
if (.Platform$OS.type == "windows") {
  system2("taskkill", args = c("/F", "/IM", "myapp.exe"),
          stdout = NULL, stderr = NULL)
} else {
  system2("pkill", args = "myapp",
          stdout = NULL, stderr = NULL)
}
```

# Advanced Topics

## S3 Object System Basics

```{r}
#| output.lines: 3
person <- list(name = 'Alice', age = 30)
class(person) <- 'Person'
print.Person <- function(x, ...) {
  cat(sprintf("Person: %s (age %d)\n", x$name, x$age))
}
person
```

---

## Formula Objects

```{r}
#| output.lines: 3
f <- y ~ x1 + x2
class(f); typeof(f); all.vars(f)
```

```{r, eval=FALSE}
y ~ x                 # simple regression
y ~ x1 + x2           # multiple regression
y ~ x1 * x2           # with interaction (= x1 + x2 + x1:x2)
y ~ .                  # all other columns
y ~ poly(x, 3)        # polynomial
y ~ log(x)            # transformed
```

---

## Linear Regression Example

```{r, out.width="40%"}
#| output.lines: 3
model <- lm(dist ~ speed, data = cars)
model$coefficients
plot(cars$speed, cars$dist, pch = 19)
abline(model, col = 'red', lwd = 2)
```

---

## Random Number Generation

```{r}
#| output.lines: 3
set.seed(42)
runif(5); rnorm(5, mean = 0, sd = 1)
rbinom(10, size = 1, prob = 0.5)
rpois(10, lambda = 3)
rexp(5, rate = 1)
```


## Ex 136: S3 Classes

- Create S3 class `Circle`: `list(radius=r)` with `class="Circle"`
- Write `print.Circle` and `area.Circle`
- Create instance; dispatch `print` and `area`


## Sol 136

```{r}
#| output.lines: 3
Circle <- function(r) structure(list(radius=r), class="Circle")
print.Circle <- function(x, ...)
  cat(sprintf("Circle(r=%.2f)\n", x$radius))
area <- function(x, ...) UseMethod("area")
area.Circle <- function(x, ...) pi * x$radius^2
c1 <- Circle(5); c1
area(c1)
```


## Ex 137: Formula and `lm`

- `f <- y ~ x`; show `class()`, `typeof()`, `all.vars()`
- Fit `lm(dist ~ speed, data=cars)`. Extract coefficients
- Plot data with regression line


## Sol 137

```{r}
#| output.lines: 2
f <- y ~ x; class(f); typeof(f); all.vars(f)
model <- lm(dist ~ speed, data=cars)
coef(model)
```


## Sol 137 Cont'd

```{r, out.width="45%", echo=FALSE}
model <- lm(dist ~ speed, data=cars)
plot(cars$speed, cars$dist, pch=19)
abline(model, col="red", lwd=2)
```


## Ex 138: Plotting

- Plot `sin(x)` and `cos(x)` for `x` from 0 to $2\pi$ with legend
- `par(mfrow=c(1,2))`: histogram and boxplot side by side
- Add annotation with `text()` and horizontal line with `abline(h=...)`


## Sol 138

```{r, out.width="35%"}
#| output.lines: 3
x <- seq(0, 2*pi, length=100)
plot(x, sin(x), type="l", col="blue", lwd=2, ylab="y")
lines(x, cos(x), col="red", lwd=2, lty=2)
legend("topright", c("sin","cos"), col=c("blue","red"),
       lty=1:2, lwd=2)
```


## Ex 139: Random Number Generation

- `set.seed(42)` then generate 5 uniform, 5 normal, 10 Bernoulli
- Simulate 1000 dice rolls and tabulate
- Estimate $\pi$ via Monte Carlo: sample `(x,y)` uniform in $[0,1]^2$, count proportion inside unit circle quadrant


## Sol 139

```{r}
#| output.lines: 3
set.seed(42)
runif(5); rnorm(5); rbinom(10, 1, 0.5)
table(sample(1:6, 1000, replace=TRUE))
n <- 1e5; x <- runif(n); y <- runif(n)
4 * mean(x^2 + y^2 <= 1)  # ≈ π
```


## Ex 140: Options and Benchmarking

- Save current `digits` option, set to 15, print `pi`, restore
- Time `crossprod(x)` for `x <- matrix(rnorm(1e6),1000)` with `system.time`
- Use `proc.time` for manual timing of `sum(1:1e7)`


## Sol 140

```{r}
#| output.lines: 3
old <- options(digits=15); pi
options(old)
system.time({
  x <- matrix(rnorm(1e6), 1000, 1000)
  y <- crossprod(x)
})
pt <- proc.time(); s <- sum(1:1e7); proc.time()-pt
```



## Ex 141: Binary Search

Write `binary_search(x, target)` that returns the index of `target` in sorted vector `x`, or `NA` if not found. Use a `while` loop with `lo`/`hi` pointers.

- Test: `binary_search(c(2,5,8,12,16,23,38,56,72,91), 23)` → 6
- Test: `binary_search(c(2,5,8,12,16,23,38,56,72,91), 10)` → NA
- What is the time complexity?


## Sol 141

```{r}
#| output.lines: 3
binary_search <- function(x, target) {
  lo <- 1; hi <- length(x)
  while (lo <= hi) {
    mid <- (lo + hi) %/% 2
    if (x[mid] == target) return(mid)
    if (x[mid] < target) lo <- mid + 1 else hi <- mid - 1
  }
  NA
}
binary_search(c(2,5,8,12,16,23,38,56,72,91), 23)
binary_search(c(2,5,8,12,16,23,38,56,72,91), 10)
# O(log n) time complexity
```


## Ex 142: Merge Sort

Implement `merge_sort(x)` recursively:

- Base case: length 1 → return as-is
- Split in half, sort each half, merge
- Write `merge_two(a, b)` helper that merges two sorted vectors
- Test: `merge_sort(c(38, 27, 43, 3, 9, 82, 10))`


## Sol 142

```{r, eval=FALSE}
merge_two <- function(a, b) {
  res <- numeric(length(a) + length(b))
  i <- j <- k <- 1
  while (i <= length(a) && j <= length(b)) {
    if (a[i] <= b[j]) { res[k] <- a[i]; i <- i+1 }
    else { res[k] <- b[j]; j <- j+1 }
    k <- k + 1
  }
  if (i <= length(a)) res[k:(k+length(a)-i)] <- a[i:length(a)]
  if (j <= length(b)) res[k:(k+length(b)-j)] <- b[j:length(b)]
  res
}
```


## Sol 142 Cont'd

```{r, include=FALSE}
merge_two <- function(a, b) {
  res <- numeric(length(a) + length(b))
  i <- j <- k <- 1
  while (i <= length(a) && j <= length(b)) {
    if (a[i] <= b[j]) { res[k] <- a[i]; i <- i+1 }
    else { res[k] <- b[j]; j <- j+1 }
    k <- k + 1
  }
  if (i <= length(a)) res[k:(k+length(a)-i)] <- a[i:length(a)]
  if (j <= length(b)) res[k:(k+length(b)-j)] <- b[j:length(b)]
  res
}
merge_sort <- function(x) {
  if (length(x) <= 1) return(x)
  mid <- length(x) %/% 2
  merge_two(merge_sort(x[1:mid]),
            merge_sort(x[(mid+1):length(x)]))
}
```

```{r, eval=FALSE}
merge_sort <- function(x) {
  if (length(x) <= 1) return(x)
  mid <- length(x) %/% 2
  merge_two(merge_sort(x[1:mid]),
            merge_sort(x[(mid+1):length(x)]))
}
merge_sort(c(38, 27, 43, 3, 9, 82, 10))
```

```{r, echo=FALSE}
#| output.lines: 3
merge_sort(c(38, 27, 43, 3, 9, 82, 10))
```

## Ex 143: Function Composition

Write `compose(...)` that takes $n$ functions and returns their composition: `compose(f, g, h)(x)` = `f(g(h(x)))`.

- Test: `compose(sqrt, abs, sum)(c(-4, -5, -7))` → 4
- Compose `toupper`, `trimws`, `paste0` to build a string cleaner
- Show it works with `Reduce`


## Sol 143

```{r}
#| output.lines: 3
compose <- function(...) {
  fns <- rev(list(...))
  function(x) Reduce(\(v, f) f(v), fns, init = x)
}
compose(sqrt, abs, sum)(c(-4, -5, -7))  # sqrt(16) = 4
clean <- compose(toupper, trimws)
clean("  hello world  ")
```


## Ex 144: Currying

Write `curry(f, ...)` that partially applies arguments to `f`, returning a new function that accepts the remaining arguments.

- `add <- function(a, b) a + b`; `add5 <- curry(add, 5)`; `add5(3)` → 8
- `curry(paste, sep = "-")("a", "b", "c")` → "a-b-c"
- Use currying to create `log2 <- curry(log, base = 2)`


## Sol 144

```{r}
#| output.lines: 3
curry <- function(f, ...) {
  captured <- list(...)
  function(...) do.call(f, c(captured, list(...)))
}
add <- function(a, b) a + b
add5 <- curry(add, 5); add5(3)
dash_paste <- curry(paste, sep = "-")
dash_paste("a", "b", "c")
log2 <- curry(log, base = 2); log2(8)
```


## Ex 145: Stack Data Structure

Implement a stack using closures with operations: `push(x)`, `pop()`, `peek()`, `size()`, `is_empty()`.

- Push 10, 20, 30; peek → 30; pop → 30; size → 2
- Handle `pop()` on empty stack gracefully (return NULL + warning)
- Use the stack to reverse a string


## Sol 145

```{r, eval=FALSE}
make_stack <- function() {
  items <- list()
  list(
    push = function(x) items[[length(items)+1]] <<- x,
    pop = function() {
      if (length(items) == 0) {
        warning("empty stack"); return(NULL) }
      val <- items[[length(items)]]
      items[[length(items)]] <<- NULL; val },
    peek = function() items[[length(items)]],
    size = function() length(items),
    is_empty = function() length(items) == 0)
}
```


## Sol 145 Cont'd

```{r, include=FALSE}
make_stack <- function() {
  items <- list()
  list(
    push = function(x) items[[length(items)+1]] <<- x,
    pop = function() {
      if (length(items) == 0) {
        warning("empty stack"); return(NULL) }
      val <- items[[length(items)]]
      items[[length(items)]] <<- NULL; val },
    peek = function() items[[length(items)]],
    size = function() length(items),
    is_empty = function() length(items) == 0)
}
```

```{r, eval=FALSE}
s <- make_stack()
s$push(10); s$push(20); s$push(30)
s$peek(); s$pop(); s$size()
# reverse string
chars <- strsplit("Hello", "")[[1]]
rs <- make_stack()
for (ch in chars) rs$push(ch)
paste0(replicate(rs$size(), rs$pop()), collapse = "")
```

```{r, echo=FALSE}
#| output.lines: 2
s <- make_stack()
s$push(10); s$push(20); s$push(30)
s$peek(); s$pop(); s$size()
# reverse string
chars <- strsplit("Hello", "")[[1]]
rs <- make_stack()
for (ch in chars) rs$push(ch)
paste0(replicate(rs$size(), rs$pop()), collapse = "")
```

## Ex 146: Queue Data Structure

Implement a queue (FIFO) using closures: `enqueue(x)`, `dequeue()`, `front()`, `size()`.

- Enqueue "A", "B", "C"; dequeue → "A"; front → "B"
- Use the queue to implement breadth-first traversal of a nested list
- What is the time complexity of your `dequeue`? How could it be improved?


## Sol 146

```{r, eval=FALSE}
make_queue <- function() {
  items <- list()
  list(
    enqueue = function(x)
      items[[length(items)+1]] <<- x,
    dequeue = function() {
      val <- items[[1]]; items[[1]] <<- NULL; val },
    front = function() items[[1]],
    size = function() length(items),
    is_empty = function() length(items) == 0)
}
```


## Sol 146 Cont'd

```{r, include=FALSE}
make_queue <- function() {
  items <- list()
  list(
    enqueue = function(x)
      items[[length(items)+1]] <<- x,
    dequeue = function() {
      val <- items[[1]]; items[[1]] <<- NULL; val },
    front = function() items[[1]],
    size = function() length(items),
    is_empty = function() length(items) == 0)
}
```

```{r, eval=FALSE}
q <- make_queue()
q$enqueue("A"); q$enqueue("B"); q$enqueue("C")
q$dequeue(); q$front(); q$size()
# dequeue is O(n) due to list shifting
# improvement: use two-stack or index-based approach
```

```{r, echo=FALSE}
#| output.lines: 3
q <- make_queue()
q$enqueue("A"); q$enqueue("B"); q$enqueue("C")
q$dequeue(); q$front(); q$size()
# dequeue is O(n) due to list shifting
# improvement: use two-stack or index-based approach
```

## Ex 147: Matrix Chain Multiplication Order

Given dimensions of $n$ matrices, find minimum scalar multiplications using dynamic programming. Matrices $A_i$ has dimension $p_{i-1} \times p_i$.

- Write `mcm(p)` for dimension vector `p` of length $n+1$
- Test: `mcm(c(10, 30, 5, 60))` → 4500
- Return both the minimum cost and the optimal split


## Sol 147

```{r, eval=FALSE}
mcm <- function(p) {
  n <- length(p) - 1
  m <- matrix(0, n, n)
  for (len in 2:n)
    for (i in 1:(n - len + 1)) {
      j <- i + len - 1; m[i,j] <- Inf
      for (k in i:(j-1))
        m[i,j] <- min(m[i,j],
          m[i,k] + m[k+1,j] + p[i]*p[k+1]*p[j+1])
    }
  m[1, n]
}
```


## Sol 147 Cont'd

```{r, include=FALSE}
mcm <- function(p) {
  n <- length(p) - 1
  m <- matrix(0, n, n)
  for (len in 2:n)
    for (i in 1:(n - len + 1)) {
      j <- i + len - 1; m[i,j] <- Inf
      for (k in i:(j-1))
        m[i,j] <- min(m[i,j],
          m[i,k] + m[k+1,j] + p[i]*p[k+1]*p[j+1])
    }
  m[1, n]
}
```

```{r}
#| output.lines: 3
mcm(c(10, 30, 5, 60))         # 4500
mcm(c(40, 20, 30, 10, 30))    # 26000
```

## Ex 148: Bisection Method

Write `bisection(f, a, b, tol=1e-8)` that finds a root of $f$ in $[a, b]$.

- Require $f(a) \cdot f(b) < 0$ (else `stop`)
- Test: root of $x^3 - x - 2$ near $[1, 2]$
- Compare with `uniroot` result


## Sol 148

```{r, eval=FALSE}
bisection <- function(f, a, b, tol = 1e-8) {
  stopifnot(f(a) * f(b) < 0)
  n <- 0
  while (b - a > tol) {
    mid <- (a + b) / 2; n <- n + 1
    if (f(mid) == 0) return(list(root=mid, iter=n))
    if (f(a) * f(mid) < 0) b <- mid else a <- mid
  }
  list(root = (a + b) / 2, iter = n)
}
```


## Sol 148 Cont'd

```{r, include=FALSE}
bisection <- function(f, a, b, tol = 1e-8) {
  stopifnot(f(a) * f(b) < 0)
  n <- 0
  while (b - a > tol) {
    mid <- (a + b) / 2; n <- n + 1
    if (f(mid) == 0) return(list(root=mid, iter=n))
    if (f(a) * f(mid) < 0) b <- mid else a <- mid
  }
  list(root = (a + b) / 2, iter = n)
}
```

```{r}
#| output.lines: 3
f <- function(x) x^3 - x - 2
bisection(f, 1, 2)
uniroot(f, c(1, 2))$root
```


## Ex 149: Run-Length Encoding

Write `my_rle(x)` that returns a list with `lengths` and `values` (like `rle()`).

- Test: `my_rle(c(1,1,1,2,2,3,3,3,3,1))` → lengths: 3,2,4,1; values: 1,2,3,1
- Write `my_inverse_rle(r)` that reconstructs the original vector
- Verify: `my_inverse_rle(my_rle(x))` equals `x` for various inputs


## Sol 149

```{r}
#| output.lines: 2
my_rle <- function(x) {
  n <- length(x); vals <- x[1]; lens <- 1L; k <- 1
  for (i in 2:n) {
    if (x[i] == vals[k]) lens[k] <- lens[k] + 1L
    else { k <- k+1; vals[k] <- x[i]; lens[k] <- 1L }
  }
  list(lengths = lens, values = vals)
}
my_inverse_rle <- function(r) rep(r$values, r$lengths)
x <- c(1,1,1,2,2,3,3,3,3,1)
my_rle(x)
identical(my_inverse_rle(my_rle(x)), x)
```


## Ex 150: Environment Inspector

Write `env_info(fn)` that prints the enclosing environment chain of function `fn` (up to `.GlobalEnv` or `emptyenv()`).

- Test on a closure created by `make_counter()`
- Show `environment(fn)` differs from `parent.env(environment(fn))`
- List all objects in the closure's enclosing environment using `ls(envir = ...)`


## Sol 150

```{r, eval=FALSE}
env_info <- function(fn) {
  e <- environment(fn)
  while (!identical(e, emptyenv())) {
    cat(environmentName(e), ": ",
        paste(ls(e), collapse=", "), "\n")
    e <- parent.env(e)
    if (identical(e, globalenv())) {
      cat("(global)\n"); break }
  }
}
```


## Sol 150 Cont'd

```{r, include=FALSE}
env_info <- function(fn) {
  e <- environment(fn)
  while (!identical(e, emptyenv())) {
    cat(environmentName(e), ": ",
        paste(ls(e), collapse=", "), "\n")
    e <- parent.env(e)
    if (identical(e, globalenv())) {
      cat("(global)\n"); break }
  }
}
make_counter <- function() {
  n <- 0
  list(inc = function() { n <<- n+1; n })
}
```

```{r, eval=FALSE}
make_counter <- function() {
  n <- 0
  list(inc = function() { n <<- n+1; n })
}
ctr <- make_counter()
env_info(ctr$inc)
ls(envir = environment(ctr$inc))
```

```{r, echo=FALSE}
#| output.lines: 3
ctr <- make_counter()
env_info(ctr$inc)
ls(envir = environment(ctr$inc))
```

## Ex 151: Trampoline --- Tail-Call Optimization

R has no TCO. Implement a trampoline: a loop that keeps calling thunks (zero-arg functions) until a non-function result is returned.

- Write `trampoline(f, ...)`
- Rewrite `factorial` in trampolined form
- Show it can handle `tramp_fact(10000)` without stack overflow


## Sol 151

```{r}
#| output.lines: 3
trampoline <- function(f, ...) {
  result <- f(...)
  while (is.function(result)) result <- result()
  result
}
tramp_fact <- function(n, acc = 1) {
  if (n <= 1) return(acc)
  function() tramp_fact(n - 1, acc * n)
}
trampoline(tramp_fact, 10)    # 3628800
# trampoline(tramp_fact, 10000) works without stack overflow
# (result is Inf due to numeric overflow, but no crash)
```


## Ex 152: Retry with Exponential Backoff

Write `retry(fn, max_attempts = 3, base_delay = 0.1)` that:

- Calls `fn()`; on error, waits `base_delay * 2^(attempt-1)` seconds and retries
- Returns the result on success, or re-throws the last error after `max_attempts`
- Test with a function that fails randomly: `function() { if (runif(1) < 0.7) stop("fail"); 42 }`


## Sol 152

```{r}
#| output.lines: 3
retry <- function(fn, max_attempts = 3, base_delay = 0.1) {
  for (i in seq_len(max_attempts)) {
    result <- tryCatch(fn(), error = function(e) e)
    if (!inherits(result, "error")) return(result)
    if (i < max_attempts)
      Sys.sleep(base_delay * 2^(i - 1))
  }
  stop(result$message)
}
set.seed(42)
flaky <- function() { if (runif(1) < 0.5) stop("fail"); 42 }
retry(flaky, max_attempts = 5)
```


## Ex 153: Linked List

Implement a singly linked list using nested lists: `node(value, next_node)`.

- Write `ll_push(ll, val)`, `ll_to_vector(ll)`, `ll_length(ll)`, `ll_reverse(ll)`
- Build list 1→2→3, reverse it, convert to vector
- Why are linked lists rarely used in R?


## Sol 153

```{r, eval=FALSE}
node <- function(val, nxt = NULL) list(val = val, nxt = nxt)
ll_push <- function(ll, val) node(val, ll)
ll_length <- function(ll) {
  n <- 0; while (!is.null(ll)) { n <- n+1; ll <- ll$nxt }; n
}
ll_to_vector <- function(ll) {
  v <- c(); while(!is.null(ll)) { v<-c(v,ll$val); ll<-ll$nxt }; v
}
```


## Sol 153 Cont'd

```{r, include=FALSE}
node <- function(val, nxt = NULL) list(val = val, nxt = nxt)
ll_push <- function(ll, val) node(val, ll)
ll_length <- function(ll) {
  n <- 0; while (!is.null(ll)) { n <- n+1; ll <- ll$nxt }; n
}
ll_to_vector <- function(ll) {
  v <- c(); while(!is.null(ll)) { v<-c(v,ll$val); ll<-ll$nxt }; v
}
ll_reverse <- function(ll) {
  prev <- NULL
  while (!is.null(ll)) {
    nxt <- ll$nxt; ll$nxt <- prev; prev <- ll; ll <- nxt }
  prev
}
```

```{r, eval=FALSE}
ll_reverse <- function(ll) {
  prev <- NULL
  while (!is.null(ll)) {
    nxt <- ll$nxt; ll$nxt <- prev; prev <- ll; ll <- nxt }
  prev
}
ll <- Reduce(\(acc, x) ll_push(acc, x), 3:1, init = NULL)
ll_to_vector(ll)                    # 1 2 3
ll_to_vector(ll_reverse(ll))       # 3 2 1
# R copies on modify; linked lists lose their O(1) advantage
```

```{r, echo=FALSE}
#| output.lines: 2
ll <- Reduce(\(acc, x) ll_push(acc, x), 3:1, init = NULL)
ll_to_vector(ll)                    # 1 2 3
ll_to_vector(ll_reverse(ll))       # 3 2 1
# R copies on modify; linked lists lose their O(1) advantage
```

## Ex 154: Input Validator Factory

Write `make_validator(...)` that takes named validation rules and returns a function checking all rules.

- Rules: `list(name = is.character, age = \(x) is.numeric(x) && x > 0)`
- Validator returns `TRUE` if all pass, else a character vector of failures
- Test with valid and invalid inputs


## Sol 154

```{r, eval=FALSE}
make_validator <- function(rules) {
  function(data) {
    fails <- character()
    for (nm in names(rules)) {
      if (!nm %in% names(data))
        fails <- c(fails, paste(nm, "missing"))
      else if (!rules[[nm]](data[[nm]]))
        fails <- c(fails, paste(nm, "invalid"))
    }
    if (length(fails) == 0) TRUE else fails
  }
}
```


## Sol 154 Cont'd

```{r, include=FALSE}
make_validator <- function(rules) {
  function(data) {
    fails <- character()
    for (nm in names(rules)) {
      if (!nm %in% names(data))
        fails <- c(fails, paste(nm, "missing"))
      else if (!rules[[nm]](data[[nm]]))
        fails <- c(fails, paste(nm, "invalid"))
    }
    if (length(fails) == 0) TRUE else fails
  }
}
```

```{r, eval=FALSE}
v <- make_validator(list(name = is.character,
  age = \(x) is.numeric(x) && x > 0))
v(list(name = "Alice", age = 30))       # TRUE
v(list(name = 42, age = -1))            # failures
v(list(name = "Bob"))              # age missing
```

```{r, echo=FALSE}
#| output.lines: 3
v <- make_validator(list(name = is.character,
  age = \(x) is.numeric(x) && x > 0))
v(list(name = "Alice", age = 30))       # TRUE
v(list(name = 42, age = -1))            # failures
v(list(name = "Bob"))              # age missing
```

## Ex 155: Recursive Flatten

Write `flatten(x)` that recursively flattens arbitrarily nested lists into a single flat list.

- Test: `flatten(list(1, list(2, list(3, 4)), 5))` → `list(1, 2, 3, 4, 5)`
- Compare with `rapply(x, identity, how = "unlist")`
- Handle mixed types: `list("a", list(1, list(TRUE)))`


## Sol 155

```{r}
#| output.lines: 1
flatten <- function(x) {
  if (!is.list(x)) return(list(x))
  result <- list()
  for (item in x) result <- c(result, flatten(item))
  result
}
str(flatten(list(1, list(2, list(3, 4)), 5)))
rapply(list(1, list(2, list(3, 4)), 5), identity,
       how = "unlist")
str(flatten(list("a", list(1, list(TRUE)))))
```


## Ex 156: Memoize Generic

Write `memoize(f)` that wraps any single-argument function with a hash-map cache.

- Test with expensive `slow_square <- function(x) { Sys.sleep(0.1); x^2 }`
- Show second call is instant
- Add `cache_info()` method returning hit/miss counts


## Sol 156

```{r, eval=FALSE}
memoize <- function(f) {
  cache <- new.env(parent = emptyenv())
  hits <- 0L; misses <- 0L
  wrapped <- function(x) {
    key <- as.character(x)
    if (exists(key, envir = cache)) {
      hits <<- hits + 1L; return(get(key, envir = cache))
    }
    misses <<- misses + 1L
```


## Sol 156 Cont'd

```{r, include=FALSE}
memoize <- function(f) {
  cache <- new.env(parent = emptyenv())
  hits <- 0L; misses <- 0L
  wrapped <- function(x) {
    key <- as.character(x)
    if (exists(key, envir = cache)) {
      hits <<- hits + 1L; return(get(key, envir = cache))
    }
    misses <<- misses + 1L
    val <- f(x); assign(key, val, envir = cache); val
  }
  attr(wrapped, "cache_info") <-
    function() c(hits = hits, misses = misses)
  wrapped
}
```

```{r, eval=FALSE}
    val <- f(x); assign(key, val, envir = cache); val
  }
  attr(wrapped, "cache_info") <-
    function() c(hits = hits, misses = misses)
  wrapped
}
sq <- memoize(\(x) x^2)
sq(5); sq(5); sq(3)
attr(sq, "cache_info")()          # hits:1, misses:2
```

```{r, echo=FALSE}
#| output.lines: 2
sq <- memoize(\(x) x^2)
sq(5); sq(5); sq(3)
attr(sq, "cache_info")()          # hits:1, misses:2
```

## Ex 157: State Machine

Implement a finite state machine for a turnstile: states `{locked, unlocked}`, events `{coin, push}`.

- `coin` in `locked` → `unlocked`; `push` in `locked` → `locked` (+ "blocked" message)
- `coin` in `unlocked` → `unlocked` (+ "thank you"); `push` → `locked` (+ "open")
- Use closure to track state. Return a list with `event(e)` and `state()` methods


## Sol 157

```{r, eval=FALSE}
make_turnstile <- function() {
  st <- "locked"
  event <- function(e) {
    msg <- if (st == "locked" && e == "coin") {
      st <<- "unlocked"; "unlocked"
    } else if (st == "locked" && e == "push") {
      "blocked"
    } else if (st == "unlocked" && e == "push") {
```


## Sol 157 Cont'd

```{r, include=FALSE}
make_turnstile <- function() {
  st <- "locked"
  event <- function(e) {
    msg <- if (st == "locked" && e == "coin") {
      st <<- "unlocked"; "unlocked"
    } else if (st == "locked" && e == "push") {
      "blocked"
    } else if (st == "unlocked" && e == "push") {
      st <<- "locked"; "open"
    } else { "thank you" }
    msg
  }
  list(event = event, state = function() st)
}
```

```{r, eval=FALSE}
      st <<- "locked"; "open"
    } else { "thank you" }
    msg
  }
  list(event = event, state = function() st)
}
ts <- make_turnstile()
ts$event("push")              # blocked
ts$event("coin"); ts$state()  # unlocked
```

```{r, echo=FALSE}
#| output.lines: 2
ts <- make_turnstile()
ts$event("push")              # blocked
ts$event("coin"); ts$state()  # unlocked
```

## Ex 158: Matrix Rotation

Write functions to rotate an $n \times n$ matrix:

- `rotate_90(m)`: 90° clockwise
- `rotate_180(m)`: 180°
- `rotate_270(m)`: 270° clockwise (= 90° counter-clockwise)
- Verify: `rotate_90` applied 4 times returns the original


## Sol 158

```{r}
#| output.lines: 2
rotate_90 <- function(m) t(m)[, ncol(m):1]
rotate_180 <- function(m) m[nrow(m):1, ncol(m):1]
rotate_270 <- function(m) t(m)[nrow(m):1, ]
m <- matrix(1:9, 3, 3)
m
rotate_90(m)
identical(Reduce(\(x, f) f(x), rep(list(rotate_90), 4),
          init = m), m)   # TRUE
```


## Ex 159: Towers of Hanoi

Write `hanoi(n, from="A", to="C", aux="B")` that prints all moves.

- Test `hanoi(3)` — should produce 7 moves
- Count total moves for $n$ disks (verify: $2^n - 1$)
- Return moves as a data frame with columns `disk`, `from`, `to`


## Sol 159

```{r, eval=FALSE}
hanoi <- function(n, from="A", to="C", aux="B") {
  moves <- data.frame(disk=integer(), from=character(),
                      to=character(), stringsAsFactors=FALSE)
  solve <- function(n, from, to, aux) {
    if (n == 0) return()
    solve(n-1, from, aux, to)
    moves[nrow(moves)+1,] <<- list(n, from, to)
    solve(n-1, aux, to, from)
  }
  solve(n, from, to, aux); moves
}
```


## Sol 159 Cont'd

```{r, include=FALSE}
hanoi <- function(n, from="A", to="C", aux="B") {
  moves <- data.frame(disk=integer(), from=character(),
                      to=character(), stringsAsFactors=FALSE)
  solve <- function(n, from, to, aux) {
    if (n == 0) return()
    solve(n-1, from, aux, to)
    moves[nrow(moves)+1,] <<- list(n, from, to)
    solve(n-1, aux, to, from)
  }
  solve(n, from, to, aux); moves
}
```

```{r}
#| output.lines: 3
h <- hanoi(3); cat("Moves:", nrow(h), "\n")  # 7 = 2^3-1
h
```


## Ex 160: Custom `Reduce` Variants

Write `reduce_right(f, x, init)` that folds from the right: $f(x_1, f(x_2, f(x_3, \text{init})))$.

- Test: `reduce_right("-", 1:4, 0)` vs `Reduce("-", 1:4, 0)`
- Write `scan_left(f, x, init)` that returns all intermediate results (like `Reduce` with `accumulate=TRUE`)
- Use `scan_left` to compute running balance from transactions


## Sol 160

```{r}
#| output.lines: 3
reduce_right <- function(f, x, init) {
  Reduce(f, rev(x), init, right = TRUE)
}
Reduce("-", 1:4, 0)                   # ((((0-1)-2)-3)-4)
reduce_right("-", 1:4, 0)             # 1-(2-(3-(4-0)))
scan_left <- function(f, x, init)
  Reduce(f, x, init, accumulate = TRUE)
txns <- c(100, -30, 50, -20, -10)
scan_left("+", txns, 1000)            # running balance
```


## Ex 161: Pipeline Builder

Write `make_pipeline(...)` that accepts functions and returns a pipeline function. Each step is named for debugging.

- `pipe <- make_pipeline(abs, sqrt, round)`; `pipe(-16)` → 4
- Add `describe()` method listing all steps
- Add `remove_step(name)` and `add_step(fn, name)` for dynamic modification


## Sol 161

```{r, eval=FALSE}
make_pipeline <- function(...) {
  steps <- list(...)
  nms <- vapply(substitute(list(...))[-1], deparse,
                character(1))
  names(steps) <- nms
  pipe <- function(x) Reduce(\(v, f) f(v), steps, init=x)
  pipe_env <- environment(pipe)
  attr(pipe, "describe") <- function()
    cat("Steps:", paste(names(steps), collapse=" -> "), "\n")
  attr(pipe, "steps") <- function() names(steps)
  pipe
}
```


## Sol 161 Cont'd

```{r, include=FALSE}
make_pipeline <- function(...) {
  steps <- list(...)
  nms <- vapply(substitute(list(...))[-1], deparse,
                character(1))
  names(steps) <- nms
  pipe <- function(x) Reduce(\(v, f) f(v), steps, init=x)
  pipe_env <- environment(pipe)
  attr(pipe, "describe") <- function()
    cat("Steps:", paste(names(steps), collapse=" -> "), "\n")
  attr(pipe, "steps") <- function() names(steps)
  pipe
}
```

```{r, eval=FALSE}
p <- make_pipeline(abs, sqrt, round)
p(-16)
attr(p, "describe")()
```

```{r, echo=FALSE}
#| output.lines: 3
p <- make_pipeline(abs, sqrt, round)
p(-16)
attr(p, "describe")()
```

## Ex 162: Safe Division Pipeline

Write `safe(fn)` that wraps any function: on success returns `list(ok = TRUE, value = result)`, on error returns `list(ok = FALSE, error = msg)`.

- Chain safe operations: `safe(log)` then `safe(sqrt)`
- Write `and_then(result, fn)` that applies `safe(fn)` only if `result$ok`
- Build: `safe(as.numeric)("3.14") |> and_then(log) |> and_then(sqrt)`


## Sol 162

```{r}
#| output.lines: 1
safe <- function(fn) {
  function(...) tryCatch(
    list(ok = TRUE, value = fn(...)),
    error = function(e) list(ok = FALSE, error = e$message))
}
and_then <- function(result, fn) {
  if (!result$ok) return(result); safe(fn)(result$value)
}
safe(as.numeric)("3.14") |> and_then(log) |> and_then(sqrt)
safe(as.numeric)("abc")  |> and_then(log) |> and_then(sqrt)
```


## Ex 163: Accumulator Pattern Variants

Write three accumulator closures:

- `make_running_mean()`: accepts new values, always returns current mean
- `make_running_sd()`: online standard deviation (Welford's algorithm)
- `make_ema(alpha)`: exponential moving average

Test all three with the stream `c(10, 12, 8, 15, 11, 9, 13)`.


## Sol 163

```{r}
#| output.lines: 3
make_running_mean <- function() {
  total <- 0; n <- 0
  function(x) { total <<- total + x; n <<- n + 1; total/n }
}
make_ema <- function(alpha = 0.3) {
  val <- NULL
  function(x) { val <<- if (is.null(val)) x
    else alpha*x + (1-alpha)*val; val }
}
rm <- make_running_mean(); ema <- make_ema(0.3)
stream <- c(10, 12, 8, 15, 11, 9, 13)
round(sapply(stream, rm), 2)
round(sapply(stream, ema), 2)
```


## Ex 164: Trie (Prefix Tree)

Implement a trie for string lookup using nested environments:

- `trie_insert(trie, word)`: insert a word
- `trie_search(trie, word)`: exact match → TRUE/FALSE
- `trie_starts_with(trie, prefix)`: prefix match → TRUE/FALSE
- Test with words: "apple", "app", "application", "bat"


## Sol 164

```{r, eval=FALSE}
make_trie <- function() new.env(parent = emptyenv())
trie_insert <- function(trie, word) {
  node <- trie
  for (ch in strsplit(word, "")[[1]]) {
    if (!exists(ch, envir=node))
      assign(ch, new.env(parent=emptyenv()), envir=node)
    node <- get(ch, envir = node) }
  assign("$end", TRUE, envir = node)
}
```


## Sol 164 Cont'd

```{r, include=FALSE}
make_trie <- function() new.env(parent = emptyenv())
trie_insert <- function(trie, word) {
  node <- trie
  for (ch in strsplit(word, "")[[1]]) {
    if (!exists(ch, envir=node))
      assign(ch, new.env(parent=emptyenv()), envir=node)
    node <- get(ch, envir = node) }
  assign("$end", TRUE, envir = node)
}
trie_search <- function(trie, word) {
  node <- trie
  for (ch in strsplit(word, "")[[1]]) {
    if (!exists(ch, envir=node)) return(FALSE)
    node <- get(ch, envir=node) }
  isTRUE(tryCatch(get("$end",envir=node), error=\(e) FALSE))
}
```

```{r, eval=FALSE}
trie_search <- function(trie, word) {
  node <- trie
  for (ch in strsplit(word, "")[[1]]) {
    if (!exists(ch, envir=node)) return(FALSE)
    node <- get(ch, envir=node) }
  isTRUE(tryCatch(get("$end",envir=node), error=\(e) FALSE))
}
tr <- make_trie()
for (w in c("apple","app","bat")) trie_insert(tr, w)
c(app=trie_search(tr,"app"), ap=trie_search(tr,"ap"))
```

```{r, echo=FALSE}
#| output.lines: 2
tr <- make_trie()
for (w in c("apple","app","bat")) trie_insert(tr, w)
c(app=trie_search(tr,"app"), ap=trie_search(tr,"ap"))
```

## Ex 165: Map/Filter/Reduce from Scratch

Implement `my_map(x, f)`, `my_filter(x, pred)`, `my_reduce(x, f, init)` without using any apply functions or `Reduce`.

- Use only `for` loops
- Test: pipe `1:20` through filter(even) → map(square) → reduce(sum)
- Compare results with `sum(Filter(\(x) x%%2==0, 1:20)^2)`


## Sol 165

```{r, eval=FALSE}
my_map <- function(x, f) {
  out <- vector("list", length(x))
  for (i in seq_along(x)) out[[i]] <- f(x[[i]])
  out
}
```


## Sol 165 Cont'd

```{r, include=FALSE}
my_map <- function(x, f) {
  out <- vector("list", length(x))
  for (i in seq_along(x)) out[[i]] <- f(x[[i]])
  out
}
my_filter <- function(x, pred) {
  out <- list()
  for (item in x) if (pred(item)) out[[length(out)+1]] <- item
  out
}
my_reduce <- function(x, f, init) {
  acc <- init; for (item in x) acc <- f(acc, item); acc
}
```

```{r, eval=FALSE}
my_filter <- function(x, pred) {
  out <- list()
  for (item in x) if (pred(item)) out[[length(out)+1]] <- item
  out
}
my_reduce <- function(x, f, init) {
  acc <- init; for (item in x) acc <- f(acc, item); acc
}
result <- my_filter(1:20, \(x) x %% 2 == 0) |>
  my_map(\(x) x^2) |> my_reduce(`+`, 0)
result  # 2^2 + 4^2 + ... + 20^2 = 1540
sum(Filter(\(x) x%%2==0, 1:20)^2)   # same
```

```{r, echo=FALSE}
#| output.lines: 2
result <- my_filter(1:20, \(x) x %% 2 == 0) |>
  my_map(\(x) x^2) |> my_reduce(`+`, 0)
result  # 2^2 + 4^2 + ... + 20^2 = 1540
sum(Filter(\(x) x%%2==0, 1:20)^2)   # same
```

## Ex 166: String Calculator

Write `calc(expr)` that evaluates simple arithmetic strings: `"3 + 5 * 2"` → 13.

- Support `+`, `-`, `*`, `/` with standard precedence
- Handle parentheses: `"(3 + 5) * 2"` → 16
- Hint: tokenize with `strsplit` and `gsub`, then use recursive descent or `parse`


## Sol 166

```{r}
#| output.lines: 3
calc <- function(expr) {
  # Safe: only allow digits, operators, parens, spaces, dots
  stopifnot(grepl("^[0-9+\\-*/().^\\s]+$", expr, perl=TRUE))
  eval(parse(text = expr))
}
calc("3 + 5 * 2")        # 13
calc("(3 + 5) * 2")      # 16
calc("10 / 3")            # 3.333...
calc("2^10")              # 1024
```


## Ex 167: Data Frame Toolkit

Without using any packages, write:

- `my_select(df, ...)`: select columns by name (like `dplyr::select`)
- `my_mutate(df, ...)`: add/modify columns (like `dplyr::mutate`)
- `my_summarize(df, by, ...)`: group-by summarize
- Test all on `mtcars`


## Sol 167

```{r}
#| output.lines: 2
my_select <- function(df, ...) {
  cols <- as.character(substitute(list(...)))[-1]
  df[, cols, drop = FALSE]
}
my_mutate <- function(df, ...) {
  exprs <- as.list(substitute(list(...)))[-1]
  for (nm in names(exprs))
    df[[nm]] <- eval(exprs[[nm]], df, parent.frame())
  df
}
head(my_select(mtcars, mpg, cyl, wt), 3)
head(my_mutate(mtcars, kpl = mpg * 0.425), 3)[,"kpl"]
```


## Ex 168: Generate Permutations

Write `permutations(x)` returning all permutations of vector `x` as a list.

- Test: `permutations(1:3)` → 6 permutations
- Use recursion: pick each element as first, permute the rest
- How many permutations of 8 elements? Can your function handle it?


## Sol 168

```{r}
#| output.lines: 3
permutations <- function(x) {
  if (length(x) <= 1) return(list(x))
  result <- list()
  for (i in seq_along(x)) {
    rest <- permutations(x[-i])
    result <- c(result, lapply(rest, \(p) c(x[i], p)))
  }
  result
}
perms <- permutations(1:3)
length(perms)                       # 6
perms[[1]]; perms[[6]]
factorial(8)                        # 40320 permutations
```


## Ex 169: Event Emitter

Build an event system using closures:

- `emitter$on(event, callback)` — register a listener
- `emitter$emit(event, ...)` — fire all listeners for that event
- `emitter$off(event, callback)` — remove a listener
- Test with "click" and "hover" events


## Sol 169

```{r, eval=FALSE}
make_emitter <- function() {
  listeners <- list()
  list(
    on = function(event, cb) {
      listeners[[event]] <<- c(listeners[[event]], list(cb)) },
    emit = function(event, ...) {
      for (cb in listeners[[event]]) cb(...) },
    off = function(event) {
      listeners[[event]] <<- NULL }
  )
}
```


## Sol 169 Cont'd

```{r, include=FALSE}
make_emitter <- function() {
  listeners <- list()
  list(
    on = function(event, cb) {
      listeners[[event]] <<- c(listeners[[event]], list(cb)) },
    emit = function(event, ...) {
      for (cb in listeners[[event]]) cb(...) },
    off = function(event) {
      listeners[[event]] <<- NULL }
  )
}
```

```{r, eval=FALSE}
em <- make_emitter()
em$on("click", \(x) cat("Click at", x, "\n"))
em$on("click", \(x) cat("Log:", x, "\n"))
em$emit("click", 42); em$off("click")
```

```{r, echo=FALSE}
#| output.lines: 3
em <- make_emitter()
em$on("click", \(x) cat("Click at", x, "\n"))
em$on("click", \(x) cat("Log:", x, "\n"))
em$emit("click", 42); em$off("click")
```

## Ex 170: Topological Sort

Write `topo_sort(edges)` for a DAG given as list of `c(from, to)` pairs.

- Test: `list(c("A","B"), c("A","C"), c("B","D"), c("C","D"))` → A, B or C, then D
- Detect cycles: `list(c("A","B"), c("B","A"))` → error
- Use Kahn's algorithm (in-degree based)


## Sol 170

```{r, eval=FALSE}
topo_sort <- function(edges) {
  nodes <- unique(unlist(edges))
  indeg <- setNames(rep(0L, length(nodes)), nodes)
  adj   <- setNames(vector("list", length(nodes)), nodes)
  for (e in edges) {
    adj[[e[1]]] <- c(adj[[e[1]]], e[2])
    indeg[e[2]] <- indeg[e[2]] + 1L }
  queue <- names(indeg[indeg == 0]); result <- character()
  while (length(queue) > 0) { n <- queue[1]; queue <- queue[-1]
    result <- c(result, n)
    for (m in adj[[n]]) { indeg[m] <- indeg[m] - 1L
      if (indeg[m] == 0) queue <- c(queue, m) } }
  if (length(result) != length(nodes)) stop("cycle!")
  result }
topo_sort(list(c("A","B"),c("A","C"),c("B","D"),c("C","D")))
```
