---
title: "Base R: Exercises with Solutions"
author: ""
date: ""
format:
  beamer:
    theme: Madrid
    colortheme: beaver
    fontsize: 9pt
    slide-level: 2
    aspectratio: 169
    pdf-engine: xelatex
    include-in-header:
      text: |
        \usepackage{booktabs}
        \usepackage{etoolbox}
        \setbeamertemplate{footline}[text line]{%
          \parbox{\linewidth}{\vspace*{-8pt}%
          \hfill\insertframenumber / \inserttotalframenumber~~~~~~~~~}}
        \setbeamertemplate{navigation symbols}{}
        \setbeamercolor{item projected}{fg=white}
        \setbeamercolor{frametitle}{fg=black}
        \setbeamercolor{block title}{fg=white}
        \definecolor{foo}{rgb}{.2,.2,.7}
        \newcommand{\comm}[1]{\textcolor[rgb]{0.737,0.353,0.396}{{\tt #1}}}
        \AtBeginSection[]{
          \begin{frame}
          \vfill\centering
          \begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{section page}
            \usebeamerfont{title}{\color{foo}\insertsectionhead}\par
          \end{beamercolorbox}
          \vfill
          \end{frame}
        }
        \AtBeginSubsection{}
        %% ── tighten spacing around code blocks and output ──
        \setlength{\parskip}{0pt}
        \setlength{\OuterFrameSep}{0pt}
        \makeatletter
        \preto{\@verbatim}{\topsep=2pt \partopsep=0pt}
        \makeatother
        \AtBeginEnvironment{Shaded}{\vspace{0pt}}
        \AfterEndEnvironment{Shaded}{\vspace{0pt}}
        \AtBeginEnvironment{verbatim}{\vspace{0pt}}
        \AfterEndEnvironment{verbatim}{\vspace{0pt}}
execute:
  echo: true
  warning: false
knitr:
  opts_chunk:
    comment: "#"
    fig.align: center
---

## Outline

\tableofcontents

# Vectors and Types (Ex 1--18)

## Ex 1: Logical Indexing

::: {.alertblock}
### Problem
Let `v <- c(14, -3, 22, 7, -8, 0, 31, 5, -1, 18)`.

- Extract all elements strictly between 5 and 25
- Replace every negative element with `0` in a copy `w`
- How many elements are negative or exceed 20?
- Find indices where $|v_i|$ exceeds its rank
:::


## Sol 1

```{r}
v <- c(14, -3, 22, 7, -8, 0, 31, 5, -1, 18)
v[v > 5 & v < 25]
w <- v; w[w < 0] <- 0; w
sum(v < 0 | v > 20)
which(abs(v) > rank(v))
```


## Ex 2: Named Vectors

::: {.alertblock}
### Problem
Create `stock <- c(AAPL=189.5, MSFT=415.2, GOOG=172.3, AMZN=185.7)`.

- Access `GOOG` by name; add `NVDA = 875.4`
- Compute percentage change from base 150; round to 1 dp
- Return stocks above mean price, sorted descending
- Rename all tickers to lowercase
:::


## Sol 2

```{r}
stock <- c(AAPL=189.5, MSFT=415.2, GOOG=172.3, AMZN=185.7)
stock["GOOG"]; stock["NVDA"] <- 875.4
round((stock - 150) / 150 * 100, 1)
sort(stock[stock > mean(stock)], decreasing=TRUE)
names(stock) <- tolower(names(stock)); names(stock)
```


## Ex 3: `which`, `which.min`, `which.max`

::: {.alertblock}
### Problem
Let `x <- c(3, 7, 2, 9, 1, 5, 8, 4, 6, 10)`.

- Indices of elements > 6
- Index of min and max
- Second-largest value (no `sort`)
- Local maxima: elements greater than both immediate neighbours (ignore endpoints)
:::


## Sol 3

```{r}
x <- c(3, 7, 2, 9, 1, 5, 8, 4, 6, 10)
which(x > 6)
which.min(x); which.max(x)
max(x[x != max(x)])
n <- length(x); idx <- 2:(n-1)
x[idx][x[idx] > x[idx-1] & x[idx] > x[idx+1]]
```


## Ex 4: Cumulative and Parallel Functions

::: {.alertblock}
### Problem
Let `x <- c(2,5,3,8,1,7,4)` and `y <- c(6,2,9,1,4,3,8)`.

- Running maximum of `x`
- Running sum and product
- Element-wise min and max
- Verify $\sum\text{pmin}(x,y) + \sum\text{pmax}(x,y) = \sum x + \sum y$
:::


## Sol 4

```{r}
x <- c(2,5,3,8,1,7,4); y <- c(6,2,9,1,4,3,8)
cummax(x)
cumsum(x); cumprod(x)
pmin(x,y); pmax(x,y)
sum(pmin(x,y)) + sum(pmax(x,y)) == sum(x) + sum(y)
```


## Ex 5: Set Operations

::: {.alertblock}
### Problem
Let `a <- c(1,3,5,7,9,3,5)` and `b <- c(2,3,5,6,7,2)`.

- Unique elements of `a`
- Positions of duplicated values in `a`
- $a \cup b$, $a \cap b$, $a \setminus b$
- `match(c(3,5,4), b)` — explain the `NA`
:::


## Sol 5

```{r}
a <- c(1,3,5,7,9,3,5); b <- c(2,3,5,6,7,2)
unique(a)
which(duplicated(a))
union(a,b); intersect(a,b); setdiff(a,b)
match(c(3,5,4), b)  # 4 not found -> NA
```


## Ex 6: Recycling Rule

::: {.alertblock}
### Problem

- What does `c(1,2,3,4,5,6) + c(10,20)` return? Why?
- Let `m <- matrix(1:12, nrow=3)`. What does `m + c(100,200,300)` do?
- Extract odd-indexed and even-indexed elements of `1:12` using `c(TRUE,FALSE)`
- Generate `1 0 1 0 1 0 1 0 1 0` using `rep`
:::


## Sol 6

```{r}
c(1,2,3,4,5,6) + c(10,20)                         # recycled
m <- matrix(1:12, nrow=3)
m + c(100,200,300)    # adds per column (col-major)
x <- 1:12; x[c(T,F)]; x[c(F,T)]
rep(c(1,0), 5)
```


## Ex 7: Sorting and Ranking

::: {.alertblock}
### Problem
Let `x <- c(40,15,30,10,50,25)`.

- Sort ascending and descending
- Find `order(x)` and verify `x[order(x)] == sort(x)`
- What rank does `30` receive?
- Sort `c("banana","apple","cherry")` alphabetically
:::


## Sol 7

```{r}
x <- c(40,15,30,10,50,25)
sort(x); sort(x, decreasing=TRUE)
ord <- order(x); all(x[ord] == sort(x))
rank(x)  # 30 -> rank 4
sort(c("banana","apple","cherry"))
```


## Ex 8: `NA`, `NULL`, `NaN`, `Inf`

::: {.alertblock}
### Problem

- Given `x <- c(3, NA, -1, NaN, Inf, -Inf, 0)`, apply `is.na`, `is.nan`, `is.finite`, `is.infinite`. Why is `is.na(NaN)` TRUE?
- Remove non-finite values and compute mean
- Show `c(1, NULL, 3)` vs `c(1, NA, 3)` lengths
- Compute `NA^0`, `NA | TRUE`, `NA & FALSE`
:::


## Sol 8

```{r}
x <- c(3, NA, -1, NaN, Inf, -Inf, 0)
is.na(x); is.nan(x); is.finite(x)
mean(x[is.finite(x)])
length(c(1, NULL, 3)); length(c(1, NA, 3))        # 2 vs 3
NA^0; NA | TRUE; NA & FALSE
```


## Ex 9: Type Coercion

::: {.alertblock}
### Problem

- What does `c(TRUE, 1L, 2.5, "x")` produce? What type?
- Coerce `c("3","1.5","NA","7")` to numeric. What happens to `"NA"`?
- Predict `as.integer(c(3.9, -2.7, TRUE, "5"))`
- Why does `1 == "1"` return `TRUE`?
:::


## Sol 9

```{r}
x <- c(TRUE, 1L, 2.5, "x"); x; typeof(x)         # character
as.numeric(c("3","1.5","NA","7"))                  # NA + warning
as.integer(c(3.9, -2.7, TRUE, "5"))               # 3,-2,1,5
1 == "1"  # 1 coerced to "1", then string compare
```


## Ex 10: `seq` and `seq_along`

::: {.alertblock}
### Problem

- Generate `1.0 2.5 4.0 5.5 7.0 8.5 10.0` using `seq`
- Show `1:length(x)` is unsafe when `x <- NULL`; fix with `seq_along`
- 6 equally-spaced values from 0 to $2\pi$
- What does `seq_along(integer(0))` return?
:::


## Sol 10

```{r}
seq(1, 10, by=1.5)
x <- NULL; 1:length(x)                            # c(1,0) WRONG
seq_along(x)                                       # integer(0) SAFE
seq(0, 2*pi, length.out=6)
seq_along(integer(0))                              # integer(0)
```


## Ex 11: `rep` Patterns

::: {.alertblock}
### Problem
Using `rep`, generate:

- `1 1 2 2 3 3 4 4 5 5`
- `1 2 3 1 2 3 1 2 3 1 2 3`
- `"A" "A" "A" "B" "B" "C"`
- Alternating `TRUE FALSE` of length 20
:::


## Sol 11

```{r}
rep(1:5, each=2)
rep(1:3, times=4)
rep(c("A","B","C"), c(3,2,1))
rep(c(TRUE,FALSE), 10)
```


## Ex 12: `%in%` and `match`

::: {.alertblock}
### Problem
`haystack <- c(3,1,4,1,5,9,2,6,5,3)`; `needles <- c(5,7,2)`.

- Which needles are present?
- First occurrence position of each
- Difference between `match("a", c("b","a","a"))` and `which(c("b","a","a") == "a")`
- Filter `mtcars` rows where `cyl %in% c(4,6)`
:::


## Sol 12

```{r}
haystack <- c(3,1,4,1,5,9,2,6,5,3); needles <- c(5,7,2)
needles %in% haystack
match(needles, haystack)                           # NA for 7
match("a", c("b","a","a"))                         # 2 (first)
which(c("b","a","a") == "a")                       # c(2,3) (all)
nrow(mtcars[mtcars$cyl %in% c(4,6), ])
```


## Ex 13: Vector Arithmetic

::: {.alertblock}
### Problem

- Compute $H_{100} = \sum_{k=1}^{100}\frac{1}{k}$ in one line
- Compute $\sum_{k=0}^{50}\frac{(-1)^k}{k!}$. Compare with `exp(-1)`
- For `x <- seq(0.1,2,by=0.1)`, compute $e^x\sin x - \ln x$, rounded to 4 dp
- Compute `diff(cumsum(1:10))` and explain
:::


## Sol 13

```{r}
sum(1 / 1:100)                                     # H_100
k <- 0:50; sum((-1)^k / factorial(k)); exp(-1)
x <- seq(0.1, 2, by=0.1)
round(exp(x)*sin(x) - log(x), 4)
diff(cumsum(1:10))  # recovers 2:10
```


## Ex 14: Attributes and `structure`

::: {.alertblock}
### Problem

- Create `a <- 1:4`; attach attributes `"unit"="kg"` and `"source"="lab"`. Print
- Recreate using `structure()` in one call
- `x <- 1:3` vs `y <- c(1,2,3)`: are they `identical`? `all.equal`? Why?
- When to use `isTRUE(all.equal())` in an `if`?
:::


## Sol 14

```{r}
a <- 1:4; attr(a,"unit") <- "kg"; attr(a,"source") <- "lab"
attributes(a)
b <- structure(1:4, unit="kg", source="lab")
x <- 1:3; y <- c(1,2,3)
identical(x, y); all.equal(x, y)                   # F; T
isTRUE(all.equal(0.1+0.2, 0.3))                   # TRUE
```


## Ex 15: Factors

::: {.alertblock}
### Problem

- Create factor `f <- factor(c("low","med","high","med","low"))`. What are the levels?
- Create an ordered factor with levels `low < med < high`. Test `f > "low"`
- Use `table(f)` to count occurrences
- What does `as.integer(f)` return and why?
:::


## Sol 15

```{r}
f <- factor(c("low","med","high","med","low"))
levels(f)                                          # alphabetical
fo <- factor(f, levels=c("low","med","high"), ordered=TRUE)
fo > "low"
table(f)
as.integer(f)  # underlying integer codes
```


## Ex 16: Floating Point

::: {.alertblock}
### Problem

- Is `0.1 + 0.2 == 0.3` TRUE or FALSE? Why?
- Use `all.equal` to compare correctly
- What is `trunc(-3.7)` vs `floor(-3.7)`?
- Demonstrate banker's rounding: `round(0.5)`, `round(1.5)`, `round(2.5)`
:::


## Sol 16

```{r}
0.1 + 0.2 == 0.3  # FALSE (floating point)
all.equal(0.1 + 0.2, 0.3)                         # TRUE
trunc(-3.7); floor(-3.7)  # -3 vs -4
round(0.5); round(1.5); round(2.5)  # 0; 2; 2
```


## Ex 17: Appending and Inserting

::: {.alertblock}
### Problem

- Start with `v <- 1:5`. Append `c(10,11)` to the end
- Insert `99` after position 3 using `append`
- What happens if you assign `v[10] <- 42` when `length(v)` is 7?
- Remove the 3rd element without using negative indexing (use logical)
:::


## Sol 17

```{r}
v <- 1:5; v <- c(v, c(10,11)); v
append(v, 99, after=3)
v[10] <- 42; v   # NAs fill gaps
v[seq_along(v) != 3]
```


## Ex 18: `sample` and `sequence`

::: {.alertblock}
### Problem

- Simulate 20 dice rolls with `sample`
- Use `sample` to shuffle `letters[1:10]`
- Using `sequence`, generate `1 2 3 4  1 2 3  1 2  1`
- Generate `5 4 3 2 1  6 5 4 3 2 1` using `sequence`
:::


## Sol 18

```{r}
set.seed(42)
sample(1:6, 20, replace=TRUE)
sample(letters[1:10])
sequence(4:1)
sequence(5:6, from=5:6, by=-1)
```


# Operators and Common Functions (Ex 19--26)


## Ex 19: Operator Precedence

::: {.alertblock}
### Problem

- Without running, predict: `2 + 3 * 4`, `2^3^2`, `1:3 + 1`, `1:(3+1)`
- Why does `-2^2` give `-4` instead of `4`?
- Predict: `TRUE | FALSE & FALSE`
- Write `x <- 5; x %% 3 == 2 & x > 3` and trace operator precedence
:::


## Sol 19

```{r}
2 + 3 * 4     # 14: * before +
2^3^2          # 512: ^ right-associative = 2^9
1:3 + 1        # 2 3 4: : before +
1:(3+1)        # 1 2 3 4
-2^2           # -4: ^ before unary -
TRUE | FALSE & FALSE  # TRUE: & before |
x <- 5; x %% 3 == 2 & x > 3  # TRUE
```


## Ex 20: Math Functions

::: {.alertblock}
### Problem

- Compute `floor`, `ceiling`, `trunc`, `round` of `-5.7`
- Compute $\binom{10}{3}$ and verify with `factorial`
- Compute `119 %/% 13` and `119 %% 13`. Verify `119 == 13 * (119 %/% 13) + 119 %% 13`
- `signif(12345678, 4)`
:::


## Sol 20

```{r}
floor(-5.7); ceiling(-5.7); trunc(-5.7); round(-5.7)
choose(10, 3); factorial(10)/(factorial(3)*factorial(7))
119 %/% 13; 119 %% 13
119 == 13 * (119 %/% 13) + 119 %% 13              # TRUE
signif(12345678, 4)
```


## Ex 21: Statistical Summary

::: {.alertblock}
### Problem
`v <- c(4, 7, 2, 9, 1, 5, 8, 3, 6, 10)`.

- mean, median, var, sd, range
- `quantile(v, c(0.25, 0.75))` and IQR
- Compute z-scores: $(v - \bar v)/s$. How many exceed $|z| > 1$?
- Use `diff(v)` to find the largest jump
:::


## Sol 21

```{r}
v <- c(4, 7, 2, 9, 1, 5, 8, 3, 6, 10)
mean(v); median(v); var(v); sd(v); range(v)
quantile(v, c(0.25, 0.75)); IQR(v)
z <- (v - mean(v)) / sd(v); sum(abs(z) > 1)
max(abs(diff(v)))                                  # largest jump
```


## Ex 22: `table` and `prop.table`

::: {.alertblock}
### Problem
`x <- c("A","B","A","C","B","A","C","C","B","A")`.

- Frequency table of `x`
- Proportion table
- Create `y <- c(1,1,2,1,2,2,1,2,1,2)`. Cross-tabulate `x` and `y`; add margins
- Which category of `x` is most frequent?
:::


## Sol 22

```{r}
x <- c("A","B","A","C","B","A","C","C","B","A")
table(x)
prop.table(table(x))
y <- c(1,1,2,1,2,2,1,2,1,2)
addmargins(table(x, y))
names(which.max(table(x)))                         # "A"
```


## Ex 23: Harmonic Series Convergence

::: {.alertblock}
### Problem

- Compute $H_n = \sum_{k=1}^{n}\frac{1}{k}$ for $n \in \{10, 100, 1000, 10000\}$
- Compare each $H_n$ with $\ln(n) + \gamma$ where $\gamma \approx 0.5772$
- Plot $H_n - \ln n$ for $n = 1, \ldots, 200$
:::


## Sol 23

```{r, out.width="45%"}
ns <- c(10, 100, 1000, 10000)
Hn <- sapply(ns, function(n) sum(1/1:n))
data.frame(n=ns, Hn=round(Hn,4),
           approx=round(log(ns)+0.5772,4))
x <- 1:200; y <- cumsum(1/x) - log(x)
plot(x, y, type="l", ylab=expression(H[n]-ln(n)))
abline(h=0.5772, col="red", lty=2)
```


## Ex 24: `pmin`/`pmax` Application

::: {.alertblock}
### Problem

- Clip vector `v <- c(-5,3,12,-2,8,25)` to range $[0, 10]$ using `pmin` and `pmax`
- Given two portfolios `a <- c(5,8,3,9)` and `b <- c(7,2,6,4)`, compute the "best of both" portfolio
- Compute $\sum_{k=1}^{100}\min(2^k, k^4)$ without loops
:::


## Sol 24

```{r}
v <- c(-5, 3, 12, -2, 8, 25)
pmin(pmax(v, 0), 10)                               # clip to [0,10]
a <- c(5,8,3,9); b <- c(7,2,6,4)
pmax(a, b)
k <- 1:100; sum(pmin(2^k, k^4))
```


## Ex 25: Cumulative Functions

::: {.alertblock}
### Problem
`x <- c(3, -1, 4, -1, 5, -9, 2, 6)`.

- Compute `cumsum(x)`. At which index does it first become negative?
- Maximum drawdown: $\max_i(\text{cummax}_i - \text{cumsum}_i)$
- Compute `cumprod(1 + c(0.05, -0.02, 0.03, -0.01))` (portfolio growth)
:::


## Sol 25

```{r}
x <- c(3, -1, 4, -1, 5, -9, 2, 6)
cs <- cumsum(x); cs
which(cs < 0)[1]                                   # first negative
cm <- cummax(cs); max(cm - cs)                     # max drawdown
cumprod(1 + c(0.05, -0.02, 0.03, -0.01))          # growth
```


## Ex 26: `order` and Data Sorting

::: {.alertblock}
### Problem
`d <- data.frame(name=c("Zara","Amy","Mike","Bob"), score=c(88,95,72,95), age=c(22,20,25,21))`.

- Sort by score ascending
- Sort by score descending, ties broken by age ascending
- Find the name of the person with the highest score (youngest if tied)
:::


## Sol 26

```{r}
d <- data.frame(name=c("Zara","Amy","Mike","Bob"),
                score=c(88,95,72,95), age=c(22,20,25,21))
d[order(d$score), ]
d[order(-d$score, d$age), ]
d[order(-d$score, d$age), ]$name[1]                # Bob
```


# Data Frames (Ex 27--40)


## Ex 27: Creating and Inspecting

::: {.alertblock}
### Problem

- Create a data frame with columns `name` (Alice, Bob, Cathy), `age` (25, 30, 28), `score` (88.5, 92.3, 76.8)
- Show `str`, `dim`, `names`, `summary`
- What class is `df$name`?
- Access `score` of row 2 by name and by position
:::


## Sol 27

```{r}
df <- data.frame(name=c("Alice","Bob","Cathy"),
                 age=c(25,30,28), score=c(88.5,92.3,76.8))
str(df); dim(df); names(df)
class(df$name)
df$score[2]; df[2, "score"]
```


## Ex 28: Row Filtering

::: {.alertblock}
### Problem
Using `tg <- ToothGrowth`:

- Rows where `supp == "OJ"` and `dose >= 1`
- Mean tooth length for VC vs OJ
- Rows where `len` is above its overall median
- Same as (a) but using `subset()`
:::


## Sol 28

```{r}
tg <- ToothGrowth
head(tg[tg$supp == "OJ" & tg$dose >= 1, ], 4)
tapply(tg$len, tg$supp, mean)
head(tg[tg$len > median(tg$len), ], 4)
head(subset(tg, supp=="OJ" & dose >= 1), 4)
```


## Ex 29: Adding and Removing Columns

::: {.alertblock}
### Problem
Start with `df <- data.frame(x=1:5, y=c(10,20,30,40,50))`.

- Add column `z = x * y`
- Add column `label` using `paste0("item", 1:5)`
- Remove column `z`
- Rename `y` to `value`
:::


## Sol 29

```{r}
df <- data.frame(x=1:5, y=c(10,20,30,40,50))
df$z <- df$x * df$y
df$label <- paste0("item", 1:5)
df$z <- NULL
names(df)[names(df)=="y"] <- "value"; df
```


## Ex 30: `with()` and `within()`

::: {.alertblock}
### Problem
Using `tg <- ToothGrowth`:

- Use `with()` to compute `mean(len[dose == 2])`
- Use `within()` to add `log_len <- log(len)` and `len_centered <- len - mean(len)`
:::


## Sol 30

```{r}
tg <- ToothGrowth
with(tg, mean(len[dose == 2]))
tg2 <- within(tg, {
  log_len <- log(len)
  len_centered <- len - mean(len)
})
head(tg2, 3)
```


## Ex 31: Handling NAs

::: {.alertblock}
### Problem
`df <- data.frame(x=c(1,NA,3,4,NA), y=c(NA,2,3,NA,5), z=c(1,2,3,4,5))`.

- Which rows are complete?
- Remove all incomplete rows
- Replace NAs in column `x` with the mean of `x` (ignoring NAs)
- Count NAs per column
:::


## Sol 31

```{r}
df <- data.frame(x=c(1,NA,3,4,NA), y=c(NA,2,3,NA,5), z=1:5)
complete.cases(df)
df[complete.cases(df), ]
df$x[is.na(df$x)] <- mean(df$x, na.rm=TRUE); df
colSums(is.na(df))
```


## Ex 32: `merge()`

::: {.alertblock}
### Problem
`students <- data.frame(id=1:4, name=c("A","B","C","D"))`.
`grades <- data.frame(sid=c(1,2,2,4), course=c("Math","Math","Eng","Eng"), grade=c(85,90,78,92))`.

- Inner join on id
- Left join (keep all students)
- How many rows in each result? Why?
:::


## Sol 32

```{r}
students <- data.frame(id=1:4, name=c("A","B","C","D"))
grades <- data.frame(sid=c(1,2,2,4),
  course=c("Math","Math","Eng","Eng"), grade=c(85,90,78,92))
merge(students, grades, by.x="id", by.y="sid")     # inner
merge(students, grades, by.x="id", by.y="sid",
      all.x=TRUE)                                   # left
# inner: 4 rows; left: 5 rows (C has NA)
```


## Ex 33: `aggregate()`

::: {.alertblock}
### Problem
Using `mtcars`:

- Mean `mpg` by `cyl`
- Mean and sd of `mpg` by `cyl` and `am`
- Max `hp` by `cyl`
:::


## Sol 33

```{r}
aggregate(mpg ~ cyl, data=mtcars, FUN=mean)
aggregate(mpg ~ cyl + am, data=mtcars,
  FUN=function(x) c(m=mean(x), s=sd(x)))
aggregate(hp ~ cyl, data=mtcars, FUN=max)
```


## Ex 34: `cbind` and `rbind`

::: {.alertblock}
### Problem

- Create two 1-column data frames and `cbind` them
- Create two data frames with same columns and `rbind`
- Use `do.call(rbind, list_of_dfs)` with 3 data frames
:::


## Sol 34

```{r}
d1 <- data.frame(a=1:3); d2 <- data.frame(b=4:6)
cbind(d1, d2)
e1 <- data.frame(x="a", y=1); e2 <- data.frame(x="b", y=2)
rbind(e1, e2)
dfs <- list(data.frame(x=1,y=2), data.frame(x=3,y=4),
            data.frame(x=5,y=6))
do.call(rbind, dfs)
```


## Ex 35: Data Frame as Named List

::: {.alertblock}
### Problem

- Show `is.list(mtcars)` is TRUE. What does `typeof(mtcars)` return?
- Access the `mpg` column using `[[` notation
- What is the difference between `mtcars["mpg"]` and `mtcars[["mpg"]]`?
- Convert `mtcars[1:3, 1:3]` to a list and back
:::


## Sol 35

```{r}
is.list(mtcars); typeof(mtcars)                    # TRUE; "list"
head(mtcars[["mpg"]])
class(mtcars["mpg"]); class(mtcars[["mpg"]])       # df vs numeric
l <- as.list(mtcars[1:3, 1:3]); as.data.frame(l)
```


## Ex 36: `split()` on Data Frames

::: {.alertblock}
### Problem

- Split `iris` by `Species`. What type is the result?
- Compute mean `Sepal.Length` per species using `split` + `sapply`
- Recombine with `do.call(rbind, ...)`
:::


## Sol 36

```{r}
sp <- split(iris, iris$Species)
class(sp); names(sp)                               # list
sapply(sp, function(d) mean(d$Sepal.Length))
head(do.call(rbind, sp), 4)
```


## Ex 37: `subset()` with `select`

::: {.alertblock}
### Problem
Using `mtcars`:

- Select only `mpg`, `cyl`, `hp` for cars with `mpg > 25`
- Exclude `drat` and `wt` columns
- Why can you write `subset(mtcars, mpg > 25)` without quoting `mpg`?
:::


## Sol 37

```{r}
subset(mtcars, mpg > 25, select=c(mpg, cyl, hp))
head(subset(mtcars, select=-c(drat, wt)), 3)
# subset uses non-standard evaluation: it captures
# the expression and evaluates it within the data frame
```


## Ex 38: Sorting with Multiple Keys

::: {.alertblock}
### Problem
`df <- data.frame(dept=c("A","B","A","B","A"), name=c("Zoe","Amy","Bob","Cal","Amy"), salary=c(50,60,70,55,65))`.

- Sort by dept ascending, then salary descending
- Find the highest-paid person in each department (using `order` + `!duplicated`)
:::


## Sol 38

```{r}
df <- data.frame(dept=c("A","B","A","B","A"),
  name=c("Zoe","Amy","Bob","Cal","Amy"), salary=c(50,60,70,55,65))
df[order(df$dept, -df$salary), ]
sorted <- df[order(df$dept, -df$salary), ]
sorted[!duplicated(sorted$dept), ]
```


## Ex 39: Column-wise Operations

::: {.alertblock}
### Problem

- Use `colMeans` on `mtcars[,1:4]`
- Use `sapply` to find `max` of each column
- Standardize (z-score) all numeric columns using `scale()`. Compare with manual `sweep`
:::


## Sol 39

```{r}
colMeans(mtcars[,1:4])
sapply(mtcars[,1:4], max)
head(scale(mtcars[,1:2]), 3)
m <- as.matrix(mtcars[,1:2])
head(sweep(sweep(m, 2, colMeans(m)), 2, apply(m,2,sd), "/"), 3)
```


## Ex 40: Building a Gradebook

::: {.alertblock}
### Problem

- Create `grades <- data.frame(name=c("A","B","C","D","E"), hw=c(85,92,78,95,88), exam=c(90,85,92,88,76))`
- Add `final = 0.4*hw + 0.6*exam`
- Add `letter` using nested `ifelse` (A/B/C/F)
- Sort by `final` descending
:::


## Sol 40

```{r}
grades <- data.frame(name=c("A","B","C","D","E"),
  hw=c(85,92,78,95,88), exam=c(90,85,92,88,76))
grades$final <- 0.4*grades$hw + 0.6*grades$exam
grades$letter <- ifelse(grades$final>=90,"A",
  ifelse(grades$final>=80,"B",
  ifelse(grades$final>=70,"C","F")))
grades[order(-grades$final), ]
```


# Lists (Ex 41--52)


## Ex 41: Creating and Accessing Lists

::: {.alertblock}
### Problem

- Create `l <- list(name="Alice", scores=c(85,92,78), passed=TRUE)`
- Access `scores` with `$`, `[[]]`, and `[]`. Show the different classes
- Add element `grade = "B"`. Remove `passed`
- What is `length(l)` before and after?
:::


## Sol 41

```{r}
l <- list(name="Alice", scores=c(85,92,78), passed=TRUE)
l$scores; l[["scores"]]; l["scores"]
class(l$scores); class(l["scores"])                # vector vs list
l$grade <- "B"; l$passed <- NULL
length(l)  # now 3
```


## Ex 42: `unlist()` Behavior

::: {.alertblock}
### Problem

- `l <- list(a=1:3, b=4:5)`. What does `unlist(l)` produce? Explain the names
- `unlist(l, use.names=FALSE)`
- `l2 <- list(x="hello", y=1:3)`. What type is `unlist(l2)`? Why?
- Mean of a list of numbers: `list(10, 20, 30)`
:::


## Sol 42

```{r}
l <- list(a=1:3, b=4:5)
unlist(l)                # named: a1 a2 a3 b1 b2
unlist(l, use.names=FALSE)
unlist(list(x="hello", y=1:3))  # all character
mean(unlist(list(10, 20, 30)))                     # 20
```


## Ex 43: Nested Lists

::: {.alertblock}
### Problem

- Create `student <- list(info=list(name="Bob", age=20), grades=list(math=90, eng=85))`
- Access Bob's math grade
- Add `grades$sci = 92`
- Use `str()` to visualize the structure
:::


## Sol 43

```{r}
student <- list(info=list(name="Bob", age=20),
                grades=list(math=90, eng=85))
student$grades$math                                # 90
student$grades$sci <- 92
str(student)
```


## Ex 44: `split()` and Recombine

::: {.alertblock}
### Problem

- `split(mtcars$mpg, mtcars$cyl)` — what type?
- Compute `mean` per group using `sapply`
- Same using `tapply` in one line
- `split(1:10, rep(c("odd","even"), 5))`
:::


## Sol 44

```{r}
sp <- split(mtcars$mpg, mtcars$cyl); class(sp)    # list
sapply(sp, mean)
tapply(mtcars$mpg, mtcars$cyl, mean)
split(1:10, rep(c("odd","even"), 5))
```


## Ex 45: List Concatenation

::: {.alertblock}
### Problem

- Concatenate `list(a=1)` and `list(b=2)` with `c()`
- What happens with `c(list(a=1), 5)`?
- Build a list incrementally in a loop: for i=1:5, append `i^2`
:::


## Sol 45

```{r}
c(list(a=1), list(b=2))
c(list(a=1), 5)                                    # 5 becomes element
l <- list()
for (i in 1:5) l[[i]] <- i^2; l
```


## Ex 46: Conversion Table

::: {.alertblock}
### Problem

- Vector to list: `as.list(1:3)` vs `list(1:3)`. Explain difference
- List to data frame: `as.data.frame(list(a=1:3, b=4:6))`
- Data frame to matrix: what happens with `as.matrix(data.frame(x=1:3, y=c("a","b","c")))`?
- List to vector: why use `unlist` not `as.vector`?
:::


## Sol 46

```{r}
as.list(1:3)    # 3 elements
list(1:3)       # 1 element (the vector)
as.data.frame(list(a=1:3, b=4:6))
as.matrix(data.frame(x=1:3, y=c("a","b","c")))    # all char!
unlist(list(a=1, b=2))
```


## Ex 47: Named List as Lookup Table

::: {.alertblock}
### Problem

- Create a lookup: `grade_points <- list(A=4.0, B=3.0, C=2.0, D=1.0, F=0)`
- Look up GPA for grades `c("A","B","F","A","C")` using `sapply` + `[[`
- Compute mean GPA
:::


## Sol 47

```{r}
grade_points <- list(A=4.0, B=3.0, C=2.0, D=1.0, F=0)
g <- c("A","B","F","A","C")
gpa <- sapply(g, function(x) grade_points[[x]])
gpa
mean(gpa)                                          # 2.6
```


## Ex 48: `lapply`/`sapply` on Lists

::: {.alertblock}
### Problem
`l <- list(a=1:10, b=rnorm(20), c=c(TRUE,FALSE,TRUE,TRUE))`.

- Compute `mean` of each element with `lapply` and `sapply`
- Compute `length` of each
- Apply custom function: range (max - min)
:::


## Sol 48

```{r}
set.seed(1)
l <- list(a=1:10, b=rnorm(20), c=c(TRUE,FALSE,TRUE,TRUE))
lapply(l, mean); sapply(l, mean)
sapply(l, length)
sapply(l, function(x) diff(range(x)))
```


## Ex 49: `Filter`, `Find`, `Position`

::: {.alertblock}
### Problem
`l <- list(1:3, "hello", 4:6, TRUE, NULL, c(1.5, 2.5))`.

- Keep only numeric elements
- Find first character element
- Position of first character element
- Filter integers from `1:30` that are divisible by both 3 and 7
:::


## Sol 49

```{r}
l <- list(1:3, "hello", 4:6, TRUE, NULL, c(1.5, 2.5))
Filter(is.numeric, l)
Find(is.character, l)                              # "hello"
Position(is.character, l)                          # 2
Filter(\(x) x%%3==0 & x%%7==0, 1:30)              # 21
```


## Ex 50: `Reduce()`

::: {.alertblock}
### Problem

- Compute `1+2+3+4+5` using `Reduce("+", 1:5)`. Show accumulated steps
- Find intersection of 3 vectors via `Reduce(intersect, ...)`
- Merge 3 data frames pairwise using `Reduce`
:::


## Sol 50

```{r}
Reduce("+", 1:5)                                   # 15
Reduce("+", 1:5, accumulate=TRUE)                  # steps
vecs <- list(c(1,2,3,4), c(2,3,4,5), c(3,4,5,6))
Reduce(intersect, vecs)                            # 3 4
dfs <- list(data.frame(id=1:3, a=10:12),
            data.frame(id=2:4, b=20:22))
Reduce(\(x,y) merge(x,y,by="id",all=TRUE), dfs)
```


## Ex 51: List Modification Patterns

::: {.alertblock}
### Problem

- Given `l <- list(x=1, y=2, z=3)`, remove `x` and `z` in one operation
- Replace all numeric elements with their squares using `lapply`
- Deep copy a nested list and modify without affecting original
:::


## Sol 51

```{r}
l <- list(x=1, y=2, z=3)
l[c("x","z")] <- NULL; l                          # only y
l2 <- list(a=3, b="hi", c=4)
lapply(l2, function(x) if(is.numeric(x)) x^2 else x)
orig <- list(inner=list(val=10))
copy <- orig; copy$inner$val <- 99
orig$inner$val  # still 10 (R copies on modify)
```


## Ex 52: `rapply()` --- Recursive Apply

::: {.alertblock}
### Problem
`l <- list(a=1:3, b=list(c=4:6, d=list(e=7:9)))`.

- Apply `sum` to all leaf elements using `rapply(l, sum)`
- Use `rapply(l, sum, how="unlist")` vs `how="list"`
- Double all values with `rapply(l, function(x) x*2, how="replace")`
:::


## Sol 52

```{r}
l <- list(a=1:3, b=list(c=4:6, d=list(e=7:9)))
rapply(l, sum)
rapply(l, sum, how="unlist")                       # named vector
rapply(l, sum, how="list")                         # list structure
str(rapply(l, \(x) x*2, how="replace"))            # doubled
```


# Matrices (Ex 53--58)


## Ex 53: Creating Matrices

::: {.alertblock}
### Problem

- Create 3×4 matrix filled column-wise from `1:12`
- Create same matrix filled row-wise
- Name rows `r1,r2,r3` and columns `c1,...,c4`. Access `m["r2","c3"]`
- Create a 3×3 identity matrix with `diag`
:::


## Sol 53

```{r}
matrix(1:12, nrow=3)
matrix(1:12, nrow=3, byrow=TRUE)
m <- matrix(1:12, 3, 4)
rownames(m) <- paste0("r",1:3); colnames(m) <- paste0("c",1:4)
m["r2","c3"]                                       # 8
diag(3)
```


## Ex 54: Matrix Arithmetic

::: {.alertblock}
### Problem
`A <- matrix(1:4, 2, 2); B <- matrix(5:8, 2, 2)`.

- Element-wise `*` vs matrix `%*%`
- Compute `t(A)`, `det(A)`, `solve(A)`
- Verify `A %*% solve(A)` is the identity
- `crossprod(A,B)` vs `t(A) %*% B`
:::


## Sol 54

```{r}
A <- matrix(1:4, 2, 2); B <- matrix(5:8, 2, 2)
A * B; A %*% B
t(A); det(A); solve(A)
round(A %*% solve(A))                              # identity
all.equal(crossprod(A,B), t(A) %*% B)             # TRUE
```


## Ex 55: `apply()` on Matrices

::: {.alertblock}
### Problem
`m <- matrix(1:12, nrow=3)`.

- Row sums and column sums via `apply`
- Per-column range
- Standardize each column (subtract mean, divide by sd)
- Compare `apply(m,2,sum)` speed with `colSums(m)`
:::


## Sol 55

```{r}
m <- matrix(1:12, nrow=3)
apply(m, 1, sum); apply(m, 2, sum)
apply(m, 2, range)
apply(m, 2, \(x) (x-mean(x))/sd(x))
all.equal(apply(m,2,sum), colSums(m))              # TRUE
```


## Ex 56: `outer()` Product

::: {.alertblock}
### Problem

- Multiplication table 1..5 × 1..5 using `outer`
- Compute $f(x,y)=x^2+y^2$ for `x=1:4`, `y=1:3`
- All pairwise `paste` of `c("a","b")` and `c("1","2","3")`
:::


## Sol 56

```{r}
outer(1:5, 1:5, "*")
outer(1:4, 1:3, \(x,y) x^2 + y^2)
outer(c("a","b"), c("1","2","3"), paste0)
```


## Ex 57: Matrix Row/Column Selection

::: {.alertblock}
### Problem
`m <- matrix(1:12, 3, 4)`.

- Extract row 2 as a vector vs as a 1-row matrix (`drop`)
- Extract column 3 both ways
- Select submatrix: rows 1-2, columns 2-3
- Replace all elements > 8 with `NA`
:::


## Sol 57

```{r}
m <- matrix(1:12, 3, 4)
m[2,]; m[2,,drop=FALSE]
m[,3]; m[,3,drop=FALSE]
m[1:2, 2:3]
m[m > 8] <- NA; m
```


## Ex 58: Linear System

::: {.alertblock}
### Problem
Solve $2x + 3y = 7, \; 4x - y = 1$.

- Set up as `A %*% x = b` and solve with `solve(A, b)`
- Verify by multiplying back
- Compute eigenvalues of `A`
:::


## Sol 58

```{r}
A <- matrix(c(2,4,3,-1), 2, 2); b <- c(7,1)
x <- solve(A, b); x
A %*% x  # should be c(7,1)
eigen(A)$values
```


# Programming (Ex 59--84)


## Ex 59: Default Arguments

::: {.alertblock}
### Problem

- Write `power(x, n=2)` that computes `x^n`. Test with 1 and 2 arguments
- Write `greet(name, greeting="Hello")` that returns `"Hello, Alice!"`
- What happens if you call `power(n=3)` without `x`?
:::


## Sol 59

```{r, error=TRUE}
power <- function(x, n=2) x^n
power(3); power(3, 3); power(x=5, n=2)
greet <- function(name, greeting="Hello")
  paste0(greeting, ", ", name, "!")
greet("Alice"); greet("Bob", "Hi")
power(n=3)  # error: x is missing
```


## Ex 60: Functions as Arguments

::: {.alertblock}
### Problem

- Write `apply_fn(x, fn)` that applies `fn` to `x`. Test with `sqrt`, `log`, `abs`
- Extend to `apply_fn(x, fn, ...)` supporting extra arguments. Test: `apply_fn(c(1,NA,3), mean, na.rm=TRUE)`
:::


## Sol 60

```{r}
apply_fn <- function(x, fn, ...) fn(x, ...)
apply_fn(c(4,9,16), sqrt)
apply_fn(c(1,NA,3), mean, na.rm=TRUE)             # 2
```


## Ex 61: Implicit Return

::: {.alertblock}
### Problem

- Write two versions of average: one with explicit `return()`, one without
- Write `bad_avg` that assigns to a local variable as last line. Show it returns silently
- Capture the invisible result with `x <- bad_avg(1:10)`
:::


## Sol 61

```{r}
avg_explicit <- function(x) { return(sum(x)/length(x)) }
avg_implicit <- function(x) { sum(x)/length(x) }
avg_explicit(1:10); avg_implicit(1:10)             # same
bad_avg <- function(x) { result <- sum(x)/length(x) }
bad_avg(1:10)  # no output!
x <- bad_avg(1:10); x  # but value is captured
```


## Ex 62: Anonymous Functions

::: {.alertblock}
### Problem

- Use `sapply(1:5, function(x) x^2 + 1)`
- Rewrite with `\(x)` shorthand
- Use anonymous function in `Filter` to keep even numbers from `1:20`
- Nested: `sapply(1:3, \(i) sapply(1:3, \(j) i*j))`
:::


## Sol 62

```{r}
sapply(1:5, function(x) x^2+1)
sapply(1:5, \(x) x^2+1)
Filter(\(x) x%%2==0, 1:20)
sapply(1:3, \(i) sapply(1:3, \(j) i*j))            # 3x3 matrix
```


## Ex 63: Closures

::: {.alertblock}
### Problem

- Write `make_adder(n)` that returns a function adding `n`. Test `add5 <- make_adder(5); add5(3)`
- Write `make_counter()` returning a list with `increment`, `get`, `reset` functions using `<<-`
- Verify the counter works across multiple calls
:::


## Sol 63

```{r}
make_adder <- function(n) function(x) x + n
add5 <- make_adder(5); add5(3)                     # 8
make_counter <- function() {
  n <- 0
  list(inc = function() {n <<- n+1; n},
       get = function() n,
       reset = function() {n <<- 0})
}
ctr <- make_counter()
ctr$inc(); ctr$inc(); ctr$get()                    # 2
ctr$reset(); ctr$get()                             # 0
```


## Ex 64: Dot-Prefix Naming

::: {.alertblock}
### Problem

- Define `.helper <- function(x) x+1` and `public <- function(x) .helper(x)*2`
- Show `.helper` doesn't appear in `ls()` but does in `ls(all.names=TRUE)`
- Name 3 built-in R objects starting with `.`
:::


## Sol 64

```{r}
e <- new.env()
e$.helper <- function(x) x+1
e$public  <- function(x) e$.helper(x)*2
e$public(5)  # 12
ls(e); ls(e, all.names=TRUE)
# .Machine, .Platform, .GlobalEnv
```


## Ex 65: `if (FALSE)` Idiom

::: {.alertblock}
### Problem

- Wrap code in `if (FALSE) \{ ... \`} and show it doesn't execute
- Change to `if (TRUE)` and show it runs
- Why is this better than commenting with `#` for multi-line blocks?
:::


## Sol 65

```{r}
if (FALSE) {
  cat("This won't print\n")
  x <- 999
}
exists("x")  # from previous chunks, may be TRUE
if (TRUE) { cat("This prints\n") }
# Unlike #, the code inside is still parsed by R,
# so syntax errors are caught immediately.
```


## Ex 66: Local Helper Functions

::: {.alertblock}
### Problem

- Write `outer_fn(x)` that defines inner helper `validate(v)` checking `is.numeric(v) && length(v) > 0`, then uses it
- Show that `validate` is not accessible outside `outer_fn`
- Why is this pattern useful?
:::


## Sol 66

```{r, error=TRUE}
outer_fn <- function(x) {
  validate <- function(v) is.numeric(v) && length(v) > 0
  if (!validate(x)) stop("invalid input")
  mean(x)
}
outer_fn(1:10)                                     # 5.5
outer_fn("abc")                                    # error
# validate is local, not in global env
# keeps global namespace clean; helper has access to
#     outer function's variables via lexical scoping
```


## Ex 67: `if`/`else` and `ifelse`

::: {.alertblock}
### Problem

- Write scalar `classify(x)` returning "pos", "neg", or "zero" using `if/else`
- Vectorize: classify `c(-3,0,5,2,-1)` using `ifelse`
- Nested `ifelse` to map scores to grades: $\geq 90$ A, $\geq 80$ B, $\geq 70$ C, else F
:::


## Sol 67

```{r}
classify <- function(x) {
  if (x > 0) "pos" else if (x < 0) "neg" else "zero"
}
classify(5); classify(-1); classify(0)
ifelse(c(-3,0,5,2,-1)>0, "pos",
  ifelse(c(-3,0,5,2,-1)<0, "neg", "zero"))
s <- c(95,82,71,65,90)
ifelse(s>=90,"A",ifelse(s>=80,"B",ifelse(s>=70,"C","F")))
```


## Ex 68: `switch()`

::: {.alertblock}
### Problem

- Write `calc(op, a, b)` using `switch(op, "+"=a+b, "-"=a-b, "*"=a*b, "/"=a/b, stop("unknown"))`
- Test with all 4 operators and an invalid one
- Write `day_type(day)` classifying weekdays vs weekends
:::


## Sol 68

```{r, error=TRUE}
calc <- function(op, a, b)
  switch(op, "+"=a+b, "-"=a-b, "*"=a*b, "/"=a/b,
         stop("unknown op"))
calc("+",3,4); calc("*",5,6); calc("/",10,3)
calc("^",2,3)  # error
day_type <- function(d) switch(d, Monday=,Tuesday=,
  Wednesday=,Thursday=,Friday="Weekday",
  Saturday=,Sunday="Weekend","Unknown")
day_type("Monday"); day_type("Sunday")
```


## Ex 69: `&&`/`||` vs `&`/`|`

::: {.alertblock}
### Problem

- Show `&&` short-circuits: `FALSE && stop("boom")` doesn't error
- Show `&` doesn't short-circuit: `c(FALSE,TRUE) & c(TRUE,TRUE)` is vectorized
- Write safe null-check: `if (!is.null(x) && x > 0)` and explain why `&&` is required
:::


## Sol 69

```{r}
FALSE && stop("boom")  # no error: short-circuit
c(FALSE,TRUE) & c(TRUE,TRUE)                      # vectorized
x <- NULL
if (!is.null(x) && x > 0) cat("pos") else cat("safe")
# && stops after !is.null(x) is FALSE; & would try x>0
```


## Ex 70: `for` Loop --- FizzBuzz

::: {.alertblock}
### Problem

- Classic FizzBuzz 1-20: print "Fizz" if divisible by 3, "Buzz" if by 5, "FizzBuzz" if both, else the number
- Vectorized version using `ifelse` (no loop)
- Compare with `paste0` trick
:::


## Sol 70

```{r}
for (i in 1:20) {
  if (i%%15==0) cat("FizzBuzz ")
  else if (i%%3==0) cat("Fizz ")
  else if (i%%5==0) cat("Buzz ")
  else cat(i, "")
}
cat("\n")
x <- 1:20
ifelse(x%%15==0,"FizzBuzz",ifelse(x%%3==0,"Fizz",
  ifelse(x%%5==0,"Buzz",as.character(x))))
```


## Ex 71: `while` Loop --- Newton's Method

::: {.alertblock}
### Problem
Implement Newton's method for $\sqrt{a}$: $x_{n+1} = \frac{1}{2}(x_n + a/x_n)$.

- Write `my_sqrt(a, tol=1e-10)` using `while`
- Test on 2, 9, 100
- Count iterations needed for `a=2`
:::


## Sol 71

```{r}
my_sqrt <- function(a, tol=1e-10) {
  x <- a; n <- 0
  while (TRUE) {
    x_new <- 0.5 * (x + a/x); n <- n + 1
    if (abs(x_new - x) < tol) break
    x <- x_new
  }
  list(root=x_new, iterations=n)
}
my_sqrt(2); my_sqrt(9)$root; my_sqrt(100)$root
my_sqrt(2)$iterations
```


## Ex 72: `repeat`/`break` --- Collatz

::: {.alertblock}
### Problem
Collatz: if $n$ even, $n/2$; if odd, $3n+1$. Conjecture: always reaches 1.

- Write `collatz(n)` returning the sequence using `repeat`/`break`
- Find the starting number 1--100 with the longest sequence
- Plot the sequence for `n=27`
:::


## Sol 72

```{r, out.width="35%"}
collatz <- function(n) {
  seq <- n
  repeat { if (n==1) break; n <- if(n%%2==0) n/2 else 3*n+1
           seq <- c(seq, n) }
  seq
}
lengths <- sapply(1:100, \(n) length(collatz(n)))
which.max(lengths)  # starting number
plot(collatz(27), type="l", main="Collatz(27)")
```


## Ex 73: `next` --- Filtered Sum

::: {.alertblock}
### Problem

- Sum integers 1-100 that are NOT multiples of 7, using `for`+`next`
- Same with vectorized filtering (no loop)
- Sum only prime numbers 1-50. Write `is_prime(n)` helper
:::


## Sol 73

```{r}
s <- 0; for(i in 1:100) {if(i%%7==0) next; s<-s+i}; s
sum((1:100)[(1:100)%%7 != 0])                      # same
is_prime <- function(n) {
  if (n < 2) return(FALSE)
  if (n == 2) return(TRUE)
  all(n %% 2:floor(sqrt(n)) != 0)
}
sum(Filter(is_prime, 1:50))
```


## Ex 74: Vectorization Speed

::: {.alertblock}
### Problem

- Write loop version of `sum(x^2)` for `x <- 1:1e5`
- Compare `system.time` with vectorized `sum(x^2)`
- Pre-allocate vs grow-in-loop: time `v <- c(); for(...) v <- c(v,i)` vs `v <- numeric(n); for(...) v[i] <- i`
:::


## Sol 74

```{r}
n <- 1e5; x <- 1:n
loop_sum <- function(x) {s<-0; for(v in x) s<-s+v^2; s}
system.time(loop_sum(x))                           # slow
system.time(sum(x^2))                              # fast
system.time({v<-c(); for(i in 1:n) v<-c(v,i)})    # very slow
system.time({v<-numeric(n); for(i in 1:n) v[i]<-i}) # faster
```


## Ex 75: Sieve of Eratosthenes

::: {.alertblock}
### Problem

- Implement `sieve(n)` returning all primes up to `n` using a logical vector and nested loops
- Test with `sieve(50)`
- How many primes below 1000?
:::


## Sol 75

```{r}
sieve <- function(n) {
  is_p <- rep(TRUE, n); is_p[1] <- FALSE
  for (i in 2:floor(sqrt(n)))
    if (is_p[i]) is_p[seq(i*2, n, i)] <- FALSE
  which(is_p)
}
sieve(50)
length(sieve(1000))                                # 168
```


## Ex 76: Recursive Factorial and Fibonacci

::: {.alertblock}
### Problem

- Write `my_fact(n)` using `Recall`
- Write `fib(n)` using `Recall`
- Compute `sapply(1:10, fib)`
- Why is naive recursive `fib` slow for large `n`?
:::


## Sol 76

```{r}
my_fact <- function(n) {
  if (n <= 1) return(1); n * Recall(n-1)
}
my_fact(5); my_fact(10)
fib <- function(n) {
  if (n <= 2) return(1); Recall(n-1) + Recall(n-2)
}
sapply(1:10, fib)
# exponential time: each call branches into 2
```


## Ex 77: `match.arg` and `missing`

::: {.alertblock}
### Problem

- Write `summary_stat(x, type=c("mean","median","sd"))` using `match.arg`. Test partial matching
- Write `f(x, y)` that checks `missing(y)` and uses default `y=0`
- Show `nargs()` inside a function
:::


## Sol 77

```{r}
summary_stat <- function(x, type=c("mean","median","sd")) {
  type <- match.arg(type)
  switch(type, mean=mean(x), median=median(x), sd=sd(x))
}
summary_stat(1:10, "mea")                          # mean
f <- function(x, y) {
  if (missing(y)) y <- 0; x + y
}
f(5); f(5, 3)
g <- function(a,b,c) nargs(); g(1); g(1,2,3)
```


## Ex 78: `invisible` and Side Effects

::: {.alertblock}
### Problem

- Write `my_print(x)` that `cat`s and returns `invisible(x)`. Show it doesn't print the return
- Capture the invisible return
- Write `log_and_return(x)` that prints a message and returns `invisible(NULL)`
:::


## Sol 78

```{r}
my_print <- function(x) { cat("Value:", x, "\n"); invisible(x) }
my_print(42)          # prints message, no [1] 42
val <- my_print(42); val                           # captured
log_return <- function(x) {
  cat("Processed:", x, "\n"); invisible(NULL)
}
r <- log_return(5); is.null(r)                     # TRUE
```


## Ex 79: `on.exit` with `add=TRUE`

::: {.alertblock}
### Problem

- Write function that sets `options(digits=3)`, uses `on.exit` to restore
- Write function with two `on.exit(add=TRUE)` calls. Show both run
- Show cleanup runs even on error
:::


## Sol 79

```{r}
show_digits <- function() {
  old <- options(digits=3)
  on.exit(options(old))
  cat("Inside:", getOption("digits"), "\n")
}
show_digits(); cat("Outside:", getOption("digits"), "\n")
two_exits <- function() {
  on.exit(cat("exit1\n"), add=TRUE)
  on.exit(cat("exit2\n"), add=TRUE)
  cat("body\n")
}
two_exits()
tryCatch({
  f <- function() { on.exit(cat("cleanup!\n")); stop("err") }
  f()
}, error = function(e) cat("caught:", e$message, "\n"))
```


## Ex 80: `Vectorize()`

::: {.alertblock}
### Problem

- Write scalar `classify(x)` using `if/else` (not vectorized)
- Use `Vectorize` to create vector version. Test on `c(-2, 0, 5)`
- Why is `Vectorize` just a wrapper around `mapply`?
:::


## Sol 80

```{r}
classify <- function(x) {
  if (x > 0) "pos" else if (x < 0) "neg" else "zero"
}
classify_v <- Vectorize(classify)
classify_v(c(-2, 0, 5))
# Vectorize calls mapply internally, applying the
# scalar function element-wise. Not truly vectorized (still loops).
```


## Ex 81: Native Pipe `|>`

::: {.alertblock}
### Problem

- Rewrite `round(mean(abs(c(-3,4,-1,2))), 2)` using `|>`
- Pipe `mtcars` through `subset` then `head`
- Use `_` placeholder: `mtcars |> lm(mpg ~ wt, data = _)`
:::


## Sol 81

```{r}
c(-3,4,-1,2) |> abs() |> mean() |> round(2)
mtcars |> subset(cyl == 4) |> head(3)
mtcars |> lm(mpg ~ wt, data = _) |>
  coef()
```


## Ex 82: `do.call`

::: {.alertblock}
### Problem

- Use `do.call(paste, c(as.list(c("a","b","c")), sep="-"))`
- Use `do.call(rbind, list_of_dfs)` to stack 3 data frames
- Build a call to `seq` dynamically from a list of arguments
:::


## Sol 82

```{r}
do.call(paste, c(as.list(c("a","b","c")), sep="-"))
dfs <- list(data.frame(x=1,y=2), data.frame(x=3,y=4))
do.call(rbind, dfs)
do.call(seq, list(from=1, to=10, by=2))
```


## Ex 83: `tryCatch` and `try`

::: {.alertblock}
### Problem

- Use `tryCatch(log("a"), error=function(e) NA)`
- Loop over `list(1,"a",4)`, wrapping each in `try`. Keep only successes
- Write `safe_div(a,b)` that `stop`s on `b==0` and test with `tryCatch`
:::


## Sol 83

```{r}
tryCatch(log("a"), error=function(e) NA)
inputs <- list(1, "a", 4)
results <- lapply(inputs, \(x) try(sqrt(x), silent=TRUE))
Filter(\(r) !inherits(r,"try-error"), results)
safe_div <- function(a,b) {if(b==0) stop("div0!"); a/b}
tryCatch(safe_div(10,0), error=function(e) Inf)
```


## Ex 84: `stopifnot` and `warning`

::: {.alertblock}
### Problem

- Write `my_mean(x)` with `stopifnot(is.numeric(x), length(x)>0)`
- Write `safe_sqrt(x)` that issues `warning` for negative input and returns `sqrt(abs(x))`
- Use `suppressWarnings` to silence it
:::


## Sol 84

```{r, error=TRUE}
my_mean <- function(x) {
  stopifnot(is.numeric(x), length(x) > 0)
  sum(x)/length(x)
}
my_mean(1:5); my_mean("abc")
safe_sqrt <- function(x) {
  if (any(x < 0)) warning("Negative values!")
  sqrt(abs(x))
}
safe_sqrt(c(4, -9, 16))
suppressWarnings(safe_sqrt(-1))
```


# Apply Family (Ex 85--96)


## Ex 85: `apply` --- Row and Column

::: {.alertblock}
### Problem
`m <- matrix(c(3,1,4,1,5,9,2,6,5,3,5,8), nrow=3)`.

- Column means via `apply`
- Row ranges
- Apply custom function returning a named vector per column
- Apply on a data frame's numeric columns
:::


## Sol 85

```{r}
m <- matrix(c(3,1,4,1,5,9,2,6,5,3,5,8), nrow=3)
apply(m, 2, mean)
apply(m, 1, range)                                 # 2x3 matrix
apply(m, 2, \(x) c(mu=mean(x), sd=sd(x)))
apply(mtcars[1:3, 1:4], 2, mean)
```


## Ex 86: `lapply` and `sapply`

::: {.alertblock}
### Problem
`l <- list(a=1:10, b=seq(0,1,by=0.1), c=c(T,F,T,T,F))`.

- `sapply(l, length)`
- `sapply(l, range)` — what shape?
- `vapply(l, mean, numeric(1))`
- Why prefer `vapply` over `sapply`?
:::


## Sol 86

```{r}
l <- list(a=1:10, b=seq(0,1,by=0.1), c=c(T,F,T,T,F))
sapply(l, length)                                  # named vector
sapply(l, range)                                   # 2x3 matrix
vapply(l, mean, numeric(1))                        # named vector
# vapply guarantees return type; sapply may return list
# unexpectedly if elements have different lengths
```


## Ex 87: `tapply`

::: {.alertblock}
### Problem

- Mean `mpg` by `cyl` in `mtcars`
- Two-way: mean `mpg` by `cyl` × `am`
- Use `tapply` on `ToothGrowth`: mean `len` by `supp` and `dose`
:::


## Sol 87

```{r}
tapply(mtcars$mpg, mtcars$cyl, mean)
tapply(mtcars$mpg, list(mtcars$cyl, mtcars$am), mean)
tapply(ToothGrowth$len,
  list(ToothGrowth$supp, ToothGrowth$dose), mean)
```


## Ex 88: `mapply` and `Map`

::: {.alertblock}
### Problem

- `mapply(function(x,y) x+y, 1:5, 10:14)`
- `mapply(rep, 1:4, 4:1)`
- `Map("+", 1:3, 10:12)` — what type?
- Use `Map` to apply `paste` element-wise to two vectors
:::


## Sol 88

```{r}
mapply(function(x,y) x+y, 1:5, 10:14)
mapply(rep, 1:4, 4:1)                              # list
Map("+", 1:3, 10:12)                               # list
Map(paste, c("a","b","c"), 1:3, sep="-")
```


## Ex 89: `Reduce` Applications

::: {.alertblock}
### Problem

- Factorial via `Reduce("*", 1:6)`
- Cumulative products via `accumulate=TRUE`
- Find the common elements of 4 vectors using `Reduce(intersect, ...)`
:::


## Sol 89

```{r}
Reduce("*", 1:6)                                   # 720
Reduce("*", 1:6, accumulate=TRUE)
Reduce(intersect, list(1:10, 3:12, 5:15, 7:20))   # 7:10
```


## Ex 90: Functional Pipeline

::: {.alertblock}
### Problem
Process `x <- c(-5, 3, -1, 8, -2, 7)`:

- Filter positive → square → sum. Do with nested calls
- Same with pipe `|>`
- Same with `Reduce` composing 3 functions
:::


## Sol 90

```{r}
x <- c(-5, 3, -1, 8, -2, 7)
sum(x[x>0]^2)                                     # nested
x[x>0] |> (\(v) v^2)() |> sum()                    # pipe
fns <- list(\(v) v[v>0], \(v) v^2, sum)
Reduce(\(v,f) f(v), fns, init=x)
```


## Ex 91: `tapply` + `aggregate` Pipeline

::: {.alertblock}
### Problem
Using `iris`:

- Mean `Sepal.Length` by `Species` with `tapply`
- Same with `aggregate`
- Mean of all numeric columns by `Species` with `aggregate(. ~ Species, ...)`
:::


## Sol 91

```{r}
tapply(iris$Sepal.Length, iris$Species, mean)
aggregate(Sepal.Length ~ Species, data=iris, mean)
aggregate(. ~ Species, data=iris, mean)
```


## Ex 92: Memoization via Closure

::: {.alertblock}
### Problem

- Write `make_memo(fn)` that returns a memoized version storing results in a list environment
- Memoize `fib`
- Show second call is instant
:::


## Sol 92

```{r}
make_memo <- function(fn) {
  cache <- list()
  function(...) {
    key <- paste(..., sep=",")
    if (is.null(cache[[key]])) cache[[key]] <<- fn(...)
    cache[[key]]
  }
}
slow_fn <- function(x) { Sys.sleep(0.1); x^2 }
fast_fn <- make_memo(slow_fn)
system.time(fast_fn(5))                            # slow
system.time(fast_fn(5))                            # instant
```


## Ex 93: `lapply` with Index

::: {.alertblock}
### Problem

- Given `l <- list(a=1:3, b=4:6)`, use `lapply(seq_along(l), ...)` to create `"a: 1 2 3"`, `"b: 4 5 6"`
- Same with `Map` and `names(l)`
- Same with `mapply`
:::


## Sol 93

```{r}
l <- list(a=1:3, b=4:6)
lapply(seq_along(l), \(i)
  paste0(names(l)[i], ": ", paste(l[[i]], collapse=" ")))
Map(\(nm, v) paste0(nm, ": ", paste(v,collapse=" ")),
    names(l), l)
mapply(\(nm,v) paste0(nm,": ",paste(v,collapse=" ")),
       names(l), l)
```


## Ex 94: Nested `lapply`

::: {.alertblock}
### Problem

- Generate a 3×3 multiplication table using `lapply(1:3, \(i) sapply(1:3, \(j) i*j))`
- Convert to a matrix with `do.call(rbind, ...)`
- Compare with `outer(1:3, 1:3)`
:::


## Sol 94

```{r}
l <- lapply(1:3, \(i) sapply(1:3, \(j) i*j))
do.call(rbind, l)
outer(1:3, 1:3)                                    # same
```


## Ex 95: `Map` for Parallel Processing

::: {.alertblock}
### Problem
`prices <- list(c(100,102,98), c(50,55,52), c(200,210,205))`.
`names_v <- c("AAPL", "GOOG", "MSFT")`.

- Compute return `diff(p)/p[-length(p)]` for each using `Map`
- Compute mean return per stock
:::


## Sol 95

```{r}
prices <- list(c(100,102,98), c(50,55,52), c(200,210,205))
names_v <- c("AAPL","GOOG","MSFT")
returns <- Map(\(p) diff(p)/p[-length(p)], prices)
names(returns) <- names_v
sapply(returns, mean)
```


## Ex 96: `Filter` + `Reduce` Combo

::: {.alertblock}
### Problem

- From `1:100`, filter multiples of 3, then sum with `Reduce`
- From a list of vectors, keep only those with length > 3, then find their common intersection
- Compare with direct `sum(Filter(...))`
:::


## Sol 96

```{r}
Reduce("+", Filter(\(x) x%%3==0, 1:100))
vecs <- list(1:10, 3:8, 2:15, 5:12, 4:20)
long <- Filter(\(v) length(v)>3, vecs)
Reduce(intersect, long)                            # 5:8
sum(Filter(\(x) x%%3==0, 1:100))                    # same as (a)
```


# Strings and Regex (Ex 97--110)


## Ex 97: Basic String Functions

::: {.alertblock}
### Problem
`x <- c("  Hello World  ", "R Programming", "data SCIENCE")`.

- Trim whitespace; count chars of each
- Convert to uppercase, lowercase
- `startsWith` / `endsWith` tests
- Extract first 4 chars using `substr`
:::


## Sol 97

```{r}
x <- c("  Hello World  ","R Programming","data SCIENCE")
xt <- trimws(x); nchar(xt)
toupper(xt); tolower(xt)
startsWith(xt, "Hello"); endsWith(xt, "SCIENCE")
substr(xt, 1, 4)
```


## Ex 98: `paste` and `sprintf`

::: {.alertblock}
### Problem

- `paste("item", 1:5, sep="_")`
- Same with `collapse=", "`
- `sprintf("Name: %s, Score: %.1f", "Alice", 95.678)`
- Zero-pad: `sprintf("%03d", 1:5)`
:::


## Sol 98

```{r}
paste("item", 1:5, sep="_")
paste("item", 1:5, sep="_", collapse=", ")
sprintf("Name: %s, Score: %.1f", "Alice", 95.678)
sprintf("%03d", 1:5)
```


## Ex 99: `strsplit`

::: {.alertblock}
### Problem

- Split `"a.b.c.d"` by `"."` (note the escape!)
- Split `"hello world foo"` by space
- Split each character of `"ABCDE"`
- Split `c("a-b", "c-d-e")` — what shape is the result?
:::


## Sol 99

```{r}
strsplit("a.b.c.d", "\\.")
strsplit("hello world foo", " ")
strsplit("ABCDE", "")
strsplit(c("a-b","c-d-e"), "-")                    # list of 2
```


## Ex 100: `grep` and `grepl`

::: {.alertblock}
### Problem
`w <- c("statistics","computation","algorithm","distribution","estimation")`.

- Words containing `"tion"`
- Words starting with vowel
- Words of $\geq$ 12 characters
- `grepl` to create logical mask
:::


## Sol 100

```{r}
w <- c("statistics","computation","algorithm",
       "distribution","estimation")
grep("tion", w, value=TRUE)
grep("^[aeiou]", w, value=TRUE)
w[nchar(w) >= 12]
grepl("tion", w)
```


## Ex 101: `sub` and `gsub`

::: {.alertblock}
### Problem

- Replace first space with `"_"` in `"hello world foo"`
- Replace all spaces
- Remove all digits from `"abc123def456"`
- Compress multiple spaces to one
:::


## Sol 101

```{r}
sub(" ", "_", "hello world foo")                   # first only
gsub(" ", "_", "hello world foo")                  # all
gsub("[0-9]", "", "abc123def456")
gsub(" +", " ", "too   many    spaces")
```


## Ex 102: Capture Groups

::: {.alertblock}
### Problem

- Swap first/last name: `sub("(\\w+) (\\w+)", "\\2, \\1", "John Smith")`
- Extract domain from email: `sub(".*@", "", "user@example.com")`
- Extract year from dates using capture group
:::


## Sol 102

```{r}
sub("(\\w+) (\\w+)", "\\2, \\1",
    c("John Smith","Jane Doe"))
sub(".*@", "", c("user@example.com","a@b.org"))
sub("(\\d{4})-\\d{2}-\\d{2}", "\\1",
    c("2024-01-15","2023-12-31"))                  # year
```


## Ex 103: `regexpr` and `gregexpr`

::: {.alertblock}
### Problem
`x <- "Phone: 02-1234-5678, Fax: 04-8765-4321"`.

- Extract all phone numbers with `gregexpr` + `regmatches`
- Extract first match only with `regexpr`
- Extract all words with `\\b\\w+\\b`
:::


## Sol 103

```{r}
x <- "Phone: 02-1234-5678, Fax: 04-8765-4321"
m <- gregexpr("[0-9]+-[0-9]+-[0-9]+", x)
regmatches(x, m)
m1 <- regexpr("[0-9]+-[0-9]+-[0-9]+", x)
regmatches(x, m1)
regmatches(x, gregexpr("\\b\\w+\\b", x))
```


## Ex 104: Greedy vs Lazy

::: {.alertblock}
### Problem
`x <- "<b>bold</b> and <i>italic</i>"`.

- `gregexpr("<.*>", x)` + `regmatches` — what matches? (greedy)
- `gregexpr("<.*?>", x)` — what matches? (lazy)
- Explain the difference
:::


## Sol 104

```{r}
x <- "<b>bold</b> and <i>italic</i>"
regmatches(x, gregexpr("<.*>", x))                 # one big match
regmatches(x, gregexpr("<.*?>", x))                # 4 tags
# greedy: .* matches as much as possible (all the way
#     to last >); lazy: .*? matches as little as possible
```


## Ex 105: Anchors and Quantifiers

::: {.alertblock}
### Problem
`x <- c("cat","concatenate","scat","catalog")`.

- Starts with `cat`
- Ends with `cat`
- Exactly `cat`
- `c("a","ab","abb","abbb")`: match `ab+` vs `ab*` vs `ab?`
:::


## Sol 105

```{r}
x <- c("cat","concatenate","scat","catalog")
grep("^cat", x, value=TRUE)
grep("cat$", x, value=TRUE)
grep("^cat$", x, value=TRUE)
y <- c("a","ab","abb","abbb")
grep("ab+", y, value=TRUE); grep("ab*", y, value=TRUE)
```


## Ex 106: POSIX Classes

::: {.alertblock}
### Problem
`x <- c("Hello!", "123", "abc", "A1!")`.

- Contains punctuation
- All digits
- Contains only alphabetic
- All alphanumeric
:::


## Sol 106

```{r}
x <- c("Hello!", "123", "abc", "A1!")
grep("[[:punct:]]", x, value=TRUE)
grep("^[[:digit:]]+$", x, value=TRUE)
grep("^[[:alpha:]]+$", x, value=TRUE)
grep("^[[:alnum:]]+$", x, value=TRUE)
```


## Ex 107: Lookahead/Lookbehind

::: {.alertblock}
### Problem

- Extract numbers preceded by `$`: `c("$100","200","$300")`
- Find files NOT ending in `.txt`: `c("a.txt","b.png","c.txt","d.jpg")`
- Explain why `perl=TRUE` is needed
:::


## Sol 107

```{r}
x <- c("$100","200","$300")
regmatches(x, gregexpr("(?<=\\$)\\d+", x, perl=TRUE))
y <- c("a.txt","b.png","c.txt","d.jpg")
grep("\\.(?!txt$)", y, value=TRUE, perl=TRUE)
# lookahead/lookbehind are Perl-compatible regex
# extensions not available in base (POSIX) regex
```


## Ex 108: `regexec` --- Structured Extraction

::: {.alertblock}
### Problem
`dates <- c("2024-01-15 14:30", "2023-12-31 09:00")`.

- Write pattern to capture year, month, day, hour, minute separately
- Extract with `regexec` + `regmatches`
- Build a data frame from the results
:::


## Sol 108

```{r}
dates <- c("2024-01-15 14:30", "2023-12-31 09:00")
pat <- "(\\d{4})-(\\d{2})-(\\d{2}) (\\d{2}):(\\d{2})"
m <- regexec(pat, dates)
parts <- regmatches(dates, m)
do.call(rbind, lapply(parts, \(p)
  setNames(p[-1], c("Y","M","D","h","m"))))
```


## Ex 109: camelCase ↔ snake_case

::: {.alertblock}
### Problem

- Write `camel_to_snake(s)` converting `"camelCase"` → `"camel_case"`
- Write `snake_to_camel(s)` converting `"snake_case"` → `"snakeCase"`
- Test both on `"myHTTPRequest"` / `"my_http_request"`
:::


## Sol 109

```{r}
camel_to_snake <- function(s) tolower(gsub("([A-Z])","_\\1",s))
camel_to_snake("camelCase")
camel_to_snake("myHTTPRequest")
snake_to_camel <- function(s) {
  parts <- strsplit(s, "_")[[1]]
  paste0(parts[1], paste0(toupper(substring(parts[-1],1,1)),
         substring(parts[-1],2), collapse=""))
}
snake_to_camel("my_http_request")
```


## Ex 110: Clean Column Names

::: {.alertblock}
### Problem
`cols <- c("First Name", "Last.Name", "Age (years)", "Income$")`.

- Replace all non-alphanumeric chars with `_`
- Remove leading/trailing `_`
- Collapse consecutive `_` to single `_`
- Convert to lowercase
:::


## Sol 110

```{r}
cols <- c("First Name","Last.Name","Age (years)","Income$")
x <- gsub("[^a-zA-Z0-9]", "_", cols)
x <- gsub("^_+|_+$", "", x)
x <- gsub("_+", "_", x)
tolower(x)
```


# Date/Time, I/O, Error Handling, Advanced (Ex 111--120)


## Ex 111: Date Arithmetic

::: {.alertblock}
### Problem

- Create today's date with `Sys.Date()`. Add 30 days
- Days between `2024-01-01` and `2024-12-31`
- Generate monthly sequence Jan--Jun 2024
- Extract weekday name and month name from a date
:::


## Sol 111

```{r}
today <- Sys.Date(); today + 30
as.Date("2024-12-31") - as.Date("2024-01-01")     # 365
seq(as.Date("2024-01-01"), as.Date("2024-06-01"),
    by="month")
d <- as.Date("2024-07-15")
weekdays(d); months(d); format(d, "%A %B %d, %Y")
```


## Ex 112: `POSIXct` and Timezones

::: {.alertblock}
### Problem

- Create `as.POSIXct("2024-01-15 14:30:00", tz="Asia/Taipei")`
- Extract hour, minute via `POSIXlt`
- Difference between two timestamps in hours
- Format with `strftime`
:::


## Sol 112

```{r}
t1 <- as.POSIXct("2024-01-15 14:30:00", tz="Asia/Taipei")
lt <- as.POSIXlt(t1); lt$hour; lt$min
t2 <- as.POSIXct("2024-01-15 18:45:00", tz="Asia/Taipei")
difftime(t2, t1, units="hours")
strftime(t1, "%Y/%m/%d %H:%M %Z")
```


## Ex 113: Binary I/O

::: {.alertblock}
### Problem

- `charToRaw("Hello")` — what type?
- Convert back with `rawToChar`
- `as.integer(charToRaw("ABC"))` — what are the byte values?
- Write raw bytes to a temp file with `writeBin` and read back
:::


## Sol 113

```{r}
r <- charToRaw("Hello"); r; typeof(r)             # raw
rawToChar(r)                                       # "Hello"
as.integer(charToRaw("ABC"))                       # 65 66 67
tmp <- tempfile()
writeBin(charToRaw("test123"), tmp)
rawToChar(readBin(tmp, "raw", 20))
```


## Ex 114: `readline` and `interactive`

::: {.alertblock}
### Problem

- Write a function that uses `readline` to ask for user's name (use `eval=FALSE`)
- Use `interactive()` to branch: interactive → readline, batch → default
- Why does `readline` always return character?
:::


## Sol 114

```{r, eval=FALSE}
ask_name <- function() {
  if (interactive()) {
    name <- readline("Your name: ")
  } else {
    name <- "default_user"
  }
  cat("Hello,", name, "\n")
}
# readline reads raw text from console; R has no way
# to know the intended type, so it always returns character
```


## Ex 115: `system2`

::: {.alertblock}
### Problem

- Run `system2("echo", args=c("hello","world"))`
- Capture output with `stdout=TRUE`
- Run in background with `wait=FALSE`
- Cross-platform check with `.Platform$OS.type`
:::


## Sol 115

```{r, eval=FALSE}
system2("echo", args=c("hello","world"))
out <- system2("date", stdout=TRUE); out
system2("sleep", args="1", wait=FALSE)
if (.Platform$OS.type == "unix") {
  system2("uname", args="-a", stdout=TRUE)
}
```


## Ex 116: S3 Classes

::: {.alertblock}
### Problem

- Create S3 class `Circle`: `list(radius=r)` with `class="Circle"`
- Write `print.Circle` and `area.Circle`
- Create instance; dispatch `print` and `area`
:::


## Sol 116

```{r}
Circle <- function(r) structure(list(radius=r), class="Circle")
print.Circle <- function(x, ...)
  cat(sprintf("Circle(r=%.2f)\n", x$radius))
area <- function(x, ...) UseMethod("area")
area.Circle <- function(x, ...) pi * x$radius^2
c1 <- Circle(5); c1
area(c1)
```


## Ex 117: Formula and `lm`

::: {.alertblock}
### Problem

- `f <- y ~ x`; show `class()`, `typeof()`, `all.vars()`
- Fit `lm(dist ~ speed, data=cars)`. Extract coefficients
- Plot data with regression line
:::


## Sol 117

```{r, out.width="35%"}
f <- y ~ x; class(f); typeof(f); all.vars(f)
model <- lm(dist ~ speed, data=cars)
coef(model)
plot(cars$speed, cars$dist, pch=19)
abline(model, col="red", lwd=2)
```


## Ex 118: Plotting

::: {.alertblock}
### Problem

- Plot `sin(x)` and `cos(x)` for `x` from 0 to $2\pi$ with legend
- `par(mfrow=c(1,2))`: histogram and boxplot side by side
- Add annotation with `text()` and horizontal line with `abline(h=...)`
:::


## Sol 118

```{r, out.width="35%"}
x <- seq(0, 2*pi, length=100)
plot(x, sin(x), type="l", col="blue", lwd=2, ylab="y")
lines(x, cos(x), col="red", lwd=2, lty=2)
legend("topright", c("sin","cos"), col=c("blue","red"),
       lty=1:2, lwd=2)
```


## Ex 119: Random Number Generation

::: {.alertblock}
### Problem

- `set.seed(42)` then generate 5 uniform, 5 normal, 10 Bernoulli
- Simulate 1000 dice rolls and tabulate
- Estimate $\pi$ via Monte Carlo: sample `(x,y)` uniform in $[0,1]^2$, count proportion inside unit circle quadrant
:::


## Sol 119

```{r}
set.seed(42)
runif(5); rnorm(5); rbinom(10, 1, 0.5)
table(sample(1:6, 1000, replace=TRUE))
n <- 1e5; x <- runif(n); y <- runif(n)
4 * mean(x^2 + y^2 <= 1)  # ≈ π
```


## Ex 120: Options and Benchmarking

::: {.alertblock}
### Problem

- Save current `digits` option, set to 15, print `pi`, restore
- Time `crossprod(x)` for `x <- matrix(rnorm(1e6),1000)` with `system.time`
- Use `proc.time` for manual timing of `sum(1:1e7)`
:::


## Sol 120

```{r}
old <- options(digits=15); pi
options(old)
system.time({
  x <- matrix(rnorm(1e6), 1000, 1000)
  y <- crossprod(x)
})
pt <- proc.time(); s <- sum(1:1e7); proc.time()-pt
```


# Closures and Lexical Scoping (Ex 121--128)


## Ex 121: Power Factory

::: {.alertblock}
### Problem

- Write `power_factory(n)` that returns a function computing $x^n$
- Create `square` and `cube` from it; test on 5
- Inspect the captured `n` with `environment(square)$n`
- Verify `square` and `cube` have different environments
:::


## Sol 121

```{r}
power_factory <- function(n) function(x) x^n
square <- power_factory(2); cube <- power_factory(3)
square(5); cube(5)
environment(square)$n; environment(cube)$n          # 2; 3
identical(environment(square), environment(cube))   # FALSE
```


## Ex 122: Closure Counter

::: {.alertblock}
### Problem
Write `make_counter()` returning `list(increment, get, reset)`.

- `increment()` adds 1 to internal `n` and returns it
- `get()` returns current `n`; `reset()` sets `n` back to 0
- Create two independent counters; show they don't share state
- Rewrite using `local()` for a single-instance counter
:::


## Sol 122

```{r}
make_counter <- function() {
  n <- 0
  list(increment = function() { n <<- n + 1; n },
       get = function() n,
       reset = function() { n <<- 0 })
}
c1 <- make_counter(); c2 <- make_counter()
c1$increment(); c1$increment(); c2$increment()
cat("c1:", c1$get(), " c2:", c2$get(), "\n")
c1$reset(); c1$get()
```

```{r}
# local() single-instance
counter <- local({
  n <- 0
  list(increment = function() { n <<- n + 1; n },
       get = function() n)
})
counter$increment(); counter$increment(); counter$get()
```


## Ex 123: Loop Capture Bug

::: {.alertblock}
### Problem

- Create 5 functions in a `for` loop: `funs[[i]] <- function() i`. Call all 5. What happens?
- Fix using `force()` inside a wrapper function
- Fix using `lapply`
- Explain *why* the bug occurs (lazy evaluation + shared environment)
:::


## Sol 123

```{r}
funs <- list()
for (i in 1:5) funs[[i]] <- function() i
sapply(funs, \(f) f())                              # all 5!
# Bug: all closures share the same env; i=5 after loop

make_fun <- function(val) { force(val); function() val }
funs <- list()
for (i in 1:5) funs[[i]] <- make_fun(i)
sapply(funs, \(f) f())

funs <- lapply(1:5, \(i) function() i)
sapply(funs, \(f) f())
# Lazy eval: function() i captures a promise to i,
# not its value. By call time, loop finished, i=5.
```


## Ex 124: Memoized Fibonacci

::: {.alertblock}
### Problem
Write `make_memo_fib()` returning `list(fib, cache_size, clear_cache)`.

- Use a named list as cache; `fib(n)` checks cache before computing
- Time `fib(35)` first call vs second call
- How large is the cache after `fib(35)`?
- Clear cache and verify `fib(10)` still works
:::


## Sol 124

```{r}
make_memo_fib <- function() {
  cache <- list()
  fib <- function(n) {
    key <- as.character(n)
    if (!is.null(cache[[key]])) return(cache[[key]])
    result <- if (n <= 2) 1 else fib(n-1) + fib(n-2)
    cache[[key]] <<- result; result
  }
  list(fib = fib,
       cache_size = function() length(cache),
       clear_cache = function() { cache <<- list() })
}
mf <- make_memo_fib()
system.time(mf$fib(35))                             # first
system.time(mf$fib(35))                             # second: instant
mf$cache_size()                                      # 35
mf$clear_cache(); mf$fib(10)
```


## Ex 125: Decorator --- Logging Wrapper

::: {.alertblock}
### Problem
Write `with_logging(f)` that wraps `f` to print call count on each invocation.

- Return a closure that increments a counter and calls `f(...)`
- Test: `logged_mean <- with_logging(mean)`. Call it 3 times
- Extend to also print elapsed time per call
:::


## Sol 125

```{r}
with_logging <- function(f, fname = deparse(substitute(f))) {
  count <- 0L
  function(...) {
    count <<- count + 1L
    t0 <- proc.time()["elapsed"]
    cat(sprintf("[%s] call #%d\n", fname, count))
    result <- f(...)
    cat(sprintf("  elapsed: %.4fs\n",
        proc.time()["elapsed"] - t0))
    result
  }
}
logged_mean <- with_logging(mean)
logged_mean(1:100); logged_mean(rnorm(1e5))
```


## Ex 126: Bank Account with Transaction Log

::: {.alertblock}
### Problem
Write `make_account(owner, balance)` returning `list(deposit, withdraw, statement, balance)`.

- `deposit(amount)` adds and logs. `withdraw(amount)` subtracts; `stop()` if insufficient
- `statement()` returns a data frame of all transactions
- Create two accounts; verify independence
:::


## Sol 126

```{r}
make_account <- function(owner, init = 0) {
  bal <- init
  log <- data.frame(time=Sys.time()[0], type=character(0),
                    amount=numeric(0), balance=numeric(0))
  record <- function(type, amount) {
    log <<- rbind(log, data.frame(
      time=Sys.time(), type=type, amount=amount, balance=bal))
  }
  list(
    deposit = function(amt) {
      stopifnot(amt > 0); bal <<- bal + amt
      record("deposit", amt); invisible(bal)
    },
    withdraw = function(amt) {
      stopifnot(amt > 0)
      if (amt > bal) stop("Insufficient funds")
      bal <<- bal - amt; record("withdraw", amt); invisible(bal)
    },
    balance = function() bal,
    statement = function() log)
}
alice <- make_account("Alice", 1000)
bob   <- make_account("Bob", 500)
alice$deposit(200); alice$withdraw(50); bob$deposit(100)
alice$balance(); bob$balance()
alice$statement()
```


## Ex 127: Moving Average Filter

::: {.alertblock}
### Problem
Write `make_ema(alpha)` returning a function that computes an exponential moving average: $\text{EMA}_t = \alpha x_t + (1-\alpha)\text{EMA}_{t-1}$.

- First call initializes EMA to the input value
- Test with prices `c(100, 102, 101, 105, 103, 107, 106)` and $\alpha=0.2$
- Create two EMAs with different $\alpha$; verify independence
:::


## Sol 127

```{r}
make_ema <- function(alpha = 0.3) {
  value <- NULL
  function(x) {
    if (is.null(value)) value <<- x
    else value <<- alpha * x + (1 - alpha) * value
    value
  }
}
ema <- make_ema(0.2)
sapply(c(100, 102, 101, 105, 103, 107, 106), ema)
ema_fast <- make_ema(0.5); ema_slow <- make_ema(0.1)
sapply(c(100, 110, 90), ema_fast)
sapply(c(100, 110, 90), ema_slow)
```


## Ex 128: Fibonacci Iterator

::: {.alertblock}
### Problem
Write `make_fib_iterator()` returning `list(next_val, reset)`.

- `next_val()` returns the next Fibonacci number on each call
- Use `replicate(12, fib_it$next_val())` to get first 12 values
- Reset and verify the sequence restarts
:::


## Sol 128

```{r}
make_fib_iterator <- function() {
  a <- 0; b <- 1
  list(
    next_val = function() {
      val <- a; new <- a + b; a <<- b; b <<- new; val
    },
    reset = function() { a <<- 0; b <<- 1 })
}
fib_it <- make_fib_iterator()
replicate(12, fib_it$next_val())
fib_it$reset(); replicate(5, fib_it$next_val())
```


# Split--Apply--Combine Deep Dive (Ex 129--136)


## Ex 129: Multi-Factor Split and Unsplit

::: {.alertblock}
### Problem
Using `ToothGrowth`:

- Split by both `supp` and `dose`; how many groups?
- Compute mean `len` per group using `sapply`
- Add a z-score column within each group using `split` + `lapply` + `do.call(rbind,...)`
- Use `unsplit` to restore original ordering after modifying split pieces
:::


## Sol 129

```{r}
tg <- ToothGrowth
sp <- split(tg, list(tg$supp, tg$dose))
length(sp)                                           # 6
sapply(sp, \(d) mean(d$len))
parts <- lapply(sp, \(d) {
  d$len_z <- (d$len - mean(d$len)) / sd(d$len); d })
tg2 <- do.call(rbind, parts)
head(tg2[order(as.numeric(rownames(tg2))), ], 4)
```

```{r}
# unsplit example
sg <- split(1:8, rep(c('A','B'), 4))
sg$A <- sg$A * 10; sg$B <- sg$B * (-1)
unsplit(sg, rep(c('A','B'), 4))
```


## Ex 130: `aggregate()` --- Matrix Column Gotcha

::: {.alertblock}
### Problem
Using `ToothGrowth`:

- Use `aggregate` with `FUN = function(x) c(m=mean(x), s=sd(x))`. What type is the result column?
- Expand the matrix column into a proper flat data frame
- Use `cbind(mpg, hp, wt) ~ cyl` to summarize multiple columns of `mtcars`
- Use `.` shorthand: `aggregate(. ~ cyl, ...)` on a subset of `mtcars`
:::


## Sol 130

```{r}
tg <- ToothGrowth
res <- aggregate(len ~ supp, data = tg,
  FUN = function(x) c(m=mean(x), s=sd(x)))
str(res)                                             # matrix col!
cbind(res[1], as.data.frame(res$len))                # flatten

aggregate(cbind(mpg, hp, wt) ~ cyl,
          data = mtcars, FUN = mean)
aggregate(. ~ cyl,
  data = mtcars[, c('mpg','hp','wt','cyl')], FUN = mean)
```


## Ex 131: `merge()` --- All Join Types

::: {.alertblock}
### Problem
`cust <- data.frame(id=1:4, name=c("A","B","C","D"))`.
`ord <- data.frame(cid=c(2,3,3,5), prod=c("pen","book","ink","tape"))`.

- Inner join. Which rows are dropped and why?
- Left join. What appears for customer D?
- Right join. What appears for `cid=5`?
- Full outer join
:::


## Sol 131

```{r}
cust <- data.frame(id=1:4, name=c("A","B","C","D"))
ord <- data.frame(cid=c(2,3,3,5),
                  prod=c("pen","book","ink","tape"))
merge(cust, ord, by.x="id", by.y="cid")              # inner
merge(cust, ord, by.x="id", by.y="cid", all.x=TRUE)  # left
merge(cust, ord, by.x="id", by.y="cid", all.y=TRUE)  # right
merge(cust, ord, by.x="id", by.y="cid", all=TRUE)    # full
```


## Ex 132: Multi-Table Merge with `Reduce`

::: {.alertblock}
### Problem

- Create `products(pid, name)`, `orders(oid, pid, qty)`, `prices(pid, unit_price)`
- Inner-join all three; compute `total = qty * unit_price`
- Add `regions(pid, region)`. Use `Reduce` to merge all four
:::


## Sol 132

```{r}
products <- data.frame(pid=1:5,
  name=c('Pen','Book','Ink','Paper','Tape'))
orders <- data.frame(oid=101:106,
  pid=c(2,1,3,2,5,1), qty=c(3,10,2,5,1,8))
prices <- data.frame(pid=1:5,
  unit_price=c(1.5, 12, 8, 3, 2.5))
m <- merge(merge(products, orders, by='pid'),
           prices, by='pid')
m$total <- m$qty * m$unit_price; m[, c('oid','name','total')]

regions <- data.frame(pid=1:5, region=c('N','S','N','S','N'))
Reduce(\(x, y) merge(x, y, by='pid', all=TRUE),
       list(products, orders, prices, regions))
```


## Ex 133: `vapply` vs `sapply` Safety

::: {.alertblock}
### Problem

- `sapply(list(a=1:3, b=4:6), identity)` returns a matrix. `sapply(list(a=1:3, b=4:5), identity)` returns a list. Demonstrate
- Show that `vapply(list(a=1:3, b=4:6), identity, numeric(1))` throws an error
- Use `vapply` correctly with `FUN.VALUE = numeric(3)` for same-length vectors
- Use `vapply(iris[,1:4], median, numeric(1))` to safely compute medians
:::


## Sol 133

```{r}
sapply(list(a=1:3, b=4:6), identity)                 # matrix
sapply(list(a=1:3, b=4:5), identity)                 # list!
```

```{r, error=TRUE}
vapply(list(a=1:3, b=4:6), identity, numeric(1))    # error!
vapply(list(a=1:3, b=4:6), identity, numeric(3))    # OK
vapply(iris[,1:4], median, numeric(1))               # safe
```


## Ex 134: `rapply()` --- Recursive Apply

::: {.alertblock}
### Problem
`nested <- list(a=1:3, b=list(c=4:6, d=list(e=7:9, f="text")))`.

- Use `rapply` with `sum`, `classes="integer"` to sum only integer leaves
- Same with `how="list"` to preserve structure
- Use `how="replace"` to multiply all integer leaves by 10
- What happens to the `"text"` element in each case?
:::


## Sol 134

```{r}
nested <- list(a=1:3,
  b=list(c=4:6, d=list(e=7:9, f="text")))
rapply(nested, sum, classes="integer")               # unlist
rapply(nested, sum, classes="integer", how="list")   # structure
rapply(nested, \(x) x*10, classes="integer",
       how="replace")                                # replace
# "text" is left unchanged (not class "integer")
```


## Ex 135: `Reduce` --- Running Max Drawdown

::: {.alertblock}
### Problem
\texttt{r <- c(0.02, -0.01, 0.03, -0.04, 0.01, -0.02, 0.05)}.

- Use `Reduce` with `accumulate=TRUE` to compute cumulative wealth $W_t = W_{t-1}(1+r_t)$, starting from $W_0=1$
- Compute running max of cumulative wealth using `Reduce` and `max`
- Compute drawdown: $\text{DD}_t = 1 - W_t / \max_{s \le t} W_s$
- Verify with vectorized `1 - cumprod(1+r) / cummax(cumprod(1+r))`
:::


## Sol 135

```{r}
r <- c(0.02, -0.01, 0.03, -0.04, 0.01, -0.02, 0.05)
W <- Reduce(\(w, ri) w*(1+ri), r, init=1,
            accumulate=TRUE)[-1]
Wmax <- Reduce(\(mx, w) max(mx, w), W,
               accumulate=TRUE)
DD <- 1 - W / Wmax; round(DD, 6)
DD2 <- 1 - cumprod(1+r) / cummax(cumprod(1+r))
all.equal(DD, DD2)                                   # TRUE
```


## Ex 136: `Filter` Pipeline with `|>`

::: {.alertblock}
### Problem
`words <- c("apple","Banana","cherry","Date","123","fig","GRAPE","hi")`.

- Use `Filter` to keep only all-lowercase words
- Use `Filter` + `Negate` to remove strings containing digits
- Find first word with $>5$ characters using `Find`; find its position with `Position`
- Chain with `|>`: filter lowercase $\to$ filter length $>3$ $\to$ sort
:::


## Sol 136

```{r}
words <- c("apple","Banana","cherry","Date","123",
           "fig","GRAPE","hi")
Filter(\(w) w == tolower(w), words)
Filter(Negate(\(w) grepl("[0-9]", w)), words)
Find(\(w) nchar(w) > 5, words)
Position(\(w) nchar(w) > 5, words)

words |>
  Filter(f = \(w) w == tolower(w)) |>
  Filter(f = \(w) nchar(w) > 3) |>
  sort()
```


# Advanced Closure Applications (Ex 137--140)


## Ex 137: Closure-Based ODE Solver (Euler Method)

::: {.alertblock}
### Problem
Write `make_ode_solver(f, y0, dt)` where $f(t,y)$ defines $dy/dt$.

- `step()` advances one time step; `run(n)` returns trajectory as a data frame
- Solve $dy/dt = -0.5y$, $y(0)=10$, $dt=0.1$ for 100 steps
- Plot Euler vs exact $y=10e^{-0.5t}$
:::


## Sol 137

```{r, out.width="40%"}
make_ode_solver <- function(f, y0, dt = 0.01) {
  t <- 0; y <- y0
  list(
    run = function(n) {
      traj <- matrix(NA, n+1, 2, dimnames=list(NULL, c("t","y")))
      traj[1, ] <- c(t, y)
      for (i in seq_len(n)) {
        y <<- y + f(t, y)*dt; t <<- t + dt
        traj[i+1, ] <- c(t, y)
      }
      as.data.frame(traj)
    },
    state = function() c(t=t, y=y),
    reset = function(y0_new=y0) { t <<- 0; y <<- y0_new })
}
solver <- make_ode_solver(\(t, y) -0.5*y, y0=10, dt=0.1)
traj <- solver$run(100)
plot(traj$t, traj$y, type='l', lwd=2, col='steelblue',
     xlab='t', ylab='y', main='Euler vs Exact')
curve(10*exp(-0.5*x), add=TRUE, col='red', lwd=2, lty=2)
legend('topright', c('Euler','Exact'),
       col=c('steelblue','red'), lty=1:2, lwd=2)
```


## Ex 138: Rate Limiter

::: {.alertblock}
### Problem
Write `make_rate_limiter(f, max_calls, period)` wrapping `f` so at most `max_calls` invocations are allowed per `period`-second window.

- If limit exceeded, return `NULL` with a warning
- Track timestamps in a closure variable
- Test: limiter allowing 3 calls per 2 seconds; call 5 times rapidly
:::


## Sol 138

```{r}
make_rate_limiter <- function(f, max_calls=3, period=2) {
  timestamps <- numeric(0)
  function(...) {
    now <- as.numeric(Sys.time())
    timestamps <<- timestamps[timestamps > now - period]
    if (length(timestamps) >= max_calls) {
      warning("Rate limit exceeded"); return(NULL)
    }
    timestamps <<- c(timestamps, now)
    f(...)
  }
}
limited_sqrt <- make_rate_limiter(sqrt, 3, 2)
for (i in 1:5) {
  r <- limited_sqrt(i * 10)
  cat(sprintf("Call %d: %s\n", i,
      if (is.null(r)) "BLOCKED" else sprintf("%.4f", r)))
}
```


## Ex 139: Portfolio Analytics Factory

::: {.alertblock}
### Problem
Write `make_portfolio(weights, tickers)` returning:

- `returns(price_matrix)` --- weighted portfolio returns
- `sharpe(price_matrix, rf)` --- annualized Sharpe ratio
- `drawdown(price_matrix)` --- maximum drawdown
- Test with simulated 252-day prices for 3 assets
:::


## Sol 139

```{r, out.width="40%"}
make_portfolio <- function(w, tickers) {
  w <- w / sum(w); names(w) <- tickers
  list(
    returns = function(P) {
      r <- apply(P, 2, \(x) diff(x)/head(x,-1)); drop(r %*% w)
    },
    sharpe = function(P, rf=0) {
      r <- apply(P, 2, \(x) diff(x)/head(x,-1))
      pr <- drop(r %*% w)
      (mean(pr) - rf/252) / sd(pr) * sqrt(252)
    },
    drawdown = function(P) {
      r <- apply(P, 2, \(x) diff(x)/head(x,-1))
      cum <- cumprod(1 + drop(r %*% w))
      max(1 - cum / cummax(cum))
    })
}
set.seed(42); P <- matrix(NA, 253, 3); P[1,] <- c(100,50,200)
for (i in 2:253) P[i,] <- P[i-1,]*exp(rnorm(3, 3e-4,
                                        c(.015,.02,.01)))
pf <- make_portfolio(c(.4,.3,.3), c("A","G","M"))
cat("Sharpe:", round(pf$sharpe(P), 3), "\n")
cat("MaxDD:", round(pf$drawdown(P)*100, 2), "%\n")
r <- pf$returns(P)
plot(cumprod(1+r), type='l', lwd=2, col='steelblue',
     main='Portfolio Growth', ylab='Growth of $1')
abline(h=1, lty=2, col='grey50')
```


## Ex 140: `by()` and Full EDA Report

::: {.alertblock}
### Problem
Using `mtcars`:

- Use `by(mtcars, mtcars$cyl, \(d) cor(d$mpg, d$wt))` to compute per-cylinder correlation
- Mean, sd, min, max of `mpg` by `cyl` in one `aggregate` call; expand the matrix column
- Merge with group counts from `aggregate(..., FUN=length)`
- Sort by mean `mpg` descending
:::


## Sol 140

```{r}
by(mtcars, mtcars$cyl, \(d) cor(d$mpg, d$wt))
res <- aggregate(mpg ~ cyl, data=mtcars,
  FUN=\(x) c(mean=mean(x), sd=sd(x), min=min(x), max=max(x)))
flat <- cbind(res[1], as.data.frame(res$mpg))

counts <- aggregate(mpg ~ cyl, data=mtcars, FUN=length)
names(counts)[2] <- 'n'
flat <- merge(flat, counts, by='cyl')
flat[order(-flat$mean), ]
```
